### 2017-03-01 Manuscript/ms.tex
Starting on manuscript

% Template for PLoS
% Version 3.4 January 2017
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Linking mosquito surveillance to dengue risk through Bayesian mechanistic modeling}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Clinton Leach\textsuperscript{1*},
Colleen Webb\textsuperscript{1},
Kim Pepin\textsuperscript{2},
Alvaro Eiras\textsuperscript{3},
Mevin Hooten\textsuperscript{4},
Jennifer Hoeting\textsuperscript{4}

\bigskip
\textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* clint.leach@colostate.edu

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}

We know mosquitoes spread dengue, but mosquito abundance, as estimated from trap-based surveillance data, is not causally linked to human disease risk.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}

We know mosquitoes spread dengue, but mosquito abundance, as estimated from trap-based surveillance data, is not causally linked to human disease risk.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}

\subsection*{Dengue background}
\begin{itemize}
  \item large public health burden, with millions cases/year in Brazil
  \item Complex dynamics -- vector-borne (with resulting seasonal forcing), multi-serotype (cross-immunity, ADE)
\end{itemize}

\subsection*{Dengue prevention}
\begin{itemize}
  \item Mosquito surveillance and control -- even once a vaccine is developed dengue management plans will likely still include vector control (\cite{Achee2015})
  \item General efforts -- education, source reduction, etc; National Dengue Control system
  \item Targeted efforts -- ecovec and "MI-Dengue" system -- trap monitoring and spatial prioritization of source reduction and larvicide application (\cite{Eiras2009}).
  \begin{itemize}
  \item Based on connection between mosquito abundance and disease risk
  \item \href{http://www.vitoria.es.gov.br/prefeitura/mosquito-controle-ajuda-a-reduzir-incidencia-de-viroses}{general mosquito control} (City of Vitoria)
  \item \href{http://www.vitoria.es.gov.br/prefeitura/combateadengue}{fight against dengue} (City of Vitoria)
  \end{itemize}
\end{itemize}

\subsection*{Causal inference and mechanistic models}

Previous efforts to use MosquiTRAP surveillance to predict human dengue cases in Vitoria relied on linear statistical models and asked whether models that included surveillance data performed better than models with lagged case data alone \cite{Pepin2015}.
This procedure essentially tests for Granger causality \cite{Granger1969}, which states that some variable $X$ \emph{causes} $Y$ if $X$ contains unique information about $Y$ (i.e. if a model with $X$ and $Y$ provides better predictions of $Y$ than a model with $Y$ alone).
However, as noted by Sugihara \emph{et al.} \cite{Sugihara2012a}, Granger causality assumes that $X$ and $Y$ are separable, which is not the case if $X$ and $Y$ are linked in a non-linear dynamical system.
In this case, a model with $X$ and $Y$ is unlikely to outperform the model with only $Y$, since $Y$ already contains the information from $X$, a consequence of Taken's theorem (CITATION). 
Instead, Sugihara \emph{et al.} argue that if $X$ drives $Y$, then you should be able to reconstruct $X$ from $Y$.

Applying this reasoning to the problem of establishing the relationship between mosquito abundance and human dengue risk, it is not surprising that including information from mosquito surveillance does not improve our ability to predict human dengue cases, as information about epidemiologically relevant mosquito abundance is already embedded in the time series of human case reports.
Following from \cite{Sugihara2012a} then, we ask whether we can reconstruct the mosquito surveillance time series from the time series of human case reports. 
We do this through the use of a mechanistic differential equation model embedded in a Bayesian statistical framework.
This differs from the equation-free approach of \cite{Sugihara2012a}, but the explicit use of a mechanistic model will better inform our understanding of the processes linking mosquito abundance and human disease risk and the embedding in a Bayesian statistical model will better account for uncertainty and the measurement processes in these data.

\section*{Methods}

General questions:
\begin{itemize}
\item Do we need to write out the explicit factorization of the posterior?
\end{itemize}
\subsection*{Study system and data}

Vitoria is a coastal city and the capital of the state of Espirito Santo, Brazil, with a population of 304,331 (check) in the city proper, and an additional 1.5 million people in the greater metropolitan area.  
The city has an area of approximately 93 sq. km., with most of that area taken up by a large island surrounded by the river Rio Santa Maria and bay Baia de Vitoria.  
There is also a large, mountainous natural area that occupies a large portion of the interior of the island.

Since 2008, the city has monitored mosquito abundance using 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a grid across the city.
Each trap is checked weekly, providing us with 243 weeks (2008 through week 34 of 2012) of counts of gravid female \emph{Aedes aegypti}.
Dengue is a mandatory notifiable disease, and thus the city's Ministry of Health Secretary maintains a database of weekly notified probable dengue cases (i.e. medical care sought for dengue-like symptoms).

Governador Valladares is also a city for which we have data.

Sete Lagoas is also a city for which we have data.

\subsection*{Process Model}

Dengue epidemiology is complicated considerably by the presence of four simultaneously circulating serotypes (referred to as DENV-1, DENV-2, DENV-3, DENV-4).
Infection with one serotype confers life-long immunity to that serotype, along with temporary immunity to other serotypes.  
As this cross-immunity wanes, antibodies from the previous infection can actually result in antibody-dependent enhancement (ADE), wherein hosts are more susceptible to infection with the other serotypes and more likely to develop severe symptoms (i.e. dengue haemorrhagic fever or dengue shock syndrome).
The strength and duration of these different inter-serotype interactions are not well understood, though different models suggest that temporary cross-immunity alone (without ADE) is sufficient to reproduce observed multi-annual dynamics in Thailand \cite{Wearing2006,Reich2013}.
Similarly, \cite{Aguiar2013} find that capturing both primary and secondary infections and the period of cross-immunity is crticial, but that explicitly including all four serotypes (at great cost to model complexity) does not perform much better than a two serotype model, and indeed that the number of hospital admissions caused by a third or fourth infection is very low \cite{Aguiar2011a}.

Explicitly accounting for interactions between serotypes, even only two of the four, leads to a large and complex mechanistic model.
Moreover, since dengue case reports do not include information on serotype, we do not have enough information to inform the dynamics of individual serotypes.
As such, we simplify the model of \cite{Wearing2006} to an SEIRS framework, which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
In this framework, susceptible individuals ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
Following a latent period ($\frac{1}{\rho}$), exposed individuals become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
Infectious individuals recover at rate $\gamma$ and remain cross-immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.

Similarly, susceptible mosquitoes become exposed by biting infectious humans and pass through a temperature-dependent latent period ($\frac{1}{\rho_{v}}$) before becoming infectious.
Total mosquito abundance, and thus the size of the mosquito susceptible pool, is forced by a latent process model informed by the mosquito surveillance data (see below).

The complete differential equation model is then given as:
\begin{align} 
\frac{dS}{dt} &= bN - bS - \alpha \frac{V_{I}}{N} S + \delta R\\
\frac{dE}{dt} &= \alpha \frac{V_{I}}{N} S - (\rho + b)E\\
\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
\frac{dR}{dt} &= \gamma I - (\delta + b)R\\
V_{St} & = V_{Nt} - V_E(t) - V_I(t)\\
\frac{V_E}{dt} &= \alpha \frac{I}{N} V_S - (\rho_{vt} + d_v)V_E\\
\frac{V_I}{dt} &= \rho_{vt} V_E - d_v V_I
\end{align}

\subsection*{Data model}

This mechanistic differential equation model is then embedded in a Bayesian statistical model in order to estimate the latent mosquito abundance time series and generate sample mosquito trap counts to compare with our observations.
To connect the differential equation model the the observed case reports, we add an extra state, $X$, that collects the cumulative number of transitions from the exposed to infectious class (assuming that case reporting coincides with the onset of symptoms).
We then model the number of new cases reported in week $t$ ($y_t$) as:
\begin{equation}
y_t  \sim \text{NegBin}(\phi_y (X_t - X_{t-1}), \eta_y),
\end{equation}
where $\phi_y$ is the reporting probability, and $\eta_y$ controls the overdispersion relative to the Poisson.

The unobserved total mosquito abundance is modeled as a random walk on the log-scale:
\begin{equation}
\log(V_{Nt}) \sim \text{Normal}(\log(V_{Nt-1}), \sigma)
\end{equation}
This is then connected to the number of mosquitoes trapped in week $t$ ($q_t$) with a negative binomial measurement model:
\begin{equation}
q_t \sim \text{NegBin}(\phi_q \tau_t V_{Nt}, \eta_q),
\end{equation}
where $\phi_q$ is the per-trap capture probability, $\tau_t$ is the number of traps inspected in week $t$, and $\eta_q$ controls overdispersion relative to the Poisson.

The flexibility of the mosquito process model means that we should be able to essentially fit the case data arbitrarily well.
The question is then whether or not the mosquito abundance time series estimated from the case data reproduces the observed mosquito surveillance time series.
Since the initial conditions and epidemiological parameters potentially affect the mapping of the case data to mosquito abundance, we estimate these as well.
The intention here is that the mosquito process will allow us to fit to the case data (i.e. there's essentially a best-fit trajectory that we'll always be able to find), while the epidemiological parameters will facilitate better fits to the mosquito data (while maintaining the same quality of fit at the case level).

\subsection*{Parameterization and priors}

Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito (i.e., the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
To account for this, we make $\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground (URL/citation).
We use the temperature relationship from the recent meta-analysis of \cite{Chan2012}, which incorporates a large number of records and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
We use a slightly more conservative fitted relationship from a log-normal model excludes right-censored data and produces a less extreme temperature dependence:
\begin{align}
\frac{1}{\rho_v} &= \exp \left(\mu + \frac{1}{14}\right)\\
\mu & = \exp(1.9 - 0.04 T)\\
\end{align}

This relationship is similar to those reported in \cite{Focks1995} and \cite{Tjaden2013}, though there is substantial uncertainty about the exact shape of this relationship.

\begin{table}[h]
\begin{center}
\caption{Fixed parameter values, taken from the literature.}
\begin{tabular}{llll}
Parameter & Description & Value & Citation\\
\hline
$1/d = 1/b$ & Human life-span & 76 years & City of Vitoria\\
$1/d_V = 1/b_V$ & Vector life-span & 5 days & \cite{Brady2013} \\
$1/\rho$ & Latent period in host & 6.1 days & \cite{Chan2012}\\
$1/\gamma$ & Infectious period in host & 3 days & \cite{Nguyet2013}\\
$1/\delta$ & Period of cross-immunity & 686 days &  \cite{Reich2013}\\
\end{tabular}
\end{center}
\label{fixedparms}
\end{table}

\subsection*{Implementation}

The posterior distribution can be written as:
\begin{equation}
aaaaaaah
\end{equation}
Samples from this distribution were generated using Hamiltonian Monte Carlo (CITATIONS) implemented in the rstan package for R. 
Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.

\subsection*{Posterior predictive checks}

\section*{Results}

\section*{Discussion}

\subsection*{Additional analyses}

\begin{itemize}
  \item Dig into seasonal trends in fit
  \item Look at how mosquito predictions change as data are added
\end{itemize}

\subsection*{Unmodeled processes}

Human movement \cite{Adams2009, Cosner2009a, Stoddard2009, Dalziel2013}
\begin{itemize}
  \item Local mosquito abundance does not necessarily predict local disease risk
  \begin{itemize}
    \item human movement creates complex host-vector mixing structure (depends on vector ecology and activity periods)
    \item mosquitoes may vary in their contribution to human disease burden (this effect size is likely fairly small and the model doesn't (currently) provide a way to pull this out)
  \end{itemize}
  \item Need to account for human movement in order to prioritize spatial application of control
\end{itemize}
\section*{Conclusions}

\section*{Supporting information}

\subsection*{Full model}

\begin{align}
y_t & \sim \text{NegBin}(\phi_y (X_t - X_{t-1}), \eta_y)\\
q_t & \sim \text{NegBin}(\phi_q \tau_t V_{Nt}, \eta_q)\\
\\ 
\frac{dX}{dt} &= \rho E \\
\frac{dS}{dt} &= bN - bS - \alpha \frac{V_{I}}{N} S + \delta R\\
\frac{dE}{dt} &= \alpha \frac{V_{I}}{N} S - (\rho + b)E\\
\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
\frac{dR}{dt} &= \gamma I - (\delta + b)R\\
\\
V_{St} & = V_{Nt} - V_E(t) - V_I(t)\\
\frac{dV_E}{dt} &= \alpha \frac{I}{N} V_S - (\rho_{vt} + d_v)V_E\\
\frac{dV_I}{dt} &= \rho_{vt} V_E - d_v V_I\\
\\
\log(V_{N0}) & \sim \text{Normal}(13, 2)\\
\log(V_{Nt}) & \sim \text{Normal}(\log(V_{Nt-1}), \sigma^2)\\
\sigma & \sim \text{Half-normal}(0, 1)\\
\phi_q & \sim \text{Half-normal}(-16, 2.5)\\
\eta_y &\sim \text{Half-normal}(0, 5)\\
\eta_q & \sim \text{Half-cauchy}(0, 5)\\
\alpha_{rel} & \sim \text{Half-normal}(1, 0.3)\\
\rho_{rel} & \sim \text{Half-normal}(1, 0.3)\\
\gamma_{rel} & \sim \text{Half-normal}(1, 0.3)\\
\delta_{rel} & \sim \text{Half-normal}(1, 0.3)\\
d_{v,rel} & \sim \text{Half-normal}(1, 0.3)\\
\mathbf{\pi_{0}} & \sim \text{Normal}(\mathbf{\mu}, 0.09\mathbf{I})\\
\mathbf{p_0} & = \frac{\exp(\mathbf{\pi_0})}{\sum \exp(\mathbf{\pi_0})} \\
\end{align}

\subsection*{Details of prior specification}

\begin{itemize}
\item \textbf{Period of cross-immunity} ($T_{\delta}$): Reich \emph{et al.} \cite{Reich2013} estimate a mean of 1.88 years (97 weeks) with a 95\% confidence interval of (0.88, 4.31).
We specify a gamma prior on the multiplicative scale to have a mean of one and 0.01 and 0.99 quantiles at roughly $0.88/1.88 = 0.47$ and $4.31/1.88 = 2.3$,, respectively:
\begin{equation}
T_{\delta} \sim \text{Gamma}(10, 10)
\end{equation}
where $\delta = (97 T_\delta)^{-1}$.
Note that we use a slightly more conservative interval than Reich \emph{et al.} to avoid numerical instability at extreme values.
\\
\item \textbf{Decay rate of infectiousness} ($\gamma$): Nguyet \emph{et al.} \cite{Nguyet2013} find that there is very little transmission of dengue from humans to mosquitoes after 5-6 days since the onset of fever (with roughly an extra of infectiousness prior to onset of fever).
In reality, an individual's infectiousness decays with time, but the ODE model assumes infectiousness is constant for the (exponentially distributed) duration of the infection period.    
To parameterize the model then, we assume a rough equivalence between the probability of infection from a single individual at a given time and the proportion of individuals still infectious at a given time.
Thus we choose our prior for $\gamma$ to keep the time at which only 5\% of individuals are still infectious between 5 and 7 days, with a mean 5\% threshold of 6 days.
Again we specify the prior on the multiplicative scale to have a mean of one and 0.05 and 0.95 quantiles at $2.99/3.5 = 0.85$ and $4.19 / 3.5 = 1.2$, respectively:
\begin{equation}
\gamma' \sim \text{Gamma}(100, 100)
\end{equation}
where $\gamma = 3.5\gamma'$.
\\
\item \textbf{Intrinsic incubation period} ($T_{\rho}$): Chan and Johansson \cite{Chan2012} estimate a mean of 6.1 days with a 95\% credible interval of (3, 10).
Thus, we specify a gamma prior on the multiplicative scale to have a mean of one, and 0.05 and 0.95 quantiles of $3 / 6.1 = 0.49$ and $10 / 6.1 = 1.6$, respectively:
\begin{equation}
T_{\rho} \sim \text{Gamma}(8.3, 8.3)
\end{equation}
where $\rho = (0.87 T_{\rho})^{-1}$.
\\
\item \textbf{Mosquito mortality rate} ($d_v$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $d_v$.  
We use nonlinear least squares to fit the exponential cdf with rate $d_v$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d_v = 1.47$.
We use the same procedure to estimate $d_v$ for the 0.25 and 0.75 quantiles of the estimated mortality curves (assuming symmetry around the mean), which gives 1.6 and 1.37.
To specify the prior, we again scale $d_v'$ to have a mean of one, and 0.25 and 0.75 quantiles of $1.37 / 1.47 = 0.93$ and $1.6 / 1.47 = 1.08$, respectively:
\begin{equation}
d_v' \sim \text{Gamma}(100, 100)
\end{equation}
where $d_v = 1.47d_v'$.
\end{itemize}

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 HMC Diagnostics.}
\label{S1_Diag}
{\bf Trace plots and convergence tests.}  .

\section*{Acknowledgments}

Thanks!

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
%

\bibliographystyle{plos2015}
\bibliography{dengue}

\end{document}

### 2017-03-02 Manuscript/ms.tex
Description of priors and fixed parameters

@@
 
-\subsection*{Additional analyses}
+\begin{itemize}
+\item Human movement \cite{Adams2009, Cosner2009a, Stoddard2009, Dalziel2013}
+\item Disease risk possibly driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Canyon1999}.
 
-\begin{itemize}
-  \item Dig into seasonal trends in fit
-  \item Look at how mosquito predictions change as data are added
 \end{itemize}
 
-\subsection*{Unmodeled processes}
-
-Human movement \cite{Adams2009, Cosner2009a, Stoddard2009, Dalziel2013}
-\begin{itemize}
-  \item Local mosquito abundance does not necessarily predict local disease risk
-  \begin{itemize}
-    \item human movement creates complex host-vector mixing structure (depends on vector ecology and activity periods)
-    \item mosquitoes may vary in their contribution to human disease burden (this effect size is likely fairly small and the model doesn't (currently) provide a way to pull this out)
-  \end{itemize}
-  \item Need to account for human movement in order to prioritize spatial application of control
-\end{itemize}
 \section*{Conclusions}
@@
 \begin{align}
-y_t & \sim \text{NegBin}(\phi_y (X_t - X_{t-1}), \eta_y)\\
-q_t & \sim \text{NegBin}(\phi_q \tau_t V_{Nt}, \eta_q)\\
+y_t & \sim \text{NegBin}(\phi_y  (X_t - X_{t-1})N, \eta_y)\\
+q_t & \sim \text{NegBin}(\phi_q \tau_t V_{Nt}N, \eta_q)\\
 \\ 
 \frac{dX}{dt} &= \rho E \\
-\frac{dS}{dt} &= bN - bS - \alpha \frac{V_{I}}{N} S + \delta R\\
-\frac{dE}{dt} &= \alpha \frac{V_{I}}{N} S - (\rho + b)E\\
+\frac{dS}{dt} &= b - bS - \alpha V_{I} S + \delta R\\
+\frac{dE}{dt} &= \alpha V_{I}S - (\rho + b)E\\
 \frac{dI}{dt} &= \rho E - (\gamma + b)I\\
@@
 V_{St} & = V_{Nt} - V_E(t) - V_I(t)\\
-\frac{dV_E}{dt} &= \alpha \frac{I}{N} V_S - (\rho_{vt} + d_v)V_E\\
+\frac{dV_E}{dt} &= \alpha I V_S - (\rho_{vt} + d_v)V_E\\
 \frac{dV_I}{dt} &= \rho_{vt} V_E - d_v V_I\\
 \\
-\log(V_{N0}) & \sim \text{Normal}(13, 2)\\
-\log(V_{Nt}) & \sim \text{Normal}(\log(V_{Nt-1}), \sigma^2)\\
-\sigma & \sim \text{Half-normal}(0, 1)\\
-\phi_q & \sim \text{Half-normal}(-16, 2.5)\\
+\log(V_{Nt}) & \sim \text{Normal}(\beta \log(V_{Nt-1}), \sigma^2)\\
+\log(V_{N0}) & \sim \text{Normal}(0.7, 2)\\
+\beta & \sim \text{Normal}(0, 0.4)\\
+-\log(\phi_q) & \sim \text{Half-normal}(13, 0.5)\\
 \eta_y &\sim \text{Half-normal}(0, 5)\\
-\eta_q & \sim \text{Half-cauchy}(0, 5)\\
-\alpha_{rel} & \sim \text{Half-normal}(1, 0.3)\\
-\rho_{rel} & \sim \text{Half-normal}(1, 0.3)\\
-\gamma_{rel} & \sim \text{Half-normal}(1, 0.3)\\
-\delta_{rel} & \sim \text{Half-normal}(1, 0.3)\\
-d_{v,rel} & \sim \text{Half-normal}(1, 0.3)\\
-\mathbf{\pi_{0}} & \sim \text{Normal}(\mathbf{\mu}, 0.09\mathbf{I})\\
-\mathbf{p_0} & = \frac{\exp(\mathbf{\pi_0})}{\sum \exp(\mathbf{\pi_0})} \\
+\eta_q & \sim \text{Half-normal}(0, 5)\\
+T_{\rho} & \sim \text{Gamma}(8.3, 8.3)\\
+\rho & = (0.87 T_{\rho})^{-1}\\
+T_{\delta} & \sim \text{Gamma}(10, 10)\\
+\delta & = (97 T_{\delta})^{-1}\\
+\gamma' & \sim \text{Gamma}(100, 100)\\
+\gamma & = 3.5\gamma'\\
+d_{v}' & \sim \text{Gamma}(100, 100)\\
+d_v & = 1.47 d_v'\\
+S_0' & \sim \text{Normal}(-0.4, 0.2)\\
+E_0' & \sim \text{Normal}(-9, 0.6)\\
+I_0' & \sim \text{Normal}(-9, 0.6)\\
+R_0' & = 0\\
+S_0 & = \frac{\exp(S_0)}{\exp(S_0' + E_0' + I_0' + R_0')} \\
 \end{align}
@@
 \\
+\item \textbf{Intrinsic incubation period} ($T_{\rho}$): Chan and Johansson \cite{Chan2012} estimate a mean of 6.1 days with a 95\% credible interval of (3, 10).
+Thus, we specify a gamma prior on the multiplicative scale to have a mean of one, and 0.05 and 0.95 quantiles of $3 / 6.1 = 0.49$ and $10 / 6.1 = 1.6$, respectively:
+\begin{equation}
+T_{\rho} \sim \text{Gamma}(8.3, 8.3)
+\end{equation}
+where $\rho = (0.87 T_{\rho})^{-1}$.
+\\
 \item \textbf{Decay rate of infectiousness} ($\gamma$): Nguyet \emph{et al.} \cite{Nguyet2013} find that there is very little transmission of dengue from humans to mosquitoes after 5-6 days since the onset of fever (with roughly an extra of infectiousness prior to onset of fever).
@@
 \\
-\item \textbf{Intrinsic incubation period} ($T_{\rho}$): Chan and Johansson \cite{Chan2012} estimate a mean of 6.1 days with a 95\% credible interval of (3, 10).
-Thus, we specify a gamma prior on the multiplicative scale to have a mean of one, and 0.05 and 0.95 quantiles of $3 / 6.1 = 0.49$ and $10 / 6.1 = 1.6$, respectively:
-\begin{equation}
-T_{\rho} \sim \text{Gamma}(8.3, 8.3)
-\end{equation}
-where $\rho = (0.87 T_{\rho})^{-1}$.
-\\
 \item \textbf{Mosquito mortality rate} ($d_v$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
@@
 where $d_v = 1.47d_v'$.
+\\
+\item \textbf{Initial conditions} ($S_0, E_0, I_0, R_0$): The SEIRS model assumes a constant population size and is scaled so that $S_0+E_0+I_0+R_0 = 1$.
+We specify our prior on an unconstrained scale and transform to the simplex with the softmax transformation.
+To maintain identifiability of this transformation, we fix $R_0' = 0$.
+We choose the prior for $S_0'$  so that $S_0$ has a mean of roughly  0.2, with the mass concentrated between 0.2 and 0.6.
+This range produces dynamics with recurring annual outbreaks and avoids the extremes where the disease either does not spread or spreads far too quickly and then dies out.
+This is also roughly consistent with \cite{Cardoso2011a} who report notified cases for Vitoria from 1995 to 2008. 
+They report that the 2008 outbreak (where our data begins) was driven by serotypes 2 and 3, and 2009 by serotype 2 only, with 13590 reported cases in previous serotype 2 years.
+Assuming 12 unreported cases for every reported case, 163,080 total cases, or about 50\% of the population immune to serotype 2 at the start of 2008.
+We choose the priors for $E_0'$ and $I_0'$ so that the initial sizes of these compartments were less than 100 individuals, but with a mode slightly off from zero.
+This was again chosen primarily to ensure sustained multi-annual transmission.
+\\
+\item \textbf{Mosquito population size and capture probability} ($\log(V_{N0}), \phi_q$): Wearing and Rohani \cite{Wearing2006} assume 2 mosquitoes per person in their model of dengue transmission, so we use that to inform our prior on $\log(V_{N0})$.
+Since $V_{N}$ has been scaled by the human population size, we center the prior for $\log(V_{N0})$ at $\log(2) = 0.7$ and set the variance to concentrate mass between 0.5 and 5 mosquitoes per person.
+Maybe see Focks (2000) for further justification of range of mosquitoes/person.
+The per-trap mosquito capture probability, $\phi_q$ follows $V_{N0}$ closely and scales the estimated mosquito abundance down to magnitude of the trap counts.
+Since $ E[q_t] = \phi_q \tau V_{Nt} N$, we set the prior on $\log(\phi_q)$ to have a mean of $\log(q_1) - \log(\tau_1) - \log(V_{N0}) - \log(N) = -13$ and a variance to match the prior on $\log(V_{N0})$.
+\\
+\item \textbf{Overdispersion parameters} ($\eta_y, \eta_q$): The negative binomial likelihoods were parameterized as:
+\begin{equation}
+\text{NegBin}(y | \mu, \eta) = {y + 1/\eta -1 \choose y}\left(\frac{\mu\eta}{1 + \mu\eta}\right)^y\left(\frac{1}{1+\eta\mu}\right)^{1/\eta}
+\end{equation}
+such that
+\begin{align}
+E[y] & = \mu\\
+\text{Var}[y] &= \mu + \eta\mu^2
+\end{align}
+Thus, $\eta = 0$ corresponds to Poisson, and $\eta > 0$ leads to overdispersion.
+We put a relatively conservative truncated normal prior on both $\eta_y$ and $\eta_q$ that concentrates mass near zero but allows for larger values.
+\\
+\item \textbf{Autoregressive parameter} ($\beta$): Though the autoregressive framework requires that $|\beta| < 1$, for computational efficiency, we avoid hard bounds and instead use a normal prior centered at zero, with variance such that the majority of mass falls between -1 and 1.
+\end{itemize}
+
+\subsection*{Fixed parameters}
+
+\begin{itemize}
+\item \textbf{Transmission rate} ($\alpha$): The transmission rate only appears multiplied by a term from the mosquito model (either $V_I$ or $V_S$) and is thus effectively non-identifiable when we are also estimating the overall size of the mosquito population.
+Because of this, we fix $\alpha$ at a reasonable value from the literature (though our results should not be sensitive to the exact value used).
+The transmission rate is the product of the mosquito feeding rate and the probability of transmission given feeding.
+Scott \textit{et al.} \cite{Scott2000} estimate that \textit{Aedes aegypti} females take 0.63 blood meals per day in Puerto Rico, and 0.76 blood meals per day in Thailand.
+We take the average of these and set the biting rate at 0.69 per day, or 4.87 per week.
+There are relatively few reliable estimates of the transmission probability of DENV, but Lambrechts \textit{et al.} \cite{Lambrechts2011} estimate transmission probabilities close to one  for several related flaviviruses at 25 degrees C (roughly the mean temperature for Vitoria).
+\\
+\item \textbf{Reporting probability} ($\phi_y$): Silva \textit{et al.} \cite{Silva2016} estimate that there are 12 cases of dengue for every case reported to Brazil's Notifiable Diseases Information System (SINAN, the source of our data), so we fix $\phi_y = 1/12$.
+\\
+\item \textbf{Variance of mosquito process} ($\sigma$): We found that trying to estimate $\sigma$ introduces substantial computational difficulties, so we choose to fix $\sigma = 0.2$ which allows for maximum weekly fluctuations of roughly 50\%.
+This is sufficiently large to allow enough flexibility to fit both the case and mosquito data while guarding against unreasonably large fluctuations in modeled mosquito abundance.
 \end{itemize}
@@
 \label{S1_Diag}
-{\bf Trace plots and convergence tests.}  .
+{\bf Trace plots and convergence tests.}
 

### 2017-03-20 Manuscript/ms.tex
Updates to methods
Updating parameter table and reorganizing sections

@@
 
-General questions:
-\begin{itemize}
-\item Do we need to write out the explicit factorization of the posterior?
-\end{itemize}
 \subsection*{Study system and data}
 
-Vitoria is a coastal city and the capital of the state of Espirito Santo, Brazil, with a population of 304,331 (check) in the city proper, and an additional 1.5 million people in the greater metropolitan area.
+Vitoria is a coastal city and the capital of the state of Espirito Santo, Brazil, with a population of 327,801 in the city proper, and an additional 1.5 million people in the greater metropolitan area.
 The city has an area of approximately 93 sq. km., with most of that area taken up by a large island surrounded by the river Rio Santa Maria and bay Baia de Vitoria.
-There is also a large, mountainous natural area that occupies a large portion of the interior of the island.
 
@@
 The strength and duration of these different inter-serotype interactions are not well understood, though different models suggest that temporary cross-immunity alone (without ADE) is sufficient to reproduce observed multi-annual dynamics in Thailand \cite{Wearing2006,Reich2013}.
-Similarly, \cite{Aguiar2013} find that capturing both primary and secondary infections and the period of cross-immunity is crticial, but that explicitly including all four serotypes (at great cost to model complexity) does not perform much better than a two serotype model, and indeed that the number of hospital admissions caused by a third or fourth infection is very low \cite{Aguiar2011a}.
+Similarly, \cite{Aguiar2013} find that capturing both primary and secondary infections and the period of cross-immunity is critical, but that explicitly including all four serotypes (at great cost to model complexity) does not perform much better than a two serotype model, and indeed that the number of hospital admissions caused by a third or fourth infection is very low \cite{Aguiar2011a}.
 
@@
 Following a latent period ($\frac{1}{\rho}$), exposed individuals become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
-Infectious individuals recover at rate $\gamma$ and remain cross-immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
+Infectious individuals recover at rate $\gamma$ and remain immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
 
-Similarly, susceptible mosquitoes become exposed by biting infectious humans and pass through a temperature-dependent latent period ($\frac{1}{\rho_{v}}$) before becoming infectious.
-Total mosquito abundance, and thus the size of the mosquito susceptible pool, is forced by a latent process model informed by the mosquito surveillance data (see below).
+Similarly, susceptible mosquitoes ($V_S$) become exposed by biting infectious humans and pass through a temperature-dependent latent period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
 
@@
 
-\subsection*{Data model}
-
-This mechanistic differential equation model is then embedded in a Bayesian statistical model in order to estimate the latent mosquito abundance time series and generate sample mosquito trap counts to compare with our observations.
-To connect the differential equation model the the observed case reports, we add an extra state, $X$, that collects the cumulative number of transitions from the exposed to infectious class (assuming that case reporting coincides with the onset of symptoms).
-We then model the number of new cases reported in week $t$ ($y_t$) as:
+The unobserved total mosquito abundance $V_N$, and thus the size of the available susceptible pool, is modeled as an AR(1) process on the log-scale:
 \begin{equation}
-y_t  \sim \text{NegBin}(\phi_y (X_t - X_{t-1}), \eta_y),
+\log(V_{Nt}) \sim \text{Normal}(\beta \log(V_{Nt-1}), \sigma).
 \end{equation}
-where $\phi_y$ is the reporting probability, and $\eta_y$ controls the overdispersion relative to the Poisson.
-
-The unobserved total mosquito abundance is modeled as a random walk on the log-scale:
-\begin{equation}
-\log(V_{Nt}) \sim \text{Normal}(\log(V_{Nt-1}), \sigma)
-\end{equation}
-This is then connected to the number of mosquitoes trapped in week $t$ ($q_t$) with a negative binomial measurement model:
-\begin{equation}
-q_t \sim \text{NegBin}(\phi_q \tau_t V_{Nt}, \eta_q),
-\end{equation}
-where $\phi_q$ is the per-trap capture probability, $\tau_t$ is the number of traps inspected in week $t$, and $\eta_q$ controls overdispersion relative to the Poisson.
 
@@
 there's essentially a best-fit trajectory that we'll always be able to find), while the epidemiological parameters will facilitate better fits to the mosquito data (while maintaining the same quality of fit at the case level).
+\subsection*{Data model}
+
+To connect the differential equation model the the observed case reports, we add an extra state, $X$, that collects the cumulative number of transitions from the exposed to infectious class (assuming that case reporting coincides with the onset of symptoms).
+We then model the number of new cases reported in week $t$ ($y_t$) as:
+\begin{equation}
+y_t  \sim \text{NegBin}(\phi_y (X_t - X_{t-1}), \eta_y),
+\end{equation}
+where $\phi_y$ is the reporting probability, $X_t - X_{t-1}$ is the number of new infectious individuals in week $t$, and $\eta_y$ controls the overdispersion relative to the Poisson.
+
+We also model the number of mosquitoes trapped in week $t$ ($q_t$) as:
+\begin{equation}
+q_t \sim \text{NegBin}(\phi_q \tau_t V_{Nt}, \eta_q),
+\end{equation}
+where $\phi_q$ is the per-trap capture probability, $\tau_t$ is the number of traps inspected in week $t$, and $\eta_q$ controls overdispersion relative to the Poisson.
+
+\subsection*{Posterior predictive checks}
 
@@
 
-Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito (i.e., the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
-To account for this, we make $\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground (URL/citation).
-We use the temperature relationship from the recent meta-analysis of \cite{Chan2012}, which incorporates a large number of records and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
-We use a slightly more conservative fitted relationship from a log-normal model excludes right-censored data and produces a less extreme temperature dependence:
-\begin{align}
-\frac{1}{\rho_v} &= \exp \left(\mu + \frac{1}{14}\right)\\
-\mu & = \exp(1.9 - 0.04 T)\\
-\end{align}
-
-This relationship is similar to those reported in \cite{Focks1995} and \cite{Tjaden2013}, though there is substantial uncertainty about the exact shape of this relationship.
-
-\begin{table}[h]
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} 
 \begin{center}
-\caption{Fixed parameter values, taken from the literature.}
+\caption{Parameters.}
 \begin{tabular}{llll}
@@
 \hline
-$1/d = 1/b$ & Human life-span & 76 years & City of Vitoria\\
-$1/d_V = 1/b_V$ & Vector life-span & 5 days & \cite{Brady2013} \\
-$1/\rho$ & Latent period in host & 6.1 days & \cite{Chan2012}\\
-$1/\gamma$ & Infectious period in host & 3 days & \cite{Nguyet2013}\\
-$1/\delta$ & Period of cross-immunity & 686 days &  \cite{Reich2013}\\
+$N$ & Population size & 327801 & Vitoria Census\\
+$1/d$ & Human life-span & 76 years & Vitoria Census\\
+$\alpha$ & Transmission rate & 4.87 week$^{-1}$ & \cite{Scott2000}\\
+$\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
+$1/\rho_{vt}$ & Extrinsic incubation period & $\exp \left(\exp(1.9 - 0.04 T_t) + \frac{1}{14}\right)$ & \cite{Chan2012}\\
+$\sigma$  & Standard dev. of AR(1) process & 0.2 & \\
+\hline
+$d_v$ & Vector death rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
+$1/\rho$ & Latent period in host & 0.87 weeks  & \cite{Chan2012}\\
+$\gamma$ & Rate of loss of infectiousness & 3.5 week$^{-1}$ & \cite{Nguyet2013}\\
+$1/\delta$ & Period of cross-immunity & 97 weeks &  \cite{Reich2013}\\
+$\phi_q$ & Per-trap mosquito capture probability & $2 \times 10^{-6}$ & 
 \end{tabular}
@@
 \label{fixedparms}
+\end{adjustwidth}
 \end{table}
@@
 Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.
-
-\subsection*{Posterior predictive checks}
 
@@
 To maintain identifiability of this transformation, we fix $R_0' = 0$.
-We choose the prior for $S_0'$  so that $S_0$ has a mean of roughly  0.2, with the mass concentrated between 0.2 and 0.6.
+We choose the prior for $S_0'$  so that $S_0$ has a mean of roughly  0.4, with the mass concentrated between 0.2 and 0.6.
 This range produces dynamics with recurring annual outbreaks and avoids the extremes where the disease either does not spread or spreads far too quickly and then dies out.
@@
 This is sufficiently large to allow enough flexibility to fit both the case and mosquito data while guarding against unreasonably large fluctuations in modeled mosquito abundance.
+\\
+\item \textbf{Extrinsic incubation period} ($1/\rho_{vt}$): Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito  (the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
+To account for this, we make $1/\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground (URL/citation).
+We use the temperature relationship from the recent meta-analysis of \cite{Chan2012}, which incorporates a large number of records and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
+We use a slightly more conservative fitted relationship from a log-normal model that excludes right-censored data and produces a less extreme temperature dependence:
+\begin{align}
+\frac{1}{\rho_v} &= \exp \left(\mu + \frac{1}{14}\right)\\
+\mu & = \exp(1.9 - 0.04 T)
+\end{align}
+This relationship is similar to those reported in \cite{Focks1995} and \cite{Tjaden2013}, though there is uncertainty about the exact shape of this relationship.
 \end{itemize}

### 2017-03-20 Analysis/citystan.r
Adding code

@@ -1,0 +1,80 @@
+library(rstan)
+library(plyr)
+library(magrittr)
+
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+#===============================================================================
+# Loading data and setting up model
+  
+tseries <- read.csv("Data/Vitoria.data.csv")
+obs <- tapply(tseries$Cases, list(tseries$tot.week), sum, na.rm=TRUE)
+
+# Population size
+pop <- 327801
+
+# Pull out mosquito counts and traps inspected
+q <- tapply(tseries$Mosquitoes, list(tseries$tot.week), sum, na.rm = TRUE)
+tau <- tapply(tseries$Trap, list(tseries$tot.week), sum, na.rm = TRUE)
+
+# Computing eip forcing
+weather <- read.csv("Data/Vitoria.weather.csv")
+temp <- weather$Mean.TemperatureC
+
+week <- rep(c(1:243), each = 7)
+meantemp <- tapply(temp[1:1701], week, mean, na.rm = T)
+
+rov <- 7 / exp(exp(1.9 - 0.04 * meantemp) + 1 / (2 * 7))
+
+#===============================================================================
+# Running stan
+
+dat.stan <- list(T = 243,
+                 y = obs,
+                 q = q,
+                 tau = tau,
+                 rov = rov,
+                 pop = pop)
+
+inits = list(list(p0_raw = c(-0.4, -9, -9),
+                  log_phi_q = -13,
+                  eta_inv_y = 5,
+                  eta_inv_q = 5,
+                  ro_c = 1,
+                  gamma_c = 1,
+                  dv_c = 1,
+                  delta_c = 1,
+                  logNv0 = 0.7,
+                  beta = 1,
+                  epsilon = rep(0, 242)),
+             list(p0 = c(-0.2, -8, -8),
+                  log_phi_q = -12,
+                  eta_inv_y = 1,
+                  eta_inv_q = 10,
+                  ro_c = 1,
+                  gamma_c = 1,
+                  dv_c = 1,
+                  delta_c = 1,
+                  logNv0 = 0,
+                  beta = 0.5,
+                  epsilon = rep(0, 242)),
+             list(p0_raw = c(-0.5, -7, -7),
+                  log_phi_q = -14,
+                  eta_inv_y = 0.1,
+                  eta_inv_q = 0.1,
+                  ro_c = 1,
+                  gamma_c = 1,
+                  dv_c = 1,
[…]

### 2017-04-06 Manuscript/ms.tex
Cleaning up introduction

@@
 
-\subsection*{Dengue background}
+\subsection*{Dengue background and prevention}
 \begin{itemize}
   \item large public health burden, with millions cases/year in Brazil
-  \item Complex dynamics -- vector-borne (with resulting seasonal forcing), multi-serotype (cross-immunity, ADE)
+  \item spread is prevented primarily through mosquito control -- even once a vaccine is developed dengue management plans will likely still include vector control \cite{Achee2015}
+  \item Targeted efforts do reduce disease burden 
+  \begin{itemize}
+    \item contact tracing and residential spraying \cite{Vazquez-Prokopec2017}
+  	\item ecovec and "MI-Dengue" system -- trap monitoring and spatial prioritization of source reduction and larvicide application \cite{Eiras2009, Pepin2013}
   \end{itemize}
-
-\subsection*{Dengue prevention}
-\begin{itemize}
-  \item Mosquito surveillance and control -- even once a vaccine is developed dengue management plans will likely still include vector control (\cite{Achee2015})
-  \item General efforts -- education, source reduction, etc; National Dengue Control system
-  \item Targeted efforts -- ecovec and "MI-Dengue" system -- trap monitoring and spatial prioritization of source reduction and larvicide application (\cite{Eiras2009}).
-  \begin{itemize}
-  \item Based on connection between mosquito abundance and disease risk
-  \item \href{http://www.vitoria.es.gov.br/prefeitura/mosquito-controle-ajuda-a-reduzir-incidencia-de-viroses}{general mosquito control} (City of Vitoria)
-  \item \href{http://www.vitoria.es.gov.br/prefeitura/combateadengue}{fight against dengue} (City of Vitoria)
-  \end{itemize}
+  \item Using mosquito surveillance to efficiently target control efforts requires understanding the relationship between mosquito abundance and disease risk
 \end{itemize}
@@
 
-Previous efforts to use MosquiTRAP surveillance to predict human dengue cases in Vitoria relied on linear statistical models and asked whether models that included surveillance data performed better than models with lagged case data alone \cite{Pepin2015}.
-This procedure essentially tests for Granger causality \cite{Granger1969}, which states that some variable $X$ \emph{causes} $Y$ if $X$ contains unique information about $Y$ (i.e.
+Even though mosquitoes are necessary for dengue transmission, establishing a clear link between mosquito abundance (as measured through various surveillance-based indices) and disease risk has proven difficult thus far (CITATIONS).
+Many of the attempts to establish this link have relied on tests of the ability of mosquito abundance measures to predict human disease risk in regression models.
+For example, \cite{Pepin2015} found that models that included MosquiTRAP surveillance data failed to predict weekly dengue cases any better than autoregressive models that include lagged case data alone.
+Evaluating the influence of mosquito abundance in this way is analogous to a test of Granger causality \cite{Granger1969}, which states that some variable $X$ (e.g.
+mosquito abundance) \emph{causes} $Y$ (e.g.
+human dengue cases) if $X$ contains unique information about $Y$ (i.e.
 if a model with $X$ and $Y$ provides better predictions of $Y$ than a model with $Y$ alone).
+By this definition, the results of \cite{Pepin2015} suggest that mosquito abundance does not cause dengue cases (in the Granger sense).
+
 However, as noted by Sugihara \emph{et al.} \cite{Sugihara2012a}, Granger causality assumes that $X$ and $Y$ are separable, which is not the case if $X$ and $Y$ are linked in a non-linear dynamical system.
-In this case, a model with $X$ and $Y$ is unlikely to outperform the model with only $Y$, since $Y$ already contains the information from $X$, a consequence of Taken's theorem (CITATION). 
-Instead, Sugihara \emph{et al.} argue that if $X$ drives $Y$, then you should be able to reconstruct $X$ from $Y$.
+In this case, a model with $X$ and $Y$ will not outperform the model with only $Y$, as the information contained in $X$ is already embedded in $Y$, a consequence of Taken's theorem (CITATION). 
+Since dengue transmission is governed by such a non-linear dynamical system, information about the driving component of mosquito abundance should be already embedded in the time series of human case reports.
+As a result, predictive utility is not an appropriate way to test the causal influence of mosquito abundance on dengue risk.
+Instead, Sugihara \emph{et al.} argue that if $X$ is a driver $Y$, then you should be able to reconstruct $X$ from $Y$.
 
-Applying this reasoning to the problem of establishing the relationship between mosquito abundance and human dengue risk, it is not surprising that including information from mosquito surveillance does not improve our ability to predict human dengue cases, as information about epidemiologically relevant mosquito abundance is already embedded in the time series of human case reports.
-Following from \cite{Sugihara2012a} then, we ask whether we can reconstruct the mosquito surveillance time series from the time series of human case reports. 
-We do this through the use of a mechanistic differential equation model embedded in a Bayesian statistical framework.
-This differs from the equation-free approach of \cite{Sugihara2012a}, but the explicit use of a mechanistic model will better inform our understanding of the processes linking mosquito abundance and human disease risk and the embedding in a Bayesian statistical model will better account for uncertainty and the measurement processes in these data.
+In this paper, we apply this perspective to the problem of linking mosquito abundance to disease risk.
+To test whether or not mosquito abundance, as measured through surveillance efforts, drives dengue dynamics, we ask whether we can reconstruct a time series of mosquito surveillance data from a time series of case reports.
+We approach this as an inverse problem: we first posit a mechanistic model of dengue transmission and then develop a Bayesian statistical framework to estimate the mosquito abundance time series that is consistent with that model and the time series of case reports.
+We then evaluate the causal link between mosquito abundance and case reports by using posterior predictive checks to compare the estimated and observed mosquito time series.
 
@@
 
-The flexibility of the mosquito process model means that we should be able to essentially fit the case data arbitrarily well.
-The question is then whether or not the mosquito abundance time series estimated from the case data reproduces the observed mosquito surveillance time series.
-Since the initial conditions and epidemiological parameters potentially affect the mapping of the case data to mosquito abundance, we estimate these as well.
-The intention here is that the mosquito process will allow us to fit to the case data (i.e.
-there's essentially a best-fit trajectory that we'll always be able to find), while the epidemiological parameters will facilitate better fits to the mosquito data (while maintaining the same quality of fit at the case level).
 \subsection*{Data model}
@@
 
+\begin{figure}[!h]
+\caption{{\bf Vitoria data and model fits.}
+Weekly observations from 2008 to week 34 of 2012 (points), with corresponding median posterior prediction (line) and 90\% credible inverval (gray band).
+A: reported dengue cases.
+B: total number of trapped mosquiotes.
+}
+\label{timeseries}
+\end{figure}
 \section*{Discussion}

### 2017-04-06 Manuscript/ms.tex
Adding description of posteror predictive checks

@@
 
-\subsection*{Posterior predictive checks}
-
 \subsection*{Parameterization and priors}
@@
 
-The posterior distribution can be written as:
+Samples from the posterior distribution were generated using Hamiltonian Monte Carlo (CITATIONS) implemented in the rstan package for R. 
+We ran 3 chains with different starting values for 10,000 iterations each, discarding the first 5,000 as burn-in.
+Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.
+
+\subsection*{Posterior predictive checks}
+
+We evaluate the goodness-of-fit of the estimated mosquito abundance time series through the use of posterior predictive checks \cite{Gelman1996,Gelman2003}.
+Specifically, for each iteration, $i$, of MCMC, we sample
 \begin{equation}
-aaaaaaah
+q^{rep(i)}_t \sim \text{NegBin}(\phi^{(i)}_q \tau_t V^{(i)}_{Nt}, \eta^{(i)}_q)
 \end{equation}
-Samples from this distribution were generated using Hamiltonian Monte Carlo (CITATIONS) implemented in the rstan package for R. 
-Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.
+and compute
+\begin{equation}
+p_t = P(q^{rep}_t > q_t) = \frac{1}{M}\sum I(q^{rep(i)}_t > q_t).
+\end{equation}
+If the reconstructed mosquito trajectories are a good fit for the observed mosquito trap counts, then we would expect the time series $p_t$ to be random in time (CITATION?).
+Any systematic trends in $p_t$ indicate lack-of-fit.
+
+In addition to looking at weekly comparisons, we also develop posterior checks to compare the temporal variability and correlation structures of the observed and expected time series.
+To test whether the estimated time series capture the seasonal features of the observed mosquito data, we compute the posterior auto-correlation function of the replicate data:
+\begin{equation}
+R^{rep(i)}(l) = \text{corr}(q^{rep(i)}_t, q^{rep(i)}_{t-l}),
+\end{equation}
+for lags ranging from 1 to 55 weeks.
+We also compute posterior distributions of $\min(q^{rep}_t)$ , $\max(q^{rep}_t)$, and $\sigma(q^{rep}_t - q^{rep}_{t-1})$ as measures of the smoothness of the modeled process.
 

### 2017-04-07 Manuscript/ms.tex
Text for parameter section
And adding initial conditions to the table

@@
 
+Several of the parameters in this model are assumed to fixed and known. 
+The population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
+To maintain identifiability, the transmission rate and case reporting probability are also fixed at literature values, while the standard deviation of the mosquito process is fixed at a level that allows reasonably flexible behavior.
+Lastly, the extrinsic incubation period in mosquitoes is modeled as a function of weekly mean temperature and forced with data pulled from WeatherUnderground.
+
+The remaining parameters include the epidemiological parameters that control the flows between the compartments of the differential equation model ($d_v$, $\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the autoregressive parameter in the mosquito process model ($\beta$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
+Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table \ref{fixedparms} for means and Supplemental Material for detailed explanations).
+
+For any given values of these parameters, the mosquito process model should be flexible enough to produce a good fit to the case reports data.
+As a result, we expect that these parameters will be informed primarily by the mosquito surveillance data (i.e.
+these parameters add degrees of freedom that let the mosquito process model fit the surveillance data better while maintaining a good fit to the case data).
+
 \begin{table}[!ht]
@@
 \begin{center}
-\caption{Parameters.}
+\caption{Model parameters and their values.  Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the prior mean.}
 \begin{tabular}{llll}
@@
 $\sigma$  & Standard dev. of AR(1) process & 0.2 & \\
+$V_{E0}$ & Initial exposed mosquitoes &  0 & \\
+$V_{I0}$ & Initial infectious mosquitoes & 0 & \\
 \hline
@@
 $1/\delta$ & Period of cross-immunity & 97 weeks &  \cite{Reich2013}\\
+$S_0$ & Proportion initially susceptible & 0.4 & \cite{Cardoso2011a} \\
+$E_0$ & Proportion initially exposed & $8\times 10 ^ {-5}$ & \\
+$I_0$ & Proportion initially infectious & $8\times 10 ^ {-5}$ & \\
 $\phi_q$ & Per-trap mosquito capture probability & $2 \times 10^{-6}$ & 

### 2017-04-10 Manuscript/ms.tex
Preliminary fig captions and plot code

@@
 \begin{table}[!ht]
+\label{fixedparms}
 \begin{adjustwidth}{-2.25in}{0in} 
@@
 \end{center}
-\label{fixedparms}
 \end{adjustwidth}
@@
 We evaluate the goodness-of-fit of the estimated mosquito abundance time series through the use of posterior predictive checks \cite{Gelman1996,Gelman2003}.
-Specifically, for each iteration, $i$, of MCMC, we sample
+Specifically, for each iteration, $i$, of MCMC, we sample replicate mosquito surveillance data:
 \begin{equation}
@@
 
-In addition to looking at weekly comparisons, we also develop posterior checks to compare the temporal variability and correlation structures of the observed and expected time series.
-To test whether the estimated time series capture the seasonal features of the observed mosquito data, we compute the posterior auto-correlation function of the replicate data:
+In addition to looking at weekly comparisons, we also develop posterior checks to compare the temporal variability and correlation structures of the observed and estimated time series.
+To test whether the modeled mosquito time series capture the seasonal features of the observed mosquito data, we compute the posterior auto-correlation function of the replicate data:
 \begin{equation}
@@
 \begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
 \caption{{\bf Vitoria data and model fits.}
@@
 \end{figure}
+
+\begin{figure}[!h]
+\caption{{\bf Weekly posterior predictive p-values.}
+$P(q^{rep}_t > q_t)$ for each week $t$ based on 15000 replicate mosquito surveillance data sets estimated from case reports.
+}
+\label{pseries}
+\end{figure}
+
+\begin{figure}[!h]
+\caption{{\bf Posterior autocorrelation function.}
+Autocorrelation of observed mosquito trap counts (points) and replicate time series estimated from case reports (vertical lines, indicating 90\% posterior credible interval). 
+}
+\label{autocorr}
+\end{figure}
+
 \section*{Discussion}

### 2017-04-10 Analysis/plots.r

@@ -1,0 +1,84 @@
+# Code to generate plots for manuscript
+
+library(rstan)
+library(plyr)
+library(magrittr)
+library(ggplot2)
+
+# Loading and processing data
+tseries <- read.csv("Data/Vitoria.data.csv")
+
+post <- ddply(tseries, .(tot.week), summarise,
+              qobs = sum(Mosquitoes, na.rm = T),
+              yobs = sum(Cases, na.rm = T),
+              tau = sum(Trap, na.rm = T),
+              week = unique(Week),
+              year = unique(Year))
+
+# Loading MCMC results
+sim <- readRDS("Results/mcmc.rds")
+
+#===============================================================================
+# Figure 1
+
+# Computing posterior median and 80% credible interval
+yhat <- rstan::extract(sim, c("y_hat"), permute = F) %>% apply(3, quantile, c(0.1, 0.5, 0.9))
+qhat <- rstan::extract(sim, c("q_hat"), permute = F) %>% apply(3, quantile, c(0.1, 0.5, 0.9))
+
+# Assembling data frame for ggplot
+post$yhat = yhat[2, ]
+post$ymin = yhat[1, ]
+post$ymax = yhat[3, ]
+post$qhat = qhat[2, ]
+post$qmin = qhat[1, ]
+post$qmax = qhat[3, ]
+
+# Plotting
+ggplot(post, aes(tot.week, yobs)) + 
+  geom_ribbon(aes(ymin = ymin, ymax = ymax), fill = "grey70") +
+  geom_point(size = 0.5) + 
+  geom_line(aes(tot.week, yhat)) + 
+  theme_classic() +
+  scale_y_continuous(expand = c(0, 0)) + 
+  scale_x_continuous(expand = c(0, 1)) +
+  ylab("case reports") + 
+  xlab("week") +
+  ggtitle("A") -> fig1.a
+
+ggplot(post, aes(tot.week, qobs)) + 
+  geom_ribbon(aes(ymin = qmin, ymax = qmax), fill = "grey70") +
+  geom_point(size = 0.5) + 
+  geom_line(aes(tot.week, qhat)) + 
+  theme_classic() +
+  scale_y_continuous(expand = c(0.05, 0)) + 
+  scale_x_continuous(expand = c(0, 1)) +
+  ylab("mosquitoes trapped") +
+  xlab("week") +
+  ggtitle("B") -> fig1.b
+
+postscript("Manuscript/figures/fig1.eps",
+     width = 5.2, height = 3,
[…]

### 2017-04-11 Analysis/plots.r
Adding figures 2 and 3.

@@ -19,7 +19,7 @@ post <- ddply(tseries, .(tot.week), summarise,
 sim <- readRDS("Results/mcmc.rds")
 
 #===============================================================================
-# Figure 1
+# Figure 1: Observed and estimated time series
 
 # Computing posterior median and 80% credible interval
 yhat <- rstan::extract(sim, c("y_hat"), permute = F) %>% apply(3, quantile, c(0.1, 0.5, 0.9))
@@ -65,20 +65,64 @@ grid.arrange(fig1.a, fig1.b, ncol = 2)
 dev.off()
 
 #===============================================================================
-# Figure 2
+# Figure 2: weekly posterior predictive p-value for replicate mosquito
+# surveillance data
 
+# Extracting full MCMC chain of replicate data
 mcmc <- rstan::extract(sim, "q_hat", permute = T)[[1]]
 nmcmc <- dim(mcmc)[1]
 
+# Computing P(q^{rep}_t > q_t)
 for(i in 1:243){
   post$p[i] <- sum(mcmc[, i] > post$qobs[i]) / nmcmc
 }
 
+# Plot
+postscript("Manuscript/figures/fig2.eps",
+           width = 4, height = 3,
+           family = "ArialMT")
+
 ggplot(post, aes(tot.week, p)) +
   geom_line() + 
   theme_classic() + 
+  scale_y_continuous(expand = c(0, 0.01)) + 
+  scale_x_continuous(expand = c(0, 1)) +
   xlab("week") + 
   ylab("p")
 
+dev.off()
 #===============================================================================
-# Figure 3
+# Figure 3: observed and posterior autocorrelation functions
+
+# Calculating the observed autocorrelation function
+obsacf <- acf(post$qobs, lag.max = 55, plot = F)$acf
+
+# Calculating the posterior acf function
+acfs <- apply(mcmc, 1, function(x){acf(x, lag.max = 55, plot = F)$acf})
+
+# Median and 80% credible interval
+acfrange <- apply(acfs, 1, quantile, c(0.1, 0.5, 0.9))
+
+# Assembling data frame for ggplot
+acfdf <- data.frame("lag" = c(0:55), 
+                    "obs" = obsacf, 
+                    "med" = acfrange[2, ], 
+                    "min" = acfrange[1, ], 
+                    "max" = acfrange[3, ])
+
+# Plot
+postscript("Manuscript/figures/fig3.eps",
[…]

### 2017-04-11 Manuscript/ms.tex
Adding figures 2 and 3.
Intro motivation text

@@
 
-\subsection*{Dengue background and prevention}
-\begin{itemize}
-  \item large public health burden, with millions cases/year in Brazil
-  \item spread is prevented primarily through mosquito control -- even once a vaccine is developed dengue management plans will likely still include vector control \cite{Achee2015}
-  \item Targeted efforts do reduce disease burden 
-  \begin{itemize}
-    \item contact tracing and residential spraying \cite{Vazquez-Prokopec2017}
-  	\item ecovec and "MI-Dengue" system -- trap monitoring and spatial prioritization of source reduction and larvicide application \cite{Eiras2009, Pepin2013}
-  \end{itemize}
-  \item Using mosquito surveillance to efficiently target control efforts requires understanding the relationship between mosquito abundance and disease risk
-\end{itemize}
+Dengue fever is a massive global public health burden, with millions of cases per year in Brazil alone (CITATION).  
+Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
+Even if a vaccine is developed, mosquito control is likely to remain an important component of dengue management \cite{Achee2015}.
+Studies have shown that targeted mosquito control efforts can be effective in reducing disease burden.
+For example contact tracing and residential spraying reduced disease risk in Cairns, Australia \cite{Vazquez-Prokopec2017}.
+Though this approach was successful, it requires detailed data following individual cases, potentially limiting its utility for large-scale deployment.
 
-\subsection*{Causal inference and mechanistic models}
+Other efforts have sought to guide the deployment of control efforts through the use of fine-scale mosquito surveillance.
+For instance, the "MI-Dengue" system deployed throughout several cities in Brazil uses a grid of mosquito sticky-traps (MosquiTRAPs) for spatial prioritization of source reduction and larvicide application \cite{Eiras2009}.
+Comparing cities with and without this system, \cite{Pepin2013} estimated that it prevented X cases and saved X dollars in economic losses from 20XX - 20XX.
+Though this system has been effective, the mosquito indices and thresholds used to prioritize control efforts remain somewhat ad-hoc.
+Using such surveillance data to efficiently target limited control resources requires a detailed understanding of the relationship between the components of mosquito abundances measured by the trap and disease risk.
 
-Even though mosquitoes are necessary for dengue transmission, establishing a clear link between mosquito abundance (as measured through various surveillance-based indices) and disease risk has proven difficult thus far (CITATIONS).
+However, even though mosquitoes are necessary for dengue transmission, establishing a clear link between surveillance-based indices of mosquito abundance and disease risk has proven difficult thus far (CITATIONS).
 Many of the attempts to establish this link have relied on tests of the ability of mosquito abundance measures to predict human disease risk in regression models.
@@
 \begin{figure}[!h]
+\includegraphics[angle = 270, scale=1]{figures/fig2.eps}
 \caption{{\bf Weekly posterior predictive p-values.}
@@
 \begin{figure}[!h]
+\includegraphics[angle = 270, scale=1]{figures/fig3.eps}
 \caption{{\bf Posterior autocorrelation function.}

### 2017-04-25 Manuscript/ms.tex
Adding text to results section

@@
 
+Questions:
+\begin{itemize}
+	\item To what extent do the case reports inform our estimates of mosquito abundance?
+How do we separate the information coming from the case reports from the information coming from the mosquito data itself?
+	\item Do our estimates of epidemiological parameters differ from their prior (literature) values?
+	\item Do the case reports reproduce the observed mosquito surveillance time series?
+	\item If no, how does it fail?
+\end{itemize}
+
+The model is able to effectively capture the observed weekly case reports for Vitoria (Fig. \ref{timeseries}A), as expected given the degrees of freedom in the mosquito process model.
+In addition, most of the observed mosquito trap counts fall within the 80\% posterior credible interval (Fig. \ref{timeseries}B).
+We expect the posterior distribution of mosquito abundance time series to be informed primarily by the observed case data.
+However, the mapping from the case reports to mosquito abundance is influenced by the parameters of the mechanistic model and deviations of the posterior distribution of these parameters from their prior reflect the influence of the mosquito trap count data.
+Of these parameters, the posterior estimate of mosquito mortality rate (blah) was lower than its prior mean (1.47 week$^{-1}$), while the posterior estimate of the intrinsic incubation period (blah) was larger than its prior mean (0.87 weeks).
+That these two parameters differ from their literature values reflects the fact that they are critical in controlling the interaction between mosquito abundance and human disease risk, as the intrinsic incubation period controls how quickly humans are able to begin transmitting to mosquitoes, and the mortality rate controls for how long mosquitoes are able to remain infectious and transmit to humans.
+
+However, there is a seasonal trend in the weekly posterior p-values for the replicate mosquito surveillance series (Fig.
+\ref{pseries}), indicating a tendency to overestimate the number of mosquitoes trapped during the downswing of the annual epidemic and underestimate the number of mosquitoes trapped during the upswing.
+This seasonal trend indicates a systematic lack-of-fit.
+
+This lack-of-fit is due in large part to the failure of the estimated mosquito trajectories to capture the periodic seasonal structure in the observed mosquito surveillance data (Fig. \ref{autocorr}).
+In particular, the autocorrelation in the posterior replicate mosquito surveillance series decays much faster than in the observed data and remains near zero for lags near a year, whereas the observed data have positive autocorrelation at yearly lags (\ref{autocorr}).
+Moreover, the posterior standard deviations of 1 week differences are substantially larger than the observed (Supplemental?), indicating that the observed time series is much smoother than the posterior replicates.
+These posterior checks indicate that the case data and mechanistic model are not able to reproduce the observed time series of mosquito trap counts.
+
 \begin{figure}[!h]
@@
 \caption{{\bf Vitoria data and model fits.}
-Weekly observations from 2008 to week 34 of 2012 (points), with corresponding median posterior prediction (line) and 90\% credible inverval (gray band).
+Weekly observations from 2008 to week 34 of 2012 (points), with corresponding median posterior prediction (line) and 80\% credible inverval (gray band).
 A: reported dengue cases.
@@
 \begin{figure}[!h]
+%\includegraphics[angle = 270, scale=1]{figures/fig2.eps}
+\caption{{\bf Posterior densities of epidemiological parameters.}
+A: Intrinsic incubation period. B: Period of cross-immunity. C: Decay rate of infectiousness. D: Mosquito mortality rate.
+}
+\label{pseries}
+\end{figure}
+
+\begin{figure}[!h]
 \includegraphics[angle = 270, scale=1]{figures/fig2.eps}
 \caption{{\bf Weekly posterior predictive p-values.}
-$P(q^{rep}_t > q_t)$ for each week $t$ based on 15000 replicate mosquito surveillance data sets estimated from case reports.
+$P(q^{rep}_t > q_t)$ for each week of the year for 2008 through week 34 of 2012.
+Gray lines show loess-smoothed curves of the p-values for each year.
+Based on 15000 replicate mosquito surveillance data sets estimated from case reports.
 }

### 2017-04-25 Analysis/citystan.r
Modeling mosquito death rate in time
And moving model from mosquito abundance to net growth rate

@@ -43,38 +43,42 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   eta_inv_q = 5,
                   ro_c = 1,
                   gamma_c = 1,
-                  dv_c = 1,
                   delta_c = 1,
                   logNv0 = 0.7,
-                  beta = 1,
-                  epsilon = rep(0, 242)),
-             list(p0 = c(-0.2, -8, -8),
-                  log_phi_q = -12,
-                  eta_inv_y = 1,
-                  eta_inv_q = 10,
-                  ro_c = 1,
-                  gamma_c = 1,
-                  dv_c = 1,
-                  delta_c = 1,
-                  logNv0 = 0,
-                  beta = 0.5,
-                  epsilon = rep(0, 242)),
-             list(p0_raw = c(-0.5, -7, -7),
-                  log_phi_q = -14,
-                  eta_inv_y = 0.1,
-                  eta_inv_q = 0.1,
-                  ro_c = 1,
-                  gamma_c = 1,
-                  dv_c = 1,
-                  delta_c = 1,
-                  logNv0 = 2,
-                  beta = 0,
-                  epsilon = rep(0, 242)))
+                  betar = 1,
+                  betad = 1,
+                  sigmar = 0.1,
+                  sigmad = 0.1,
+                  rv = rep(0, 243),
+                  logdv = rep(0.39, 243)))
+
+             # list(p0 = c(-0.2, -8, -8),
+             #      log_phi_q = -12,
+             #      eta_inv_y = 1,
+             #      eta_inv_q = 10,
+             #      ro_c = 1,
+             #      gamma_c = 1,
+             #      dv_c = 1,
+             #      delta_c = 1,
+             #      logNv0 = 0,
+             #      beta = 0.5,
+             #      epsilon = rep(0, 242)),
+             # list(p0_raw = c(-0.5, -7, -7),
+             #      log_phi_q = -14,
+             #      eta_inv_y = 0.1,
+             #      eta_inv_q = 0.1,
+             #      ro_c = 1,
+             #      gamma_c = 1,
+             #      dv_c = 1,
+             #      delta_c = 1,
+             #      logNv0 = 2,
+             #      beta = 0,
+             #      epsilon = rep(0, 242)))
 
 sim <- stan(file = "Code/seirs.stan", 
             data = dat.stan, 
             init = inits, 
-            iter = 10000, 
-            chains = 3,
+            iter = 1000, 
+            chains = 1,
             control = list(adapt_delta = 0.9))
 

### 2017-05-08 Manuscript/ms.tex
Updating text to new model structure
Estimating time series of both mosquito growth rate and death rate.

@@
 Similarly, susceptible mosquitoes ($V_S$) become exposed by biting infectious humans and pass through a temperature-dependent latent period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
+Total mosquito population size ($V_N$) is controlled by a time-varying growth rate, $r_t$, and captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.
 
@@
 \frac{dR}{dt} &= \gamma I - (\delta + b)R\\
-V_{St} & = V_{Nt} - V_E(t) - V_I(t)\\
-\frac{V_E}{dt} &= \alpha \frac{I}{N} V_S - (\rho_{vt} + d_v)V_E\\
-\frac{V_I}{dt} &= \rho_{vt} V_E - d_v V_I
+\frac{dV_N}{dt} & = r_t V_N - \phi_q V_N\\
+\frac{dV_E}{dt} &= \alpha \frac{I}{N} V_S - (\rho_{vt} + d)V_E\\
+\frac{dV_I}{dt} &= \rho_{vt} V_E - d V_I\\
+\frac{dV_C}{dt} & = \phi_q V_N\\
+V_S &= V_N - V_E - V_I
 \end{align}
 
-The unobserved total mosquito abundance $V_N$, and thus the size of the available susceptible pool, is modeled as an AR(1) process on the log-scale:
-\begin{equation}
-\log(V_{Nt}) \sim \text{Normal}(\beta \log(V_{Nt-1}), \sigma).
-\end{equation}
+The mosquito demographic parameters (net grown rate, $r_t$, and death rate, $d_t$) are modeled as an AR(1) process with a periodic regression component to account for unmodeled seasonal forcing (e.g.
+from temperature, precipitation, etc.):
+\begin{align}
+r_t & = \alpha_{1}r_{t-1} + \alpha_2 \sin(2\pi t / 52) + \alpha_3 \cos(2\pi t / 52) + \epsilon_t\\
+\log(d_{t}) & = \beta_{1}\log(d_{t-1}) + \beta_2 \sin(2\pi t / 52) + \beta_3 \cos(2\pi t / 52) + \xi_t,
+\end{align}
+where
+\begin{align}
+\epsilon_t & \sim \text{Normal}(0, \sigma_{\epsilon})\\
+\xi_t & \sim \text{Normal}(0, \sigma_{\xi}).
+\end{align}
 
@@
 
-We also model the number of mosquitoes trapped in week $t$ ($q_t$) as:
+We similarly model the number of mosquitoes trapped in week $t$ ($q_t$) as:
 \begin{equation}
-q_t \sim \text{NegBin}(\phi_q \tau_t V_{Nt}, \eta_q),
+q_t \sim \text{NegBin}(V_{Ct} - V_{Ct-1}, \eta_q),
 \end{equation}
-where $\phi_q$ is the per-trap capture probability, $\tau_t$ is the number of traps inspected in week $t$, and $\eta_q$ controls overdispersion relative to the Poisson.
+where $V_{Ct} - V_{Ct-1}$ is the number of new mosquitoes captured in week $t$, and $\eta_q$ controls overdispersion relative to the Poisson.
 
@@
 The population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
-To maintain identifiability, the transmission rate and case reporting probability are also fixed at literature values, while the standard deviation of the mosquito process is fixed at a level that allows reasonably flexible behavior.
+To maintain identifiability, the transmission rate and case reporting probability are also fixed at literature values.
 Lastly, the extrinsic incubation period in mosquitoes is modeled as a function of weekly mean temperature and forced with data pulled from WeatherUnderground.
 
-The remaining parameters include the epidemiological parameters that control the flows between the compartments of the differential equation model ($d_v$, $\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the autoregressive parameter in the mosquito process model ($\beta$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
+The remaining parameters include the epidemiological parameters that control the flows between the compartments of the differential equation model ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the regression parameters and process noise in the models for mosquito growth and death rates ($\mathbf{\alpha}$, $\mathbf{\beta}$, and $\sigma_{\epsilon}$ and $\sigma_{\xi}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
 Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table \ref{fixedparms} for means and Supplemental Material for detailed explanations).
@@
 \begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
+%\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
 \caption{{\bf Vitoria data and model fits.}
@@
 \begin{figure}[!h]
-\includegraphics[angle = 270, scale=1]{figures/fig2.eps}
+%\includegraphics[angle = 270, scale=1]{figures/fig2.eps}
 \caption{{\bf Weekly posterior predictive p-values.}
@@
 \begin{figure}[!h]
-\includegraphics[angle = 270, scale=1]{figures/fig3.eps}
+%\includegraphics[angle = 270, scale=1]{figures/fig3.eps}
 \caption{{\bf Posterior autocorrelation function.}
@@
 y_t & \sim \text{NegBin}(\phi_y  (X_t - X_{t-1})N, \eta_y)\\
-q_t & \sim \text{NegBin}(\phi_q \tau_t V_{Nt}N, \eta_q)\\
+q_t & \sim \text{NegBin}(V_{Ct} - V_{Ct-1}, \eta_q),
 \\ 
@@
 \\
-V_{St} & = V_{Nt} - V_E(t) - V_I(t)\\
-\frac{dV_E}{dt} &= \alpha I V_S - (\rho_{vt} + d_v)V_E\\
-\frac{dV_I}{dt} &= \rho_{vt} V_E - d_v V_I\\
+\frac{dV_N}{dt} & = r_t V_N - \phi_q V_N\\
+\frac{dV_E}{dt} &= \alpha \frac{I}{N} V_S - (\rho_{vt} + d)V_E\\
+\frac{dV_I}{dt} &= \rho_{vt} V_E - d V_I\\
+\frac{dV_C}{dt} & = \phi_q V_N\\
+V_S &= V_N - V_E - V_I\\
 \\
-\log(V_{Nt}) & \sim \text{Normal}(\beta \log(V_{Nt-1}), \sigma^2)\\
-\log(V_{N0}) & \sim \text{Normal}(0.7, 2)\\
-\beta & \sim \text{Normal}(0, 0.4)\\
+r_t & = \alpha_{1}r_{t-1} + \alpha_2 \sin(2\pi t / 52) + \alpha_3 \cos(2\pi t / 52) + \epsilon_t\\
+\log(d_{t}) & = \beta_{1}\log(d_{t-1}) + \beta_2 \sin(2\pi t / 52) + \beta_3 \cos(2\pi t / 52) + \xi_t\\
+\epsilon_t & \sim \text{Normal}(0, \sigma_{\epsilon})\\
+\xi_t & \sim \text{Normal}(0, \sigma_{\xi})
+\end{align}
+
+\begin{align}
+\mathbf{\beta} & \sim \text{Normal}(0, 0.4)\\
+\mathbf{\alpha} & \sim \text{Normal}(0, 0.4)\\
+\sigma_{\epsilon} & \sim \text{Normal}(0, 0.3)\\
+\sigma_{\xi} & \sim \text{Normal}(0, 0.3)\\
 -\log(\phi_q) & \sim \text{Half-normal}(13, 0.5)\\
@@
 \\
-\item \textbf{Mosquito mortality rate} ($d_v$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
-They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $d_v$.  
-We use nonlinear least squares to fit the exponential cdf with rate $d_v$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d_v = 1.47$.
-We use the same procedure to estimate $d_v$ for the 0.25 and 0.75 quantiles of the estimated mortality curves (assuming symmetry around the mean), which gives 1.6 and 1.37.
-To specify the prior, we again scale $d_v'$ to have a mean of one, and 0.25 and 0.75 quantiles of $1.37 / 1.47 = 0.93$ and $1.6 / 1.47 = 1.08$, respectively:
+\item \textbf{Initial mosquito mortality rate} ($d_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
+They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $d$.  
+We use nonlinear least squares to fit the exponential cdf with rate $d_v$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d = 1.47$.
+We use the same procedure to estimate $d$ for the 0.25 and 0.75 quantiles of the estimated mortality curves (assuming symmetry around the mean), which gives 1.6 and 1.37.
+We then specify the prior on $\log(d)$ to induce the above quantiles on $d$, which gives us:
 \begin{equation}
-d_v' \sim \text{Gamma}(100, 100)
+\log(d) \sim \text{Normal}(0.39, 0.12).
 \end{equation}
-where $d_v = 1.47d_v'$.
+\\
+\item \textbf{Initial mosquito growth rate} ($r_0$):
 \\
@@
 \\
-\item \textbf{Autoregressive parameter} ($\beta$): Though the autoregressive framework requires that $|\beta| < 1$, for computational efficiency, we avoid hard bounds and instead use a normal prior centered at zero, with variance such that the majority of mass falls between -1 and 1.
+\item \textbf{Regression parameters} ($\mathbf{\beta}$, $\mathbf{\alpha}$): Though the autoregressive framework requires that $|\beta| < 1$, for computational efficiency, we avoid hard bounds and instead use a normal prior centered at zero, with variance such that the majority of mass falls between -1 and 1.
+\\
+\item \textbf{Process variances} ($\sigma_{\epsilon}$, $\sigma_{\xi}$):
 \end{itemize}
@@
 \item \textbf{Reporting probability} ($\phi_y$): Silva \textit{et al.} \cite{Silva2016} estimate that there are 12 cases of dengue for every case reported to Brazil's Notifiable Diseases Information System (SINAN, the source of our data), so we fix $\phi_y = 1/12$.
-\\
-\item \textbf{Variance of mosquito process} ($\sigma$): We found that trying to estimate $\sigma$ introduces substantial computational difficulties, so we choose to fix $\sigma = 0.2$ which allows for maximum weekly fluctuations of roughly 50\%.
-This is sufficiently large to allow enough flexibility to fit both the case and mosquito data while guarding against unreasonably large fluctuations in modeled mosquito abundance.
 \\

### 2017-05-08 Analysis/citystan.r
Adding periodic regression to mosq rates

@@ -35,6 +35,9 @@ dat.stan <- list(T = 243,
                  q = q,
                  tau = tau,
                  rov = rov,
+                 sincos = matrix(c(sin(2 * pi * c(1:242) / 52),
+                                   cos(2 * pi * c(1:242) / 52)),
+                                 ncol = 2),
                  pop = pop)
 
 inits = list(list(p0_raw = c(-0.4, -9, -9),
@@ -45,40 +48,51 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   gamma_c = 1,
                   delta_c = 1,
                   logNv0 = 0.7,
-                  betar = 1,
-                  betad = 1,
+                  beta0_r = 1,
+                  beta0_d = 1,
+                  beta_r = c(0, 0),
+                  beta_d = c(0, 0),
                   sigmar = 0.1,
                   sigmad = 0.1,
                   rv = rep(0, 243),
+                  logdv = rep(0.39, 243)),
+             list(p0_raw = c(-0.2, -8, -8),
+                  log_phi_q = -12,
+                  eta_inv_y = 1,
+                  eta_inv_q = 10,
+                  ro_c = 1,
+                  gamma_c = 1,
+                  delta_c = 1,
+                  logNv0 = 1,
+                  beta0_r = 0,
+                  beta0_d = 0,
+                  beta_r = c(0, 0),
+                  beta_d = c(0, 0),
+                  sigmar = 0.5,
+                  sigmad = 0.5,
+                  rv = rep(0, 243),
+                  logdv = rep(0.39, 243)),
+             list(p0_raw = c(-0.5, -7, -7),
+                  log_phi_q = -14,
+                  eta_inv_y = 0.1,
+                  eta_inv_q = 0.1,
+                  ro_c = 1,
+                  gamma_c = 1,
+                  delta_c = 1,
+                  logNv0 = 2,
+                  beta0_r = 0.5,
+                  beta0_d = 0.5,
+                  beta_r = c(0, 0),
+                  beta_d = c(0.5, 0.5),
+                  sigmar = 0.05,
+                  sigmad = 0.05,
+                  rv = rep(0, 243),
                   logdv = rep(0.39, 243)))
 
-             # list(p0 = c(-0.2, -8, -8),
-             #      log_phi_q = -12,
-             #      eta_inv_y = 1,
-             #      eta_inv_q = 10,
-             #      ro_c = 1,
-             #      gamma_c = 1,
[…]

### 2017-05-10 Analysis/citystan.r
GP-ifying mosquito demographic rates

@@ -35,9 +35,7 @@ dat.stan <- list(T = 243,
                  q = q,
                  tau = tau,
                  rov = rov,
-                 sincos = matrix(c(sin(2 * pi * c(1:242) / 52),
-                                   cos(2 * pi * c(1:242) / 52)),
-                                 ncol = 2),
+                 sinmat = (sinpi(as.matrix(dist(c(1:243))) / 52)) ^ 2,
                  pop = pop)
 
 inits = list(list(p0_raw = c(-0.4, -9, -9),
@@ -48,51 +46,50 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   gamma_c = 1,
                   delta_c = 1,
                   logNv0 = 0.7,
-                  beta0_r = 1,
-                  beta0_d = 1,
-                  beta_r = c(0, 0),
-                  beta_d = c(0, 0),
-                  sigmar = 0.1,
-                  sigmad = 0.1,
-                  rv = rep(0, 243),
-                  logdv = rep(0.39, 243)),
-             list(p0_raw = c(-0.2, -8, -8),
-                  log_phi_q = -12,
-                  eta_inv_y = 1,
-                  eta_inv_q = 10,
-                  ro_c = 1,
-                  gamma_c = 1,
-                  delta_c = 1,
-                  logNv0 = 1,
-                  beta0_r = 0,
-                  beta0_d = 0,
-                  beta_r = c(0, 0),
-                  beta_d = c(0, 0),
-                  sigmar = 0.5,
-                  sigmad = 0.5,
-                  rv = rep(0, 243),
-                  logdv = rep(0.39, 243)),
-             list(p0_raw = c(-0.5, -7, -7),
-                  log_phi_q = -14,
-                  eta_inv_y = 0.1,
-                  eta_inv_q = 0.1,
-                  ro_c = 1,
-                  gamma_c = 1,
-                  delta_c = 1,
-                  logNv0 = 2,
-                  beta0_r = 0.5,
-                  beta0_d = 0.5,
-                  beta_r = c(0, 0),
-                  beta_d = c(0.5, 0.5),
-                  sigmar = 0.05,
-                  sigmad = 0.05,
-                  rv = rep(0, 243),
-                  logdv = rep(0.39, 243)))
+                  hsq_r = 1,
+                  omegasq_r = 1,
+                  hsq_d = 1,
+                  omegasq_d = 1,
+                  z_r = rep(0, 243),
[…]

### 2017-05-12 Analysis/citystan.r
Cleaned up GP model

@@ -46,50 +46,19 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   gamma_c = 1,
                   delta_c = 1,
                   logNv0 = 0.7,
-                  hsq_r = 1,
-                  omegasq_r = 1,
-                  hsq_d = 1,
-                  omegasq_d = 1,
+                  h_r = 1,
+                  omega_r = 1,
+                  jitter_r = 0.1,
+                  h_d = 1,
+                  omega_d = 1,
+                  jitter_d = 0.1,
                   z_r = rep(0, 243),
                   z_d = rep(0, 243)))
-# ,
-#              list(p0_raw = c(-0.2, -8, -8),
-#                   log_phi_q = -12,
-#                   eta_inv_y = 1,
-#                   eta_inv_q = 10,
-#                   ro_c = 1,
-#                   gamma_c = 1,
-#                   delta_c = 1,
-#                   logNv0 = 1,
-#                   beta0_r = 0,
-#                   beta0_d = 0,
-#                   beta_r = c(0, 0),
-#                   beta_d = c(0, 0),
-#                   sigmar = 0.5,
-#                   sigmad = 0.5,
-#                   rv = rep(0, 243),
-#                   logdv = rep(0.39, 243)),
-#              list(p0_raw = c(-0.5, -7, -7),
-#                   log_phi_q = -14,
-#                   eta_inv_y = 0.1,
-#                   eta_inv_q = 0.1,
-#                   ro_c = 1,
-#                   gamma_c = 1,
-#                   delta_c = 1,
-#                   logNv0 = 2,
-#                   beta0_r = 0.5,
-#                   beta0_d = 0.5,
-#                   beta_r = c(0, 0),
-#                   beta_d = c(0.5, 0.5),
-#                   sigmar = 0.05,
-#                   sigmad = 0.05,
-#                   rv = rep(0, 243),
-#                   logdv = rep(0.39, 243)))
 
 fit <- stan(file = "Code/seirs.stan", 
             data = dat.stan, 
             init = inits, 
-            iter = 100, 
+            iter = 500, 
             chains = 1,
             control = list(adapt_delta = 0.9))
 

### 2017-05-15 Manuscript/ms.tex
Deleting old and outlining new results

@@
 \begin{itemize}
-	\item To what extent do the case reports inform our estimates of mosquito abundance?
-How do we separate the information coming from the case reports from the information coming from the mosquito data itself?
-	\item Do our estimates of epidemiological parameters differ from their prior (literature) values?
-	\item Do the case reports reproduce the observed mosquito surveillance time series?
-	\item If no, how does it fail?
+\item How wells does the model fit the case reports and mosquito surveillance data?
+\item What do the estimated time series of demographic rates look like? How smooth are they? How strong is the periodic signal relative to noise?
+\item How far/how well can we make predictions?
+\item What do the estimated demographic rates tell us about efficient implementation of control?
 \end{itemize}
-
-The model is able to effectively capture the observed weekly case reports for Vitoria (Fig. \ref{timeseries}A), as expected given the degrees of freedom in the mosquito process model.
-In addition, most of the observed mosquito trap counts fall within the 80\% posterior credible interval (Fig. \ref{timeseries}B).
-We expect the posterior distribution of mosquito abundance time series to be informed primarily by the observed case data.
-However, the mapping from the case reports to mosquito abundance is influenced by the parameters of the mechanistic model and deviations of the posterior distribution of these parameters from their prior reflect the influence of the mosquito trap count data.
-Of these parameters, the posterior estimate of mosquito mortality rate (blah) was lower than its prior mean (1.47 week$^{-1}$), while the posterior estimate of the intrinsic incubation period (blah) was larger than its prior mean (0.87 weeks).
-That these two parameters differ from their literature values reflects the fact that they are critical in controlling the interaction between mosquito abundance and human disease risk, as the intrinsic incubation period controls how quickly humans are able to begin transmitting to mosquitoes, and the mortality rate controls for how long mosquitoes are able to remain infectious and transmit to humans.
-
-However, there is a seasonal trend in the weekly posterior p-values for the replicate mosquito surveillance series (Fig.
-\ref{pseries}), indicating a tendency to overestimate the number of mosquitoes trapped during the downswing of the annual epidemic and underestimate the number of mosquitoes trapped during the upswing.
-This seasonal trend indicates a systematic lack-of-fit.
-
-This lack-of-fit is due in large part to the failure of the estimated mosquito trajectories to capture the periodic seasonal structure in the observed mosquito surveillance data (Fig. \ref{autocorr}).
-In particular, the autocorrelation in the posterior replicate mosquito surveillance series decays much faster than in the observed data and remains near zero for lags near a year, whereas the observed data have positive autocorrelation at yearly lags (\ref{autocorr}).
-Moreover, the posterior standard deviations of 1 week differences are substantially larger than the observed (Supplemental?), indicating that the observed time series is much smoother than the posterior replicates.
-These posterior checks indicate that the case data and mechanistic model are not able to reproduce the observed time series of mosquito trap counts.
 
@@
 
-\begin{figure}[!h]
-%\includegraphics[angle = 270, scale=1]{figures/fig2.eps}
-\caption{{\bf Posterior densities of epidemiological parameters.}
-A: Intrinsic incubation period. B: Period of cross-immunity. C: Decay rate of infectiousness. D: Mosquito mortality rate.
-}
-\label{pseries}
-\end{figure}
-
-\begin{figure}[!h]
-%\includegraphics[angle = 270, scale=1]{figures/fig2.eps}
-\caption{{\bf Weekly posterior predictive p-values.}
-$P(q^{rep}_t > q_t)$ for each week of the year for 2008 through week 34 of 2012.
-Gray lines show loess-smoothed curves of the p-values for each year.
-Based on 15000 replicate mosquito surveillance data sets estimated from case reports.
-}
-\label{pseries}
-\end{figure}
-
-\begin{figure}[!h]
-%\includegraphics[angle = 270, scale=1]{figures/fig3.eps}
-\caption{{\bf Posterior autocorrelation function.}
-Autocorrelation of observed mosquito trap counts (points) and replicate time series estimated from case reports (vertical lines, indicating 90\% posterior credible interval). 
-}
-\label{autocorr}
-\end{figure}
-
 \section*{Discussion}
@@
 \item Disease risk possibly driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Canyon1999}.
-
 \end{itemize}

### 2017-05-15 Manuscript/ms.tex
Reframing the introduction

@@
 Dengue fever is a massive global public health burden, with millions of cases per year in Brazil alone (CITATION).  
-Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
-Even if a vaccine is developed, mosquito control is likely to remain an important component of dengue management \cite{Achee2015}.
-Studies have shown that targeted mosquito control efforts can be effective in reducing disease burden.
+Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito surveillance and control programs \cite{Achee2015}.
+Deploying these mosquito control measures effectively and efficiently requires a detailed understanding of how mosquito surveillance indices map to human disease risk.
 For example contact tracing and residential spraying reduced disease risk in Cairns, Australia \cite{Vazquez-Prokopec2017}.
-Though this approach was successful, it requires detailed data following individual cases, potentially limiting its utility for large-scale deployment.
+This approach ensures that the mosquitoes targeted by control are epidemiologically relevant, but the detailed data it requires potentially limits its utility for large-scale deployment.
 
-Other efforts have sought to guide the deployment of control efforts through the use of fine-scale mosquito surveillance.
+More commonly, cities deploy mosquito control based on trap-based mosquito surveillance data.
 For instance, the "MI-Dengue" system deployed throughout several cities in Brazil uses a grid of mosquito sticky-traps (MosquiTRAPs) for spatial prioritization of source reduction and larvicide application \cite{Eiras2009}.
@@
 Though this system has been effective, the mosquito indices and thresholds used to prioritize control efforts remain somewhat ad-hoc.
-Using such surveillance data to efficiently target limited control resources requires a detailed understanding of the relationship between the components of mosquito abundances measured by the trap and disease risk.
+Using such surveillance data to efficiently target limited control resources requires a more detailed understanding of the relationship between the components of mosquito abundance measured by traps and disease risk.
 
-However, even though mosquitoes are necessary for dengue transmission, establishing a clear link between surveillance-based indices of mosquito abundance and disease risk has proven difficult thus far (CITATIONS).
-Many of the attempts to establish this link have relied on tests of the ability of mosquito abundance measures to predict human disease risk in regression models.
-For example, \cite{Pepin2015} found that models that included MosquiTRAP surveillance data failed to predict weekly dengue cases any better than autoregressive models that include lagged case data alone.
-Evaluating the influence of mosquito abundance in this way is analogous to a test of Granger causality \cite{Granger1969}, which states that some variable $X$ (e.g.
-mosquito abundance) \emph{causes} $Y$ (e.g.
-human dengue cases) if $X$ contains unique information about $Y$ (i.e.
-if a model with $X$ and $Y$ provides better predictions of $Y$ than a model with $Y$ alone).
-By this definition, the results of \cite{Pepin2015} suggest that mosquito abundance does not cause dengue cases (in the Granger sense).
+Many of the attempts to establish this relationship statistically have relied on tests of the ability of mosquito abundance measures to predict human disease risk in regression models.
+For example, \cite{Pepin2015} found that models that included MosquiTRAP surveillance data failed to predict weekly dengue cases any better than models that include lagged case data alone.
+However, these models fail to account for the mechanistic, non-linear relationship between mosquito population dynamics and dengue epidemiology.
+In particular, the cycle of transmission between humans and dengue is controlled not just by mosquito population size, but also by the emergence of new susceptible adults and the mortality rate.
 
-However, as noted by Sugihara \emph{et al.} \cite{Sugihara2012a}, Granger causality assumes that $X$ and $Y$ are separable, which is not the case if $X$ and $Y$ are linked in a non-linear dynamical system.
-In this case, a model with $X$ and $Y$ will not outperform the model with only $Y$, as the information contained in $X$ is already embedded in $Y$, a consequence of Taken's theorem (CITATION). 
-Since dengue transmission is governed by such a non-linear dynamical system, information about the driving component of mosquito abundance should be already embedded in the time series of human case reports.
-As a result, predictive utility is not an appropriate way to test the causal influence of mosquito abundance on dengue risk.
-Instead, Sugihara \emph{et al.} argue that if $X$ is a driver $Y$, then you should be able to reconstruct $X$ from $Y$.
-
-In this paper, we apply this perspective to the problem of linking mosquito abundance to disease risk.
-To test whether or not mosquito abundance, as measured through surveillance efforts, drives dengue dynamics, we ask whether we can reconstruct a time series of mosquito surveillance data from a time series of case reports.
-We approach this as an inverse problem: we first posit a mechanistic model of dengue transmission and then develop a Bayesian statistical framework to estimate the mosquito abundance time series that is consistent with that model and the time series of case reports.
-We then evaluate the causal link between mosquito abundance and case reports by using posterior predictive checks to compare the estimated and observed mosquito time series.
+In this paper, we seek to connect mosquito surveillance data with human case reports through the use of a Bayesian, mechanistic model.
+We first propose a differential equation model for dengue epidemiology.
+We then embed this model into a Bayesian statistical framework that allows us to estimate latent time series of mosquito demographic parameters from time series of mosquito trap counts and reported cases of dengue fever.
+Lastly, we explore the ability of this fitted model to predict annual dengue outbreaks and its utility for targeting control efforts.
 

### 2017-05-17 Analysis/citystan.r
Dropping GP for periodic regression

@@ -35,7 +35,10 @@ dat.stan <- list(T = 243,
                  q = q,
                  tau = tau,
                  rov = rov,
-                 sinmat = (sinpi(as.matrix(dist(c(1:243))) / 52)) ^ 2,
+                 sincos = matrix(c(rep(1, 243),
+                                   sin(2 * pi * c(1:243) / 52),
+                                   cos(2 * pi * c(1:243) / 52)),
+                                 ncol = 3),
                  pop = pop)
 
 inits = list(list(p0_raw = c(-0.4, -9, -9),
@@ -46,12 +49,38 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   gamma_c = 1,
                   delta_c = 1,
                   logNv0 = 0.7,
-                  h_r = 1,
-                  omega_r = 1,
-                  jitter_r = 0.1,
-                  h_d = 1,
-                  omega_d = 1,
-                  jitter_d = 0.1,
+                  alpha = c(0, 0, 0),
+                  beta = c(0.38, 0, 0),
+                  sigmar = 0.1,
+                  sigmad = 0.1,
+                  z_r = rep(0, 243),
+                  z_d = rep(0, 243)),
+             list(p0_raw = c(-0.4, -9, -9),
+                  log_phi_q = -13,
+                  eta_inv_y = 2,
+                  eta_inv_q = 10,
+                  ro_c = 1,
+                  gamma_c = 1,
+                  delta_c = 1,
+                  logNv0 = 0.7,
+                  alpha = c(0, 0, 0),
+                  beta = c(0.38, 0.05, 0.05),
+                  sigmar = 0.01,
+                  sigmad = 0.01,
+                  z_r = rep(0, 243),
+                  z_d = rep(0, 243)),
+             list(p0_raw = c(-0.4, -9, -9),
+                  log_phi_q = -13,
+                  eta_inv_y = 10,
+                  eta_inv_q = 2,
+                  ro_c = 1,
+                  gamma_c = 1,
+                  delta_c = 1,
+                  logNv0 = 0.7,
+                  alpha = c(0, 0.05, 0.05),
+                  beta = c(0.38, 0, 0),
+                  sigmar = 0.5,
+                  sigmad = 0.5,
                   z_r = rep(0, 243),
                   z_d = rep(0, 243)))
 
@@ -59,6 +88,7 @@ fit <- stan(file = "Code/seirs.stan",
             data = dat.stan, 
             init = inits, 
[…]

### 2017-05-18 Analysis/citystan.r
Modeling net birth rate instead of growth rate

@@ -35,10 +35,9 @@ dat.stan <- list(T = 243,
                  q = q,
                  tau = tau,
                  rov = rov,
-                 sincos = matrix(c(rep(1, 243),
-                                   sin(2 * pi * c(1:243) / 52),
+                 sincos = matrix(c(sin(2 * pi * c(1:243) / 52),
                                    cos(2 * pi * c(1:243) / 52)),
-                                 ncol = 3),
+                                 ncol = 2),
                  pop = pop)
 
 inits = list(list(p0_raw = c(-0.4, -9, -9),
@@ -49,46 +48,46 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   gamma_c = 1,
                   delta_c = 1,
                   logNv0 = 0.7,
-                  alpha = c(0, 0, 0),
-                  beta = c(0.38, 0, 0),
-                  sigmar = 0.1,
+                  alpha0 = 1,
+                  alpha = c(0, 0),
+                  beta0 = 0.39,
+                  beta = c(0, 0),
+                  sigmab = 0.1,
                   sigmad = 0.1,
-                  z_r = rep(0, 243),
-                  z_d = rep(0, 243)),
-             list(p0_raw = c(-0.4, -9, -9),
-                  log_phi_q = -13,
-                  eta_inv_y = 2,
-                  eta_inv_q = 10,
-                  ro_c = 1,
-                  gamma_c = 1,
-                  delta_c = 1,
-                  logNv0 = 0.7,
-                  alpha = c(0, 0, 0),
-                  beta = c(0.38, 0.05, 0.05),
-                  sigmar = 0.01,
-                  sigmad = 0.01,
-                  z_r = rep(0, 243),
-                  z_d = rep(0, 243)),
-             list(p0_raw = c(-0.4, -9, -9),
-                  log_phi_q = -13,
-                  eta_inv_y = 10,
-                  eta_inv_q = 2,
-                  ro_c = 1,
-                  gamma_c = 1,
-                  delta_c = 1,
-                  logNv0 = 0.7,
-                  alpha = c(0, 0.05, 0.05),
-                  beta = c(0.38, 0, 0),
-                  sigmar = 0.5,
-                  sigmad = 0.5,
-                  z_r = rep(0, 243),
+                  z_b = rep(0, 243),
                   z_d = rep(0, 243)))
+             # list(p0_raw = c(-0.4, -9, -9),
+             #      log_phi_q = -13,
[…]

### 2017-05-19 Analysis/citystan.r
Deterministic emergence rate

@@ -52,42 +52,40 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   alpha = c(0, 0),
                   beta0 = 0.39,
                   beta = c(0, 0),
-                  sigmab = 0.1,
                   sigmad = 0.1,
-                  z_b = rep(0, 243),
+                  z_d = rep(0, 243)),
+             list(p0_raw = c(-0.4, -9, -9),
+                  log_phi_q = -13,
+                  eta_inv_y = 10,
+                  eta_inv_q = 10,
+                  ro_c = 1,
+                  gamma_c = 1,
+                  delta_c = 1,
+                  logNv0 = 0.7,
+                  alpha0 = 0.5,
+                  alpha = c(0.01, 0.01),
+                  beta0 = 0.39,
+                  beta = c(0.01, 0.01),
+                  sigmad = 0.1,
+                  z_d = rep(0, 243)),
+             list(p0_raw = c(-0.4, -9, -9),
+                  log_phi_q = -13,
+                  eta_inv_y = 2,
+                  eta_inv_q = 2,
+                  ro_c = 1,
+                  gamma_c = 1,
+                  delta_c = 1,
+                  logNv0 = 0.7,
+                  alpha0 = 0,
+                  alpha = c(0.5, 0.5),
+                  beta0 = 0.1,
+                  beta = c(0, 0),
+                  sigmad = 0.5,
                   z_d = rep(0, 243)))
-             # list(p0_raw = c(-0.4, -9, -9),
-             #      log_phi_q = -13,
-             #      eta_inv_y = 2,
-             #      eta_inv_q = 10,
-             #      ro_c = 1,
-             #      gamma_c = 1,
-             #      delta_c = 1,
-             #      logNv0 = 0.7,
-             #      alpha = c(0.7, 0, 0),
-             #      beta = c(0.38, 0.05, 0.05),
-             #      sigmar = 0.01,
-             #      sigmad = 0.01,
-             #      z_r = rep(0, 243),
-             #      z_d = rep(0, 243)),
-             # list(p0_raw = c(-0.4, -9, -9),
-             #      log_phi_q = -13,
-             #      eta_inv_y = 10,
-             #      eta_inv_q = 2,
-             #      ro_c = 1,
-             #      gamma_c = 1,
-             #      delta_c = 1,
-             #      logNv0 = 0.7,
-             #      alpha = c(0.7, 0.05, 0.05),
-             #      beta = c(0.38, 0, 0),
[…]

### 2017-05-19 Analysis/plots.r

@@ -4,6 +4,7 @@ library(rstan)
 library(plyr)
 library(magrittr)
 library(ggplot2)
+library(gridExtra)
 
 # Loading and processing data
 tseries <- read.csv("Data/Vitoria.data.csv")
@@ -65,6 +66,35 @@ grid.arrange(fig1.a, fig1.b, ncol = 2)
 dev.off()
 
 #===============================================================================
+# Figure 2: estimated demographic rates
+
+b <- rstan::extract(sim, "bv", permute = T)[[1]] %>% apply(2, quantile, c(0.1, 0.5, 0.9))
+d <- rstan::extract(sim, "dv", permute = T)[[1]] %>% apply(2, quantile, c(0.1, 0.5, 0.9))
+
+post[, c("bmin", "b", "bmax")] <- t(b)
+post[, c("dmin", "d", "dmax")] <- t(d)
+
+ggplot(post, aes(tot.week, b)) + 
+  geom_ribbon(aes(ymin = bmin, ymax = bmax), fill = "grey70") +
+  geom_line() + 
+  theme_classic() +
+  scale_y_continuous(expand = c(0.05, 0)) + 
+  scale_x_continuous(expand = c(0, 1)) +
+  ylab("mosquito net emergence rate") +
+  xlab("week") +
+  ggtitle("A")
+
+ggplot(post, aes(tot.week, d)) + 
+  geom_ribbon(aes(ymin = dmin, ymax = dmax), fill = "grey70") +
+  geom_line() + 
+  theme_classic() +
+  scale_y_continuous(expand = c(0.05, 0)) + 
+  scale_x_continuous(expand = c(0, 1)) +
+  ylab("mosquito per-capita death rate") +
+  xlab("week") +
+  ggtitle("B")
+
+#===============================================================================
 # Figure 2: weekly posterior predictive p-value for replicate mosquito
 # surveillance data
 
@@ -82,8 +112,9 @@ postscript("Manuscript/figures/fig2.eps",
            width = 4, height = 3,
            family = "ArialMT")
 
-ggplot(post, aes(tot.week, p)) +
-  geom_line() + 
+ggplot(post, aes(week, p, group = factor(year))) + 
+  geom_point() + 
+  geom_smooth(se = F, color = "gray40") +
   theme_classic() + 
   scale_y_continuous(expand = c(0, 0.01)) + 
   scale_x_continuous(expand = c(0, 1)) +

### 2017-05-19 Analysis/citystan.r
Tweaking priors and initial values

@@ -62,10 +62,10 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   gamma_c = 1,
                   delta_c = 1,
                   logNv0 = 0.7,
-                  alpha0 = 0.5,
-                  alpha = c(0.01, 0.01),
-                  beta0 = 0.39,
-                  beta = c(0.01, 0.01),
+                  alpha0 = -1,
+                  alpha = c(4, 3),
+                  beta0 = 0.1,
+                  beta = c(0.2, 0.1),
                   sigmad = 0.1,
                   z_d = rep(0, 243)),
              list(p0_raw = c(-0.4, -9, -9),
@@ -77,7 +77,7 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   delta_c = 1,
                   logNv0 = 0.7,
                   alpha0 = 0,
-                  alpha = c(0.5, 0.5),
+                  alpha = c(1, 1),
                   beta0 = 0.1,
                   beta = c(0, 0),
                   sigmad = 0.5,

### 2017-05-19 Manuscript/ms.tex
Working on restructuring intro

@@
 Dengue fever is a massive global public health burden, with millions of cases per year in Brazil alone (CITATION).  
-Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito surveillance and control programs \cite{Achee2015}.
-Deploying these mosquito control measures effectively and efficiently requires a detailed understanding of how mosquito surveillance indices map to human disease risk.
-For example contact tracing and residential spraying reduced disease risk in Cairns, Australia \cite{Vazquez-Prokopec2017}.
-This approach ensures that the mosquitoes targeted by control are epidemiologically relevant, but the detailed data it requires potentially limits its utility for large-scale deployment.
+Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
+Though eradication of \textit{Aedes aegypti} is infeasible, effectively targeted control interventions may still substantially reduce disease risk.
 
-More commonly, cities deploy mosquito control based on trap-based mosquito surveillance data.
+Cities deploy mosquito control based on trap-based mosquito surveillance data.
 For instance, the "MI-Dengue" system deployed throughout several cities in Brazil uses a grid of mosquito sticky-traps (MosquiTRAPs) for spatial prioritization of source reduction and larvicide application \cite{Eiras2009}.
@@
 \item Disease risk possibly driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Canyon1999}.
+\item Explicitly targeting mosquitoes that have been implicated in transmission through the use of contact tracing \cite{Vazquez-Prokopec2017}
 \end{itemize}

### 2017-05-22 Analysis/predict.r
Code to use model for predictions

@@ -1,0 +1,68 @@
+library(rstan)
+library(plyr)
+library(magrittr)
+
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+#===============================================================================
+# Loading data and setting up model
+
+tseries <- read.csv("Data/Vitoria.data.csv")
+obs <- tapply(tseries$Cases, list(tseries$tot.week), sum, na.rm=TRUE)
+
+# Population size
+pop <- 327801
+
+# Pull out mosquito counts and traps inspected
+q <- tapply(tseries$Mosquitoes, list(tseries$tot.week), sum, na.rm = TRUE)
+tau <- tapply(tseries$Trap, list(tseries$tot.week), sum, na.rm = TRUE)
+
+# Computing eip forcing
+weather <- read.csv("Data/Vitoria.weather.csv")
+temp <- weather$Mean.TemperatureC
+
+week <- rep(c(1:243), each = 7)
+meantemp <- tapply(temp[1:1701], week, mean, na.rm = T)
+
+rov <- 7 / exp(exp(1.9 - 0.04 * meantemp) + 1 / (2 * 7))
+
+#===============================================================================
+# Running stan
+
+T_pred <- 52
+T_fit <- 243 - T_pred
+
+dat.stan <- list(T = T_fit,
+                 T_pred = T_pred,
+                 y = head(obs, T_fit),
+                 q = head(q, T_fit),
+                 tau = head(tau, T_fit),
+                 rov = rov,
+                 sincos = matrix(c(sin(2 * pi * c(1:243) / 52),
+                                   cos(2 * pi * c(1:243) / 52)),
+                                 ncol = 2),
+                 pop = pop)
+
+inits = list(list(p0_raw = c(-0.4, -9, -9),
+                  log_phi_q = -13,
+                  eta_inv_y = 5,
+                  eta_inv_q = 5,
+                  ro_c = 1,
+                  gamma_c = 1,
+                  delta_c = 1,
+                  logNv0 = 0.7,
+                  alpha0 = 1,
+                  alpha = c(0, 0),
+                  beta0 = 0.39,
+                  beta = c(0, 0),
+                  sigmad = 0.1,
+                  z_d = rep(0, T_fit)))
+
+fit <- stan(file = "Code/predict.stan", 
+            data = dat.stan, 
+            init = inits, 
+            iter = 500, 
+            chains = 1,
+            control = list(adapt_delta = 0.9))
+

### 2017-05-22 Analysis/citystan.r
Cleaning up weather data for potential use as covariates

@@ -19,25 +19,32 @@ q <- tapply(tseries$Mosquitoes, list(tseries$tot.week), sum, na.rm = TRUE)
 tau <- tapply(tseries$Trap, list(tseries$tot.week), sum, na.rm = TRUE)
 
 # Computing eip forcing
-weather <- read.csv("Data/Vitoria.weather.csv")
-temp <- weather$Mean.TemperatureC
+weather <- read.csv("Data/Vitoria.weather.csv") %>% 
+  mutate(date = ymd(BRST), year = year(date), week = week(date))
 
-week <- rep(c(1:243), each = 7)
-meantemp <- tapply(temp[1:1701], week, mean, na.rm = T)
+weather <- subset(weather, date < date[1] + weeks(243))
+weather$tot.week <- rep(c(1:243), each = 7)
 
-rov <- 7 / exp(exp(1.9 - 0.04 * meantemp) + 1 / (2 * 7))
+covars <- ddply(weather, .(tot.week), summarise,
+                temp = mean(Mean.TemperatureC, na.rm = T),
+                humidity = mean(Mean.Humidity, na.rm = T)
+                )
+
+rov <- 7 / exp(exp(1.9 - 0.04 * covars$temp) + 1 / (2 * 7))
 
 #===============================================================================
 # Running stan
 
 dat.stan <- list(T = 243,
+                 D = 2,
                  y = obs,
                  q = q,
                  tau = tau,
                  rov = rov,
-                 sincos = matrix(c(sin(2 * pi * c(1:243) / 52),
-                                   cos(2 * pi * c(1:243) / 52)),
+                 sincos = matrix(c(sin(2 * pi * c(1:243) / 52), 
+                                   cos(2 * pi * c(1:243) / 52)), 
                                  ncol = 2),
+                 covars = scale(covars[, -1]),
                  pop = pop)
 
 inits = list(list(p0_raw = c(-0.4, -9, -9),
@@ -62,7 +69,7 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   gamma_c = 1,
                   delta_c = 1,
                   logNv0 = 0.7,
-                  alpha0 = -1,
+                  alpha0 = 0,
                   alpha = c(4, 3),
                   beta0 = 0.1,
                   beta = c(0.2, 0.1),
@@ -76,7 +83,7 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   gamma_c = 1,
                   delta_c = 1,
[…]

### 2017-05-22 Analysis/predict.r

@@ -9,6 +9,7 @@ options(mc.cores = parallel::detectCores())
 # Loading data and setting up model
 
 tseries <- read.csv("Data/Vitoria.data.csv")
+
 obs <- tapply(tseries$Cases, list(tseries$tot.week), sum, na.rm=TRUE)
 
 # Population size
@@ -18,23 +19,38 @@ pop <- 327801
 q <- tapply(tseries$Mosquitoes, list(tseries$tot.week), sum, na.rm = TRUE)
 tau <- tapply(tseries$Trap, list(tseries$tot.week), sum, na.rm = TRUE)
 
-# Computing eip forcing
-weather <- read.csv("Data/Vitoria.weather.csv")
-temp <- weather$Mean.TemperatureC
+# Wrangling weather data and computing EIP
+weather <- read.csv("Data/Vitoria.weather.csv") %>% 
+  mutate(date = ymd(BRST), year = year(date), week = week(date))
 
-week <- rep(c(1:243), each = 7)
-meantemp <- tapply(temp[1:1701], week, mean, na.rm = T)
+start <- weather$date[1]
+T_pred <- 34
+T_fit <- 243 - T_pred
 
-rov <- 7 / exp(exp(1.9 - 0.04 * meantemp) + 1 / (2 * 7))
+weather <- subset(weather, date < date[1] + weeks(T_fit))
+weather$tot.week <- rep(c(1:T_fit), each = 7)
+
+covars <- ddply(weather, .(tot.week), summarise,
+                temp = mean(Mean.TemperatureC, na.rm = T), 
+                humidity = mean(Mean.Humidity, na.rm = T)
+                )
+
+annual <- ddply(weather, .(week), summarise,
+                temp = mean(Mean.TemperatureC, na.rm = T),
+                humidity = mean(Mean.Humidity, na.rm = T)
+                )
+
+predvars <- rbind(covars[, -1],
+                  annual[week(start + weeks((T_fit + 1):243)), -1])
+
+rov <- 7 / exp(exp(1.9 - 0.04 * predvars$temp) + 1 / (2 * 7))
 
 #===============================================================================
 # Running stan
 
-T_pred <- 52
-T_fit <- 243 - T_pred
-
 dat.stan <- list(T = T_fit,
                  T_pred = T_pred,
+                 D = 2,
                  y = head(obs, T_fit),
                  q = head(q, T_fit),
                  tau = head(tau, T_fit),
@@ -42,6 +58,7 @@ dat.stan <- list(T = T_fit,
[…]

### 2017-05-22 Analysis/citystan.r
Fixing error in stan code and initial conditions

@@ -68,9 +68,9 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   ro_c = 1,
                   gamma_c = 1,
                   delta_c = 1,
-                  logNv0 = 0.7,
+                  logNv0 = 1,
                   alpha0 = 0,
-                  alpha = c(4, 3),
+                  alpha = c(2, 3),
                   beta0 = 0.1,
                   beta = c(0.2, 0.1),
                   sigmad = 0.1,

### 2017-05-23 Manuscript/ms.tex
Separating supplement into own document
And updating description of model structure to reflect seasonally forced emergence rate and latent stochastic death rate.

@@
 Similarly, susceptible mosquitoes ($V_S$) become exposed by biting infectious humans and pass through a temperature-dependent latent period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
-Total mosquito population size ($V_N$) is controlled by a time-varying growth rate, $r_t$, and captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.
+Total mosquito population size ($V_N$) is controlled by a time-varying net emergence rate, $b_t$, and captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.
 
@@
 \begin{align} 
-\frac{dS}{dt} &= bN - bS - \alpha \frac{V_{I}}{N} S + \delta R\\
-\frac{dE}{dt} &= \alpha \frac{V_{I}}{N} S - (\rho + b)E\\
+\frac{dS}{dt} &= bN - bS - \lambda \frac{V_{I}}{N} S + \delta R\\
+\frac{dE}{dt} &= \lambda \frac{V_{I}}{N} S - (\rho + b)E\\
 \frac{dI}{dt} &= \rho E - (\gamma + b)I\\
 \frac{dR}{dt} &= \gamma I - (\delta + b)R\\
-\frac{dV_N}{dt} & = r_t V_N - \phi_q V_N\\
-\frac{dV_E}{dt} &= \alpha \frac{I}{N} V_S - (\rho_{vt} + d)V_E\\
-\frac{dV_I}{dt} &= \rho_{vt} V_E - d V_I\\
-\frac{dV_C}{dt} & = \phi_q V_N\\
-V_S &= V_N - V_E - V_I
+\frac{dV_S}{dt} & = b_t - \lambda \frac{I}{N} V_S - (d_t + \phi_q \tau_t) V_S \\
+\frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{vt} + d_t + \phi_q \tau_t)V_E\\
+\frac{dV_I}{dt} &= \rho_{vt} V_E - (d_t + \phi_q \tau_t) V_I\\
+\frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
+V_N &= V_S + V_E + V_I
 \end{align}
 
-The mosquito demographic parameters (net grown rate, $r_t$, and death rate, $d_t$) are modeled as an AR(1) process with a periodic regression component to account for unmodeled seasonal forcing (e.g.
+The mosquito demographic parameters (net emergence rate, $b_t$, and per capita death rate, $d_t$) are modeled with a periodic regression component to account for unmodeled seasonal forcing (e.g.
 from temperature, precipitation, etc.):
 \begin{align}
-r_t & = \alpha_{1}r_{t-1} + \alpha_2 \sin(2\pi t / 52) + \alpha_3 \cos(2\pi t / 52) + \epsilon_t\\
-\log(d_{t}) & = \beta_{1}\log(d_{t-1}) + \beta_2 \sin(2\pi t / 52) + \beta_3 \cos(2\pi t / 52) + \xi_t,
+\log(b_t) & = \alpha_{0} + \alpha_1 \sin(2\pi t / 52) + \alpha_2 \cos(2\pi t / 52) \\
+\log(d_{t}) & = \beta_{0} + \beta_1 \sin(2\pi t / 52) + \beta_2 \cos(2\pi t / 52) + \epsilon_t,
 \end{align}
@@
 \begin{align}
-\epsilon_t & \sim \text{Normal}(0, \sigma_{\epsilon})\\
-\xi_t & \sim \text{Normal}(0, \sigma_{\xi}).
+\epsilon_t & \sim \text{Normal}(0, \sigma)\\
 \end{align}
@@
 
-The remaining parameters include the epidemiological parameters that control the flows between the compartments of the differential equation model ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the regression parameters and process noise in the models for mosquito growth and death rates ($\mathbf{\alpha}$, $\mathbf{\beta}$, and $\sigma_{\epsilon}$ and $\sigma_{\xi}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
+The remaining parameters include the epidemiological parameters that control the flows between the compartments of the differential equation model ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the regression parameters and process noise in the models for mosquito growth and death rates ($\mathbf{\alpha}$, $\mathbf{\beta}$, and $\sigma$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
 Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table \ref{fixedparms} for means and Supplemental Material for detailed explanations).
@@
 
-\subsection*{Full model}
-
-\begin{align}
-y_t & \sim \text{NegBin}(\phi_y  (X_t - X_{t-1})N, \eta_y)\\
-q_t & \sim \text{NegBin}(V_{Ct} - V_{Ct-1}, \eta_q),
-\\ 
-\frac{dX}{dt} &= \rho E \\
-\frac{dS}{dt} &= b - bS - \alpha V_{I} S + \delta R\\
-\frac{dE}{dt} &= \alpha V_{I}S - (\rho + b)E\\
-\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
-\frac{dR}{dt} &= \gamma I - (\delta + b)R\\
-\\
-\frac{dV_N}{dt} & = r_t V_N - \phi_q V_N\\
-\frac{dV_E}{dt} &= \alpha \frac{I}{N} V_S - (\rho_{vt} + d)V_E\\
-\frac{dV_I}{dt} &= \rho_{vt} V_E - d V_I\\
-\frac{dV_C}{dt} & = \phi_q V_N\\
-V_S &= V_N - V_E - V_I\\
-\\
-r_t & = \alpha_{1}r_{t-1} + \alpha_2 \sin(2\pi t / 52) + \alpha_3 \cos(2\pi t / 52) + \epsilon_t\\
-\log(d_{t}) & = \beta_{1}\log(d_{t-1}) + \beta_2 \sin(2\pi t / 52) + \beta_3 \cos(2\pi t / 52) + \xi_t\\
-\epsilon_t & \sim \text{Normal}(0, \sigma_{\epsilon})\\
-\xi_t & \sim \text{Normal}(0, \sigma_{\xi})
-\end{align}
-
-\begin{align}
-\mathbf{\beta} & \sim \text{Normal}(0, 0.4)\\
-\mathbf{\alpha} & \sim \text{Normal}(0, 0.4)\\
-\sigma_{\epsilon} & \sim \text{Normal}(0, 0.3)\\
-\sigma_{\xi} & \sim \text{Normal}(0, 0.3)\\
--\log(\phi_q) & \sim \text{Half-normal}(13, 0.5)\\
-\eta_y &\sim \text{Half-normal}(0, 5)\\
-\eta_q & \sim \text{Half-normal}(0, 5)\\
-T_{\rho} & \sim \text{Gamma}(8.3, 8.3)\\
-\rho & = (0.87 T_{\rho})^{-1}\\
-T_{\delta} & \sim \text{Gamma}(10, 10)\\
-\delta & = (97 T_{\delta})^{-1}\\
-\gamma' & \sim \text{Gamma}(100, 100)\\
-\gamma & = 3.5\gamma'\\
-d_{v}' & \sim \text{Gamma}(100, 100)\\
-d_v & = 1.47 d_v'\\
-S_0' & \sim \text{Normal}(-0.4, 0.2)\\
-E_0' & \sim \text{Normal}(-9, 0.6)\\
-I_0' & \sim \text{Normal}(-9, 0.6)\\
-R_0' & = 0\\
-S_0 & = \frac{\exp(S_0)}{\exp(S_0' + E_0' + I_0' + R_0')} \\
-\end{align}
-
-\subsection*{Details of prior specification}
-
-\begin{itemize}
-\item \textbf{Period of cross-immunity} ($T_{\delta}$): Reich \emph{et al.} \cite{Reich2013} estimate a mean of 1.88 years (97 weeks) with a 95\% confidence interval of (0.88, 4.31).
-We specify a gamma prior on the multiplicative scale to have a mean of one and 0.01 and 0.99 quantiles at roughly $0.88/1.88 = 0.47$ and $4.31/1.88 = 2.3$,, respectively:
-\begin{equation}
-T_{\delta} \sim \text{Gamma}(10, 10)
-\end{equation}
-where $\delta = (97 T_\delta)^{-1}$.
-Note that we use a slightly more conservative interval than Reich \emph{et al.} to avoid numerical instability at extreme values.
-\\
-\item \textbf{Intrinsic incubation period} ($T_{\rho}$): Chan and Johansson \cite{Chan2012} estimate a mean of 6.1 days with a 95\% credible interval of (3, 10).
-Thus, we specify a gamma prior on the multiplicative scale to have a mean of one, and 0.05 and 0.95 quantiles of $3 / 6.1 = 0.49$ and $10 / 6.1 = 1.6$, respectively:
-\begin{equation}
-T_{\rho} \sim \text{Gamma}(8.3, 8.3)
-\end{equation}
-where $\rho = (0.87 T_{\rho})^{-1}$.
-\\
-\item \textbf{Decay rate of infectiousness} ($\gamma$): Nguyet \emph{et al.} \cite{Nguyet2013} find that there is very little transmission of dengue from humans to mosquitoes after 5-6 days since the onset of fever (with roughly an extra of infectiousness prior to onset of fever).
-In reality, an individual's infectiousness decays with time, but the ODE model assumes infectiousness is constant for the (exponentially distributed) duration of the infection period.    
-To parameterize the model then, we assume a rough equivalence between the probability of infection from a single individual at a given time and the proportion of individuals still infectious at a given time.
-Thus we choose our prior for $\gamma$ to keep the time at which only 5\% of individuals are still infectious between 5 and 7 days, with a mean 5\% threshold of 6 days.
-Again we specify the prior on the multiplicative scale to have a mean of one and 0.05 and 0.95 quantiles at $2.99/3.5 = 0.85$ and $4.19 / 3.5 = 1.2$, respectively:
-\begin{equation}
-\gamma' \sim \text{Gamma}(100, 100)
-\end{equation}
-where $\gamma = 3.5\gamma'$.
-\\
-\item \textbf{Initial mosquito mortality rate} ($d_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
-They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $d$.  
-We use nonlinear least squares to fit the exponential cdf with rate $d_v$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d = 1.47$.
-We use the same procedure to estimate $d$ for the 0.25 and 0.75 quantiles of the estimated mortality curves (assuming symmetry around the mean), which gives 1.6 and 1.37.
-We then specify the prior on $\log(d)$ to induce the above quantiles on $d$, which gives us:
-\begin{equation}
-\log(d) \sim \text{Normal}(0.39, 0.12).
-\end{equation}
-\\
-\item \textbf{Initial mosquito growth rate} ($r_0$):
-\\
-\item \textbf{Initial conditions} ($S_0, E_0, I_0, R_0$): The SEIRS model assumes a constant population size and is scaled so that $S_0+E_0+I_0+R_0 = 1$.
-We specify our prior on an unconstrained scale and transform to the simplex with the softmax transformation.
-To maintain identifiability of this transformation, we fix $R_0' = 0$.
-We choose the prior for $S_0'$  so that $S_0$ has a mean of roughly  0.4, with the mass concentrated between 0.2 and 0.6.
-This range produces dynamics with recurring annual outbreaks and avoids the extremes where the disease either does not spread or spreads far too quickly and then dies out.
-This is also roughly consistent with \cite{Cardoso2011a} who report notified cases for Vitoria from 1995 to 2008. 
-They report that the 2008 outbreak (where our data begins) was driven by serotypes 2 and 3, and 2009 by serotype 2 only, with 13590 reported cases in previous serotype 2 years.
-Assuming 12 unreported cases for every reported case, 163,080 total cases, or about 50\% of the population immune to serotype 2 at the start of 2008.
-We choose the priors for $E_0'$ and $I_0'$ so that the initial sizes of these compartments were less than 100 individuals, but with a mode slightly off from zero.
-This was again chosen primarily to ensure sustained multi-annual transmission.
-\\
-\item \textbf{Mosquito population size and capture probability} ($\log(V_{N0}), \phi_q$): Wearing and Rohani \cite{Wearing2006} assume 2 mosquitoes per person in their model of dengue transmission, so we use that to inform our prior on $\log(V_{N0})$.
-Since $V_{N}$ has been scaled by the human population size, we center the prior for $\log(V_{N0})$ at $\log(2) = 0.7$ and set the variance to concentrate mass between 0.5 and 5 mosquitoes per person.
-Maybe see Focks (2000) for further justification of range of mosquitoes/person.
-The per-trap mosquito capture probability, $\phi_q$ follows $V_{N0}$ closely and scales the estimated mosquito abundance down to magnitude of the trap counts.
-Since $ E[q_t] = \phi_q \tau V_{Nt} N$, we set the prior on $\log(\phi_q)$ to have a mean of $\log(q_1) - \log(\tau_1) - \log(V_{N0}) - \log(N) = -13$ and a variance to match the prior on $\log(V_{N0})$.
-\\
-\item \textbf{Overdispersion parameters} ($\eta_y, \eta_q$): The negative binomial likelihoods were parameterized as:
-\begin{equation}
-\text{NegBin}(y | \mu, \eta) = {y + 1/\eta -1 \choose y}\left(\frac{\mu\eta}{1 + \mu\eta}\right)^y\left(\frac{1}{1+\eta\mu}\right)^{1/\eta}
-\end{equation}
-such that
-\begin{align}
-E[y] & = \mu\\
-\text{Var}[y] &= \mu + \eta\mu^2
-\end{align}
-Thus, $\eta = 0$ corresponds to Poisson, and $\eta > 0$ leads to overdispersion.
-We put a relatively conservative truncated normal prior on both $\eta_y$ and $\eta_q$ that concentrates mass near zero but allows for larger values.
-\\
-\item \textbf{Regression parameters} ($\mathbf{\beta}$, $\mathbf{\alpha}$): Though the autoregressive framework requires that $|\beta| < 1$, for computational efficiency, we avoid hard bounds and instead use a normal prior centered at zero, with variance such that the majority of mass falls between -1 and 1.
-\\
-\item \textbf{Process variances} ($\sigma_{\epsilon}$, $\sigma_{\xi}$):
-\end{itemize}
-
-\subsection*{Fixed parameters}
-
-\begin{itemize}
-\item \textbf{Transmission rate} ($\alpha$): The transmission rate only appears multiplied by a term from the mosquito model (either $V_I$ or $V_S$) and is thus effectively non-identifiable when we are also estimating the overall size of the mosquito population.
-Because of this, we fix $\alpha$ at a reasonable value from the literature (though our results should not be sensitive to the exact value used).
-The transmission rate is the product of the mosquito feeding rate and the probability of transmission given feeding.
-Scott \textit{et al.} \cite{Scott2000} estimate that \textit{Aedes aegypti} females take 0.63 blood meals per day in Puerto Rico, and 0.76 blood meals per day in Thailand.
-We take the average of these and set the biting rate at 0.69 per day, or 4.87 per week.
-There are relatively few reliable estimates of the transmission probability of DENV, but Lambrechts \textit{et al.} \cite{Lambrechts2011} estimate transmission probabilities close to one  for several related flaviviruses at 25 degrees C (roughly the mean temperature for Vitoria).
-\\
-\item \textbf{Reporting probability} ($\phi_y$): Silva \textit{et al.} \cite{Silva2016} estimate that there are 12 cases of dengue for every case reported to Brazil's Notifiable Diseases Information System (SINAN, the source of our data), so we fix $\phi_y = 1/12$.
-\\
-\item \textbf{Extrinsic incubation period} ($1/\rho_{vt}$): Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito  (the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
-To account for this, we make $1/\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground (URL/citation).
-We use the temperature relationship from the recent meta-analysis of \cite{Chan2012}, which incorporates a large number of records and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
-We use a slightly more conservative fitted relationship from a log-normal model that excludes right-censored data and produces a less extreme temperature dependence:
-\begin{align}
-\frac{1}{\rho_v} &= \exp \left(\mu + \frac{1}{14}\right)\\
-\mu & = \exp(1.9 - 0.04 T)
-\end{align}
-This relationship is similar to those reported in \cite{Focks1995} and \cite{Tjaden2013}, though there is uncertainty about the exact shape of this relationship.
-\end{itemize}
-
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 HMC Diagnostics.}
+\paragraph*{S1 Text.}
 \label{S1_Diag}
-{\bf Trace plots and convergence tests.}
+{\bf Full description of model and priors.}
 

### 2017-05-23 Manuscript/supplement.tex
Separating supplement into own document
And updating description of model structure to reflect seasonally forced emergence rate and latent stochastic death rate.

\documentclass[12pt,letterpaper]{article}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\title{Supplementary material}
\author{Clinton Leach}

\begin{document}

\maketitle

\section*{Full model}

\begin{align}
y_t & \sim \text{NegBin}(\phi_y  (X_t - X_{t-1})N, \eta_y)\\
q_t & \sim \text{NegBin}((V_{Ct} - V_{Ct-1})N, \eta_q)\\
\\ 
\frac{dX}{dt} &= \rho E \\
\frac{dS}{dt} &= b - bS - \lambda V_{I}S + \delta R\\
\frac{dE}{dt} &= \lambda V_{I}S - (\rho + b)E\\
\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
\frac{dR}{dt} &= \gamma I - (\delta + b)R\\
\\
\frac{dV_S}{dt} & = b_t - \lambda I V_S - (d_t + \phi_q \tau_t) V_S \\
\frac{dV_E}{dt} &= \lambda IV_S - (\rho_{vt} + d_t + \phi_q \tau_t)V_E\\
\frac{dV_I}{dt} &= \rho_{vt} V_E - (d_t + \phi_q \tau_t) V_I\\
\frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
V_N &= V_S + V_E + V_I\\
\\
\log(b_t) & = \alpha_{0} + \alpha_1 \sin(2\pi t / 52) + \alpha_2 \cos(2\pi t / 52) \\
\log(d_{t}) & = \beta_{0} + \beta_1 \sin(2\pi t / 52) + \beta_2 \cos(2\pi t / 52) + \epsilon_t\\
\epsilon_t & \sim \text{Normal}(0, \sigma)
\end{align}

\begin{align}
\alpha_0 & \sim \text{Normal}(-1, 1)\\ 
\alpha_{1,2} & \sim \text{Normal}(0, 2)\\
\beta_0 & \sim \text{Normal}(0.39, 0.12)\\
\beta_{1,2} & \sim \text{Normal}(0, 0.2)\\
\sigma & \sim \text{Normal}(0, 1)\\
-\log(\phi_q) & \sim \text{Half-normal}(13, 0.5)\\
\eta_y &\sim \text{Half-normal}(0, 5)\\
\eta_q & \sim \text{Half-normal}(0, 5)\\
T_{\rho} & \sim \text{Gamma}(8.3, 8.3)\\
\rho & = (0.87 T_{\rho})^{-1}\\
T_{\delta} & \sim \text{Gamma}(10, 10)\\
\delta & = (97 T_{\delta})^{-1}\\
\gamma' & \sim \text{Gamma}(100, 100)\\
\gamma & = 3.5\gamma'\\
S_0' & \sim \text{Normal}(-0.4, 0.2)\\
E_0' & \sim \text{Normal}(-9, 0.6)\\
I_0' & \sim \text{Normal}(-9, 0.6)\\
R_0' & = 0\\
S_0 & = \frac{\exp(S_0)}{\exp(S_0' + E_0' + I_0' + R_0')} \\
\end{align}

\subsection*{Details of prior specification}

\begin{itemize}
\item \textbf{Period of cross-immunity} ($T_{\delta}$): Reich \emph{et al.} \cite{Reich2013} estimate a mean of 1.88 years (97 weeks) with a 95\% confidence interval of (0.88, 4.31).
We specify a gamma prior on the multiplicative scale to have a mean of one and 0.01 and 0.99 quantiles at roughly $0.88/1.88 = 0.47$ and $4.31/1.88 = 2.3$,, respectively:
\begin{equation}
T_{\delta} \sim \text{Gamma}(10, 10)
\end{equation}
where $\delta = (97 T_\delta)^{-1}$.
Note that we use a slightly more conservative interval than Reich \emph{et al.} to avoid numerical instability at extreme values.
\\
\item \textbf{Intrinsic incubation period} ($T_{\rho}$): Chan and Johansson \cite{Chan2012} estimate a mean of 6.1 days with a 95\% credible interval of (3, 10).
Thus, we specify a gamma prior on the multiplicative scale to have a mean of one, and 0.05 and 0.95 quantiles of $3 / 6.1 = 0.49$ and $10 / 6.1 = 1.6$, respectively:
\begin{equation}
T_{\rho} \sim \text{Gamma}(8.3, 8.3)
\end{equation}
where $\rho = (0.87 T_{\rho})^{-1}$.
\\
\item \textbf{Decay rate of infectiousness} ($\gamma$): Nguyet \emph{et al.} \cite{Nguyet2013} find that there is very little transmission of dengue from humans to mosquitoes after 5-6 days since the onset of fever (with roughly an extra of infectiousness prior to onset of fever).
In reality, an individual's infectiousness decays with time, but the ODE model assumes infectiousness is constant for the (exponentially distributed) duration of the infection period.    
To parameterize the model then, we assume a rough equivalence between the probability of infection from a single individual at a given time and the proportion of individuals still infectious at a given time.
Thus we choose our prior for $\gamma$ to keep the time at which only 5\% of individuals are still infectious between 5 and 7 days, with a mean 5\% threshold of 6 days.
Again we specify the prior on the multiplicative scale to have a mean of one and 0.05 and 0.95 quantiles at $2.99/3.5 = 0.85$ and $4.19 / 3.5 = 1.2$, respectively:
\begin{equation}
\gamma' \sim \text{Gamma}(100, 100)
\end{equation}
where $\gamma = 3.5\gamma'$.
\\
\item \textbf{Mean mosquito mortality rate} ($\beta_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $\beta_0$.  
We use nonlinear least squares to fit the exponential cdf with rate $d$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d = 1.47$.
We use the same procedure to estimate $d$ for the 0.25 and 0.75 quantiles of the estimated mortality curves (assuming symmetry around the mean), which gives 1.6 and 1.37.
We then specify the prior on $beta_0$ to induce the above quantiles on $d$, which gives us:
\begin{equation}
\beta_0 \sim \text{Normal}(0.39, 0.12).
\end{equation}
\\
\item \textbf{Mean mosquito emergence rate} ($\alpha_0$):
\\
\item \textbf{Regression parameters} ($\alpha_1,\alpha_2,\beta_1,\beta_2$): 
\\
\item \textbf{Death rate variance} ($\sigma$):
\\
\item \textbf{Initial conditions} ($S_0, E_0, I_0, R_0$): The SEIRS model assumes a constant population size and is scaled so that $S_0+E_0+I_0+R_0 = 1$.
We specify our prior on an unconstrained scale and transform to the simplex with the softmax transformation.
To maintain identifiability of this transformation, we fix $R_0' = 0$.
We choose the prior for $S_0'$  so that $S_0$ has a mean of roughly  0.4, with the mass concentrated between 0.2 and 0.6.
This range produces dynamics with recurring annual outbreaks and avoids the extremes where the disease either does not spread or spreads far too quickly and then dies out.
This is also roughly consistent with \cite{Cardoso2011a} who report notified cases for Vitoria from 1995 to 2008. 
They report that the 2008 outbreak (where our data begins) was driven by serotypes 2 and 3, and 2009 by serotype 2 only, with 13590 reported cases in previous serotype 2 years.
Assuming 12 unreported cases for every reported case, 163,080 total cases, or about 50\% of the population immune to serotype 2 at the start of 2008.
We choose the priors for $E_0'$ and $I_0'$ so that the initial sizes of these compartments were less than 100 individuals, but with a mode slightly off from zero.
This was again chosen primarily to ensure sustained multi-annual transmission.
\\
\item \textbf{Mosquito population size and capture probability} ($\log(V_{N0}), \phi_q$): Wearing and Rohani \cite{Wearing2006} assume 2 mosquitoes per person in their model of dengue transmission, so we use that to inform our prior on $\log(V_{N0})$.
Since $V_{N}$ has been scaled by the human population size, we center the prior for $\log(V_{N0})$ at $\log(2) = 0.7$ and set the variance to concentrate mass between 0.5 and 5 mosquitoes per person.
Maybe see Focks (2000) for further justification of range of mosquitoes/person.
The per-trap mosquito capture probability, $\phi_q$ follows $V_{N0}$ closely and scales the estimated mosquito abundance down to magnitude of the trap counts.
Since $ E[q_t] = \phi_q \tau V_{Nt} N$, we set the prior on $\log(\phi_q)$ to have a mean of $\log(q_1) - \log(\tau_1) - \log(V_{N0}) - \log(N) = -13$ and a variance to match the prior on $\log(V_{N0})$.
\\
\item \textbf{Overdispersion parameters} ($\eta_y, \eta_q$): The negative binomial likelihoods were parameterized as:
\begin{equation}
\text{NegBin}(y | \mu, \eta) = {y + 1/\eta -1 \choose y}\left(\frac{\mu\eta}{1 + \mu\eta}\right)^y\left(\frac{1}{1+\eta\mu}\right)^{1/\eta}
\end{equation}
such that
\begin{align}
E[y] & = \mu\\
\text{Var}[y] &= \mu + \eta\mu^2
\end{align}
Thus, $\eta = 0$ corresponds to Poisson, and $\eta > 0$ leads to overdispersion.
We put a relatively conservative truncated normal prior on both $\eta_y$ and $\eta_q$ that concentrates mass near zero but allows for larger values.
\end{itemize}

\subsection*{Fixed parameters}

\begin{itemize}
\item \textbf{Transmission rate} ($\lambda$): The transmission rate only appears multiplied by a term from the mosquito model (either $V_I$ or $V_S$) and is thus effectively non-identifiable when we are also estimating the overall size of the mosquito population.
Because of this, we fix $\lambda$ at a reasonable value from the literature (though our results should not be sensitive to the exact value used).
The transmission rate is the product of the mosquito feeding rate and the probability of transmission given feeding.
Scott \textit{et al.} \cite{Scott2000} estimate that \textit{Aedes aegypti} females take 0.63 blood meals per day in Puerto Rico, and 0.76 blood meals per day in Thailand.
We take the average of these and set the biting rate at 0.69 per day, or 4.87 per week.
There are relatively few reliable estimates of the transmission probability of DENV, but Lambrechts \textit{et al.} \cite{Lambrechts2011} estimate transmission probabilities close to one  for several related flaviviruses at 25 degrees C (roughly the mean temperature for Vitoria).
\\
\item \textbf{Reporting probability} ($\phi_y$): Silva \textit{et al.} \cite{Silva2016} estimate that there are 12 cases of dengue for every case reported to Brazil's Notifiable Diseases Information System (SINAN, the source of our data), so we fix $\phi_y = 1/12$.
\\
\item \textbf{Extrinsic incubation period} ($1/\rho_{vt}$): Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito  (the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
To account for this, we make $1/\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground (URL/citation).
We use the temperature relationship from the recent meta-analysis of \cite{Chan2012}, which incorporates a large number of records and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
We use a slightly more conservative fitted relationship from a log-normal model that excludes right-censored data and produces a less extreme temperature dependence:
\begin{align}
\frac{1}{\rho_v} &= \exp \left(\mu + \frac{1}{14}\right)\\
\mu & = \exp(1.9 - 0.04 T)
\end{align}
This relationship is similar to those reported in \cite{Focks1995} and \cite{Tjaden2013}, though there is uncertainty about the exact shape of this relationship.
\end{itemize}

\bibliographystyle{plos2015}
\bibliography{dengue}

\end{document}

### 2017-05-23 Analysis/citystan.r
Loading lubridate package

@@ -1,6 +1,7 @@
 library(rstan)
 library(plyr)
 library(magrittr)
+library(lubridate)
 
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())

### 2017-05-23 Analysis/plots.r
Figures for prediction and death rate estimation

@@ -6,7 +6,7 @@ library(magrittr)
 library(ggplot2)
 library(gridExtra)
 
-# Loading and processing data
+# Loading and processing mosquito and case data
 tseries <- read.csv("Data/Vitoria.data.csv")
 
 post <- ddply(tseries, .(tot.week), summarise,
@@ -16,8 +16,19 @@ post <- ddply(tseries, .(tot.week), summarise,
               week = unique(Week),
               year = unique(Year))
 
+# Loading and processing the weather data
+weather <- read.csv("Data/Vitoria.weather.csv") %>% 
+  mutate(date = ymd(BRST), year = year(date), week = week(date))
+
+weather <- subset(weather, date < date[1] + weeks(243))
+weather$tot.week <- rep(c(1:243), each = 7)
+
+covars <- ddply(weather, .(tot.week), summarise,
+                temp = mean(Mean.TemperatureC, na.rm = T),
+                humidity = mean(Mean.Humidity, na.rm = T))
+
 # Loading MCMC results
-sim <- readRDS("Results/mcmc.rds")
+sim <- readRDS("Results/seasonal.rds")
 
 #===============================================================================
 # Figure 1: Observed and estimated time series
@@ -82,7 +93,7 @@ ggplot(post, aes(tot.week, b)) +
   scale_x_continuous(expand = c(0, 1)) +
   ylab("mosquito net emergence rate") +
   xlab("week") +
-  ggtitle("A")
+  ggtitle("A") -> fig2.a
 
 ggplot(post, aes(tot.week, d)) + 
   geom_ribbon(aes(ymin = dmin, ymax = dmax), fill = "grey70") +
@@ -92,36 +103,63 @@ ggplot(post, aes(tot.week, d)) +
   scale_x_continuous(expand = c(0, 1)) +
   ylab("mosquito per-capita death rate") +
   xlab("week") +
-  ggtitle("B")
+  ggtitle("B") -> fig2.b
 
-#===============================================================================
-# Figure 2: weekly posterior predictive p-value for replicate mosquito
-# surveillance data
-
-# Extracting full MCMC chain of replicate data
-mcmc <- rstan::extract(sim, "q_hat", permute = T)[[1]]
-nmcmc <- dim(mcmc)[1]
-
-# Computing P(q^{rep}_t > q_t)
-for(i in 1:243){
[…]

### 2017-05-24 Analysis/plots.r
Plotting epsilon along with dv

@@ -3,8 +3,10 @@
 library(rstan)
 library(plyr)
 library(magrittr)
+library(lubridate)
 library(ggplot2)
 library(gridExtra)
+
 
 # Loading and processing mosquito and case data
 tseries <- read.csv("Data/Vitoria.data.csv")
@@ -28,7 +30,7 @@ covars <- ddply(weather, .(tot.week), summarise,
                 humidity = mean(Mean.Humidity, na.rm = T))
 
 # Loading MCMC results
-sim <- readRDS("Results/seasonal.rds")
+sim <- readRDS("Results/covars.rds")
 
 #===============================================================================
 # Figure 1: Observed and estimated time series
@@ -79,21 +81,11 @@ dev.off()
 #===============================================================================
 # Figure 2: estimated demographic rates
 
-b <- rstan::extract(sim, "bv", permute = T)[[1]] %>% apply(2, quantile, c(0.1, 0.5, 0.9))
 d <- rstan::extract(sim, "dv", permute = T)[[1]] %>% apply(2, quantile, c(0.1, 0.5, 0.9))
+epsilon <- rstan::extract(sim, "z_d", permute = T)[[1]] %>% apply(2, quantile, c(0.1, 0.5, 0.9))
 
-post[, c("bmin", "b", "bmax")] <- t(b)
 post[, c("dmin", "d", "dmax")] <- t(d)
-
-ggplot(post, aes(tot.week, b)) + 
-  geom_ribbon(aes(ymin = bmin, ymax = bmax), fill = "grey70") +
-  geom_line() + 
-  theme_classic() +
-  scale_y_continuous(expand = c(0.05, 0)) + 
-  scale_x_continuous(expand = c(0, 1)) +
-  ylab("mosquito net emergence rate") +
-  xlab("week") +
-  ggtitle("A") -> fig2.a
+post[, c("epsmin", "eps", "epsmax")] <- t(epsilon)
 
 ggplot(post, aes(tot.week, d)) + 
   geom_ribbon(aes(ymin = dmin, ymax = dmax), fill = "grey70") +
@@ -102,6 +94,17 @@ ggplot(post, aes(tot.week, d)) +
   scale_y_continuous(expand = c(0.05, 0)) + 
   scale_x_continuous(expand = c(0, 1)) +
   ylab("mosquito per-capita death rate") +
+  xlab("week") +
[…]

### 2017-05-24 Manuscript/ms.tex
Updating and cleaning methods

@@
 Vitoria is a coastal city and the capital of the state of Espirito Santo, Brazil, with a population of 327,801 in the city proper, and an additional 1.5 million people in the greater metropolitan area.
-The city has an area of approximately 93 sq. km., with most of that area taken up by a large island surrounded by the river Rio Santa Maria and bay Baia de Vitoria.
-
-Since 2008, the city has monitored mosquito abundance using 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a grid across the city.
-Each trap is checked weekly, providing us with 243 weeks (2008 through week 34 of 2012) of counts of gravid female \emph{Aedes aegypti}.
+Since 2008, the company Ecovec has monitored mosquito abundance for the city using approximately 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a grid across the city.
+Each trap is checked weekly and the mosquitoes inside counted and identified, providing us with 243 weeks (2008 through week 34 of 2012) of counts of gravid female \emph{Aedes aegypti}.
 Dengue is a mandatory notifiable disease, and thus the city's Ministry of Health Secretary maintains a database of weekly notified probable dengue cases (i.e.
-medical care sought for dengue-like symptoms).
-
-Governador Valladares is also a city for which we have data.
-
-Sete Lagoas is also a city for which we have data.
+medical care sought for dengue-like symptoms) for this same time period.
 
@@
 The strength and duration of these different inter-serotype interactions are not well understood, though different models suggest that temporary cross-immunity alone (without ADE) is sufficient to reproduce observed multi-annual dynamics in Thailand \cite{Wearing2006,Reich2013}.
-Similarly, \cite{Aguiar2013} find that capturing both primary and secondary infections and the period of cross-immunity is critical, but that explicitly including all four serotypes (at great cost to model complexity) does not perform much better than a two serotype model, and indeed that the number of hospital admissions caused by a third or fourth infection is very low \cite{Aguiar2011a}.
+Similarly, \cite{Aguiar2013} find that capturing both primary and secondary infections and the period of cross-immunity is critical, but that explicitly including all four serotypes (at great cost to model complexity) does not perform much better than a two serotype model.
 
@@
 Following a latent period ($\frac{1}{\rho}$), exposed individuals become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
-Infectious individuals recover at rate $\gamma$ and remain immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
+Infectious individuals recover at rate $\gamma$ and subsequenty remain immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
 
 Similarly, susceptible mosquitoes ($V_S$) become exposed by biting infectious humans and pass through a temperature-dependent latent period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
-Total mosquito population size ($V_N$) is controlled by a time-varying net emergence rate, $b_t$, and captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.
+Total mosquito population size ($V_N$) is controlled by a periodically varying net emergence rate, $b_t$, and a stochastic, weather dependent death rate, $d_t$.
+Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.
 
@@
 \frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
-V_N &= V_S + V_E + V_I
-\end{align}
-
-The mosquito demographic parameters (net emergence rate, $b_t$, and per capita death rate, $d_t$) are modeled with a periodic regression component to account for unmodeled seasonal forcing (e.g.
-from temperature, precipitation, etc.):
-\begin{align}
-\log(b_t) & = \alpha_{0} + \alpha_1 \sin(2\pi t / 52) + \alpha_2 \cos(2\pi t / 52) \\
-\log(d_{t}) & = \beta_{0} + \beta_1 \sin(2\pi t / 52) + \beta_2 \cos(2\pi t / 52) + \epsilon_t,
+V_N &= V_S + V_E + V_I,
 \end{align}
 where
+\begin{equation}
+\log(b_t) = \alpha_{0} + \alpha_1 \sin(2\pi t / 52) + \alpha_2 \cos(2\pi t / 52),
+\end{equation}
+and
 \begin{align}
-\epsilon_t & \sim \text{Normal}(0, \sigma)\\
+\log(d_{t}) & = \beta_{0} + X\beta + \epsilon_t\\
+\epsilon_t & \sim \text{Normal}(0, \sigma).
 \end{align}
@@
 
-To connect the differential equation model the the observed case reports, we add an extra state, $X$, that collects the cumulative number of transitions from the exposed to infectious class (assuming that case reporting coincides with the onset of symptoms).
+To connect the differential equation model the the observed case reports, we add an extra state, $C$, that collects the cumulative number of transitions from the exposed to infectious class (assuming that case reporting coincides with the onset of symptoms).
 We then model the number of new cases reported in week $t$ ($y_t$) as:
 \begin{equation}
-y_t  \sim \text{NegBin}(\phi_y (X_t - X_{t-1}), \eta_y),
+y_t  \sim \text{NegBin}(\phi_y (C_t - C_{t-1}), \eta_y),
 \end{equation}
-where $\phi_y$ is the reporting probability, $X_t - X_{t-1}$ is the number of new infectious individuals in week $t$, and $\eta_y$ controls the overdispersion relative to the Poisson.
+where $\phi_y$ is the reporting probability, $C_t - C_{t-1}$ is the number of new infectious individuals in week $t$, and $\eta_y$ controls the overdispersion relative to the Poisson.
 
@@
 The population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
-To maintain identifiability, the transmission rate and case reporting probability are also fixed at literature values.
-Lastly, the extrinsic incubation period in mosquitoes is modeled as a function of weekly mean temperature and forced with data pulled from WeatherUnderground.
+To maintain identifiability, the transmission rate ($\lambda$) and case reporting probability ($\phi_y$) are also fixed at literature values.
+Lastly, the extrinsic incubation period in mosquitoes is modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground.
 
-The remaining parameters include the epidemiological parameters that control the flows between the compartments of the differential equation model ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the regression parameters and process noise in the models for mosquito growth and death rates ($\mathbf{\alpha}$, $\mathbf{\beta}$, and $\sigma$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
+The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the regression parameters controlling mean, phase, and amplitude of the emergence rate ($\alpha_0,\alpha_1,\alpha_2$), the regression and variance parameters for the mosquito death rate process ($\beta_0,\mathbf{\beta},\sigma$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
 Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table \ref{fixedparms} for means and Supplemental Material for detailed explanations).
@@
 $1/d$ & Human life-span & 76 years & Vitoria Census\\
-$\alpha$ & Transmission rate & 4.87 week$^{-1}$ & \cite{Scott2000}\\
+$\lambda$ & Transmission rate & 4.87 week$^{-1}$ & \cite{Scott2000}\\
 $\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
 $1/\rho_{vt}$ & Extrinsic incubation period & $\exp \left(\exp(1.9 - 0.04 T_t) + \frac{1}{14}\right)$ & \cite{Chan2012}\\
-$\sigma$  & Standard dev. of AR(1) process & 0.2 & \\
 $V_{E0}$ & Initial exposed mosquitoes &  0 & \\
@@
 \hline
-$d_v$ & Vector death rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
+$\exp(\beta_0)$ & Mean vector death rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
+$\exp(\alpha_0)$ & Mean vector net emergence rate & ?? & \\
 $1/\rho$ & Latent period in host & 0.87 weeks  & \cite{Chan2012}\\
@@
 $I_0$ & Proportion initially infectious & $8\times 10 ^ {-5}$ & \\
-$\phi_q$ & Per-trap mosquito capture probability & $2 \times 10^{-6}$ & 
+$\phi_q$ & Per-trap mosquito capture rate & $2 \times 10^{-6}$ week$^{-1}$ & 
 \end{tabular}
@@
 
+\subsection*{Model validation}
+
+We evaluate the model's predictive ability by fitting to the first 191 weeks of data (up to the trough following the fourth outbreak), and predicting case reports for the subsequent year.
+In order to make these predictions, we project weather covariates into the future using weekly averages of the first 191 weeks.
+
 \subsection*{Implementation}
 
-Samples from the posterior distribution were generated using Hamiltonian Monte Carlo (CITATIONS) implemented in the rstan package for R. 
+Samples from the posterior distribution were generated using Hamiltonian Monte Carlo (CITATIONS) implemented in the rstan package for R (CITATIONS). 
 We ran 3 chains with different starting values for 10,000 iterations each, discarding the first 5,000 as burn-in.
 Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.
-
-\subsection*{Posterior predictive checks}
-
-We evaluate the goodness-of-fit of the estimated mosquito abundance time series through the use of posterior predictive checks \cite{Gelman1996,Gelman2003}.
-Specifically, for each iteration, $i$, of MCMC, we sample replicate mosquito surveillance data:
-\begin{equation}
-q^{rep(i)}_t \sim \text{NegBin}(\phi^{(i)}_q \tau_t V^{(i)}_{Nt}, \eta^{(i)}_q)
-\end{equation}
-and compute
-\begin{equation}
-p_t = P(q^{rep}_t > q_t) = \frac{1}{M}\sum I(q^{rep(i)}_t > q_t).
-\end{equation}
-If the reconstructed mosquito trajectories are a good fit for the observed mosquito trap counts, then we would expect the time series $p_t$ to be random in time (CITATION?).
-Any systematic trends in $p_t$ indicate lack-of-fit.
-
-In addition to looking at weekly comparisons, we also develop posterior checks to compare the temporal variability and correlation structures of the observed and estimated time series.
-To test whether the modeled mosquito time series capture the seasonal features of the observed mosquito data, we compute the posterior auto-correlation function of the replicate data:
-\begin{equation}
-R^{rep(i)}(l) = \text{corr}(q^{rep(i)}_t, q^{rep(i)}_{t-l}),
-\end{equation}
-for lags ranging from 1 to 55 weeks.
-We also compute posterior distributions of $\min(q^{rep}_t)$ , $\max(q^{rep}_t)$, and $\sigma(q^{rep}_t - q^{rep}_{t-1})$ as measures of the smoothness of the modeled process.
 

### 2017-05-25 Analysis/predict.r
Tweaking priors and adding covars in an attempt to improve predictions

@@ -44,6 +44,16 @@ annual <- ddply(weather, .(week), summarise,
 predvars <- rbind(covars[, -1],
                   annual[week(start + weeks((T_fit + 1):243)), -1])
 
+predvars <- cbind(predvars, matrix(c(sin(2 * pi * c(1:243) / 52), 
+                                 cos(2 * pi * c(1:243) / 52),
+                                 sin(4 * pi * c(1:243) / 52), 
+                                 cos(4 * pi * c(1:243) / 52),
+                                 sin(8 * pi * c(1:243) / 52), 
+                                 cos(8 * pi * c(1:243) / 52),
+                                 sin(16 * pi * c(1:243) / 52), 
+                                 cos(16 * pi * c(1:243) / 52)), 
+                               ncol = 8))
+
 rov <- 7 / exp(exp(1.9 - 0.04 * predvars$temp) + 1 / (2 * 7))
 
 #===============================================================================
@@ -51,7 +61,7 @@ rov <- 7 / exp(exp(1.9 - 0.04 * predvars$temp) + 1 / (2 * 7))
 
 dat.stan <- list(T = T_fit,
                  T_pred = T_pred,
-                 D = 2,
+                 D = dim(predvars)[2],
                  y = head(obs, T_fit),
                  q = head(q, T_fit),
                  tau = head(tau, T_fit),
@@ -73,14 +83,14 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   alpha0 = 1,
                   alpha = c(0, 0),
                   beta0 = 0.39,
-                  beta = c(0, 0),
+                  beta = rep(0, 10),
                   sigmad = 0.1,
                   z_d = rep(0, T_fit)))
 
 fit <- stan(file = "Code/predict.stan", 
             data = dat.stan, 
             init = inits, 
-            iter = 500, 
+            iter = 5000, 
             chains = 1,
             control = list(adapt_delta = 0.9))
 

### 2017-05-26 Analysis/citystan.r
Implementing two serotype model

@@ -31,13 +31,16 @@ covars <- ddply(weather, .(tot.week), summarise,
                 humidity = mean(Mean.Humidity, na.rm = T)
                 )
 
+covars <- covars[, -1]
+
 rov <- 7 / exp(exp(1.9 - 0.04 * covars$temp) + 1 / (2 * 7))
 
 #===============================================================================
 # Running stan
 
 dat.stan <- list(T = 243,
-                 D = 2,
+                 T_pred = 0,
+                 D = dim(covars)[2],
                  y = obs,
                  q = q,
                  tau = tau,
@@ -45,10 +48,10 @@ dat.stan <- list(T = 243,
                  sincos = matrix(c(sin(2 * pi * c(1:243) / 52), 
                                    cos(2 * pi * c(1:243) / 52)), 
                                  ncol = 2),
-                 covars = scale(covars[, -1]),
+                 covars = scale(covars),
                  pop = pop)
 
-inits = list(list(p0_raw = c(-0.4, -9, -9),
+inits = list(list(p0_raw = c(0.8, -9, -9, -11, 0.3, -11, -11,-11, -11, -2, 0.3, -9, -9),
                   log_phi_q = -13,
                   eta_inv_y = 5,
                   eta_inv_q = 5,
@@ -57,12 +60,12 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   delta_c = 1,
                   logNv0 = 0.7,
                   alpha0 = 1,
-                  alpha = c(0, 0),
+                  alpha = rep(0, 2),
                   beta0 = 0.39,
-                  beta = c(0, 0),
+                  beta = rep(0, 2),
                   sigmad = 0.1,
                   z_d = rep(0, 243)),
-             list(p0_raw = c(-0.4, -9, -9),
+             list(p0_raw = c(0.6, -9, -9, -11, 0.4, -11, -11,-11, -11, -2, 0.4, -9, -9),
                   log_phi_q = -13,
                   eta_inv_y = 10,
                   eta_inv_q = 10,
@@ -76,7 +79,7 @@ inits = list(list(p0_raw = c(-0.4, -9, -9),
                   beta = c(0.2, 0.1),
                   sigmad = 0.1,
                   z_d = rep(0, 243)),
-             list(p0_raw = c(-0.4, -9, -9),
+             list(p0_raw = c(0.5, -9, -9, -11, 0.3, -11, -11,-11, -11, -2, 0.3, -9, -9),
[…]

### 2017-07-03 Analysis/citystan.r
Code to run SEIRS seasonal AR model

@@ -8,95 +8,118 @@ options(mc.cores = parallel::detectCores())
 
 #===============================================================================
 # Loading data and setting up model
-  
+
 tseries <- read.csv("Data/Vitoria.data.csv")
-obs <- tapply(tseries$Cases, list(tseries$tot.week), sum, na.rm=TRUE)
+
+data <- ddply(tseries, .(tot.week), summarise,
+              obs = sum(Cases, na.rm = T),
+              q = sum(Mosquitoes, na.rm = T),
+              tau = sum(Trap, na.rm = T),
+              year = unique(Year),
+              week = unique(Week)
+              )
 
 # Population size
 pop <- 327801
 
-# Pull out mosquito counts and traps inspected
-q <- tapply(tseries$Mosquitoes, list(tseries$tot.week), sum, na.rm = TRUE)
-tau <- tapply(tseries$Trap, list(tseries$tot.week), sum, na.rm = TRUE)
-
-# Computing eip forcing
+# Wrangling weather data and computing EIP
 weather <- read.csv("Data/Vitoria.weather.csv") %>% 
   mutate(date = ymd(BRST), year = year(date), week = week(date))
 
 weather <- subset(weather, date < date[1] + weeks(243))
 weather$tot.week <- rep(c(1:243), each = 7)
 
-covars <- ddply(weather, .(tot.week), summarise,
-                temp = mean(Mean.TemperatureC, na.rm = T),
-                humidity = mean(Mean.Humidity, na.rm = T)
-                )
+temp <- ddply(weather, .(tot.week), summarise, temp = mean(Mean.TemperatureC, na.rm = T))
 
-covars <- covars[, -1]
-
-rov <- 7 / exp(exp(1.9 - 0.04 * covars$temp) + 1 / (2 * 7))
-
+rov <- 7 * exp(0.2 * temp$temp - 8)
 #===============================================================================
 # Running stan
 
 dat.stan <- list(T = 243,
                  T_pred = 0,
-                 D = dim(covars)[2],
-                 y = obs,
-                 q = q,
-                 tau = tau,
+                 y =data$ obs,
+                 q = data$q,
+                 tau = data$tau,
                  rov = rov,
-                 covars = scale(covars),
                  pop = pop)
 
-inits = list(list(p0_raw = c(0.6, -1, 0.3, -15, 0.2),
+inits = list(list(S0 = 0.4,
+                  E0 = 80,
+                  I0 = 40,
[…]

### 2017-07-05 Manuscript/ms.tex
Updates to intro and model structure

@@
 
-We know mosquitoes spread dengue, but mosquito abundance, as estimated from trap-based surveillance data, is not causally linked to human disease risk.
+Mosquito abundance alone is not a strong predictor of dengue fever risk, but provides information that, when coupled with case data, allows us to estimate mosquito demographic rates.
 
@@
 
-We know mosquitoes spread dengue, but mosquito abundance, as estimated from trap-based surveillance data, is not causally linked to human disease risk.
+Mosquitoes are dumb.
 
@@
 
-Dengue fever is a massive global public health burden, with millions of cases per year in Brazil alone (CITATION).  
+Dengue fever is a massive global public health burden, with millions of cases per year in Brazil alone \cite{Bhatt2013}.  
 Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
-Though eradication of \textit{Aedes aegypti} is infeasible, effectively targeted control interventions may still substantially reduce disease risk.
+To be effective, control efforts should be guided by consistent, high resolution mosquito surveillance \cite{Morrison2008}.
+The "MI-Dengue" system, implemented by the company Ecovec and deployed throughout cities in Brazil, uses a city-wide grid of mosquito sticky-traps (MosquiTRAPs) for spatial prioritization of source reduction and larvicide application \cite{Eiras2009}.
+Comparing cities with and without this system, \cite{Pepin2013} estimated that it prevented 27,191 cases of dengue fever from 2007 - 2011.
+Though this system has been effective, the mosquito indices and thresholds used to prioritize control efforts remain somewhat ad-hoc.
+As noted in \cite{Morrison2008}, entomological indices are only relevant in the context of local epidemiology, and thus control policies will be most effective when they synthesize both entomological and clinical surveillance.
 
-Cities deploy mosquito control based on trap-based mosquito surveillance data.
-For instance, the "MI-Dengue" system deployed throughout several cities in Brazil uses a grid of mosquito sticky-traps (MosquiTRAPs) for spatial prioritization of source reduction and larvicide application \cite{Eiras2009}.
-Comparing cities with and without this system, \cite{Pepin2013} estimated that it prevented X cases and saved X dollars in economic losses from 20XX - 20XX.
-Though this system has been effective, the mosquito indices and thresholds used to prioritize control efforts remain somewhat ad-hoc.
-Using such surveillance data to efficiently target limited control resources requires a more detailed understanding of the relationship between the components of mosquito abundance measured by traps and disease risk.
+Many of the attempts to link these two sources of information have found a weak relationship between mosquito surveillance disease risk (CITATIONS).
+These efforts have often relied on tests of the ability of mosquito abundance measures to predict human disease risk in regression models.
+For example, \cite{Pepin2015} found that models that included MosquiTRAP surveillance data failed to predict weekly dengue cases any better than models that included lagged case data alone.
+However, such tests fail to account for the mechanistic, non-linear relationship between mosquito population dynamics and dengue epidemiology.
+The cycle of transmission between humans and dengue is controlled not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period \cite{Achee2015}.
+In fact, the probability that an exposed mosquito will survive long enough to become infectious provides a more direct measure of transmission potential than mosquito abundance alone.
 
-Many of the attempts to establish this relationship statistically have relied on tests of the ability of mosquito abundance measures to predict human disease risk in regression models.
-For example, \cite{Pepin2015} found that models that included MosquiTRAP surveillance data failed to predict weekly dengue cases any better than models that include lagged case data alone.
-However, these models fail to account for the mechanistic, non-linear relationship between mosquito population dynamics and dengue epidemiology.
-In particular, the cycle of transmission between humans and dengue is controlled not just by mosquito population size, but also by the emergence of new susceptible adults and the mortality rate.
+Adult surveillance data (e.g. time series of mosquito trap counts) is insufficient on its own to estimate mosquito survival rates and thus transmission potential.
+However, we can borrow from the field of 'integrated population modeling' from wildlife ecology and use multiple sources of information to estimate mosquito demographic rates (CITATIONS).
+Since mosquito survival is a critical bottleneck in the transmission process, it can be informed from the observed transmission dynamics (i.e.
+reports of human cases), given an appropriate underlying mechanistic model.
 
-In this paper, we seek to connect mosquito surveillance data with human case reports through the use of a Bayesian, mechanistic model.
-We first propose a differential equation model for dengue epidemiology.
+In this paper, we propose a mechanistic transmission model that connects mosquito abundance and survival with dengue epidemiology and human case reports.
 We then embed this model into a Bayesian statistical framework that allows us to estimate latent time series of mosquito demographic parameters from time series of mosquito trap counts and reported cases of dengue fever.
-Lastly, we explore the ability of this fitted model to predict annual dengue outbreaks and its utility for targeting control efforts.
+Lastly, we explore the utility of these estimated transmission risks for guiding targeted control efforts relative to using mosquito abundance alone.
 
@@
 Following a latent period ($\frac{1}{\rho}$), exposed individuals become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
-Infectious individuals recover at rate $\gamma$ and subsequenty remain immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
+Infectious individuals recover at rate $\gamma$ and subsequently remain immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
 
 Similarly, susceptible mosquitoes ($V_S$) become exposed by biting infectious humans and pass through a temperature-dependent latent period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
-Total mosquito population size ($V_N$) is controlled by a periodically varying net emergence rate, $b_t$, and a stochastic, weather dependent death rate, $d_t$.
+Total mosquito population size ($V_N$) is controlled by stochastic, seasonally varying growth rate ($r_t$) and death rate ($d_t$).
 Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.
@@
 \frac{dR}{dt} &= \gamma I - (\delta + b)R\\
-\frac{dV_S}{dt} & = b_t - \lambda \frac{I}{N} V_S - (d_t + \phi_q \tau_t) V_S \\
+\frac{dV_N}{dt} & = r_t V_N \\
 \frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{vt} + d_t + \phi_q \tau_t)V_E\\
@@
 \frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
-V_N &= V_S + V_E + V_I,
+V_S &= V_N - V_E - V_I,
 \end{align}
 where
-\begin{equation}
-\log(b_t) = \alpha_{0} + \alpha_1 \sin(2\pi t / 52) + \alpha_2 \cos(2\pi t / 52),
-\end{equation}
+\begin{align}
+d_t & = \rho_{vt} \exp(-\nu_t)\\
+\nu_{t} & = \alpha_{\nu} \nu_{t-1} + \theta_{\nu} (\nu_{t-52} - \alpha_{\nu} \nu_{t-53}) +\epsilon_{\nu t}\\
+r_t &= \alpha_r r_{t-1} + \theta_r (r_{t-52} - \alpha_r r_{t-53}) + \epsilon_{rt},\\
+\end{align}
 and
 \begin{align}
-\log(d_{t}) & = \beta_{0} + X\beta + \epsilon_t\\
-\epsilon_t & \sim \text{Normal}(0, \sigma).
+\epsilon_{rt} & \sim \text{Normal}(0, \sigma_r)\\
+\epsilon_{\nu t} & \sim \text{Normal}(0, \sigma_{\nu}).
 \end{align}

### 2017-07-05 Manuscript/ms.tex
Cleaning up parameterization

@@
 
-Adult surveillance data (e.g. time series of mosquito trap counts) is insufficient on its own to estimate mosquito survival rates and thus transmission potential.
-However, we can borrow from the field of 'integrated population modeling' from wildlife ecology and use multiple sources of information to estimate mosquito demographic rates (CITATIONS).
+Adult \textit{Aedes aegypti} surveillance data (e.g. time series of mosquito trap counts) is insufficient on its own to estimate mosquito survival rates and thus transmission potential.
+However, we can borrow from the field of 'integrated population modeling' and combine multiple sources of information to estimate mosquito demographic rates (CITATIONS).
 Since mosquito survival is a critical bottleneck in the transmission process, it can be informed from the observed transmission dynamics (i.e.
@@
 
-In this paper, we propose a mechanistic transmission model that connects mosquito abundance and survival with dengue epidemiology and human case reports.
+In this paper, we propose a mechanistic model to connect mosquito abundance and survival with dengue epidemiology.
 We then embed this model into a Bayesian statistical framework that allows us to estimate latent time series of mosquito demographic parameters from time series of mosquito trap counts and reported cases of dengue fever.
@@
 \frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
-V_S &= V_N - V_E - V_I,
+V_S &= V_N - V_E - V_I.
 \end{align}
-where
+
+We model the weekly mosquito mortality rate, $d_{vt}$, relative to $\rho_{vt}$:
+\begin{equation}
+d_{vt} = \rho_{vt} \exp(-\mu - \nu_t).
+\end{equation}
+Assuming mosquito lifespan and latent periods are exponentially distributed (as the differential equation framework implicitly does), the probability that a newly exposed mosquito will survive until it becomes infectious is then
+\begin{equation}
+\psi_t = \frac{\rho_{vt}}{\rho_{vt} + d_{vt}} = \text{logit}^{-1}(\mu + \nu_t).
+\end{equation}
+We then model $\nu_t$ and the mosquito population growth rate, $r_t$, using a (Box-Jenkins?) seasonal auto-regressive model (CITATION):
 \begin{align}
-d_t & = \rho_{vt} \exp(-\nu_t)\\
 \nu_{t} & = \alpha_{\nu} \nu_{t-1} + \theta_{\nu} (\nu_{t-52} - \alpha_{\nu} \nu_{t-53}) +\epsilon_{\nu t}\\
@@
 \end{align}
-and
+where
 \begin{align}
@@
 
-Several of the parameters in this model are assumed to fixed and known. 
+Several of the parameters in this model are assumed to be fixed and known. 
 The population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
@@
 
-The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the regression parameters controlling mean, phase, and amplitude of the emergence rate ($\alpha_0,\alpha_1,\alpha_2$), the regression and variance parameters for the mosquito death rate process ($\beta_0,\mathbf{\beta},\sigma$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
+The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the latent mosquito process autoregressive ($\alpha_r$, $\alpha_{\nu}$, $\theta_r$, $\theta_{\nu}$) and variance ($\sigma_r$, $\sigma_{\nu}$) parameters, and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
 Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table \ref{fixedparms} for means and Supplemental Material for detailed explanations).
-
-For any given values of these parameters, the mosquito process model should be flexible enough to produce a good fit to the case reports data.
-As a result, we expect that these parameters will be informed primarily by the mosquito surveillance data (i.e.
-these parameters add degrees of freedom that let the mosquito process model fit the surveillance data better while maintaining a good fit to the case data).
 
@@
 $\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
-$1/\rho_{vt}$ & Extrinsic incubation period & $\exp \left(\exp(1.9 - 0.04 T_t) + \frac{1}{14}\right)$ & \cite{Chan2012}\\
+$1/\rho_{vt}$ & Extrinsic incubation period & $7\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
+$\exp(\mu)$ & Mean ratio of mortality to incubation rates & 3.6 & \cite{Chan2012,Brady2013} \\
 $V_{E0}$ & Initial exposed mosquitoes &  0 & \\
@@
 \hline
-$\exp(\beta_0)$ & Mean vector death rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
-$\exp(\alpha_0)$ & Mean vector net emergence rate & ?? & \\
 $1/\rho$ & Latent period in host & 0.87 weeks  & \cite{Chan2012}\\
@@
 \end{table}
-
-\subsection*{Model validation}
-
-We evaluate the model's predictive ability by fitting to the first 191 weeks of data (up to the trough following the fourth outbreak), and predicting case reports for the subsequent year.
-In order to make these predictions, we project weather covariates into the future using weekly averages of the first 191 weeks.
 
@@
 \item What do the estimated time series of demographic rates look like? How smooth are they? How strong is the periodic signal relative to noise?
-\item How far/how well can we make predictions?
-\item What do the estimated demographic rates tell us about efficient implementation of control?
+\item How do peaks in $\psi$ line up with peaks in $q_t$?
+\item How effective is control informed by $\psi$ compared to control informed by $q_t$?
 \end{itemize}

### 2017-07-07 Analysis/citystan.r
Rolling back to RW model

@@ -37,7 +37,7 @@ rov <- 7 * exp(0.2 * temp$temp - 8)
 
 dat.stan <- list(T = 243,
                  T_pred = 0,
-                 y =data$ obs,
+                 y =data$obs,
                  q = data$q,
                  tau = data$tau,
                  rov = rov,
@@ -53,10 +53,7 @@ inits = list(list(S0 = 0.4,
                   gamma_c = 1,
                   delta_c = 0.8,
                   logNv = 0.7,
-                  ar_psi = 0,
-                  ar_rv = 0,
-                  seas_psi = 0,
-                  seas_rv = 0,
+                  ar_rv = 0.8,
                   sigmapsi = 0.2,
                   sigmarv = 0.2,
                   eps_psi = rep(0, 243),
@@ -71,10 +68,7 @@ inits = list(list(S0 = 0.4,
                   gamma_c = 1.2,
                   delta_c = 1,
                   logNv = 1,
-                  ar_psi = 0.8,
-                  ar_rv = 0.8,
-                  seas_psi = 0.8,
-                  seas_rv = 0.8,
+                  ar_rv = 0.5,
                   sigmapsi = 0.5,
                   sigmarv = 0.5,
                   eps_psi = rep(0, 243),
@@ -89,10 +83,7 @@ inits = list(list(S0 = 0.4,
                   gamma_c = 1,
                   delta_c = 1,
                   logNv = 0.5,
-                  ar_psi = 0.5,
-                  ar_rv = 0.5,
-                  seas_psi = 0,
-                  seas_rv = 0,
+                  ar_rv = 0,
                   sigmapsi = 0.01,
                   sigmarv = 0.01,
                   eps_psi = rep(0, 243),
@@ -101,25 +92,30 @@ inits = list(list(S0 = 0.4,
 fit <- stan(file = "Code/seirs.stan", 
             data = dat.stan, 
             init = inits, 
-            iter = 5000, 
+            iter = 1000, 
             chains = 3,
             control = list(adapt_delta = 0.9,
                            max_treedepth = 12))
+
+pairs(fit, pars = c("eps_psi", "eps_rv", "rv", "dv", "psi_raw", "y_hat", "q_hat", 
[…]

### 2017-09-04 Manuscript/ms.tex
Updating ms and supp methods

@@
 Similarly, susceptible mosquitoes ($V_S$) become exposed by biting infectious humans and pass through a temperature-dependent latent period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
-Total mosquito population size ($V_N$) is controlled by stochastic, seasonally varying growth rate ($r_t$) and death rate ($d_t$).
+Total mosquito population size ($V_N$) is controlled by stochastic, seasonally varying growth rate ($r(t)$) and death rate ($d(t)$).
 Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.
@@
 \frac{dR}{dt} &= \gamma I - (\delta + b)R\\
-\frac{dV_N}{dt} & = r_t V_N \\
-\frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{vt} + d_t + \phi_q \tau_t)V_E\\
-\frac{dV_I}{dt} &= \rho_{vt} V_E - (d_t + \phi_q \tau_t) V_I\\
+\frac{dV_N}{dt} & = r V_N - \phi_q \tau_t V_N \\
+\frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{vt} + d + \phi_q \tau_t)V_E\\
+\frac{dV_I}{dt} &= \rho_{vt} V_E - (d + \phi_q \tau_t) V_I\\
 \frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
@@
 
-We model the weekly mosquito mortality rate, $d_{vt}$, relative to $\rho_{vt}$:
-\begin{equation}
-d_{vt} = \rho_{vt} \exp(-\mu - \nu_t).
-\end{equation}
-Assuming mosquito lifespan and latent periods are exponentially distributed (as the differential equation framework implicitly does), the probability that a newly exposed mosquito will survive until it becomes infectious is then
-\begin{equation}
-\psi_t = \frac{\rho_{vt}}{\rho_{vt} + d_{vt}} = \text{logit}^{-1}(\mu + \nu_t).
-\end{equation}
-We then model $\nu_t$ and the mosquito population growth rate, $r_t$, using a (Box-Jenkins?) seasonal auto-regressive model (CITATION):
+We model the centered and log-transformed mosquito mortality rate ($\nu$) and the per-capita growth rate ($r$) as forced harmonic oscillators with natural periods of one year:
 \begin{align}
-\nu_{t} & = \alpha_{\nu} \nu_{t-1} + \theta_{\nu} (\nu_{t-52} - \alpha_{\nu} \nu_{t-53}) +\epsilon_{\nu t}\\
-r_t &= \alpha_r r_{t-1} + \theta_r (r_{t-52} - \alpha_r r_{t-53}) + \epsilon_{rt},\\
+\frac{d^2\nu}{dt^2} &= -\omega^2 \nu + \epsilon_{\nu t}\\
+\frac{d^2 r}{dt^2} &= -\omega^2 r + \epsilon_{rt},
 \end{align}
-where
+where $\omega = 2\pi / 52$, $d(t) = d_0 \exp(\nu(t)$, and
 \begin{align}
-\epsilon_{rt} & \sim \text{Normal}(0, \sigma_r)\\
-\epsilon_{\nu t} & \sim \text{Normal}(0, \sigma_{\nu}).
+\epsilon_{\nu i} & \sim \text{Normal}(0, \sigma_{\nu})\\
+\epsilon_{ri} & \sim \text{Normal}(0, \sigma_r),
 \end{align}
+for $i = 1 \dots 243$.
+These stochastically-forced harmonic oscillators provide a flexible framework for generating smooth seasonal oscillations in the latent mosquito processes (Ramsay citation?).
 
@@
 
-Several of the parameters in this model are assumed to be fixed and known. 
+Several of the parameters in this model are assumed to be fixed and known (\ref{parameters}).
 The population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
@@
 
-The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}$), the latent mosquito process autoregressive ($\alpha_r$, $\alpha_{\nu}$, $\theta_r$, $\theta_{\nu}$) and variance ($\sigma_r$, $\sigma_{\nu}$) parameters, and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
-Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table \ref{fixedparms} for means and Supplemental Material for detailed explanations).
+The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
+Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table \ref{parameters} for means and Supplemental Material for detailed explanations).
 
 \begin{table}[!ht]
-\label{fixedparms}
+\label{parameters}
 \begin{adjustwidth}{-2.25in}{0in} 
@@
 $1/\rho_{vt}$ & Extrinsic incubation period & $7\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
-$\exp(\mu)$ & Mean ratio of mortality to incubation rates & 3.6 & \cite{Chan2012,Brady2013} \\
+$\exp(\mu)$ & Mean mosquito death rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
 $V_{E0}$ & Initial exposed mosquitoes &  0 & \\
@@
 Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.
+
+\subsection*{Control simulations}
+
+Given samples from the posterior as obtained above, we can simulate the efficacy of different control strategies.
+In particular, we investigate the optimal timing of a single pulse of control within the year (e.g., if the city only has enough resource to fog for adult mosquitoes once a year, in what week is that control going to prevent the most cases?).
+To obtain the posterior distribution of the number of cases prevented by implementing control in week $i$ each year, for each posterior sample $k$, we simulate an increase in the latent mosquito death rate during each week $i$, e.g.,
+$\hat{d}^{(k)}_{yi} = (1 + \Delta) d^{(k)}_{yi}$, where $y$ indexes the year, and $i$ indexes the week of the year.
+Then we sum the total number of cases reported in this scenario and compare to the data to get an estimate of the number of cases prevented.
+To keep our simulations conservative, and to avoid pushing the model into the unrealistic range of dengue eradication, we set $\Delta = 0.05$.
+We also only simulate the effect of control in the middle three years, as the effectiveness of control in the first year is strongly affected by initial conditions, and we lack complete data for the last year.
 

### 2017-09-04 Manuscript/supplement.tex
Updates to intro and model structure
Cleaning up parameterization
Updating ms and supp methods

@@
 
+Note that in practice, we rescale the model described in the main text by human population size to keep state variables on a unit scale.
+This also rescales mosquito population size to number of mosquitoes per person.
+Adopting this rescaling, we can express the model as:
+
 \begin{align}
-y_t & \sim \text{NegBin}(\phi_y  (X_t - X_{t-1})N, \eta_y)\\
+y_t & \sim \text{NegBin}(\phi_y  (C_t - C_{t-1})N, \eta_y)\\
 q_t & \sim \text{NegBin}((V_{Ct} - V_{Ct-1})N, \eta_q)\\
 \\
-\frac{dX}{dt} &= \rho E \\
+\frac{dC}{dt} &= \rho E \\
 \frac{dS}{dt} &= b - bS - \lambda V_{I}S + \delta R\\
@@
 \\
-\frac{dV_S}{dt} & = b_t - \lambda I V_S - (d_t + \phi_q \tau_t) V_S \\
-\frac{dV_E}{dt} &= \lambda IV_S - (\rho_{vt} + d_t + \phi_q \tau_t)V_E\\
-\frac{dV_I}{dt} &= \rho_{vt} V_E - (d_t + \phi_q \tau_t) V_I\\
+\frac{dV_N}{dt} & = r V_N - \phi_q \tau_t V_N\\
+\frac{dV_E}{dt} &= \lambda IV_S - (\rho_{vt} + d + \phi_q \tau_t)V_E\\
+\frac{dV_I}{dt} &= \rho_{vt} V_E - (d + \phi_q \tau_t) V_I\\
 \frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
-V_N &= V_S + V_E + V_I\\
+V_S &= V_N - V_E - V_I\\
 \\
-\log(b_t) & = \alpha_{0} + \alpha_1 \sin(2\pi t / 52) + \alpha_2 \cos(2\pi t / 52) \\
-\log(d_{t}) & = \beta_{0} + \beta_1 \sin(2\pi t / 52) + \beta_2 \cos(2\pi t / 52) + \epsilon_t\\
-\epsilon_t & \sim \text{Normal}(0, \sigma)
+d(t) &= d_0 \exp(\nu(t))\\
+\frac{d^2\nu}{dt^2} &= -\omega^2 \nu + \epsilon_{\nu t}\\
+\frac{d^2 r}{dt^2} &= -\omega^2 r + \epsilon_{rt},
 \end{align}
+where $\omega = 2\pi / 52$, and
+\begin{align}
+\epsilon_{\nu i} & \sim \text{Normal}(0, \sigma_{\nu})\\
+\epsilon_{ri} & \sim \text{Normal}(0, \sigma_r),
+\end{align}
+for $i = 1 \dots 243$.
 
+The following priors complete the model specification:
 \begin{align}
-\alpha_0 & \sim \text{Normal}(-1, 1)\\ 
-\alpha_{1,2} & \sim \text{Normal}(0, 2)\\
-\beta_0 & \sim \text{Normal}(0.39, 0.12)\\
-\beta_{1,2} & \sim \text{Normal}(0, 0.2)\\
-\sigma & \sim \text{Normal}(0, 1)\\
+\sigma_r & \sim \text{Half-normal}(0, 0.2)\\
+\sigma_d & \sim \text{Half-normal}(0, 0.2)\\
 -\log(\phi_q) & \sim \text{Half-normal}(13, 0.5)\\
@@
 \\
-\item \textbf{Mean mosquito mortality rate} ($\beta_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
-They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $\beta_0$.  
-We use nonlinear least squares to fit the exponential cdf with rate $d$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d = 1.47$.
-We use the same procedure to estimate $d$ for the 0.25 and 0.75 quantiles of the estimated mortality curves (assuming symmetry around the mean), which gives 1.6 and 1.37.
-We then specify the prior on $beta_0$ to induce the above quantiles on $d$, which gives us:
-\begin{equation}
-\beta_0 \sim \text{Normal}(0.39, 0.12).
-\end{equation}
-\\
-\item \textbf{Mean mosquito emergence rate} ($\alpha_0$):
-\\
-\item \textbf{Regression parameters} ($\alpha_1,\alpha_2,\beta_1,\beta_2$): 
-\\
-\item \textbf{Death rate variance} ($\sigma$):
-\\
 \item \textbf{Initial conditions} ($S_0, E_0, I_0, R_0$): The SEIRS model assumes a constant population size and is scaled so that $S_0+E_0+I_0+R_0 = 1$.
@@
 \\
+\item \textbf{Mosquito process variances} ($\sigma_r, \sigma_d$): For the purposes of prediction and generalization, we want latent growth and death time series to be smooth and regularly oscillating.
+To achieve this, we put a relatively strong, half-normal prior on the process standard deviations.
+A prior standard deviation of 0.2 provides flexibility while still pulling the $\epsilon$ terms toward zero.
+\\
 \item \textbf{Overdispersion parameters} ($\eta_y, \eta_q$): The negative binomial likelihoods were parameterized as:
@@
 \begin{itemize}
+\item \textbf{Mean mosquito mortality rate} ($d_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
+They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $d_0$.  
+We use nonlinear least squares to fit the exponential cdf with rate $d_0$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d_0 = 1.47$.
+\\
 \item \textbf{Transmission rate} ($\lambda$): The transmission rate only appears multiplied by a term from the mosquito model (either $V_I$ or $V_S$) and is thus effectively non-identifiable when we are also estimating the overall size of the mosquito population.

### 2017-09-04 Analysis/control.r
Updating control code to use full posterior
With added parallelization to handle computational burden

@@ -1,4 +1,5 @@
 library(rstan)
+library(doParallel)
 library(plyr)
 library(magrittr)
 library(lubridate)
@@ -28,232 +29,42 @@ weather <- subset(weather, date < date[1] + weeks(243))
 weather$tot.week <- rep(c(1:243), each = 7)
 
 covars <- ddply(weather, .(tot.week), summarise, 
-                temp = mean(Mean.TemperatureC, na.rm = T),
-                humid = mean(Mean.Humidity, na.rm = T))
+                temp = mean(Mean.TemperatureC, na.rm = T))
 
 rov <- 7 * exp(0.2 * covars$temp - 8)
 
 #===============================================================================
-# Posterior means
+# Loading mcmc samples
 
 fit <- readRDS("Results/oscillator.rds")
+samples <- rstan::extract(fit)[1:23]
 
-means = list(list(S0 = get_posterior_mean(fit, pars = "S0")[4],
-                  E0 = get_posterior_mean(fit, pars = "E0")[4],
-                  I0 = get_posterior_mean(fit, pars = "I0")[4],
-                  log_phi_q = get_posterior_mean(fit, pars = "log_phi_q")[4],
-                  eta_inv_y = get_posterior_mean(fit, pars = "eta_inv_y")[4],
-                  eta_inv_q = get_posterior_mean(fit, pars = "eta_inv_q")[4],
-                  ro_c = get_posterior_mean(fit, pars = "ro_c")[4],
-                  gamma_c = get_posterior_mean(fit, pars = "gamma_c")[4],
-                  delta_c = get_posterior_mean(fit, pars = "delta_c")[4],
-                  logNv = get_posterior_mean(fit, pars = "logNv")[4],
-                  psi0 = get_posterior_mean(fit, pars = "psi0")[4],
-                  rv0 = get_posterior_mean(fit, pars = "rv0")[4],
-                  sigmapsi = 1, #get_posterior_mean(fit, pars = "sigmapsi")[4],
-                  sigmarv = 1, #get_posterior_mean(fit, pars = "sigmarv")[4],
-                  eps_psi = get_posterior_mean(fit, pars = "eps_psi")[, 4],
-                  eps_rv = get_posterior_mean(fit, pars = "eps_rv")[, 4]))
+nmcmc <- dim(fit)[[1]]
 
 #===============================================================================
-# Setting control series
+# Setting up parallel
 
-data$psi <- rstan::extract(fit, "psi", permute = T)[[1]] %>% apply(2, mean)
[…]

### 2017-09-05 Analysis/control.r
Compiling stan code outside foreach loop

@@ -54,6 +54,7 @@ extract_sample <- function(x, k){
   else return(x[k, ])
 }
 
+model <- stan_model(file = "Code/seirs.stan")
 
 reduction <- foreach(k = 1:nmcmc, .combine = "cbind", .packages = c("rstan", "magrittr")) %dopar% {
   
@@ -75,13 +76,13 @@ reduction <- foreach(k = 1:nmcmc, .combine = "cbind", .packages = c("rstan", "ma
                      control = control,
                      pop = pop)
     
-    sim <- stan(file = "Code/seirs.stan", 
-                data = dat.stan, 
-                init = init, 
-                iter = 1, 
-                chains = 1,
-                warmup = 0,
-                algorithm = "Fixed_param")
+    sim <- sampling(model,
+                    data = dat.stan, 
+                    init = init, 
+                    iter = 1, 
+                    chains = 1,
+                    warmup = 0,
+                    algorithm = "Fixed_param")
     
     cases[i] <- rstan::extract(sim, "y_hat", permute = F)[, 1, ] %>% sum()
   }
@@ -89,4 +90,8 @@ reduction <- foreach(k = 1:nmcmc, .combine = "cbind", .packages = c("rstan", "ma
   return(cases)
 }
 
+stopCluster(cl)
+
 reduction <- sum(data$obs) - reduction
+
+saveRDS(reduction, "Results/control.rds")

### 2017-10-03 Manuscript/ms.tex
Place-holder figures in ms

@@
 
-Questions:
-\begin{itemize}
-\item How wells does the model fit the case reports and mosquito surveillance data?
-\item What do the estimated time series of demographic rates look like? How smooth are they? How strong is the periodic signal relative to noise?
-\item How do peaks in $\psi$ line up with peaks in $q_t$?
-\item How effective is control informed by $\psi$ compared to control informed by $q_t$?
-\end{itemize}
-
 \begin{figure}[!h]
@@
 \label{timeseries}
+\end{figure}
+
+\begin{figure}[!h]
+%\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
+\caption{{\bf Estimated latent birth and death rates.}
+}
+\label{latent}
+\end{figure}
+
+\begin{figure}[!h]
+%\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
+\caption{{\bf Cases prevented by adult control.}
+}
+\label{adultcontrol}
 \end{figure}

### 2017-10-03 Analysis/citystan.r
Updating sampler options for full run

@@ -98,9 +98,9 @@ inits = list(list(S0 = 0.4,
 fit <- stan(file = "Code/seirs.stan", 
             data = dat.stan, 
             init = inits, 
-            iter = 5000, 
+            iter = 10000, 
             chains = 3,
-            control = list(adapt_delta = 0.9,
+            control = list(adapt_delta = 0.99,
                            max_treedepth = 15))
 
 saveRDS(fit, "Results/oscillator.rds")

### 2017-10-03 Analysis/predict.r
Updates to prediction code Adding light damping to oscillator and separating fitting from predicting

@@ -2,6 +2,7 @@ library(rstan)
 library(plyr)
 library(magrittr)
 library(lubridate)
+library(doParallel)
 
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
@@ -33,18 +34,18 @@ temp <- ddply(weather, .(tot.week), summarise, temp = mean(Mean.TemperatureC, na
 
 rov <- 7 * exp(0.2 * temp$temp - 8)
 #===============================================================================
-# Running stan
+# Fitting to first four years
 
-T_pred <- 52
-T_fit <- 243 - T_pred
+T_pred <- 0
+T_fit <- 191
 
 dat.stan <- list(T = T_fit,
                  T_pred = T_pred,
                  y = head(obs, T_fit),
                  q = head(q, T_fit),
                  tau = head(tau, T_fit),
-                 rov = rov,
-                 control = matrix(1, nrow = 243, ncol = 3),
+                 rov = head(rov, T_fit),
+                 control = matrix(1, nrow = 191, ncol = 3),
                  pop = pop)
 
 inits = list(list(S0 = 0.4,
@@ -57,9 +58,9 @@ inits = list(list(S0 = 0.4,
                   gamma_c = 1,
                   delta_c = 0.8,
                   logNv = 0.7,
-                  dv0 = 0.39,
+                  dv0 = 0,
                   rv0 = 0,
-                  sigmapsi = 0.2,
+                  sigmadv = 0.2,
                   sigmarv = 0.2,
                   eps_dv = rep(0, T_fit),
                   eps_rv = rep(0, T_fit)),
@@ -73,9 +74,9 @@ inits = list(list(S0 = 0.4,
                   gamma_c = 1.2,
                   delta_c = 1,
                   logNv = 1,
-                  dv0 = 0.39,
+                  dv0 = 0,
                   rv0 = 0,
-                  sigmapsi = 0.5,
+                  sigmadv = 0.5,
                   sigmarv = 0.5,
                   eps_dv = rep(0, T_fit),
                   eps_rv = rep(0, T_fit)),
@@ -88,10 +89,10 @@ inits = list(list(S0 = 0.4,
                   ro_c = 1,
                   gamma_c = 1,
                   delta_c = 1,
-                  logNv = 0.5,
-                  dv0 = 0.39,
+                  logNv = 0.7,
+                  dv0 = 0,
                   rv0 = 0,
[…]

### 2017-10-04 Analysis/plots.r
Processing mcmc for figures

@@ -7,7 +7,6 @@ library(lubridate)
 library(ggplot2)
 library(gridExtra)
 
-
 # Loading and processing mosquito and case data
 tseries <- read.csv("Data/Vitoria.data.csv")
 
@@ -25,12 +24,12 @@ weather <- read.csv("Data/Vitoria.weather.csv") %>%
 weather <- subset(weather, date < date[1] + weeks(243))
 weather$tot.week <- rep(c(1:243), each = 7)
 
-covars <- ddply(weather, .(tot.week), summarise,
-                temp = mean(Mean.TemperatureC, na.rm = T),
-                humidity = mean(Mean.Humidity, na.rm = T))
+temp <- ddply(weather, .(tot.week), summarise, temp = mean(Mean.TemperatureC, na.rm = T))
+
+post$rov <- 7 * exp(0.2 * temp$temp - 8)
 
 # Loading MCMC results
-sim <- readRDS("Results/covars.rds")
+sim <- readRDS("Results/oscillator.rds")
 
 #===============================================================================
 # Figure 1: Observed and estimated time series
@@ -79,118 +78,34 @@ grid.arrange(fig1.a, fig1.b, ncol = 2)
 dev.off()
 
 #===============================================================================
-# Figure 2: estimated demographic rates
+# Figure 2: Estimates of latent mosquito demographic rates
 
-d <- rstan::extract(sim, "dv", permute = T)[[1]] %>% apply(2, quantile, c(0.1, 0.5, 0.9))
-epsilon <- rstan::extract(sim, "z_d", permute = T)[[1]] %>% apply(2, quantile, c(0.1, 0.5, 0.9))
+system <- rstan::extract(sim, "system", permute = T)[[1]]
 
-post[, c("dmin", "d", "dmax")] <- t(d)
-post[, c("epsmin", "eps", "epsmax")] <- t(epsilon)
+dv <- system[, , 9] %>% + 0.39 %>% exp() %>% 
+  adply(2, quantile, c(0.1, 0.5, 0.9), .id = NULL)
+names(dv) <- c("dvmin", "dvmed", "dvmax")
 
-ggplot(post, aes(tot.week, d)) + 
-  geom_ribbon(aes(ymin = dmin, ymax = dmax), fill = "grey70") +
-  geom_line() + 
-  theme_classic() +
-  scale_y_continuous(expand = c(0.05, 0)) + 
[…]

### 2017-10-05 Analysis/plots.r
Additional supplemental plots

@@ -17,6 +17,9 @@ post <- ddply(tseries, .(tot.week), summarise,
               week = unique(Week),
               year = unique(Year))
 
+# Population size
+pop <- 327801
+
 # Loading and processing the weather data
 weather <- read.csv("Data/Vitoria.weather.csv") %>% 
   mutate(date = ymd(BRST), year = year(date), week = week(date))
@@ -39,12 +42,12 @@ yhat <- rstan::extract(sim, c("y_hat"), permute = F) %>% apply(3, quantile, c(0.
 qhat <- rstan::extract(sim, c("q_hat"), permute = F) %>% apply(3, quantile, c(0.1, 0.5, 0.9))
 
 # Assembling data frame for ggplot
-post$yhat = yhat[2, ]
-post$ymin = yhat[1, ]
-post$ymax = yhat[3, ]
-post$qhat = qhat[2, ]
-post$qmin = qhat[1, ]
-post$qmax = qhat[3, ]
+post$yhat <- yhat[2, ]
+post$ymin <- yhat[1, ]
+post$ymax <- yhat[3, ]
+post$qhat <- qhat[2, ]
+post$qmin <- qhat[1, ]
+post$qmax <- qhat[3, ]
 
 # Plotting
 ggplot(post, aes(tot.week, yobs)) + 
@@ -86,26 +89,224 @@ dv <- system[, , 9] %>% + 0.39 %>% exp() %>%
   adply(2, quantile, c(0.1, 0.5, 0.9), .id = NULL)
 names(dv) <- c("dvmin", "dvmed", "dvmax")
 
-rv <- system[, , 11] %>% apply(2, quantile, c(0.1, 0.5, 0.9), .id = NULL)
+rv <- system[, , 11] %>% adply(2, quantile, c(0.1, 0.5, 0.9), .id = NULL)
 names(rv) <- c("rvmin", "rvmed", "rvmax")
 
 post <- cbind(post, dv, rv)
 
+# Plotting
+ggplot(post, aes(tot.week, rvmed)) + 
+  geom_ribbon(aes(ymin = rvmin, ymax = rvmax), fill = "grey70") +
+  geom_line() + 
+  theme_classic() +
+  scale_y_continuous(expand = c(0, 0)) + 
+  scale_x_continuous(expand = c(0, 1)) +
+  ylab("mosquito growth rate") + 
+  xlab("week") +
+  ggtitle("A") -> fig2.a
+
[…]

### 2017-10-09 Analysis/plots.r
Adding plot of death rate vs. temperature

@@ -2,6 +2,7 @@
 
 library(rstan)
 library(plyr)
+library(reshape)
 library(magrittr)
 library(lubridate)
 library(ggplot2)
@@ -29,6 +30,7 @@ weather$tot.week <- rep(c(1:243), each = 7)
 
 temp <- ddply(weather, .(tot.week), summarise, temp = mean(Mean.TemperatureC, na.rm = T))
 
+post$temp <- temp$temp
 post$rov <- 7 * exp(0.2 * temp$temp - 8)
 
 # Loading MCMC results
@@ -122,8 +124,26 @@ postscript("Manuscript/figures/fig2.eps",
 grid.arrange(fig2.a, fig2.b, ncol = 2)
 
 dev.off()
+
 #===============================================================================
-# Figure 3: Effect of adult control implemented in different weeks
+# Figure 3: Mosquito death rate as a function of temperature
+
+postscript("Manuscript/figures/fig3.eps",
+           width = 3, height = 3,
+           family = "ArialMT")
+
+ggplot(post, aes(temp, dvmed)) +
+  geom_point() + 
+  theme_classic() + 
+  scale_y_continuous(expand = c(0.05, 0)) + 
+  scale_x_continuous(expand = c(0, 1)) +
+  ylab("mosquito death rate") +
+  xlab("weekly mean temperature")
+
+dev.off()
+
+#===============================================================================
+# Figure 4: Effect of adult control implemented in different weeks
 
 adult_reduction <- readRDS("Results/adult_control.rds") %>% 
   adply(1, quantile, probs = c(0.1, 0.5, 0.9)) %>% 
@@ -137,7 +157,7 @@ larval_reduction <- readRDS("Results/larval_control.rds") %>%
 
 names(larval_reduction) <- c("X1", "min", "med", "max", "week")
 
-postscript("Manuscript/figures/fig3.eps",
+postscript("Manuscript/figures/fig4.eps",
            width = 4, height = 3,
            family = "ArialMT")
 
@@ -149,7 +169,7 @@ ggplot(larval_reduction, aes(week, med)) +
   scale_x_continuous(expand = c(0, 1)) +
   xlab("week of control") +
   ylab("cases prevented") +
-  ggtitle("A")-> fig3.a
+  ggtitle("A")-> fig4.a
 
 ggplot(adult_reduction, aes(week, med)) + 
[…]

### 2017-10-09 Manuscript/ms.tex
Adding results text and figure captions.

@@
 \end{align}
-where $\omega = 2\pi / 52$, $d(t) = d_0 \exp(\nu(t)$, and
+where $\omega = 2\pi / 52$, $d(t) = d_0 \exp(\nu(t))$, and
 \begin{align}
@@
 
+The model is able to capture the observed dynamics of both case reports and mosquito trap counts (Figure \ref{timeseries}).
+The estimated posterior median case reports explain 91\% of the variation in the observed time series, while the posterior median mosquito trap counts explain 45\% of the variation in the observed time series.
+In addition, posterior predictive checks show that the model reproduces the the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Fig XX - XX).
+Posterior means for the three epidemiological parameters ($\rho, \gamma, \delta$) are close to the literature values used to set their prior means (Supplemental Fig XX).
+
+The estimated latent mosquito demographic rates are smooth and strongly seasonal, with fairly narrow credible intervals (Figure \ref{latent}).
+Mosquito death rates fluctuate within the range estimated by \cite{Brady2013} and generally increase with temperature, as expected from laboratory studies \cite{Brady2013}(Figure \ref{temp}).
+The underlying process variances are small ($\sigma_nu = 0.01$, $\sigma_r = 0.0005$), and there is little residual structure in the process noise series, though there is a weak periodic signal in the $\epsilon_{\nu}$ (Supplemental Fig XX).
+
+Modifying these estimated demographic rates to simulate targeted control efforts, we find that control (targeting either adults via increasing the death rate or larvae via decreasing the birth rate) applied in the third week of the year is the most effective at preventing cases of dengue fever (though weeks 2 and 4 are also very close, Figure \ref{control}).
+In fact, increasing the mosquito death rate by just 5\% in the third week of 2009, 2010, and 2011 prevents 13,988 (reported) cases of dengue fever (with 80\% credible interval of 11,130 to 15,811).
+Targeting larvae instead of adults (simulated by reducing the birth rate ($b_t = r_t + d_t$) by 5\%) produces nearly identical results.
+The third week of the year is generally very early in the dengue season, well before cases peak (Figure \ref{timing}).  
+It is also generally before the estimated peak in mosquito abundance, which occurs between weeks 6 and 9 (Figure \ref{timing}).
+Annual peaks in estimated mosquito death rate, and infection risk (i.e.
+the probability that a mosquito survives long enough to become infectious, $\frac{\rho_v}{(\rho_v + d)}$) are more variable, but also do not generally line up with the timing of optimal control.
+
 \begin{figure}[!h]
-%\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
+\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
 \caption{{\bf Vitoria data and model fits.}
-Weekly observations from 2008 to week 34 of 2012 (points), with corresponding median posterior prediction (line) and 80\% credible inverval (gray band).
+Weekly observations from 2008 to week 34 of 2012 (points), with corresponding median posterior prediction (line) and 80\% credible interval (gray band).
 A: reported dengue cases.
@@
 \begin{figure}[!h]
-%\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
-\caption{{\bf Estimated latent birth and death rates.}
+\includegraphics[angle = 270, scale = 1]{figures/fig2.eps}
+\caption{{\bf Estimated latent mosquito growth and death rates.}
+Median posterior estimate (line) and 80\% credible interval.  A: weekly mosquito population growth rate ($r_t$). B: weekly mosquito death rate ($d_t$).
 }
@@
 \begin{figure}[!h]
-%\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
-\caption{{\bf Cases prevented by adult control.}
+\includegraphics[angle = 270, scale = 1]{figures/fig3.eps}
+\caption{{\bf Estimated mosquito death rate as a function of temperature.}
+Weekly median posterior death rate ($d_t$) plotted against weekly mean temperature (Celsius) in Vitoria, Brazil.
 }
-\label{adultcontrol}
+\label{temp}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/fig4.eps}
+\caption{{\bf Cases prevented by mosquito control.}
+Number of reported cases of dengue fever prevented by control deployed during a given week each year in 2009, 2010, and 2011.
+Median posterior prediction (line) and 80\% credible interval.
+A: adult control, simulated by increasing death rate in the given week by 5\%.
+B: larval control, simulated by decreasing birth rate ($r_t + d_t$) in a given week by 5\%.
+}
+\label{control}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/fig5.eps}
+\caption{{\bf Timing of optimal control relative to disease and mosquito dynamics.}
+The vertical black line indicates week 3, during which mosquito control produces the largest decline in predicted dengue case reports.
+Each colored line represents a different year: salmon?
+= 2008, green-yellow = 2009, green = 2010, blue = 2011, purple = 2012.
+A: posterior median estimate of dengue case reports per week.
+B: posterior median estimate of trapped mosquitoes per week.
+C: posterior median estimate of weekly mosquito death rate.
+D: posterior median estimate of weekly infection risk ($\frac{\rho_v}{\rho_v + d}$).
+}
+\label{timing}
 \end{figure}

### 2017-10-10 Analysis/control.r
Plotting dynamics from third week control

@@ -37,17 +37,9 @@ rov <- 7 * exp(0.2 * covars$temp - 8)
 # Loading mcmc samples
 
 fit <- readRDS("Results/oscillator.rds")
-samples <- rstan::extract(fit)[1:23]
+samples <- rstan::extract(fit)[1:16]
 
-nmcmc <- dim(fit)[[1]]
-
-#===============================================================================
-# Setting up parallel
-
-cl <- makeCluster(2, type = "SOCK")
-registerDoParallel(cl)
-#===============================================================================
-# In which week is adult control most effective?
+nmcmc <- length(samples[[1]])
 
 extract_sample <- function(x, k){
   if(length(dim(x)) == 1) return(x[k])
@@ -56,6 +48,14 @@ extract_sample <- function(x, k){
 
 model <- stan_model(file = "Code/seirs.stan")
 
+#===============================================================================
+# In which week is adult control most effective?
+
+# Setting up parallel
+cl <- makeCluster(2, type = "SOCK")
+registerDoParallel(cl)
+
+# Parallel for-loop over mcmc iterations
 reduction <- foreach(k = 1:nmcmc, .combine = "cbind", .packages = c("rstan", "magrittr")) %dopar% {
   
   init <- list(lapply(samples, extract_sample, k))
@@ -95,3 +95,44 @@ stopCluster(cl)
 reduction <- sum(data$obs) - reduction
 
 saveRDS(reduction, "Results/control.rds")
+
+#===============================================================================
+# Dynamics when control deployed in week 3
+
+# Setting up parallel
+cl <- makeCluster(3, type = "SOCK")
+registerDoParallel(cl)
+
+# Parallel for-loop over mcmc iterations
+week3 <- foreach(k = 1:nmcmc, .packages = c("rstan", "magrittr")) %dopar% {
+  
+  init <- list(lapply(samples, extract_sample, k))
+  
+  control <- matrix(1.0, nrow = 243, ncol = 3)
+  control[(data$week == 3 & data$year > 2008 & data$year < 2012), 1] <- 1.05
+  
+  dat.stan <- list(T = 243,
+                   T_pred = 0,
+                   y = data$obs,
+                   q = data$q,
+                   tau = data$tau,
+                   rov = rov,
+                   control = control,
+                   pop = pop)
+  
+  sim <- sampling(model,
+                  data = dat.stan, 
+                  init = init, 
+                  iter = 1, 
[…]

### 2017-10-10 Analysis/plots.r

@@ -36,8 +36,7 @@ post$rov <- 7 * exp(0.2 * temp$temp - 8)
 # Loading MCMC results
 sim <- readRDS("Results/oscillator.rds")
 
-#===============================================================================
-# Figure 1: Observed and estimated time series
+# Figure 1: Observed and estimated time series =================================
 
 # Computing posterior median and 80% credible interval
 yhat <- rstan::extract(sim, c("y_hat"), permute = F) %>% apply(3, quantile, c(0.1, 0.5, 0.9))
@@ -82,8 +81,7 @@ grid.arrange(fig1.a, fig1.b, ncol = 2)
 
 dev.off()
 
-#===============================================================================
-# Figure 2: Estimates of latent mosquito demographic rates
+# Figure 2: Estimates of latent mosquito demographic rates =====================
 
 system <- rstan::extract(sim, "system", permute = T)[[1]]
 
@@ -125,8 +123,7 @@ grid.arrange(fig2.a, fig2.b, ncol = 2)
 
 dev.off()
 
-#===============================================================================
-# Figure 3: Mosquito death rate as a function of temperature
+# Figure 3: Mosquito death rate as a function of temperature ===================
 
 postscript("Manuscript/figures/fig3.eps",
            width = 3, height = 3,
@@ -142,34 +139,28 @@ ggplot(post, aes(temp, dvmed)) +
 
 dev.off()
 
-#===============================================================================
-# Figure 4: Effect of adult control implemented in different weeks
+# Figure 4: Effect of adult control implemented in different weeks =============
 
+# Processing adult control experiment
 adult_reduction <- readRDS("Results/adult_control.rds") %>% 
   adply(1, quantile, probs = c(0.1, 0.5, 0.9)) %>% 
   mutate(week = as.numeric(X1))
 
 names(adult_reduction) <- c("X1", "min", "med", "max", "week")
 
-larval_reduction <- readRDS("Results/larval_control.rds") %>% 
-  adply(1, quantile, probs = c(0.1, 0.5, 0.9)) %>% 
-  mutate(week = as.numeric(X1))
+# Dynamics resulting from control in week 3
+week3 <- readRDS("Results/optimal.rds")
 
-names(larval_reduction) <- c("X1", "min", "med", "max", "week")
[…]

### 2017-10-10 Manuscript/ms.tex
Adding description of dynamics under control

@@
 
-Modifying these estimated demographic rates to simulate targeted control efforts, we find that control (targeting either adults via increasing the death rate or larvae via decreasing the birth rate) applied in the third week of the year is the most effective at preventing cases of dengue fever (though weeks 2 and 4 are also very close, Figure \ref{control}).
-In fact, increasing the mosquito death rate by just 5\% in the third week of 2009, 2010, and 2011 prevents 13,988 (reported) cases of dengue fever (with 80\% credible interval of 11,130 to 15,811).
+Modifying these estimated demographic rates to simulate targeted control efforts, we find that control applied in the third week of the year is the most effective at preventing cases of dengue fever (though weeks 2 and 4 are also very close, Figure \ref{control}A).
+In fact, increasing the mosquito death rate by just 5\% in the third week of 2009, 2010, and 2011 would prevent 13,988 (reported) cases of dengue fever (with 80\% credible interval of 11,130 to 15,811).
+Most of these gains come by way of substantially reducing the size of the large 2011 outbreak, while also further damping the smaller 2010 and 2012 outbreaks (Figure \ref{control}B).
 Targeting larvae instead of adults (simulated by reducing the birth rate ($b_t = r_t + d_t$) by 5\%) produces nearly identical results.
+
 The third week of the year is generally very early in the dengue season, well before cases peak (Figure \ref{timing}).
-It is also generally before the estimated peak in mosquito abundance, which occurs between weeks 6 and 9 (Figure \ref{timing}).
-Annual peaks in estimated mosquito death rate, and infection risk (i.e.
-the probability that a mosquito survives long enough to become infectious, $\frac{\rho_v}{(\rho_v + d)}$) are more variable, but also do not generally line up with the timing of optimal control.
+In the two years with large outbreaks (2009 and 2011), the third week of the year is near the inflection point where the disease burden begins to grow rapidly.  
+The third week is also generally before the estimated seasonal peak in mosquito abundance, which occurs between weeks 6 and 9 (Figure \ref{timing}).
+Annual peaks in estimated mosquito death rate are more variable, but do tend to fall between weeks 2 and 7.
+Though mosquito death rates tend to be high early in the year, the transmission risk (i.e.
+the probability of an exposed mosquito surviving long enough to be infectious, $\frac{\rho_v}{\rho_v + d}$) does not pick up until later in the year.
 
@@
 A: reported dengue cases.
-B: total number of trapped mosquiotes.
+B: total number of trapped mosquitoes.
 }
@@
 \caption{{\bf Estimated latent mosquito growth and death rates.}
-Median posterior estimate (line) and 80\% credible interval.  A: weekly mosquito population growth rate ($r_t$). B: weekly mosquito death rate ($d_t$).
+Median posterior estimate (line) and 80\% credible interval (gray band).  A: weekly mosquito population growth rate ($r_t$). B: weekly mosquito death rate ($d_t$).
 }
@@
 \caption{{\bf Cases prevented by mosquito control.}
-Number of reported cases of dengue fever prevented by control deployed during a given week each year in 2009, 2010, and 2011.
+A: Number of reported cases of dengue fever prevented by increasing the mosquito death rate by 5\% during a given week each year in 2009, 2010, and 2011.
 Median posterior prediction (line) and 80\% credible interval.
-A: adult control, simulated by increasing death rate in the given week by 5\%.
-B: larval control, simulated by decreasing birth rate ($r_t + d_t$) in a given week by 5\%.
+B: Weekly predicted case reports resulting from implementing control in week 3.
+Median posterior case reports from model without (dashed line) and with control (solid line) and 80\% credible interval of controlled dynamics (gray band).
 }
@@
 C: posterior median estimate of weekly mosquito death rate.
-D: posterior median estimate of weekly infection risk ($\frac{\rho_v}{\rho_v + d}$).
+D: posterior median estimate of weekly transmission risk ($\frac{\rho_v}{\rho_v + d}$).
 }
@@
 
+\subsection*{Summarize results}
+
+Targeted mosquito control can be an effective tool in reducing the disease burden within a city.
+When faced with limited resources with which to deploy that control, these results suggest that focusing efforts early in the year, specifically in the second to fourth weeks of the year, will lead to the largest reductions in disease.
+
+\subsection*{Interpretation of results}
+
+Caveat and future work: spatial prioritization of mosquito control:
 \begin{itemize}
@@
 
-\section*{Conclusions}
+\subsection*{Conclusions}
 

### 2017-10-11 Manuscript/ms.tex
Fleshing out discussion

@@
 
-Samples from the posterior distribution were generated using Hamiltonian Monte Carlo (CITATIONS) implemented in the rstan package for R (CITATIONS). 
+Samples from the posterior distribution were generated using Hamiltonian Monte Carlo implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R (\cite{R2016}). 
 We ran 3 chains with different starting values for 10,000 iterations each, discarding the first 5,000 as burn-in.
@@
 
-\subsection*{Summarize results}
-
 Targeted mosquito control can be an effective tool in reducing the disease burden within a city.
 When faced with limited resources with which to deploy that control, these results suggest that focusing efforts early in the year, specifically in the second to fourth weeks of the year, will lead to the largest reductions in disease.
+Because adult control and larval control produce the same results, the effect of control is likely driven by reducing the standing population of adult mosquitoes, rather than by limiting the transmission potential of exposed or infectious mosquitoes.
+Targeting mosquitoes early in the year as dengue just begins to spread can disrupt the transmission cycle and prevent further amplification and feedback.
 
-\subsection*{Interpretation of results}
+The importance of reducing the population of adult mosquitoes agrees with the results of \cite{Burattini2008} whose model of dengue dynamics in Singapore suggests that killing adult mosquitoes is the most effective way to control an epidemic.
+However, they find that larval control is most effective at preventing resurgence of dengue epidemics, and thus that a mixed strategy that includes both larval and adult control is likely to be most effective in the long term.
+Fully exploring such combined strategies requires a more complete model of mosquito life history and the lags induced by the mosquito's aquatic larval stage.
+In addition, the lack of a larval phase in our model might somewhat exaggerate the effectiveness of adult control, as decreasing the adult population leads to an immediate reduction in reproduction, rather than a lagged reduction.
+Temperature-dependent development \cite{Focks1993}.
+Models of development can be very complex \cite{Marori2009}.
+\cite{Pinho2010} find that adult control alone cannot reduce $R_0$ below 1 and that aquatic control is necessary.
 
-Caveat and future work: spatial prioritization of mosquito control:
-\begin{itemize}
-\item Human movement \cite{Adams2009, Cosner2009a, Stoddard2009, Dalziel2013}
-\item Disease risk possibly driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Canyon1999}.
-\item Explicitly targeting mosquitoes that have been implicated in transmission through the use of contact tracing \cite{Vazquez-Prokopec2017}
-\end{itemize}
+As implemented, mosquito mortality is the only source of uncertainty/stochasticity in the transmission process.
+Related to the idea of empirical forcing functions of \cite{Hooker2015} and dynamic smoothing from Ramsay.
+The fact that death rate increases with temperature helps to validate this approach.
+Whether or not the death rate is actually interpretable as a mortality rate, it nonetheless captures the critical bottleneck in the transmission process.
+\cite{Maciel-de-Freitas2008}
+
+Establishing that there is likely spatial heterogeneity.
+Processes that might drive spatial heterogeneity (human movement, socioeconomic factors)
+Structured human movement within the city is likely to be a major driver of heterogeneous human-mosquito mixing (\cite{Adams2009, Cosner2009a, Stoddard2009, Dalziel2013}.
+In addition variability in socioeconomic factors within the city also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
+What are the implications of this heterogeneity and how might spatial heterogeneity modify our conclusions?
+Introducing heterogeneous mixing increases $R_0$ relative to homogeneous mixing, potentially leading us to overestimate the effectiveness of control (\cite{Dye1986, Hasibeder1988}.
+This heterogeneity suggests that disease risk might be driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Canyon1999, Yoon2012}.
+What can we gain by accounting for heterogeneity?
+Though the presence of this heterogeneity suggests that our model may overestimate the efficacy of mosquito control, it also points the way towards more efficient methods.
+Most generally, control efforts should target mosquitoes that are most directly implicated in transmission.
+One way to do this is through the use of contact-tracing \cite{Vazquez-Prokopec2017}.
+How would we incorporate that heterogeneity?
+Embedding the mechanistic model explicitly into a metapopulation framework leads to a very large, unwieldy model that likely exacerbates existing difficulties inherent in fitting ODE models to data.
+
+Statistical novelty and connections to literature
+Integrated population modeling \cite{Schaub2010}
+Posterior sampling for ode models \cite{Girolami2008, Calderhead2011}
 
 \subsection*{Conclusions}
+
+Mosquito surveillance is a valuable tool in managing dengue fever, but is most powerful when integrated into a larger modeling framework.
 
@@
 
-\section*{Acknowledgments}
+\section*{Acknowledgements}
 

### 2017-10-11 Manuscript/ms.tex
Cleaning up intro and forging discussion into actual sentences

--- 2017-10-10 Manuscript/ms.tex
@@
 
-Mosquito abundance alone is not a strong predictor of dengue fever risk, but provides information that, when coupled with case data, allows us to estimate mosquito demographic rates.
-
 % Please keep the Author Summary between 150 and 200 words
@@
 \section*{Author summary}
-
-Mosquitoes are dumb.
 
@@
 The cycle of transmission between humans and dengue is controlled not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period \cite{Achee2015}.
-In fact, the probability that an exposed mosquito will survive long enough to become infectious provides a more direct measure of transmission potential than mosquito abundance alone.
+In fact, whether or not an exposed mosquito will survive long enough to become infectious represents a critical bottleneck in the transmission process.
 
-Adult \textit{Aedes aegypti} surveillance data (e.g. time series of mosquito trap counts) is insufficient on its own to estimate mosquito survival rates and thus transmission potential.
-However, we can borrow from the field of 'integrated population modeling' and combine multiple sources of information to estimate mosquito demographic rates (CITATIONS).
-Since mosquito survival is a critical bottleneck in the transmission process, it can be informed from the observed transmission dynamics (i.e.
+Adult \textit{Aedes aegypti} surveillance data (e.g. time series of mosquito trap counts) is insufficient on its own to estimate mosquito mortality rates.
+However, we can borrow from the field of 'integrated population modeling' and combine multiple sources of information to estimate mosquito demographic rates \cite{Schaub2010}.
+Since mosquito mortality is a critical bottleneck in the transmission process, it can be informed from the observed transmission dynamics (i.e.
 reports of human cases), given an appropriate underlying mechanistic model.
 
-In this paper, we propose a mechanistic model to connect mosquito abundance and survival with dengue epidemiology.
+In this paper, we propose such a mechanistic model to connect mosquito abundance and survival with dengue epidemiology.
 We then embed this model into a Bayesian statistical framework that allows us to estimate latent time series of mosquito demographic parameters from time series of mosquito trap counts and reported cases of dengue fever.
-Lastly, we explore the utility of these estimated transmission risks for guiding targeted control efforts relative to using mosquito abundance alone.
+Lastly, we use the fitted model to explore the optimal timing of adult control interventions.
 
@@
 for $i = 1 \dots 243$.
-These stochastically-forced harmonic oscillators provide a flexible framework for generating smooth seasonal oscillations in the latent mosquito processes (Ramsay citation?).
+These stochastically-forced harmonic oscillators provide a flexible framework for generating smooth seasonal oscillations in the latent mosquito processes \cite{Ramsay2017}.
 
@@
 The vertical black line indicates week 3, during which mosquito control produces the largest decline in predicted dengue case reports.
-Each colored line represents a different year: salmon?
-= 2008, green-yellow = 2009, green = 2010, blue = 2011, purple = 2012.
+Each colored line represents a different year: dark green = 2008, orange = 2009, purple = 2010, pink = 2011, light green = 2012.
 A: posterior median estimate of dengue case reports per week.
@@
 
-\subsection*{Summarize results}
+Targeted mosquito control can be an effective tool in reducing the disease burden within a city.
+When faced with limited resources with which to deploy that control, these results suggest that focusing efforts early in the year, specifically in the second to fourth weeks, will lead to the largest reductions in disease.
+Because adult control (increased the mortality rate) and larval control (decreasing the birth rate) produce very similar reductions in disease, the effect of control is likely driven by reducing the standing population of adult mosquitoes, rather than by limiting the transmission potential of exposed or infectious mosquitoes.
+Targeting mosquitoes early in the year as dengue just begins to spread can disrupt the transmission cycle and prevent further amplification and feedback.
 
-Targeted mosquito control can be an effective tool in reducing the disease burden within a city.
-When faced with limited resources with which to deploy that control, these results suggest that focusing efforts early in the year, specifically in the second to fourth weeks of the year, will lead to the largest reductions in disease.
+The importance of reducing the population of adult mosquitoes agrees with the results of \cite{Burattini2008} whose model of dengue dynamics in Singapore suggests that killing adult mosquitoes is the most effective way to control an epidemic.
+However, they find that larval control is most effective at preventing resurgence of dengue epidemics, and thus that a mixed strategy that includes both larval and adult control is likely to be most effective in the long term.
+The utility of mixed control strategies is further supported by \cite{Pinho2010}, whose model of dengue in Salvador, Brazil suggests that adult control alone is insufficient to bring $R_0$ below 1 and eliminate spread.
+Fully exploring such combined strategies requires a more complete model of mosquito life history and the lags in population dynamics induced by the mosquito's aquatic stage.
+The lack of such an aquatic phase in our model makes it more suitable for exploring the immediate, direct effects of adult control on transmission, rather than for studying the lagged feedbacks on mosquito population dynamics (and their knock-on effects on transmission).
+Though models of mosquito development can be very complex, e.g.
+\cite{Magori2009}, our model could be extended to incorporate a compartment for the aquatic phase (as in \cite{Burattini2008, Pinho2010}), though at the cost of introducing several new parameters to the model.
 
-\subsection*{Interpretation of results}
+In our model, the relationship between mosquito abundance and human disease risk is controlled by mosquito mortality, which, together with the extrinsic incubation period ($\rho_v$), controls how  many exposed mosquitoes survive long enough to become infectious.
+As implemented, mosquito mortality is the only source of uncertainty/stochasticity in the transmission process.
+As a result, mosquito mortality could be soaking up other sources of stochasticity or model misspecification.
+Our implementation of the mortality process is conceptually similar to the empirical forcing functions developed by \cite{Hooker2015}.
+In that framework, \cite{Hooker2015} allow parameters of differential equation models to change over time and use the resulting time series to differentiate causes of lack-of-fit.
+Though we do not employ the same explicit testing framework here, we treat the estimated mosquito mortality time series in the same spirit.
+In particular, we can check whether our estimated mosquito mortality time series is mechanistically plausible.
+Our estimated mosquito mortalities generally fall within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
+The fact that our estimated mortality rate varies seasonally and increases with temperature also agrees with the empirical literature on mosquito survival.
+These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.
 
-Caveat and future work: spatial prioritization of mosquito control:
-\begin{itemize}
-\item Human movement \cite{Adams2009, Cosner2009a, Stoddard2009, Dalziel2013}
-\item Disease risk possibly driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Canyon1999}.
-\item Explicitly targeting mosquitoes that have been implicated in transmission through the use of contact tracing \cite{Vazquez-Prokopec2017}
-\end{itemize}
+Though the estimated mosquito mortality process seems biologically reasonable, we are still employing it to link mosquito abundance and human disease at a relatively large spatial scale, assuming homogeneous mixing between mosquitoes and humans throughout the city.
+However, structured human movement within the city is likely to be a major driver of heterogeneous human-mosquito mixing (\cite{Adams2009, Cosner2009a, Stoddard2009, Dalziel2013}.
+In addition spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
+
+This heterogeneity has substantial consequences for disease dyamics.
+In fact, in a modeling study \cite{Dye1986, Hasibeder1988} found that introducing heterogeneous mixing always increases $R_0$ relative to homogeneous mixing.
+Though this suggests that our model may overestimate the efficacy of mosquito control, it also points the way towards more efficient and effective control methods.
+If disease risk is driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Canyon1999, Yoon2012}, control policies that preferentially targets these subsets of the mosquito population are likely to be very effective.
+Identifying these subsets is in practice difficult.
+One way to do this is through the use of contact-tracing, which can identify and target specific premises that have been implicated in transmission \cite{Vazquez-Prokopec2017}.
+Though effective, this strategy is extremely labor-intensive and would be difficult to implement in endemic areas.
+
+On the other hand, if we can capture this heterogeneity in a framework similar to ours that makes use of currently available surveillance data, we can explore more sophisticated, spatially targeted control strategies.
+However, incorporating spatial structure into our model would involve a substantial statistical and computational burden.
+Case reports data for the city of Vitoria are available for each of the city's 75 neighborhoods, but embedding the mechanistic model explicitly into a metapopulation framework leads to a very large, unwieldy mechanistic model that likely exacerbates existing difficulties inherent in fitting ODE models to data \cite{Girolami2008, Calderhead2011}.
+Developing such spatial models and the methods necessary to deal with them is a promising area of future research.
 
 \subsection*{Conclusions}
+
+Mosquito surveillance is a valuable tool in managing dengue fever, but is most powerful when integrated into a larger modeling framework.
+We have developed a simple yet realistic mechanistic model of dengue fever spread that allows us to combine mosquito and clinical surveillance to estimate the latent mosquito demographic rates that govern the critical bottleneck between mosquito abundance and human disease.
+That we have done this in a fully hierarchical Bayesian framework allows for a thorough accounting of uncertainty that can be carried through to the evaluation of different control strategies.
+This combination of model features meets many of the recommendations of \cite{Morrison2008, Achee2015} for developing more effective, biologically grounded, and data-driven dengue control policies and offers a building block on which these tools can be developed in the future.
 

### 2017-10-11 Analysis/plots.r
Color-blind friendly figure 5

@@ -214,7 +214,7 @@ ggplot(stacked, aes(week, value, color = factor(year))) +
   facet_grid(variable~., scales = "free_y", switch = "y") + 
   geom_vline(xintercept = 3) +
   theme_classic() +
-  scale_color_discrete(guide = F) + 
+  scale_color_brewer(palette = "Dark2", type = "qual", guide = F) + 
   scale_x_continuous(expand = c(0, 1)) +
   theme(strip.placement = "outside", strip.background = element_blank()) +
   ylab("")

### 2017-10-12 Manuscript/ms.tex
Small fixes

@@
 
-Many of the attempts to link these two sources of information have found a weak relationship between mosquito surveillance disease risk (CITATIONS).
+Many of the attempts to link these two sources of information have found a weak relationship between mosquito surveillance and disease risk (CITATIONS).
 These efforts have often relied on tests of the ability of mosquito abundance measures to predict human disease risk in regression models.
@@
 
-Several of the parameters in this model are assumed to be fixed and known (\ref{parameters}).
+Several of the parameters in this model are assumed to be fixed and known (Table 1).
 The population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
@@
 The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
-Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table \ref{parameters} for means and Supplemental Material for detailed explanations).
+Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).
 
@@
 
-Samples from the posterior distribution were generated using Hamiltonian Monte Carlo implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R (\cite{R2016}). 
+Samples from the posterior distribution were generated using Hamiltonian Monte Carlo implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R \cite{R2016}. 
 We ran 3 chains with different starting values for 10,000 iterations each, discarding the first 5,000 as burn-in.
@@
 The estimated latent mosquito demographic rates are smooth and strongly seasonal, with fairly narrow credible intervals (Figure \ref{latent}).
-Mosquito death rates fluctuate within the range estimated by \cite{Brady2013} and generally increase with temperature, as expected from laboratory studies \cite{Brady2013}(Figure \ref{temp}).
-The underlying process variances are small ($\sigma_nu = 0.01$, $\sigma_r = 0.0005$), and there is little residual structure in the process noise series, though there is a weak periodic signal in the $\epsilon_{\nu}$ (Supplemental Fig XX).
+Mosquito mortality rates fluctuate within the range estimated by \cite{Brady2013} and generally increase with temperature (Figure \ref{temp}).
+The underlying process variances are small ($\sigma_{\nu} = 0.01$, $\sigma_r = 0.0005$), and there is little residual structure in the process noise series, though there is a weak periodic signal in the $\epsilon_{\nu}$ (Supplemental Fig XX).
 
 Modifying these estimated demographic rates to simulate targeted control efforts, we find that control applied in the third week of the year is the most effective at preventing cases of dengue fever (though weeks 2 and 4 are also very close, Figure \ref{control}A).
-In fact, increasing the mosquito death rate by just 5\% in the third week of 2009, 2010, and 2011 would prevent 13,988 (reported) cases of dengue fever (with 80\% credible interval of 11,130 to 15,811).
+In fact, increasing the mosquito mortality rate by just 5\% in the third week of 2009, 2010, and 2011 would prevent 13,988 (reported) cases of dengue fever (with 80\% credible interval of 11,130 to 15,811).
 Most of these gains come by way of substantially reducing the size of the large 2011 outbreak, while also further damping the smaller 2010 and 2012 outbreaks (Figure \ref{control}B).
@@
 The third week is also generally before the estimated seasonal peak in mosquito abundance, which occurs between weeks 6 and 9 (Figure \ref{timing}).
-Annual peaks in estimated mosquito death rate are more variable, but do tend to fall between weeks 2 and 7.
+Annual peaks in estimated mosquito mortality rate are more variable, but tend to fall between weeks 2 and 7.
 Though mosquito death rates tend to be high early in the year, the transmission risk (i.e.
@@
 Though the estimated mosquito mortality process seems biologically reasonable, we are still employing it to link mosquito abundance and human disease at a relatively large spatial scale, assuming homogeneous mixing between mosquitoes and humans throughout the city.
-However, structured human movement within the city is likely to be a major driver of heterogeneous human-mosquito mixing (\cite{Adams2009, Cosner2009a, Stoddard2009, Dalziel2013}.
+However, structured human movement within the city is likely to be a major driver of heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009, Dalziel2013}.
 In addition spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
+This heterogeneity can have substantial consequences for disease dynamics.
+In fact, in a modeling study \cite{Dye1986, Hasibeder1988} found that introducing heterogeneous mixing always increases $R_0$ relative to homogeneous mixing.
 
-This heterogeneity has substantial consequences for disease dyamics.
-In fact, in a modeling study \cite{Dye1986, Hasibeder1988} found that introducing heterogeneous mixing always increases $R_0$ relative to homogeneous mixing.
 Though this suggests that our model may overestimate the efficacy of mosquito control, it also points the way towards more efficient and effective control methods.
-If disease risk is driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Canyon1999, Yoon2012}, control policies that preferentially targets these subsets of the mosquito population are likely to be very effective.
-Identifying these subsets is in practice difficult.
-One way to do this is through the use of contact-tracing, which can identify and target specific premises that have been implicated in transmission \cite{Vazquez-Prokopec2017}.
+If disease risk is driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Canyon1999, Yoon2012}, control policies that preferentially target these subsets of the mosquito population are likely to be very effective.
+Identifying these subsets is difficult in practice.
+One possible strategy is through the use of contact-tracing, which can identify and target specific premises that have been directly implicated in transmission \cite{Vazquez-Prokopec2017}.
 Though effective, this strategy is extremely labor-intensive and would be difficult to implement in endemic areas.
 
-On the other hand, if we can capture this heterogeneity in a framework similar to ours that makes use of currently available surveillance data, we can explore more sophisticated, spatially targeted control strategies.
+On the other hand, if we can capture this heterogeneity in a modeling framework similar to ours, that makes use of currently available surveillance data, we can explore more sophisticated, spatially targeted control strategies.
 However, incorporating spatial structure into our model would involve a substantial statistical and computational burden.
-Case reports data for the city of Vitoria are available for each of the city's 75 neighborhoods, but embedding the mechanistic model explicitly into a metapopulation framework leads to a very large, unwieldy mechanistic model that likely exacerbates existing difficulties inherent in fitting ODE models to data \cite{Girolami2008, Calderhead2011}.
+Case reports data for the city of Vitoria are available for each of the city's 75 neighborhoods, but using those neighborhoods to embed the mechanistic model explicitly into a metapopulation framework leads to a very large, unwieldy mechanistic model that likely exacerbates existing difficulties inherent in fitting ODE models to data \cite{Girolami2008, Calderhead2011}.
 Developing such spatial models and the methods necessary to deal with them is a promising area of future research.
@@
 That we have done this in a fully hierarchical Bayesian framework allows for a thorough accounting of uncertainty that can be carried through to the evaluation of different control strategies.
-This combination of model features meets many of the recommendations of \cite{Morrison2008, Achee2015} for developing more effective, biologically grounded, and data-driven dengue control policies and offers a building block on which these tools can be developed in the future.
+This combination of model features meets many of the recommendations of \cite{Morrison2008, Achee2015} for developing more effective, biologically grounded, and data-driven dengue control policies and offers a building block on which these tools can be further developed in the future.
 

### 2017-10-12 Manuscript/supplement.tex
Adding figures to supplement.

@@
 
+\newpage
+
+\section*{Supplemental figures}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/figS1.eps}
+\caption{
+Posterior estimates of process noise forcing the harmonic oscillators for mosquito mortality (A) and growth (B) rates.  Median posterior estimate (black line) and 80\% credible interval (gray band).
+}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/figS2.eps}
+\caption{
+Posterior distribution of process noise variances for mosquito mortality (A) and growth (B) rates.
+}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/figS3.eps}
+\caption{
+Posterior distribution of epidemiological parameters.
+The black line indicates the prior mean, as set from the literature.
+A: latent period in human host ($1/\rho$), in weeks.
+B: rate of infectious decay in human host ($\gamma$) in weeks $^{-1}$.
+C: period of cross-immunity ($1/\delta$) in weeks.
+}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/figS4.eps}
+\caption{Posterior distribution of initial conditions.
+A: initial proportion of the human population that is susceptible.
+B: initial number of exposed individuals.
+C: initial number of infectious individuals.
+}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/figS5.eps}
+\caption{Posterior check of autocorrelation structure in the case reports (A) and trapped mosquitoes (B) time series.
+Points indicate the observed autocorrelation at the given lag, and vertical lines give the 80\% posterior credible interval of the autocorrelation in the estimated time series.
+}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/figS6.eps}
+\caption{
+Posterior distribution of the total number of cases reported (A) and the total number of mosquitoes captured (B).  Thick line indicates the observed totals.
+}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/figS7.eps}
+\caption{
+Posterior distributions of the minimum and maximum weekly reported cases and trapped mosquitoes. Thick line indicates the observed minima and maxima.
+}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/figS8.eps}
+\caption{
+Estimated number of (reported) cases prevented by larval control (i.e decreasing the birth rate, $b_t = r_t + d_t$, by 5\%) applied in a given week.
+Posterior median (black line) and 80\% credible interval (gray band).
+Note that this is almost identical to the results from adult control.
+}
+\end{figure}
+
 \end{document}

### 2017-10-12 Analysis/plots.r

@@ -225,8 +225,7 @@ dev.off()
 # Supplemental figures
 #===============================================================================
 
-#===============================================================================
-# epsilons and sigmas
+# Figure S1: epsilons ==========================================================
 
 eps_dv <- rstan::extract(sim, "eps_dv", permute = T)[[1]] %>% 
   adply(2, quantile, c(0.1, 0.5, 0.9)) %>% 
@@ -238,48 +237,86 @@ eps_rv <- rstan::extract(sim, "eps_rv", permute = T)[[1]] %>%
   mutate(week = as.numeric(X1))
 names(eps_rv) <- c("X1", "min", "med", "max", "week")
 
+postscript("Manuscript/figures/figS1.eps",
+           width = 5.2, height = 3,
+           family = "ArialMT")
+
 ggplot(eps_dv, aes(week, med)) + 
   geom_ribbon(aes(ymin = min, ymax = max), fill = "grey70") +
   geom_line() + 
   theme_classic() + 
-  ylab("epsilon_dv")
+  scale_x_continuous(expand = c(0, 1)) +
+  ylab(expression(epsilon[d])) +
+  ggtitle("A") -> figs1.a
 
 ggplot(eps_rv, aes(week, med)) + 
   geom_ribbon(aes(ymin = min, ymax = max), fill = "grey70") +
   geom_line() + 
   theme_classic() + 
-  ylab("epsilon_rv")
+  scale_x_continuous(expand = c(0, 1)) +
+  ylab(expression(epsilon[r])) +
+  ggtitle("B") -> figs2.b
 
+grid.arrange(figs1.a, figs2.b, nrow = 1)
+
+dev.off()
+
+# Figure S2: sigmas ============================================================
+
+postscript("Manuscript/figures/figS2.eps",
+           width = 6, height = 4,
+           family = "ArialMT")
+
+par(mfrow = c(1, 2))
 rstan::extract(sim, "sigmadv", permute = T)[[1]] %>% hist(main = "", xlab = expression(sigma[d]), freq = F)
 rstan::extract(sim, "sigmarv", permute = T)[[1]] %>% hist(main = "", xlab = expression(sigma[r]), freq = F)
 
-#===============================================================================
-# epidemiological parameters
+dev.off()
+
+# Figure S3: epidemiological parameters ===================================================
 
[…]

### 2017-10-12 Manuscript/ms.tex
References to supplemental figures

@@
 The estimated posterior median case reports explain 91\% of the variation in the observed time series, while the posterior median mosquito trap counts explain 45\% of the variation in the observed time series.
-In addition, posterior predictive checks show that the model reproduces the the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Fig XX - XX).
-Posterior means for the three epidemiological parameters ($\rho, \gamma, \delta$) are close to the literature values used to set their prior means (Supplemental Fig XX).
+In addition, posterior predictive checks show that the model reproduces the the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 5 - 7).
+Posterior means for the three epidemiological parameters ($\rho, \gamma, \delta$) are close to the literature values used to set their prior means (Supplemental Figure 3).
 
@@
 Mosquito mortality rates fluctuate within the range estimated by \cite{Brady2013} and generally increase with temperature (Figure \ref{temp}).
-The underlying process variances are small ($\sigma_{\nu} = 0.01$, $\sigma_r = 0.0005$), and there is little residual structure in the process noise series, though there is a weak periodic signal in the $\epsilon_{\nu}$ (Supplemental Fig XX).
+The underlying process variances are small ($\sigma_{\nu} = 0.01$, $\sigma_r = 0.0005$, Supplemental Figure 2), and there is little residual structure in the process noise series, though there is a weak periodic signal in the $\epsilon_{\nu}$ (Supplemental Figure 1).
 
@@
 Most of these gains come by way of substantially reducing the size of the large 2011 outbreak, while also further damping the smaller 2010 and 2012 outbreaks (Figure \ref{control}B).
-Targeting larvae instead of adults (simulated by reducing the birth rate ($b_t = r_t + d_t$) by 5\%) produces nearly identical results.
+Targeting larvae instead of adults (simulated by reducing the birth rate ($b_t = r_t + d_t$) by 5\%) produces nearly identical results (Supplemental Figure 8).
 

### 2017-10-17 Manuscript/ms.tex
Small changes from Jennifer

@@
 
-Adult \textit{Aedes aegypti} surveillance data (e.g. time series of mosquito trap counts) is insufficient on its own to estimate mosquito mortality rates.
-However, we can borrow from the field of 'integrated population modeling' and combine multiple sources of information to estimate mosquito demographic rates \cite{Schaub2010}.
+Adult \textit{Aedes aegypti} surveillance data (i.e. time series of mosquito trap counts) is insufficient on its own to estimate mosquito mortality rates.
+However, we can borrow from the field of integrated population modeling and combine multiple sources of information to estimate mosquito demographic rates \cite{Schaub2010}.
 Since mosquito mortality is a critical bottleneck in the transmission process, it can be informed from the observed transmission dynamics (i.e.
@@
 
-The complete differential equation model is then given as:
+The differential equations governing the human population are then given as:
 \begin{align} 
@@
 \frac{dI}{dt} &= \rho E - (\gamma + b)I\\
-\frac{dR}{dt} &= \gamma I - (\delta + b)R\\
+\frac{dR}{dt} &= \gamma I - (\delta + b)R
+\end{align}
+while the equations governing the mosquito (vector) population are:
+\begin{align}
 \frac{dV_N}{dt} & = r V_N - \phi_q \tau_t V_N \\
@@
 \end{align}
-where $\omega = 2\pi / 52$, $d(t) = d_0 \exp(\nu(t))$, and
+where the angular frequency of the oscillator $\omega = 2\pi / 52$, the mosquito death rate $d(t) = d_0 \exp(\nu(t))$, and
 \begin{align}
@@
 
-To connect the differential equation model the the observed case reports, we add an extra state, $C$, that collects the cumulative number of transitions from the exposed to infectious class (assuming that case reporting coincides with the onset of symptoms).
+To connect the differential equation model to the observed case reports, we add an extra state, $C$, that collects the cumulative number of transitions from the exposed to infectious class (assuming that case reporting coincides with the onset of symptoms).
 We then model the number of new cases reported in week $t$ ($y_t$) as:
@@
 \end{equation}
-where $V_{Ct} - V_{Ct-1}$ is the number of new mosquitoes captured in week $t$, and $\eta_q$ controls overdispersion relative to the Poisson.
+where $V_{Ct} - V_{Ct-1}$ is the number of new mosquitoes captured in week $t$, and $\eta_q$ controls overdispersion relative to the Poisson distribution.
 
@@
 Several of the parameters in this model are assumed to be fixed and known (Table 1).
-The population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
+The human population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
 To maintain identifiability, the transmission rate ($\lambda$) and case reporting probability ($\phi_y$) are also fixed at literature values.
@@
 \hline
-$N$ & Population size & 327801 & Vitoria Census\\
+$N$ & Human population size & 327801 & Vitoria Census\\
 $1/d$ & Human life-span & 76 years & Vitoria Census\\
@@
 
-Given samples from the posterior as obtained above, we can simulate the efficacy of different control strategies.
+Given samples from the posterior distribution as obtained above, we can simulate the efficacy of different control strategies.
 In particular, we investigate the optimal timing of a single pulse of control within the year (e.g., if the city only has enough resource to fog for adult mosquitoes once a year, in what week is that control going to prevent the most cases?).
@@
 Though mosquito death rates tend to be high early in the year, the transmission risk (i.e.
-the probability of an exposed mosquito surviving long enough to be infectious, $\frac{\rho_v}{\rho_v + d}$) does not pick up until later in the year.
+the probability of an exposed mosquito surviving long enough to be infectious, $\left(\frac{\rho_v}{\rho_v + d}\right)$ generally peaks several weeks later and again later in the year.
 
@@
 Median posterior prediction (line) and 80\% credible interval.
-B: Weekly predicted case reports resulting from implementing control in week 3.
+B: Weekly predicted case reports resulting from implementing control in week 3 of each year (timing indicated by vertical red lines).
 Median posterior case reports from model without (dashed line) and with control (solid line) and 80\% credible interval of controlled dynamics (gray band).
@@
 
-On the other hand, if we can capture this heterogeneity in a modeling framework similar to ours, that makes use of currently available surveillance data, we can explore more sophisticated, spatially targeted control strategies.
+On the other hand, if we can capture this heterogeneity in a modeling framework similar to ours, using currently available surveillance data, we can explore more sophisticated, spatially targeted control strategies.
 However, incorporating spatial structure into our model would involve a substantial statistical and computational burden.

### 2017-10-17 Analysis/plots.r

@@ -165,7 +165,6 @@ postscript("Manuscript/figures/fig4.eps",
 ggplot(adult_reduction, aes(week, med)) + 
   geom_ribbon(aes(ymin = min, ymax = max), fill = "grey70") + 
   geom_line() +
-  geom_abline(slope = 0, color = "grey20") + 
   theme_classic() + 
   scale_x_continuous(expand = c(0, 1)) +
   xlab("week of control") +
@@ -176,6 +175,7 @@ ggplot(post, aes(tot.week, ycmed)) +
   geom_ribbon(aes(ymin = ycmin, ymax = ycmax), fill = "grey70") + 
   geom_line() + 
   geom_line(aes(tot.week, yhat), linetype = 2) +
+  geom_vline(xintercept = c(56, 108, 160), color = "tomato3") + 
   theme_classic() +
   scale_y_continuous(expand = c(0, 0)) + 
   scale_x_continuous(expand = c(0, 1)) +

### 2017-10-17 Manuscript/ms.tex
Adding a few citations

@@
 Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
-To be effective, control efforts should be guided by consistent, high resolution mosquito surveillance \cite{Morrison2008}.
+To be effective, control efforts need to be guided by consistent, high resolution mosquito surveillance \cite{Morrison2008}.
 The "MI-Dengue" system, implemented by the company Ecovec and deployed throughout cities in Brazil, uses a city-wide grid of mosquito sticky-traps (MosquiTRAPs) for spatial prioritization of source reduction and larvicide application \cite{Eiras2009}.
@@
 
-Many of the attempts to link these two sources of information have found a weak relationship between mosquito surveillance and disease risk (CITATIONS).
+Many of the attempts to link these two sources of information have found a weak relationship between mosquito surveillance and disease risk \cite{Bowman2014, Cromwell2017}.
 These efforts have often relied on tests of the ability of mosquito abundance measures to predict human disease risk in regression models.

### 2017-10-18 Manuscript/ms.tex
Reorganizing introduction

--- 2017-10-17 Manuscript/ms.tex
@@
 Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
-To be effective, control efforts should be guided by consistent, high resolution mosquito surveillance \cite{Morrison2008}.
-The "MI-Dengue" system, implemented by the company Ecovec and deployed throughout cities in Brazil, uses a city-wide grid of mosquito sticky-traps (MosquiTRAPs) for spatial prioritization of source reduction and larvicide application \cite{Eiras2009}.
-Comparing cities with and without this system, \cite{Pepin2013} estimated that it prevented 27,191 cases of dengue fever from 2007 - 2011.
-Though this system has been effective, the mosquito indices and thresholds used to prioritize control efforts remain somewhat ad-hoc.
-As noted in \cite{Morrison2008}, entomological indices are only relevant in the context of local epidemiology, and thus control policies will be most effective when they synthesize both entomological and clinical surveillance.
+There is limited evidence of the success of existing control programs \cite{Esu2010}, possibly due to very low mosquito abundance thresholds for sustaining transmission \cite{Scott2010a}.
+Because of this, there is a growing recognition that effective control needs to be guided by high quality vector surveillance and quantitative tools that synthesize that vector data with clinical surveillance, account for local epidemiology, and connect easily to decision making \cite{Morrison2008}.
+Moreover, mosquito control needs to be guided by an understanding of the link between mosquito surveillance indices and disease risk so that the mosquitoes most responsible for transmission can be targeted.
 
-Many of the attempts to link these two sources of information have found a weak relationship between mosquito surveillance and disease risk (CITATIONS).
-These efforts have often relied on tests of the ability of mosquito abundance measures to predict human disease risk in regression models.
-For example, \cite{Pepin2015} found that models that included MosquiTRAP surveillance data failed to predict weekly dengue cases any better than models that included lagged case data alone.
-However, such tests fail to account for the mechanistic, non-linear relationship between mosquito population dynamics and dengue epidemiology.
-The cycle of transmission between humans and dengue is controlled not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period \cite{Achee2015}.
+Many of the attempts to establish this link have found a weak relationship between mosquito surveillance and disease risk \cite{Bowman2014, Pepin2015, Cromwell2017}.
+However, such tests fail to account for the complex, non-linear interactions that mediate the relationship between mosquito abundance and human disease.
+In particular, the ability of mosquitoes to contribute to DENV transmission depends critically on the level of prior immunity in the human population \cite{Scott2010a} and by mosquito demographic processes themselves.
+The cycle of transmission between humans and mosquitoes is controlled not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period \cite{Achee2015}.
 In fact, whether or not an exposed mosquito will survive long enough to become infectious represents a critical bottleneck in the transmission process.
 
-Adult \textit{Aedes aegypti} surveillance data (i.e. time series of mosquito trap counts) is insufficient on its own to estimate mosquito mortality rates.
-However, we can borrow from the field of integrated population modeling and combine multiple sources of information to estimate mosquito demographic rates \cite{Schaub2010}.
-Since mosquito mortality is a critical bottleneck in the transmission process, it can be informed from the observed transmission dynamics (i.e.
-reports of human cases), given an appropriate underlying mechanistic model.
+The importance of these processes in governing human disease risk highlight the need to integrate mechanistic modeling into the quantitative tools used to target control efforts.
+Such mechanistic models can often perform better than even sophisticated statistical models in describing and predicting population dynamics \cite{Reilly2005}.
+Moreover, mechanistic models can provide "what-if" tools that can be used to predict the effect of management actions \cite{Buckland2007}.
+Differential equation models provide a natural way to capture the processes above and to link human epidemiology to mosquito demographic processes.
+As noted above, mosquito mortality is a critical component of this link, but cannot be estimated from mosquito surveillance alone.
+Given its role in mediating transmission, however, mosquito mortality can be informed by a combination of mosquito and clinical surveillance, borrowing from the field of integrated population modeling \cite{Shaub2010}.
 
-In this paper, we propose such a mechanistic model to connect mosquito abundance and survival with dengue epidemiology.
-We then embed this model into a Bayesian statistical framework that allows us to estimate latent time series of mosquito demographic parameters from time series of mosquito trap counts and reported cases of dengue fever.
-Lastly, we use the fitted model to explore the optimal timing of adult control interventions.
+Many cities in Brazil generate the necessary vector surveillance data through the use of the MI-Dengue system, implemented by the company Ecovec.
+This system deploys a city-wide grid of mosquito sticky traps (MosquiTRAPs) that can be used to target control in highly infested areas \cite{Eiras2009}.
+This system has been shown to be effective, preventing an estimated 27,191 cases of dengue fever from 2007 - 2011 across the cities in which it was deployed \cite{Pepin2013}.
+Despite this, the mosquito indices and thresholds used to prioritize control efforts remain somewhat ad-hoc.
+
+In this paper, we develop a modeling framework to integrate MI-Dengue and clinical surveillance in order to provide data-driven and mechanistically-grounded guidance for targeted mosquito control.
+Specifically, we develop a mechanistic differential equation model of dengue transmission that we embed in a hierarchical Bayesian statistical framework.
+This allows us to estimate latent time series of mosquito demographic parameters from available time series of mosquito trap counts and reported cases of dengue fever.
+We then use the fitted model to explore the optimal timing of adult mosquito control interventions.
 

### 2017-10-19 Manuscript/ms.tex
Fleshing out implementation

@@
 
-Samples from the posterior distribution were generated using Hamiltonian Monte Carlo implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R \cite{R2016}. 
+Sampling from the posterior distribution in non-linear differential equation models is often very difficult, due to multimodality, variable parameter sensitivities (e.g.
+small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations among parameters \cite{Reilly2005, Girolami2008, Calderhead2011}.
+In addition, the computational burden of numerically solving the differential equation model at each iteration means that sampling efficiency is crucial.
+These problems can be mitigated by introducing stochasticity through the $\epsilon_t$ in the mosquito demographic processes.
+This allows the data to pull the latent states closer even when the parameters are far from optimal values \cite{Leander2014}.
+This reduces multimodality and allows gradient-based methods like Hamiltonain Monte Carlo (HMC) to more easily and efficiently traverse the posterior.
+Samples from the posterior distribution were thus generated using HMC implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R \cite{R2016}. 
 We ran 3 chains with different starting values for 10,000 iterations each, discarding the first 5,000 as burn-in.

### 2017-10-19 Manuscript/ms.tex
Adding some discussion about transmission risk

@@
 Annual peaks in estimated mosquito mortality rate are more variable, but tend to fall between weeks 2 and 7.
-Though mosquito death rates tend to be high early in the year, the transmission risk (i.e.
-the probability of an exposed mosquito surviving long enough to be infectious, $\left(\frac{\rho_v}{\rho_v + d}\right)$ generally peaks several weeks later and again later in the year.
+Transmission risk (i.e.
+the probability of an exposed mosquito surviving long enough to be infectious, assuming exponential wait times in the differential equation, $\left(\frac{\rho_v}{\rho_v + d}\right)$ is noisy, but is generally highest from week 40 through week 10.
 
@@
 Each colored line represents a different year: dark green = 2008, orange = 2009, purple = 2010, pink = 2011, light green = 2012.
-A: posterior median estimate of dengue case reports per week.
-B: posterior median estimate of trapped mosquitoes per week.
+A: posterior median estimate of new dengue cases per week.
+B: posterior median estimate of number of mosquitoes per person each week.
 C: posterior median estimate of weekly mosquito death rate.
@@
 
-In our model, the relationship between mosquito abundance and human disease risk is controlled by mosquito mortality, which, together with the extrinsic incubation period ($\rho_v$), controls how  many exposed mosquitoes survive long enough to become infectious.
+In our model, the relationship between mosquito abundance and human disease risk is controlled by mosquito mortality, which, together with the extrinsic incubation period ($\rho_v$), controls the probability of an exposed mosquito surviving long enough to be infectious $\left(\frac{\rho_v}{\rho_v + d}\right)$.
+Since the extrinsic incubation period is forced by temperature, the time series of transmission risk is noisy, but is highest during what is generally the dengue off-season.
+That these periods of high transmission risk do not correspond to periods of high disease is likely due to the feedbacks between mosquito population dynamics and human epidemiology.
+Even though transmission risk is high and mosquito abundance is climbing late in the year, it takes a relatively long time for enough infection to build in the human population for an outbreak to occur.
+Despite these seasonal fluctuations, transmission risk is always fairly low (less than 0.4), suggesting that throughout the year, relatively few mosquitoes may be responsible for transmission.
+
 As implemented, mosquito mortality is the only source of uncertainty/stochasticity in the transmission process.

### 2017-10-23 Manuscript/ms.tex
Fleshing out discussion of control

--- 2017-10-19 Manuscript/ms.tex
@@
 As noted above, mosquito mortality is a critical component of this link, but cannot be estimated from mosquito surveillance alone.
-Given its role in mediating transmission, however, mosquito mortality can be informed by a combination of mosquito and clinical surveillance, borrowing from the field of integrated population modeling \cite{Shaub2010}.
+Given its role in mediating transmission, however, mosquito mortality can be informed by a combination of mosquito and clinical surveillance, borrowing from the field of integrated population modeling \cite{Schaub2010}.
 
@@
 When faced with limited resources with which to deploy that control, these results suggest that focusing efforts early in the year, specifically in the second to fourth weeks, will lead to the largest reductions in disease.
-Because adult control (increased the mortality rate) and larval control (decreasing the birth rate) produce very similar reductions in disease, the effect of control is likely driven by reducing the standing population of adult mosquitoes, rather than by limiting the transmission potential of exposed or infectious mosquitoes.
-Targeting mosquitoes early in the year as dengue just begins to spread can disrupt the transmission cycle and prevent further amplification and feedback.
+Because adult control (increasing the mortality rate) and larval control (decreasing the birth rate) produce nearly identical reductions in disease, the effect of control is likely driven by reducing the standing population of adult mosquitoes, rather than by limiting the transmission potential of exposed or infectious mosquitoes.
+The lag between optimal control and both mosquito population dynamics and human disease highlights the importance of early intervention and the need to account for the nonlinear processes governing transmission.
+Targeting the mosquito population early in the year as dengue just begins to spread can disrupt the transmission cycle and prevent further amplification and feedback.
 
-The importance of reducing the population of adult mosquitoes agrees with the results of \cite{Burattini2008} whose model of dengue dynamics in Singapore suggests that killing adult mosquitoes is the most effective way to control an epidemic.
-However, they find that larval control is most effective at preventing resurgence of dengue epidemics, and thus that a mixed strategy that includes both larval and adult control is likely to be most effective in the long term.
-The utility of mixed control strategies is further supported by \cite{Pinho2010}, whose model of dengue in Salvador, Brazil suggests that adult control alone is insufficient to bring $R_0$ below 1 and eliminate spread.
+Discussion of the effectiveness of adult control often focuses on the fact that it produces only a brief decline in mosquito abundance \cite{Newton1992, Esu2010}.
+When applied at the peak of an epidemic, this reduction in abundance is likely to have a relatively small effect, as the population rebounds while there are still a large number of infectious human hosts available \cite{Newton1992, Burattini2008}.
+However, adult control can also have an effect by shortening mosquito lifespan and preventing mosquitoes from living long enough to progress through the extrinsic incubation period and become infectious.
+This forms the basis for much of the theory of adult control \cite{Burattini2008, Morrison2008, Smith2012}, but modeling this effect in a standard differential equation framework is difficult.
+In particular the assumption of exponential wait times implicit in a compartmental model is perhaps ill-suited to modeling the race between mosquito lifespan and incubation period.
+The long tails of the exponential distribution may overestimate the probability that a mosquito survives long enough to become infectious and underestimate the effect of decreasing mosquito lifespan.
+More accurate models of mosquito lifespan and incubation periods could incorporate explicit delays (i.e.
+assuming a fixed incubation period \cite{Burattini2008}), or additional chained exposed compartments (i.e.
+assuming the incubation period is gamma-distributed, \cite{Lloyd2001}).
+Including such mechanisms in our model would allow us to better differentiate the effects of larval and adult control and provide a more accurate picture of the effectiveness of adult control.
+In fact, modeling studies that assume exponential wait times suggest that adult control is likely to be ineffective when applied during an epidemic \cite{Newton1992, Pinho2010}, while similar models that include explicit delays for the extrinsic incubation period show adult control being much more effective \cite{Burattini2008}.
+
+Thus, we may underestimate the effect of mosquito control applied later in the year when infectious mosquitoes are prevalent.
+However, our relatively simple model nonetheless shows that reducing mosquito abundance early in the year can be highly effective at preventing dengue cases later in the year.
+This effectiveness is driven primarily by the timing and could be achieved by either adult or larval control strategies.
+Though the mechanisms above might suggest further advantages for adult control, a mixed strategy of adult and larval control may be most effective in the long-term (\cite{Burattini2008, Pinho2010}.
 Fully exploring such combined strategies requires a more complete model of mosquito life history and the lags in population dynamics induced by the mosquito's aquatic stage.
-The lack of such an aquatic phase in our model makes it more suitable for exploring the immediate, direct effects of adult control on transmission, rather than for studying the lagged feedbacks on mosquito population dynamics (and their knock-on effects on transmission).
-Though models of mosquito development can be very complex, e.g.
-\cite{Magori2009}, our model could be extended to incorporate a compartment for the aquatic phase (as in \cite{Burattini2008, Pinho2010}), though at the cost of introducing several new parameters to the model.
+These lags likely mean that larval control should likely be applied prior to week 3, in order to achieve the desired reduction in the adult population in week 3.
+Models of mosquito development can be very complex, e.g.
+\cite{Magori2009}, but our model could be extended to account for these lags through a compartment for the aquatic phase (as in \cite{Burattini2008, Pinho2010}), though at the cost of introducing several new parameters.
 
-In our model, the relationship between mosquito abundance and human disease risk is controlled by mosquito mortality, which, together with the extrinsic incubation period ($\rho_v$), controls how  many exposed mosquitoes survive long enough to become infectious.
 As implemented, mosquito mortality is the only source of uncertainty/stochasticity in the transmission process.
@@
 Our implementation of the mortality process is conceptually similar to the empirical forcing functions developed by \cite{Hooker2015}.
-In that framework, \cite{Hooker2015} allow parameters of differential equation models to change over time and use the resulting time series to differentiate causes of lack-of-fit.
+In that framework, \cite{Hooker2015} allow the parameters of a differential equation model to change over time and use the resulting time series to differentiate causes of lack-of-fit.
 Though we do not employ the same explicit testing framework here, we treat the estimated mosquito mortality time series in the same spirit.

### 2017-10-23 Manuscript/ms.tex
Working on discussion of Hooker 2015

@@
 As implemented, mosquito mortality is the only source of uncertainty/stochasticity in the transmission process.
-As a result, mosquito mortality could be soaking up other sources of stochasticity or model misspecification.
+As a result, the estimated mosquito mortality could be soaking up other sources of stochasticity or model misspecification.
 Our implementation of the mortality process is conceptually similar to the empirical forcing functions developed by \cite{Hooker2015}.
-In that framework, \cite{Hooker2015} allow the parameters of a differential equation model to change over time and use the resulting time series to differentiate causes of lack-of-fit.
-Though we do not employ the same explicit testing framework here, we treat the estimated mosquito mortality time series in the same spirit.
-In particular, we can check whether our estimated mosquito mortality time series is mechanistically plausible.
-Our estimated mosquito mortalities generally fall within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
-The fact that our estimated mortality rate varies seasonally and increases with temperature also agrees with the empirical literature on mosquito survival.
+In that framework, \cite{Hooker2015} estimate nonparametric functions in time that modify either the differential equations or specific model parameters to provide a good fit to the data.
+These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables.
+Though we do not employ the same explicit testing framework as \cite{Hooker2015}, we treat the estimated mosquito mortality time series in the same spirit.
+In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission.
+Despite this, our estimated mosquito mortalities generally fall within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
+The fact that our estimated mortality rate varies seasonally and increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is perhaps more extreme than usually estimated in laboratory studies.
 These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.

### 2017-10-23 Manuscript/ms.tex
Spatial heterogeneity discussion and future work

@@
 Though the estimated mosquito mortality process seems biologically reasonable, we are still employing it to link mosquito abundance and human disease at a relatively large spatial scale, assuming homogeneous mixing between mosquitoes and humans throughout the city.
-However, structured human movement within the city is likely to be a major driver of heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009, Dalziel2013}.
+However, structured human movement within the city is likely to be a major driver of heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009}.
 In addition spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
 This heterogeneity can have substantial consequences for disease dynamics.
-In fact, in a modeling study \cite{Dye1986, Hasibeder1988} found that introducing heterogeneous mixing always increases $R_0$ relative to homogeneous mixing.
+In fact, \cite{Dye1986, Hasibeder1988} found that introducing heterogeneous mixing to models of vector-borne disease spread always increases $R_0$ relative to homogeneous mixing.
 
 Though this suggests that our model may overestimate the efficacy of mosquito control, it also points the way towards more efficient and effective control methods.
-If disease risk is driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Canyon1999, Yoon2012}, control policies that preferentially target these subsets of the mosquito population are likely to be very effective.
-Identifying these subsets is difficult in practice.
+Though our results suggest that applying control early in the year is critical to disrupting city-wide disease spread, targeting the city's entire mosquito population is likely infeasible and spatial prioritization of control is also necessary.
+If disease risk is driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Yoon2012}, control policies that preferentially target these subsets of the mosquito population are likely to be very effective.
+These subsets can be difficult to identify in practice.
 One possible strategy is through the use of contact-tracing, which can identify and target specific premises that have been directly implicated in transmission \cite{Vazquez-Prokopec2017}.
@@
 
-On the other hand, if we can capture this heterogeneity in a modeling framework similar to ours, using currently available surveillance data, we can explore more sophisticated, spatially targeted control strategies.
-However, incorporating spatial structure into our model would involve a substantial statistical and computational burden.
-Case reports data for the city of Vitoria are available for each of the city's 75 neighborhoods, but using those neighborhoods to embed the mechanistic model explicitly into a metapopulation framework leads to a very large, unwieldy mechanistic model that likely exacerbates existing difficulties inherent in fitting ODE models to data \cite{Girolami2008, Calderhead2011}.
+This highlights again the role that can be played by coupled mechanistic-statistical models like the one developed here that can make use of currently available surveillance data.
+Though aggregated here, MosquiTRAP surveillance data are available at the level of the individual trap, and case reports for the city of Vitoria are available for each of the city's 75 neighborhoods.
+Spatial structure and human movement can be incorporated into models of dengue transmission, e.g.
+through the methods outlined in \cite{Cosner2015}, but at the cost of substantial statistical and computational burden.
+For instance, embedding the model presented here into a metapopulation framework based on the city's neighborhoods leads to a very large, unwieldy mechanistic model.
+Moreover, without external data on human movement \cite{Stoddard2009}, it would be very difficult to identify its role in disease dynamics.
+These dimensionality and identifiability issues would likely exacerbate existing difficulties inherent in fitting ODE models to data \cite{Girolami2008, Calderhead2011}.
+These problems could be somewhat ameliorated through the use of high-dimensional filtering techniques (Citations) or gradient matching methods.
+Intermediate methods that combine city-scale mechanistic modeling with lower-level statistical approaches may be the most promising route, e.g.
+use models like the one presented here to develop temporal targeting of control, while using more traditional methods to target control spatially within the recommended temporal window.
+
 Developing such spatial models and the methods necessary to deal with them is a promising area of future research.

### 2017-10-24 Manuscript/ms.tex
Cleaning up spatial modeling paragraph

@@
 This effectiveness is driven primarily by the timing and could be achieved by either adult or larval control strategies.
-Though the mechanisms above might suggest further advantages for adult control, a mixed strategy of adult and larval control may be most effective in the long-term (\cite{Burattini2008, Pinho2010}.
+Though the mechanisms above might suggest further advantages for adult control, a mixed strategy of adult and larval control may be most effective in the long-term \cite{Burattini2008, Pinho2010}.
 Fully exploring such combined strategies requires a more complete model of mosquito life history and the lags in population dynamics induced by the mosquito's aquatic stage.
@@
 
-This highlights again the role that can be played by coupled mechanistic-statistical models like the one developed here that can make use of currently available surveillance data.
-Though aggregated here, MosquiTRAP surveillance data are available at the level of the individual trap, and case reports for the city of Vitoria are available for each of the city's 75 neighborhoods.
-Spatial structure and human movement can be incorporated into models of dengue transmission, e.g.
+Due to these difficulties, new methods are needed that can make use of currently available surveillance data to more effectively guide spatial prioritization of control.
+Though aggregated for our model, MosquiTRAP surveillance data are available at the level of the individual trap, and case reports for the city of Vitoria are available for each of the city's 75 neighborhoods.
+To interface with these data, spatial structure and human movement could be incorporated into models of dengue transmission, e.g.
 through the methods outlined in \cite{Cosner2015}, but at the cost of substantial statistical and computational burden.
 For instance, embedding the model presented here into a metapopulation framework based on the city's neighborhoods leads to a very large, unwieldy mechanistic model.
-Moreover, without external data on human movement \cite{Stoddard2009}, it would be very difficult to identify its role in disease dynamics.
+Moreover, without external data on human movement (e.g. as discussed in \cite{Stoddard2009}), it would be very difficult to identify its role in disease dynamics.
 These dimensionality and identifiability issues would likely exacerbate existing difficulties inherent in fitting ODE models to data \cite{Girolami2008, Calderhead2011}.
-These problems could be somewhat ameliorated through the use of high-dimensional filtering techniques (Citations) or gradient matching methods.
-Intermediate methods that combine city-scale mechanistic modeling with lower-level statistical approaches may be the most promising route, e.g.
-use models like the one presented here to develop temporal targeting of control, while using more traditional methods to target control spatially within the recommended temporal window.
-
 Developing such spatial models and the methods necessary to deal with them is a promising area of future research.
+In the absence of such methods, spatial targeting of control based on more traditional methods (e.g.
+local entomological thresholds) could be combined with temporal targeting based on city-scale mechanistic modeling.
+This would allow mosquito control efforts to exploit the critical timing identified by the mechanistic modeling, while also taking advantage of the spatial richness of the surveillance data to allocate limited resources.
 

### 2017-10-24 Manuscript/ms.tex
Tweaking conclusions

@@
 We have developed a simple yet realistic mechanistic model of dengue fever spread that allows us to combine mosquito and clinical surveillance to estimate the latent mosquito demographic rates that govern the critical bottleneck between mosquito abundance and human disease.
-That we have done this in a fully hierarchical Bayesian framework allows for a thorough accounting of uncertainty that can be carried through to the evaluation of different control strategies.
+The mechanistic framework allows us to capture important lags and feedbacks in the transmission process and to identify critical intervention points that would not be apparent otherwise.
+The fully hierarchical Bayesian framework in which the mechanistic model is embedded allows for a thorough accounting of uncertainty that can be carried through to the evaluation of different control strategies.
 This combination of model features meets many of the recommendations of \cite{Morrison2008, Achee2015} for developing more effective, biologically grounded, and data-driven dengue control policies and offers a building block on which these tools can be further developed in the future.

### 2017-10-24 Manuscript/ms.tex
Addressing some of Jennifer's comments

@@
 Then we sum the total number of cases reported in this scenario and compare to the data to get an estimate of the number of cases prevented.
-To keep our simulations conservative, and to avoid pushing the model into the unrealistic range of dengue eradication, we set $\Delta = 0.05$.
+To keep our simulations conservative relative to field estimates of the mortality induced by spraying \cite{Esu2010}, and to avoid pushing the model into the unrealistic range of dengue eradication, we set $\Delta = 0.05$.
 We also only simulate the effect of control in the middle three years, as the effectiveness of control in the first year is strongly affected by initial conditions, and we lack complete data for the last year.
@@
 In addition, posterior predictive checks show that the model reproduces the the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 5 - 7).
-Posterior means for the three epidemiological parameters ($\rho, \gamma, \delta$) are close to the literature values used to set their prior means (Supplemental Figure 3).
+The posterior distributions of the three epidemiological parameters ($\rho, \gamma, \delta$) differ from their priors, though they do still cover the literature values used to set their prior means (Supplemental Figure 3).
+The estimated latent period in a human host ($1/\rho$) is influenced most strongly by the data, with a posterior mean of 1.4 weeks compared to a prior mean of 0.87 weeks.
 

### 2017-10-24 Manuscript/ms.tex
Small edits and cleanin up

@@
 Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
-There is limited evidence of the success of existing control programs \cite{Esu2010}, possibly due to very low mosquito abundance thresholds for sustaining transmission \cite{Scott2010a}.
-Because of this, there is a growing recognition that effective control needs to be guided by high quality vector surveillance and quantitative tools that synthesize that vector data with clinical surveillance, account for local epidemiology, and connect easily to decision making \cite{Morrison2008}.
+There is limited direct evidence of the success of existing control programs \cite{Esu2010}, possibly due to very low mosquito abundance thresholds for sustaining transmission \cite{Scott2010a}.
+Because of this, there is a growing recognition that effective control needs to be guided by high quality vector surveillance, together with quantitative tools that synthesize that surveillance with clinical surveillance, account for local epidemiology, and connect easily to decision making \cite{Morrison2008}.
 Moreover, mosquito control needs to be guided by an understanding of the link between mosquito surveillance indices and disease risk so that the mosquitoes most responsible for transmission can be targeted.
 
-Many of the attempts to establish this link have found a weak relationship between mosquito surveillance and disease risk \cite{Bowman2014, Pepin2015, Cromwell2017}.
-However, such tests fail to account for the complex, non-linear interactions that mediate the relationship between mosquito abundance and human disease.
-In particular, the ability of mosquitoes to contribute to DENV transmission depends critically on the level of prior immunity in the human population \cite{Scott2010a} and by mosquito demographic processes themselves.
+Many of the attempts to establish this link have found a weak relationship \cite{Bowman2014, Pepin2015, Cromwell2017}.
+However, these attempts often fail to account for the complex, non-linear interactions that mediate the relationship between mosquito abundance and human disease.
+In particular, the ability of mosquitoes to contribute to DENV transmission depends critically on the level of prior immunity in the human population \cite{Scott2010a} and on mosquito demographic processes themselves.
 The cycle of transmission between humans and mosquitoes is controlled not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period \cite{Achee2015}.
@@
 The importance of these processes in governing human disease risk highlight the need to integrate mechanistic modeling into the quantitative tools used to target control efforts.
-Such mechanistic models can often perform better than even sophisticated statistical models in describing and predicting population dynamics \cite{Reilly2005}.
+Such mechanistic models can often perform better than complex autoregressive statistical models in describing and predicting population dynamics \cite{Reilly2005}.
 Moreover, mechanistic models can provide "what-if" tools that can be used to predict the effect of management actions \cite{Buckland2007}.
-Differential equation models provide a natural way to capture the processes above and to link human epidemiology to mosquito demographic processes.
+Differential equation models provide a natural way to encode the relevant biological mechanisms and to link human epidemiology to mosquito demographic processes.
 As noted above, mosquito mortality is a critical component of this link, but cannot be estimated from mosquito surveillance alone.
@@
 
-In this paper, we develop a modeling framework to integrate MI-Dengue and clinical surveillance in order to provide data-driven and mechanistically-grounded guidance for targeted mosquito control.
+In this paper, we develop a modeling framework to integrate MI-Dengue and clinical surveillance from the city of Vitoria, Brazil in order to provide data-driven and mechanistically-grounded guidance for targeted mosquito control.
 Specifically, we develop a mechanistic differential equation model of dengue transmission that we embed in a hierarchical Bayesian statistical framework.
@@
 
-Vitoria is a coastal city and the capital of the state of Espirito Santo, Brazil, with a population of 327,801 in the city proper, and an additional 1.5 million people in the greater metropolitan area.
-Since 2008, the company Ecovec has monitored mosquito abundance for the city using approximately 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a grid across the city.
+Vitoria is a coastal city and the capital of the state of Espirito Santo, Brazil, with a population of 327,801.
+Since 2008, the Ecovec has monitored mosquito abundance for the city using approximately 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a grid across the city.
 Each trap is checked weekly and the mosquitoes inside counted and identified, providing us with 243 weeks (2008 through week 34 of 2012) of counts of gravid female \emph{Aedes aegypti}.
@@
 Infection with one serotype confers life-long immunity to that serotype, along with temporary immunity to other serotypes.  
-As this cross-immunity wanes, antibodies from the previous infection can actually result in antibody-dependent enhancement (ADE), wherein hosts are more susceptible to infection with the other serotypes and more likely to develop severe symptoms (i.e.
+As this cross-immunity wanes, antibodies from the previous infection can result in antibody-dependent enhancement (ADE), wherein human hosts are more susceptible to infection with the other serotypes and more likely to develop severe symptoms (i.e.
 dengue haemorrhagic fever or dengue shock syndrome).
 The strength and duration of these different inter-serotype interactions are not well understood, though different models suggest that temporary cross-immunity alone (without ADE) is sufficient to reproduce observed multi-annual dynamics in Thailand \cite{Wearing2006,Reich2013}.
-Similarly, \cite{Aguiar2013} find that capturing both primary and secondary infections and the period of cross-immunity is critical, but that explicitly including all four serotypes (at great cost to model complexity) does not perform much better than a two serotype model.
+Similarly, \cite{Aguiar2013} find that capturing primary and secondary infections and the period of cross-immunity is critical, but that explicitly including all four serotypes (at great cost to model complexity) does not perform much better than a two serotype model.
 
@@
 As such, we simplify the model of \cite{Wearing2006} to an SEIRS framework, which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
-In this framework, susceptible individuals ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
-Following a latent period ($\frac{1}{\rho}$), exposed individuals become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
-Infectious individuals recover at rate $\gamma$ and subsequently remain immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
+In this framework, susceptible humans ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
+Following a latent period ($\frac{1}{\rho}$), exposed humans become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
+Infectious humans recover at rate $\gamma$ and subsequently remain immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
 
-Similarly, susceptible mosquitoes ($V_S$) become exposed by biting infectious humans and pass through a temperature-dependent latent period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
+Similarly, susceptible mosquitoes ($V_S$) become exposed by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
 Total mosquito population size ($V_N$) is controlled by stochastic, seasonally varying growth rate ($r(t)$) and death rate ($d(t)$).
@@
 
-We model the centered and log-transformed mosquito mortality rate ($\nu$) and the per-capita growth rate ($r$) as forced harmonic oscillators with natural periods of one year:
+We model the centered and log-transformed mosquito mortality rate ($\nu$) and the per-capita mosquito growth rate ($r$) as forced harmonic oscillators with natural periods of one year:
 \begin{align}
@@
 \end{align}
-where the angular frequency of the oscillator $\omega = 2\pi / 52$, the mosquito death rate $d(t) = d_0 \exp(\nu(t))$, and
+where the angular frequency of the oscillator, $\omega = 2\pi / 52$, the mosquito death rate $d(t) = d_0 \exp(\nu(t))$, and
 \begin{align}
@@
 \end{equation}
-where $\phi_y$ is the reporting probability, $C_t - C_{t-1}$ is the number of new infectious individuals in week $t$, and $\eta_y$ controls the overdispersion relative to the Poisson.
+where $\phi_y$ is the reporting probability, $C_t - C_{t-1}$ is the number of new infectious humans in week $t$, and $\eta_y$ controls the overdispersion relative to the Poisson distribution.
 
@@
 $1/\rho_{vt}$ & Extrinsic incubation period & $7\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
-$\exp(\mu)$ & Mean mosquito death rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
+$d_0$ & Mean mosquito death rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
 $V_{E0}$ & Initial exposed mosquitoes &  0 & \\
@@
 
-Sampling from the posterior distribution in non-linear differential equation models is often very difficult, due to multimodality, variable parameter sensitivities (e.g.
+Sampling from the posterior distribution in non-linear differential equation models is often very difficult due to multimodality, variable parameter sensitivities (e.g.
 small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations among parameters \cite{Reilly2005, Girolami2008, Calderhead2011}.
 In addition, the computational burden of numerically solving the differential equation model at each iteration means that sampling efficiency is crucial.
-These problems can be mitigated by introducing stochasticity through the $\epsilon_t$ in the mosquito demographic processes.
+These problems can be mitigated somewhat by introducing stochasticity through the $\epsilon_t$ in the mosquito demographic processes.
 This allows the data to pull the latent states closer even when the parameters are far from optimal values \cite{Leander2014}.
@@
 To obtain the posterior distribution of the number of cases prevented by implementing control in week $i$ each year, for each posterior sample $k$, we simulate an increase in the latent mosquito death rate during each week $i$, e.g.,
-$\hat{d}^{(k)}_{yi} = (1 + \Delta) d^{(k)}_{yi}$, where $y$ indexes the year, and $i$ indexes the week of the year.
+$\hat{d}^{(k)}_{ji} = (1 + \Delta) d^{(k)}_{ji}$, where $j$ indexes the year, and $i$ indexes the week of the year.
 Then we sum the total number of cases reported in this scenario and compare to the data to get an estimate of the number of cases prevented.
@@
 Annual peaks in estimated mosquito mortality rate are more variable, but tend to fall between weeks 2 and 7.
-Transmission risk (i.e.
-the probability of an exposed mosquito surviving long enough to be infectious, assuming exponential wait times in the differential equation, $\left(\frac{\rho_v}{\rho_v + d}\right)$ is noisy, but is generally highest from week 40 through week 10.
 
@@
 C: posterior median estimate of weekly mosquito death rate.
-D: posterior median estimate of weekly transmission risk ($\frac{\rho_v}{\rho_v + d}$).
 }
@@
 The long tails of the exponential distribution may overestimate the probability that a mosquito survives long enough to become infectious and underestimate the effect of decreasing mosquito lifespan.
-More accurate models of mosquito lifespan and incubation periods could incorporate explicit delays (i.e.
+More accurate models of the extrinsic incubation period could incorporate explicit delays (i.e.
 assuming a fixed incubation period \cite{Burattini2008}), or additional chained exposed compartments (i.e.
@@
 
-Thus, we may underestimate the effect of mosquito control applied later in the year when infectious mosquitoes are prevalent.
-However, our relatively simple model nonetheless shows that reducing mosquito abundance early in the year can be highly effective at preventing dengue cases later in the year.
+Thus, our model may underestimate the effect of adult mosquito control applied later in the year when infectious mosquitoes are prevalent.
+However, our model nonetheless shows that reducing mosquito abundance early in the year can be highly effective at preventing dengue cases later in the year.
 This effectiveness is driven primarily by the timing and could be achieved by either adult or larval control strategies.
@@
 Fully exploring such combined strategies requires a more complete model of mosquito life history and the lags in population dynamics induced by the mosquito's aquatic stage.
-These lags likely mean that larval control should likely be applied prior to week 3, in order to achieve the desired reduction in the adult population in week 3.
+These lags likely mean that larval control should be applied prior to week 3, in order to achieve the desired reduction in the adult population in week 3.
 Models of mosquito development can be very complex, e.g.
@@
 Though we do not employ the same explicit testing framework as \cite{Hooker2015}, we treat the estimated mosquito mortality time series in the same spirit.
-In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission.
+In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission, either through the death rate itself or through another process.
 Despite this, our estimated mosquito mortalities generally fall within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
@@
 Though the estimated mosquito mortality process seems biologically reasonable, we are still employing it to link mosquito abundance and human disease at a relatively large spatial scale, assuming homogeneous mixing between mosquitoes and humans throughout the city.
-However, structured human movement within the city is likely to be a major driver of heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009}.
+However, structured human movement within the city is likely induce heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009}.
 In addition spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
 This heterogeneity can have substantial consequences for disease dynamics.
-In fact, \cite{Dye1986, Hasibeder1988} found that introducing heterogeneous mixing to models of vector-borne disease spread always increases $R_0$ relative to homogeneous mixing.
+In fact, \cite{Dye1986, Hasibeder1988} found that introducing heterogeneous mixing to models of vector-borne disease spread always increases $R_0$ relative to models with homogeneous mixing.
 
-Though this suggests that our model may overestimate the efficacy of mosquito control, it also points the way towards more efficient and effective control methods.
+Though this suggests that our model may overestimate the efficacy of mosquito control, it also points a way towards more efficient and effective control methods.
 Though our results suggest that applying control early in the year is critical to disrupting city-wide disease spread, targeting the city's entire mosquito population is likely infeasible and spatial prioritization of control is also necessary.
 If disease risk is driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Yoon2012}, control policies that preferentially target these subsets of the mosquito population are likely to be very effective.
-These subsets can be difficult to identify in practice.
-One possible strategy is through the use of contact-tracing, which can identify and target specific premises that have been directly implicated in transmission \cite{Vazquez-Prokopec2017}.
+However, these subsets can be difficult to identify in practice.
+One possible strategy is to employ contact-tracing, which can identify and target specific premises that have been directly implicated in transmission \cite{Vazquez-Prokopec2017}.
 Though effective, this strategy is extremely labor-intensive and would be difficult to implement in endemic areas.
@@
 For instance, embedding the model presented here into a metapopulation framework based on the city's neighborhoods leads to a very large, unwieldy mechanistic model.
-Moreover, without external data on human movement (e.g. as discussed in \cite{Stoddard2009}), it would be very difficult to identify its role in disease dynamics.
+Moreover, without external data on human movement (possible sources of which are discussed in \cite{Stoddard2009}), it would be very difficult to identify its role in disease dynamics.
 These dimensionality and identifiability issues would likely exacerbate existing difficulties inherent in fitting ODE models to data \cite{Girolami2008, Calderhead2011}.
@@
 local entomological thresholds) could be combined with temporal targeting based on city-scale mechanistic modeling.
-This would allow mosquito control efforts to exploit the critical timing identified by the mechanistic modeling, while also taking advantage of the spatial richness of the surveillance data to allocate limited resources.
+This would allow mosquito control efforts to exploit the critical timing identified by the mechanistic modeling, while also taking advantage of the spatial richness of the surveillance data to allocate limited resources within that window.
 
@@
 Mosquito surveillance is a valuable tool in managing dengue fever, but is most powerful when integrated into a larger modeling framework.
-We have developed a simple yet realistic mechanistic model of dengue fever spread that allows us to combine mosquito and clinical surveillance to estimate the latent mosquito demographic rates that govern the critical bottleneck between mosquito abundance and human disease.
+We have developed a simple yet realistic mechanistic model of dengue fever spread that allows us to combine mosquito and clinical surveillance to estimate the latent mosquito demographic rates that connect mosquito abundance and human disease.
 The mechanistic framework allows us to capture important lags and feedbacks in the transmission process and to identify critical intervention points that would not be apparent otherwise.
-The fully hierarchical Bayesian framework in which the mechanistic model is embedded allows for a thorough accounting of uncertainty that can be carried through to the evaluation of different control strategies.
-This combination of model features meets many of the recommendations of \cite{Morrison2008, Achee2015} for developing more effective, biologically grounded, and data-driven dengue control policies and offers a building block on which these tools can be further developed in the future.
+The fully hierarchical Bayesian framework in which the mechanistic model is embedded allows for a thorough accounting of uncertainty that is carried through to the evaluation of different control strategies.
+This combination of model features helps to meet the need for more effective, biologically grounded, and data-driven dengue control policies and offers a building block on which these tools can be further developed in the future.
 

### 2017-10-24 Analysis/plots.r
Dropping risk from fig 5

@@ -187,7 +187,7 @@ grid.arrange(fig4.a, fig4.b, ncol = 2)
 
 dev.off()
 
-# Figure 5: timing of control ==================================================
+# Figure 5: Timing of control ==================================================
 
 post <- mutate(post, 
                ymu = system[, , 8] %>% 
@@ -201,9 +201,9 @@ post <- mutate(post,
                  apply(1, function(x){rov / (rov + x)}) %>% 
                  apply(1, median))
 
-stacked <- melt(post, id.vars = c("week", "year"), measure.vars = c("ymu", "qmu", "dvmed", "risk"))
+stacked <- melt(post, id.vars = c("week", "year"), measure.vars = c("ymu", "qmu", "dvmed"))
 
-levels(stacked$variable) <- c("cases", "mosquitoes", "death rate", "transmission risk")
+levels(stacked$variable) <- c("cases", "mosquitoes", "death rate")
 
 postscript("Manuscript/figures/fig5.eps",
            width = 4, height = 6,

### 2017-10-24 Manuscript/ms.tex
Abstract text

@@
 
+Our ability to effectively prevent the transmission of the dengue virus through targeted control of its vector, \emph{Aedes aegypti}, depends critically on our understanding of the link between mosquito abundance and human disease risk.
+The mosquito and clinical surveillance data necessary to elucidate this link are widely collected, but have yet to be coupled with a modeling framework that accounts for the complex non-linear mechanisms involved in transmission, in particular the critical bottleneck imposed by mosquito mortality.
+Here we develop a differential equation model of dengue transmission and embed it in a Bayesian hierarchical framework that allows us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vitoria, Brazil.
+We then use the fitted model to explore the optimal timing of targeted adult control during the year.
+We find that control is most effective when deployed in the third week of the year, as mosquito abundance nears its peak and dengue transmission begins to ramp up.
+Intervening at this time disrupts transmission and prevents the amplification and feedback that leads to large outbreaks of disease, an insight not possible without the underlying mechanistic model.
+Though the ability of mosquito surveillance to predict human disease is often limited, this highlights the utility of this surveillance when integrated into a larger modeling framework.
+Grounding this modeling famework in the actual mechanisms of transmission will help to establish more effective and efficient dengue control policies that allow us to better target mosquitoes that are most responsible for continuing disease spread.
+
 % Please keep the Author Summary between 150 and 200 words
@@
 % Author Summary not valid for PLOS ONE submissions.   
-\section*{Author summary}
+%\section*{Author summary}
 

### 2017-11-08 Manuscript/supplement.tex
Updating initial conditions and EIP in supplement

@@
 \gamma & = 3.5\gamma'\\
-S_0' & \sim \text{Normal}(-0.4, 0.2)\\
-E_0' & \sim \text{Normal}(-9, 0.6)\\
-I_0' & \sim \text{Normal}(-9, 0.6)\\
-R_0' & = 0\\
-S_0 & = \frac{\exp(S_0)}{\exp(S_0' + E_0' + I_0' + R_0')} \\
+S_0' & \sim \text{Beta}(4, 6)\\
+E_0' & \sim \text{Gamma}(10, 0.1)\\
+I_0' & \sim \text{Gamma}(6, 0.1)\\
+E_0 & = E_0' / N \\
+I_0 & = I_0' / N \\ 
+S_0 & = S_0'(1 - E_0 - I_0) \\
+R_0 & = 1 - S_0 - E_0 - I_0
 \end{align}
@@
 \item \textbf{Initial conditions} ($S_0, E_0, I_0, R_0$): The SEIRS model assumes a constant population size and is scaled so that $S_0+E_0+I_0+R_0 = 1$.
-We specify our prior on an unconstrained scale and transform to the simplex with the softmax transformation.
-To maintain identifiability of this transformation, we fix $R_0' = 0$.
-We choose the prior for $S_0'$  so that $S_0$ has a mean of roughly  0.4, with the mass concentrated between 0.2 and 0.6.
+We place priors on the initial number of exposed and infectious individuals, rather than the initial proportion, as it provides a more intuitive scale.
+We set the prior mean for $E_0'$ and $I_0'$ at 100 and 60 individuals respectively, to account for the fact that the annual epidemic should be just beginning, with infectious individuals lagging behind exposed.
+These means were chosen large enough to ensure sustained multi-annual transmission, but with a relatively large variance.
+We choose the prior for $S_0'$, the proportion of remaining individuals initially susceptible, to have a mean of roughly  0.4, with the mass concentrated between 0.2 and 0.6.
 This range produces dynamics with recurring annual outbreaks and avoids the extremes where the disease either does not spread or spreads far too quickly and then dies out.
 This is also roughly consistent with \cite{Cardoso2011a} who report notified cases for Vitoria from 1995 to 2008. 
-They report that the 2008 outbreak (where our data begins) was driven by serotypes 2 and 3, and 2009 by serotype 2 only, with 13590 reported cases in previous serotype 2 years.
-Assuming 12 unreported cases for every reported case, 163,080 total cases, or about 50\% of the population immune to serotype 2 at the start of 2008.
-We choose the priors for $E_0'$ and $I_0'$ so that the initial sizes of these compartments were less than 100 individuals, but with a mode slightly off from zero.
-This was again chosen primarily to ensure sustained multi-annual transmission.
+They report that the 2008 outbreak (where our data begins) was driven by serotypes 2 and 3, and 2009 by serotype 2 only, with 13590 reported serotype 2 cases in previous years.
+Assuming 12 unreported cases for every reported case, there should be 163,080 total cases, or about 50\% of the population, immune to serotype 2 at the start of 2008.
 \\
@@
 \item \textbf{Extrinsic incubation period} ($1/\rho_{vt}$): Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito  (the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
-To account for this, we make $1/\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground (URL/citation).
+To account for this, we make $1/\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground.
 We use the temperature relationship from the recent meta-analysis of \cite{Chan2012}, which incorporates a large number of records and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
-We use a slightly more conservative fitted relationship from a log-normal model that excludes right-censored data and produces a less extreme temperature dependence:
-\begin{align}
-\frac{1}{\rho_v} &= \exp \left(\mu + \frac{1}{14}\right)\\
-\mu & = \exp(1.9 - 0.04 T)
-\end{align}
+To correspond with the assumptions of the differential equation model, we use the relationship obtained under the assumption that the incubation period at any given temperature is exponentially-distributed:
+\begin{equation}
+\rho_v = 7\exp \left(0.2T - 8\right)\\
+\end{equation}
 This relationship is similar to those reported in \cite{Focks1995} and \cite{Tjaden2013}, though there is uncertainty about the exact shape of this relationship.

### 2017-11-08 Analysis/control.r
Code simulating control in week 12

@@ -136,3 +136,46 @@ week3 <- foreach(k = 1:nmcmc, .packages = c("rstan", "magrittr")) %dopar% {
 stopCluster(cl)
 
 saveRDS(week3, "Results/optimal.rds")
+
+#===============================================================================
+
+peak_cases <- ddply(data, .(year), summarise, max = which.max(obs))
+
+# Setting up parallel
+cl <- makeCluster(3, type = "SOCK")
+registerDoParallel(cl)
+
+# Parallel for-loop over mcmc iterations
+week12 <- foreach(k = 1:nmcmc, .packages = c("rstan", "magrittr")) %dopar% {
+  
+  init <- list(lapply(samples, extract_sample, k))
+  
+  control <- matrix(1.0, nrow = 243, ncol = 3)
+  control[(data$week == 12 & data$year > 2008 & data$year < 2012), 1] <- 1.05
+  
+  dat.stan <- list(T = 243,
+                   T_pred = 0,
+                   y = data$obs,
+                   q = data$q,
+                   tau = data$tau,
+                   rov = rov,
+                   control = control,
+                   pop = pop)
+  
+  sim <- sampling(model,
+                  data = dat.stan, 
+                  init = init, 
+                  iter = 1, 
+                  chains = 1,
+                  warmup = 0,
+                  algorithm = "Fixed_param")
+  
+  system <- rstan::extract(sim, "system", permute = T)[[1]][1, , ]
+  
+  return(system)
+}
+
+stopCluster(cl)
+
+saveRDS(week12, "Results/peak.rds")
+

### 2017-11-13 Manuscript/ms.tex
Updating description of supplement
Dropping "risk" from title

@@
 {\Large
-\textbf\newline{Linking mosquito surveillance to dengue risk through Bayesian mechanistic modeling}
+\textbf\newline{Linking mosquito surveillance to dengue fever through Bayesian mechanistic modeling}
 }
@@
 \paragraph*{S1 Text.}
-\label{S1_Diag}
-{\bf Full description of model and priors.}
+\label{S1}
+{\bf Full description of model and priors, along with all supplemental plots.}
 

### 2017-11-13 Analysis/citystan.r
Adding flexible Euler step size

@@ -39,6 +39,7 @@ rov <- 7 * exp(0.2 * covars$temp - 8)
 
 dat.stan <- list(T = 243,
                  T_pred = 0,
+                 steps = 56,
                  y =data$obs,
                  q = data$q,
                  tau = data$tau,
@@ -98,7 +99,7 @@ inits = list(list(S0 = 0.4,
 fit <- stan(file = "Code/seirs.stan", 
             data = dat.stan, 
             init = inits, 
-            iter = 10000, 
+            iter = 200, 
             chains = 3,
             control = list(adapt_delta = 0.99,
                            max_treedepth = 15))

### 2017-11-13 Analysis/control.r

@@ -69,6 +69,7 @@ reduction <- foreach(k = 1:nmcmc, .combine = "cbind", .packages = c("rstan", "ma
     
     dat.stan <- list(T = 243,
                      T_pred = 0,
+                     steps = 7,
                      y = data$obs,
                      q = data$q,
                      tau = data$tau,
@@ -113,6 +114,7 @@ week3 <- foreach(k = 1:nmcmc, .packages = c("rstan", "magrittr")) %dopar% {
   
   dat.stan <- list(T = 243,
                    T_pred = 0,
+                   steps = 7,
                    y = data$obs,
                    q = data$q,
                    tau = data$tau,
@@ -155,6 +157,7 @@ week12 <- foreach(k = 1:nmcmc, .packages = c("rstan", "magrittr")) %dopar% {
   
   dat.stan <- list(T = 243,
                    T_pred = 0,
+                   steps = 7,
                    y = data$obs,
                    q = data$q,
                    tau = data$tau,

### 2017-11-14 Analysis/citystan.r
Using stan ode solver

@@ -39,12 +39,11 @@ rov <- 7 * exp(0.2 * covars$temp - 8)
 
 dat.stan <- list(T = 243,
                  T_pred = 0,
-                 steps = 56,
-                 y =data$obs,
+                 ts = array(0:243, dim = c(244, 1)),
+                 y = data$obs,
                  q = data$q,
-                 tau = data$tau,
-                 rov = rov,
-                 control = matrix(1, nrow = 243, ncol = 3),
+                 tau = matrix(data$tau, nrow = 243, ncol = 1),
+                 rov = matrix(rov, nrow = 243, ncol = 1),
                  pop = pop)
 
 inits = list(list(S0 = 0.4,
@@ -99,7 +98,7 @@ inits = list(list(S0 = 0.4,
 fit <- stan(file = "Code/seirs.stan", 
             data = dat.stan, 
             init = inits, 
-            iter = 200, 
+            iter = 1000, 
             chains = 3,
             control = list(adapt_delta = 0.99,
                            max_treedepth = 15))

### 2017-11-14 Manuscript/ms.tex
Mevin's typos and style edits

@@
 The mosquito and clinical surveillance data necessary to elucidate this link are widely collected, but have yet to be coupled with a modeling framework that accounts for the complex non-linear mechanisms involved in transmission, in particular the critical bottleneck imposed by mosquito mortality.
-Here we develop a differential equation model of dengue transmission and embed it in a Bayesian hierarchical framework that allows us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vitoria, Brazil.
-We then use the fitted model to explore the optimal timing of targeted adult control during the year.
-We find that control is most effective when deployed in the third week of the year, as mosquito abundance nears its peak and dengue transmission begins to ramp up.
+We developed a differential equation model of dengue transmission and embedded it in a Bayesian hierarchical framework that allowed us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vitoria, Brazil.
+We used the fitted model to explore the optimal timing of targeted adult control during the year.
+We found that control was most effective when deployed in the third week of the year, as mosquito abundance neared its peak and dengue transmission began to ramp up.
 Intervening at this time disrupts transmission and prevents the amplification and feedback that leads to large outbreaks of disease, an insight not possible without the underlying mechanistic model.
@@
 Dengue fever is a massive global public health burden, with millions of cases per year in Brazil alone \cite{Bhatt2013}.  
-Since the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
+Because the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
 There is limited direct evidence of the success of existing control programs \cite{Esu2010}, possibly due to very low mosquito abundance thresholds for sustaining transmission \cite{Scott2010a}.
@@
 Such mechanistic models can often perform better than complex autoregressive statistical models in describing and predicting population dynamics \cite{Reilly2005}.
-Moreover, mechanistic models can provide "what-if" tools that can be used to predict the effect of management actions \cite{Buckland2007}.
+Moreover, mechanistic models can provide scenario-based tools that can be used to predict the effect of management actions \cite{Buckland2007}.
 Differential equation models provide a natural way to encode the relevant biological mechanisms and to link human epidemiology to mosquito demographic processes.
@@
 This system has been shown to be effective, preventing an estimated 27,191 cases of dengue fever from 2007 - 2011 across the cities in which it was deployed \cite{Pepin2013}.
-Despite this, the mosquito indices and thresholds used to prioritize control efforts remain somewhat ad-hoc.
+Despite this, the mosquito indices and thresholds used to prioritize control efforts remain somewhat \emph{ad-hoc}.
 
-In this paper, we develop a modeling framework to integrate MI-Dengue and clinical surveillance from the city of Vitoria, Brazil in order to provide data-driven and mechanistically-grounded guidance for targeted mosquito control.
+In this paper, we develop a modeling framework to integrate MI-Dengue and clinical surveillance from the city of Vitoria, Brazil to provide data-driven and mechanistically-grounded guidance for targeted mosquito control.
 Specifically, we develop a mechanistic differential equation model of dengue transmission that we embed in a hierarchical Bayesian statistical framework.
@@
 Each trap is checked weekly and the mosquitoes inside counted and identified, providing us with 243 weeks (2008 through week 34 of 2012) of counts of gravid female \emph{Aedes aegypti}.
-Dengue is a mandatory notifiable disease, and thus the city's Ministry of Health Secretary maintains a database of weekly notified probable dengue cases (i.e.
-medical care sought for dengue-like symptoms) for this same time period.
+Dengue is a mandatory notifiable disease, and thus the city's Ministry of Health Secretary maintains a database of weekly notified probable dengue cases (i.e., medical care sought for dengue-like symptoms) for this same time period.
 
@@
 Infection with one serotype confers life-long immunity to that serotype, along with temporary immunity to other serotypes.  
-As this cross-immunity wanes, antibodies from the previous infection can result in antibody-dependent enhancement (ADE), wherein human hosts are more susceptible to infection with the other serotypes and more likely to develop severe symptoms (i.e.
-dengue haemorrhagic fever or dengue shock syndrome).
+As this cross-immunity wanes, antibodies from the previous infection can result in antibody-dependent enhancement (ADE), wherein human hosts are more susceptible to infection with the other serotypes and more likely to develop severe symptoms (i.e., dengue haemorrhagic fever or dengue shock syndrome).
 The strength and duration of these different inter-serotype interactions are not well understood, though different models suggest that temporary cross-immunity alone (without ADE) is sufficient to reproduce observed multi-annual dynamics in Thailand \cite{Wearing2006,Reich2013}.
-Similarly, \cite{Aguiar2013} find that capturing primary and secondary infections and the period of cross-immunity is critical, but that explicitly including all four serotypes (at great cost to model complexity) does not perform much better than a two serotype model.
+Similarly, Aguiar \emph{et al.} \cite{Aguiar2013} found that capturing primary and secondary infections and the period of cross-immunity is critical, but that explicitly including all four serotypes (at great cost to model complexity) does not perform better than a two serotype model.
 
 Explicitly accounting for interactions between serotypes, even only two of the four, leads to a large and complex mechanistic model.
-Moreover, since dengue case reports do not include information on serotype, we do not have enough information to inform the dynamics of individual serotypes.
+Moreover, because dengue case reports do not include information on serotype, we do not have enough information to inform the dynamics of individual serotypes.
 As such, we simplify the model of \cite{Wearing2006} to an SEIRS framework, which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
@@
 \begin{align}
-\epsilon_{\nu i} & \sim \text{Normal}(0, \sigma_{\nu})\\
-\epsilon_{ri} & \sim \text{Normal}(0, \sigma_r),
+\epsilon_{\nu i} & \sim \text{Normal}(0, \sigma^2_{\nu})\\
+\epsilon_{ri} & \sim \text{Normal}(0, \sigma^2_r),
 \end{align}
@@
 The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
-Where possible, we place informative priors on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).
+Where possible, we specified informative priors on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).
 
@@
 
-Sampling from the posterior distribution in non-linear differential equation models is often very difficult due to multimodality, variable parameter sensitivities (e.g.
-small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations among parameters \cite{Reilly2005, Girolami2008, Calderhead2011}.
+Sampling from the posterior distribution in non-linear differential equation models is often very difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations among parameters \cite{Reilly2005, Girolami2008, Calderhead2011}.
 In addition, the computational burden of numerically solving the differential equation model at each iteration means that sampling efficiency is crucial.
@@
 Given samples from the posterior distribution as obtained above, we can simulate the efficacy of different control strategies.
-In particular, we investigate the optimal timing of a single pulse of control within the year (e.g., if the city only has enough resource to fog for adult mosquitoes once a year, in what week is that control going to prevent the most cases?).
-To obtain the posterior distribution of the number of cases prevented by implementing control in week $i$ each year, for each posterior sample $k$, we simulate an increase in the latent mosquito death rate during each week $i$, e.g.,
-$\hat{d}^{(k)}_{ji} = (1 + \Delta) d^{(k)}_{ji}$, where $j$ indexes the year, and $i$ indexes the week of the year.
-Then we sum the total number of cases reported in this scenario and compare to the data to get an estimate of the number of cases prevented.
+In particular, we investigated the optimal timing of a single pulse of control within the year (e.g., if the city only has enough resource to fog for adult mosquitoes once a year, in what week is that control going to prevent the most cases?).
+To obtain the posterior distribution of the number of cases prevented by implementing control in week $i$ each year, for each posterior sample $k$, we simulate an increase in the latent mosquito death rate during each week $i$ (e.g.,
+$\hat{d}^{(k)}_{ji} = (1 + \Delta) d^{(k)}_{ji}$ where $j$ indexes the year, and $i$ indexes the week of the year).
+Then we summed the total number of cases reported in this scenario and compare to the data to get an estimate of the number of cases prevented.
 To keep our simulations conservative relative to field estimates of the mortality induced by spraying \cite{Esu2010}, and to avoid pushing the model into the unrealistic range of dengue eradication, we set $\Delta = 0.05$.
-We also only simulate the effect of control in the middle three years, as the effectiveness of control in the first year is strongly affected by initial conditions, and we lack complete data for the last year.
+We also only simulated the effect of control in the middle three years, because the effectiveness of control in the first year is strongly affected by initial conditions, and we lack complete data for the last year.
 
@@
 The estimated posterior median case reports explain 91\% of the variation in the observed time series, while the posterior median mosquito trap counts explain 45\% of the variation in the observed time series.
-In addition, posterior predictive checks show that the model reproduces the the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 5 - 7).
+In addition, posterior predictive checks show that the model reproduces the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 5 - 7).
 The posterior distributions of the three epidemiological parameters ($\rho, \gamma, \delta$) differ from their priors, though they do still cover the literature values used to set their prior means (Supplemental Figure 3).
@@
 Mosquito mortality rates fluctuate within the range estimated by \cite{Brady2013} and generally increase with temperature (Figure \ref{temp}).
-The underlying process variances are small ($\sigma_{\nu} = 0.01$, $\sigma_r = 0.0005$, Supplemental Figure 2), and there is little residual structure in the process noise series, though there is a weak periodic signal in the $\epsilon_{\nu}$ (Supplemental Figure 1).
+The underlying posterior process standard deviations are small ($E(\sigma_{\nu}|\mathbf{y}) = 0.01$, $E(\sigma_r|\mathbf{y}) = 0.0005$, Supplemental Figure 2), and there is little residual structure in the process noise series, though there is a weak periodic signal in the $\epsilon_{\nu}$ (Supplemental Figure 1).
 
-Modifying these estimated demographic rates to simulate targeted control efforts, we find that control applied in the third week of the year is the most effective at preventing cases of dengue fever (though weeks 2 and 4 are also very close, Figure \ref{control}A).
+Modifying these estimated demographic rates to simulate targeted control efforts, we found that control applied in the third week of the year is the most effective at preventing cases of dengue fever (though weeks 2 and 4 are also very close, Figure \ref{control}A).
 In fact, increasing the mosquito mortality rate by just 5\% in the third week of 2009, 2010, and 2011 would prevent 13,988 (reported) cases of dengue fever (with 80\% credible interval of 11,130 to 15,811).
@@
 B: Weekly predicted case reports resulting from implementing control in week 3 of each year (timing indicated by vertical red lines).
-Median posterior case reports from model without (dashed line) and with control (solid line) and 80\% credible interval of controlled dynamics (gray band).
+Median posterior case reports from the model without (dashed line) and with control (solid line) and 80\% credible interval of controlled dynamics (gray band).
 }
@@
 Discussion of the effectiveness of adult control often focuses on the fact that it produces only a brief decline in mosquito abundance \cite{Newton1992, Esu2010}.
-When applied at the peak of an epidemic, this reduction in abundance is likely to have a relatively small effect, as the population rebounds while there are still a large number of infectious human hosts available \cite{Newton1992, Burattini2008}.
+When applied at the peak of an epidemic, this reduction in abundance is likely to have a relatively small effect, because the population rebounds while there are still a large number of infectious human hosts available \cite{Newton1992, Burattini2008}.
 However, adult control can also have an effect by shortening mosquito lifespan and preventing mosquitoes from living long enough to progress through the extrinsic incubation period and become infectious.
@@
 The long tails of the exponential distribution may overestimate the probability that a mosquito survives long enough to become infectious and underestimate the effect of decreasing mosquito lifespan.
-More accurate models of the extrinsic incubation period could incorporate explicit delays (i.e.
-assuming a fixed incubation period \cite{Burattini2008}), or additional chained exposed compartments (i.e.
-assuming the incubation period is gamma-distributed, \cite{Lloyd2001}).
+More accurate models of the extrinsic incubation period could incorporate explicit delays (i.e., assuming a fixed incubation period \cite{Burattini2008}), or additional chained exposed compartments (i.e., assuming the incubation period is gamma-distributed, \cite{Lloyd2001}).
 Including such mechanisms in our model would allow us to better differentiate the effects of larval and adult control and provide a more accurate picture of the effectiveness of adult control.
@@
 This effectiveness is driven primarily by the timing and could be achieved by either adult or larval control strategies.
-Though the mechanisms above might suggest further advantages for adult control, a mixed strategy of adult and larval control may be most effective in the long-term \cite{Burattini2008, Pinho2010}.
+Although the mechanisms above might suggest further advantages for adult control, a mixed strategy of adult and larval control may be most effective in the long-term \cite{Burattini2008, Pinho2010}.
 Fully exploring such combined strategies requires a more complete model of mosquito life history and the lags in population dynamics induced by the mosquito's aquatic stage.
-These lags likely mean that larval control should be applied prior to week 3, in order to achieve the desired reduction in the adult population in week 3.
-Models of mosquito development can be very complex, e.g.
-\cite{Magori2009}, but our model could be extended to account for these lags through a compartment for the aquatic phase (as in \cite{Burattini2008, Pinho2010}), though at the cost of introducing several new parameters.
+These lags likely mean larval control should be applied prior to week 3 to achieve the desired reduction in the adult population in week 3.
+Models of mosquito development can be very complex (e.g.
+\cite{Magori2009}) but our model could be extended to account for these lags through a compartment for the aquatic phase (as in \cite{Burattini2008, Pinho2010}), though at the cost of introducing several new parameters.
 
@@
 As a result, the estimated mosquito mortality could be soaking up other sources of stochasticity or model misspecification.
-Our implementation of the mortality process is conceptually similar to the empirical forcing functions developed by \cite{Hooker2015}.
-In that framework, \cite{Hooker2015} estimate nonparametric functions in time that modify either the differential equations or specific model parameters to provide a good fit to the data.
+Our implementation of the mortality process is conceptually similar to the empirical forcing functions developed by Hooker and Ellner \cite{Hooker2015}.
+In that framework, Hooker and Ellner \cite{Hooker2015} estimated nonparametric functions in time that modify either the differential equations or specific model parameters to provide a good fit to the data.
 These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables.
-Though we do not employ the same explicit testing framework as \cite{Hooker2015}, we treat the estimated mosquito mortality time series in the same spirit.
+We do not employ the same explicit testing framework as \cite{Hooker2015}, but we treat the estimated mosquito mortality time series in the same spirit.
 In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission, either through the death rate itself or through another process.
-Despite this, our estimated mosquito mortalities generally fall within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
-The fact that our estimated mortality rate varies seasonally and increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is perhaps more extreme than usually estimated in laboratory studies.
+Despite this, our estimated mosquito mortalities generally fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
+The fact that our estimated mortality rate varied seasonally and increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is perhaps more extreme than estimated in most laboratory studies.
 These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.
 
-Though the estimated mosquito mortality process seems biologically reasonable, we are still employing it to link mosquito abundance and human disease at a relatively large spatial scale, assuming homogeneous mixing between mosquitoes and humans throughout the city.
-However, structured human movement within the city is likely induce heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009}.
-In addition spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
+Even though the estimated mosquito mortality process seems biologically reasonable, we are still employing it to link mosquito abundance and human disease at a relatively large spatial scale, assuming homogeneous mixing between mosquitoes and humans throughout the city.
+However, structured human movement within the city is likely to induce heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009}.
+In addition, spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
 This heterogeneity can have substantial consequences for disease dynamics.
@@
 
-Though this suggests that our model may overestimate the efficacy of mosquito control, it also points a way towards more efficient and effective control methods.
-Though our results suggest that applying control early in the year is critical to disrupting city-wide disease spread, targeting the city's entire mosquito population is likely infeasible and spatial prioritization of control is also necessary.
+This suggests that our model may overestimate the efficacy of mosquito control, but it also points a way toward more efficient and effective control methods.
+Our results suggest that applying control early in the year is critical to disrupting city-wide disease spread, but targeting the city's entire mosquito population is likely infeasible and spatial prioritization of control is also necessary.
 If disease risk is driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Yoon2012}, control policies that preferentially target these subsets of the mosquito population are likely to be very effective.
@@
 Due to these difficulties, new methods are needed that can make use of currently available surveillance data to more effectively guide spatial prioritization of control.
-Though aggregated for our model, MosquiTRAP surveillance data are available at the level of the individual trap, and case reports for the city of Vitoria are available for each of the city's 75 neighborhoods.
-To interface with these data, spatial structure and human movement could be incorporated into models of dengue transmission, e.g.
-through the methods outlined in \cite{Cosner2015}, but at the cost of substantial statistical and computational burden.
+MosquiTRAP surveillance data are available at the level of the individual trap, and case reports for the city of Vitoria are available for each of the city's 75 neighborhoods.
+To interface with these data, spatial structure and human movement could be incorporated into models of dengue transmission  (e.g., through the methods outlined in \cite{Cosner2015}) but at the cost of substantial statistical and computational burden.
 For instance, embedding the model presented here into a metapopulation framework based on the city's neighborhoods leads to a very large, unwieldy mechanistic model.
@@
 Developing such spatial models and the methods necessary to deal with them is a promising area of future research.
-In the absence of such methods, spatial targeting of control based on more traditional methods (e.g.
-local entomological thresholds) could be combined with temporal targeting based on city-scale mechanistic modeling.
+In the absence of such methods, spatial targeting of control based on more traditional methods (e.g., local entomological thresholds) could be combined with temporal targeting based on city-scale mechanistic modeling.
 This would allow mosquito control efforts to exploit the critical timing identified by the mechanistic modeling, while also taking advantage of the spatial richness of the surveillance data to allocate limited resources within that window.
@@
 The mechanistic framework allows us to capture important lags and feedbacks in the transmission process and to identify critical intervention points that would not be apparent otherwise.
-The fully hierarchical Bayesian framework in which the mechanistic model is embedded allows for a thorough accounting of uncertainty that is carried through to the evaluation of different control strategies.
+The fully hierarchical Bayesian framework in which we embedded the mechanistic model allows for a thorough accounting of uncertainty that is carried through to the evaluation of different control strategies.
 This combination of model features helps to meet the need for more effective, biologically grounded, and data-driven dengue control policies and offers a building block on which these tools can be further developed in the future.
@@
 \label{S1}
-{\bf Full description of model and priors, along with all supplemental plots.}
+{Full description of model and priors, along with all supplemental plots.}
 

### 2017-11-14 Manuscript/ms.tex
Kim's typos and style edits

@@
 \\
-\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\textbf{2} National Wildlife Research Center, United States Department of Agriculture, Wildlife Services, Fort Collins, CO, USA
 \\
@@
 There is limited direct evidence of the success of existing control programs \cite{Esu2010}, possibly due to very low mosquito abundance thresholds for sustaining transmission \cite{Scott2010a}.
-Because of this, there is a growing recognition that effective control needs to be guided by high quality vector surveillance, together with quantitative tools that synthesize that surveillance with clinical surveillance, account for local epidemiology, and connect easily to decision making \cite{Morrison2008}.
+Because of this, there is a growing recognition that effective control needs to be guided by high quality vector surveillance, together with quantitative tools that synthesize vector surveillance with clinical surveillance, account for local epidemiology, and connect easily to decision making \cite{Morrison2008}.
 Moreover, mosquito control needs to be guided by an understanding of the link between mosquito surveillance indices and disease risk so that the mosquitoes most responsible for transmission can be targeted.
@@
 In particular, the ability of mosquitoes to contribute to DENV transmission depends critically on the level of prior immunity in the human population \cite{Scott2010a} and on mosquito demographic processes themselves.
-The cycle of transmission between humans and mosquitoes is controlled not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period \cite{Achee2015}.
+The cycle of transmission between humans and mosquitoes is influenced not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period in mosquitoes \cite{Achee2015}.
 In fact, whether or not an exposed mosquito will survive long enough to become infectious represents a critical bottleneck in the transmission process.
 
-The importance of these processes in governing human disease risk highlight the need to integrate mechanistic modeling into the quantitative tools used to target control efforts.
-Such mechanistic models can often perform better than complex autoregressive statistical models in describing and predicting population dynamics \cite{Reilly2005}.
-Moreover, mechanistic models can provide scenario-based tools that can be used to predict the effect of management actions \cite{Buckland2007}.
+The importance of mosquito demographic processes in governing human disease risk highlight the need to integrate mechanistic modeling into the quantitative tools used to target control efforts.
+Such mechanistic models can often perform better than complex autoregressive statistical models in describing and forecasting population dynamics \cite{Reilly2005}.
+Moreover, mechanistic models can provide scenario-based tools that can be used to predict the effect of management actions \cite{Buckland2007}, which cannot be done without explicit representation of mechanistic processes or data from case-control studies.
 Differential equation models provide a natural way to encode the relevant biological mechanisms and to link human epidemiology to mosquito demographic processes.
@@
 
-The model is able to capture the observed dynamics of both case reports and mosquito trap counts (Figure \ref{timeseries}).
-The estimated posterior median case reports explain 91\% of the variation in the observed time series, while the posterior median mosquito trap counts explain 45\% of the variation in the observed time series.
-In addition, posterior predictive checks show that the model reproduces the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 5 - 7).
+The model captured the observed dynamics of both case reports and mosquito trap counts (Figure \ref{timeseries}).
+The estimated posterior median case reports explained 91\% of the variation in the observed time series, while the posterior median mosquito trap counts explained 45\% of the variation in the observed time series.
+In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 5 - 7).
 The posterior distributions of the three epidemiological parameters ($\rho, \gamma, \delta$) differ from their priors, though they do still cover the literature values used to set their prior means (Supplemental Figure 3).
@@
 
-Modifying these estimated demographic rates to simulate targeted control efforts, we found that control applied in the third week of the year is the most effective at preventing cases of dengue fever (though weeks 2 and 4 are also very close, Figure \ref{control}A).
+Modifying these estimated demographic rates to simulate targeted control efforts, we found that control applied in the third week of the year (in January) is the most effective at preventing cases of dengue fever (though weeks 2 and 4 are also very close, Figure \ref{control}A).
 In fact, increasing the mosquito mortality rate by just 5\% in the third week of 2009, 2010, and 2011 would prevent 13,988 (reported) cases of dengue fever (with 80\% credible interval of 11,130 to 15,811).
@@
 \includegraphics[angle = 270, scale = 1]{figures/fig2.eps}
-\caption{{\bf Estimated latent mosquito growth and death rates.}
+\caption{{\bf Estimated latent mosquito population growth and death rates.}
 Median posterior estimate (line) and 80\% credible interval (gray band).  A: weekly mosquito population growth rate ($r_t$). B: weekly mosquito death rate ($d_t$).
@@
 The lag between optimal control and both mosquito population dynamics and human disease highlights the importance of early intervention and the need to account for the nonlinear processes governing transmission.
-Targeting the mosquito population early in the year as dengue just begins to spread can disrupt the transmission cycle and prevent further amplification and feedback.
+Targeting the mosquito population early in the year as dengue just begins to spread can disrupt the transmission cycle and prevent further amplification and feedback later in the year.
 
@@
 However, our model nonetheless shows that reducing mosquito abundance early in the year can be highly effective at preventing dengue cases later in the year.
-This effectiveness is driven primarily by the timing and could be achieved by either adult or larval control strategies.
+This effectiveness is driven primarily by the timing of the reduction in mosquito abundance and could be achieved by either adult or larval control strategies.
 Although the mechanisms above might suggest further advantages for adult control, a mixed strategy of adult and larval control may be most effective in the long-term \cite{Burattini2008, Pinho2010}.
@@
 This suggests that our model may overestimate the efficacy of mosquito control, but it also points a way toward more efficient and effective control methods.
-Our results suggest that applying control early in the year is critical to disrupting city-wide disease spread, but targeting the city's entire mosquito population is likely infeasible and spatial prioritization of control is also necessary.
+Our results suggest that applying control early in the year is critical to disrupting city-wide disease spread, but targeting the city's entire mosquito population is likely infeasible and spatial prioritization of control is also necessary \cite{Pepin2015}.
 If disease risk is driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Yoon2012}, control policies that preferentially target these subsets of the mosquito population are likely to be very effective.
@@
 
-Due to these difficulties, new methods are needed that can make use of currently available surveillance data to more effectively guide spatial prioritization of control.
+Due to these difficulties, new methods are needed that can make use of currently available surveillance data to more effectively guide spatial prioritization of control \cite{Pepin2015}.
 MosquiTRAP surveillance data are available at the level of the individual trap, and case reports for the city of Vitoria are available for each of the city's 75 neighborhoods.
@@
 Mosquito surveillance is a valuable tool in managing dengue fever, but is most powerful when integrated into a larger modeling framework.
-We have developed a simple yet realistic mechanistic model of dengue fever spread that allows us to combine mosquito and clinical surveillance to estimate the latent mosquito demographic rates that connect mosquito abundance and human disease.
+We have developed a simple yet realistic mechanistic model of dengue fever spread that allows us to combine mosquito and clinical surveillance data to estimate the latent mosquito demographic rates that connect mosquito abundance and human disease.
 The mechanistic framework allows us to capture important lags and feedbacks in the transmission process and to identify critical intervention points that would not be apparent otherwise.

### 2017-11-14 Manuscript/ms.tex
Affiliations

@@
 \\
-Clinton Leach\textsuperscript{1*},
+Clinton Leach\textsuperscript{1,2*},
 Colleen Webb\textsuperscript{1},
-Kim Pepin\textsuperscript{2},
-Alvaro Eiras\textsuperscript{3},
-Mevin Hooten\textsuperscript{4},
-Jennifer Hoeting\textsuperscript{4}
+Kim Pepin\textsuperscript{3},
+Alvaro Eiras\textsuperscript{4},
+Mevin Hooten\textsuperscript{2},
+Jennifer Hoeting\textsuperscript{2}
 
 \bigskip
-\textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\textbf{1} Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO, USA
 \\
-\textbf{2} National Wildlife Research Center, United States Department of Agriculture, Wildlife Services, Fort Collins, CO, USA
+\textbf{2} Department of Statistics, Colorado State University, Fort Collins, CO, USA
 \\
-\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\textbf{3} National Wildlife Research Center, United States Department of Agriculture, Wildlife Services, Fort Collins, CO, USA
 \\
+\textbf{4} Departamento de Parasitologia, Universidade Federal de Minas Gerais, Belo Horizonte, MG, Brazil
 \bigskip

### 2017-11-14 Manuscript/ms.tex
Expanding Table 1 to include all parameters

@@
 \hline
-$N$ & Human population size & 327801 & Vitoria Census\\
-$1/d$ & Human life-span & 76 years & Vitoria Census\\
+$N$ & Human population size & 327801 & \cite{vitpop} \\
+$1/d$ & Human life-span & 76 years & \cite{vitlong} \\
 $\lambda$ & Transmission rate & 4.87 week$^{-1}$ & \cite{Scott2000}\\
@@
 $1/\rho_{vt}$ & Extrinsic incubation period & $7\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
-$d_0$ & Mean mosquito death rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
+$d_0$ & Mean mosquito mortality rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
 $V_{E0}$ & Initial exposed mosquitoes &  0 & \\
@@
 $1/\delta$ & Period of cross-immunity & 97 weeks &  \cite{Reich2013}\\
+$\sigma_r$ & Standard deviation of mosquito growth rate forcing & $0^*$ & \\
+$\sigma_{\nu}$ & Standard deviation of mosquito mortality rate forcing & $0^*$ & \\
 $S_0$ & Proportion initially susceptible & 0.4 & \cite{Cardoso2011a} \\
@@
 $I_0$ & Proportion initially infectious & $8\times 10 ^ {-5}$ & \\
-$\phi_q$ & Per-trap mosquito capture rate & $2 \times 10^{-6}$ week$^{-1}$ & 
+$r_0$ & Initial mosquito population growth rate & 0 & \\
+$\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & \\
+$\phi_q$ & Per-trap mosquito capture rate & $2 \times 10^{-6}$ week$^{-1}$ & \\
+$\eta_y$ & Overdispersion of case reports & $0^*$ & \\
+$\eta_q$ & Overdispersion of mosquito trap counts & $0^*$ & \\
 \end{tabular}
@@
 \end{adjustwidth}
+$^*$ indicates prior mode, rather than mean.
 \end{table}

### 2017-11-15 Manuscript/ms.tex
Time indexing of parameters
More time indices

@@
 \begin{align}
-\frac{dV_N}{dt} & = r V_N - \phi_q \tau_t V_N \\
-\frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{vt} + d + \phi_q \tau_t)V_E\\
-\frac{dV_I}{dt} &= \rho_{vt} V_E - (d + \phi_q \tau_t) V_I\\
-\frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
+\frac{dV_N}{dt} & = r(t) V_N - \phi_q \tau(t) V_N \\
+\frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_E\\
+\frac{dV_I}{dt} &= \rho_{v}(t) V_E - (d(t) + \phi_q \tau(t)) V_I\\
+\frac{dV_C}{dt} & = \phi_q \tau(t) V_N\\
 V_S &= V_N - V_E - V_I.
@@
 \begin{equation}
-y_t  \sim \text{NegBin}(\phi_y (C_t - C_{t-1}), \eta_y),
+y_t  \sim \text{NegBin}(\phi_y (C(t) - C(t-1)), \eta_y),
 \end{equation}
-where $\phi_y$ is the reporting probability, $C_t - C_{t-1}$ is the number of new infectious humans in week $t$, and $\eta_y$ controls the overdispersion relative to the Poisson distribution.
+where $\phi_y$ is the reporting probability, $C(t) - C(t-1)$ is the number of new infectious humans in week $t$, and $\eta_y$ controls the overdispersion relative to the Poisson distribution.
 
@@
 \begin{equation}
-q_t \sim \text{NegBin}(V_{Ct} - V_{Ct-1}, \eta_q),
+q_t \sim \text{NegBin}(V_{C}(t) - V_{C}(t-1), \eta_q),
 \end{equation}
-where $V_{Ct} - V_{Ct-1}$ is the number of new mosquitoes captured in week $t$, and $\eta_q$ controls overdispersion relative to the Poisson distribution.
+where $V_{C}(t) - V_{C}(t-1)$ is the number of new mosquitoes captured in week $t$, and $\eta_q$ controls overdispersion relative to the Poisson distribution.
 

### 2017-11-15 Analysis/citystan.r
Adding control back in

@@ -44,6 +44,7 @@ dat.stan <- list(T = 243,
                  q = data$q,
                  tau = matrix(data$tau, nrow = 243, ncol = 1),
                  rov = matrix(rov, nrow = 243, ncol = 1),
+                 control = matrix(1, nrow = T, ncol = 2),
                  pop = pop)
 
 inits = list(list(S0 = 0.4,

### 2017-12-06 Manuscript/ms.tex
Reworking the introduction

@@
 
-Dengue fever is a massive global public health burden, with millions of cases per year in Brazil alone \cite{Bhatt2013}.  
+Dengue fever is a massive global public health burden, with millions of cases per year \cite{Bhatt2013}.  
 Because the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
-There is limited direct evidence of the success of existing control programs \cite{Esu2010}, possibly due to very low mosquito abundance thresholds for sustaining transmission \cite{Scott2010a}.
-Because of this, there is a growing recognition that effective control needs to be guided by high quality vector surveillance, together with quantitative tools that synthesize vector surveillance with clinical surveillance, account for local epidemiology, and connect easily to decision making \cite{Morrison2008}.
-Moreover, mosquito control needs to be guided by an understanding of the link between mosquito surveillance indices and disease risk so that the mosquitoes most responsible for transmission can be targeted.
+Though there have been documented successes, there is limited evidence for the long-term sustainability and effectiveness of these control programs \cite{Morrison2008}.
+Because of this, there is a growing recognition that effective control needs to be guided by high quality vector surveillance, together with quantitative tools that synthesize vector surveillance with clinical surveillance, account for local epidemiology, and connect easily to local decision making \cite{Morrison2008, Scott2010b}.
+Moreover, mosquito control needs to be guided by an understanding of the link between mosquito surveillance abundance and disease risk so that the mosquitoes most responsible for transmission can be targeted \cite{Scott2010a, Scott2010b}.
 
-Many of the attempts to establish this link have found a weak relationship \cite{Bowman2014, Pepin2015, Cromwell2017}.
-However, these attempts often fail to account for the complex, non-linear interactions that mediate the relationship between mosquito abundance and human disease.
-In particular, the ability of mosquitoes to contribute to DENV transmission depends critically on the level of prior immunity in the human population \cite{Scott2010a} and on mosquito demographic processes themselves.
-The cycle of transmission between humans and mosquitoes is influenced not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period in mosquitoes \cite{Achee2015}.
-In fact, whether or not an exposed mosquito will survive long enough to become infectious represents a critical bottleneck in the transmission process.
+Many of the attempts to establish this link have found a weak relationship between mosquito surveillance and incidence of dengue fever\cite{Bowman2014, Pepin2015, Cromwell2017}.
+However, these attempts often do not account for the complex, non-linear interactions that mediate the relationship between mosquito abundance and human disease.
+In particular, host immunity is a key intrinsic driver of infectious disease dynamics, and conditions favorable for transmission can only lead to an outbreak of disease when there is a sufficiently large population of susceptible hosts \cite{Koelle2004, Koelle2005}.
+As such, the ability of mosquitoes to contribute to DENV transmission depends critically on the level of prior immunity in the human population \cite{Scott2010a}.
+Further, the cycle of transmission between humans and mosquitoes is influenced not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period in mosquitoes \cite{Smith2012}.
+In fact, whether or not an exposed mosquito will survive long enough to become infectious represents a critical bottleneck in the transmission process and leads to nonlinear dependence of transmission on mosquito survival \cite{Smith2012}.
 
-The importance of mosquito demographic processes in governing human disease risk highlight the need to integrate mechanistic modeling into the quantitative tools used to target control efforts.
+If the goal of modern dengue control is to deploy limited mosquito control resources when and where they will be most effective at preventing disease \cite{Scott2010a, Scott2010b}, then the quantitative tools used to guide that deployment need to account for the influence of the critical processes governing transmission.
+This includes the intrinsic human epidemiological and mosquito demographic processes above, but infectious disease dynamics are often further influenced by seasonality and stochastic forcing \cite{Ellner1998,Koelle2004, Grenfell2002}.
+In fact, the interplay of these factors is one of the unique features of ecological dynamics and is often responsible for generating the complex inter-annual cycles observed therein \cite{Bjornstad2001}.
 Such mechanistic models can often perform better than complex autoregressive statistical models in describing and forecasting population dynamics \cite{Reilly2005}.
-Moreover, mechanistic models can provide scenario-based tools that can be used to predict the effect of management actions \cite{Buckland2007}, which cannot be done without explicit representation of mechanistic processes or data from case-control studies.
-Differential equation models provide a natural way to encode the relevant biological mechanisms and to link human epidemiology to mosquito demographic processes.
-As noted above, mosquito mortality is a critical component of this link, but cannot be estimated from mosquito surveillance alone.
-Given its role in mediating transmission, however, mosquito mortality can be informed by a combination of mosquito and clinical surveillance, borrowing from the field of integrated population modeling \cite{Schaub2010}.
+Moreover, in the absence of case-control studies, mechanistic models can provide scenario-based tools that can be used to predict the effect of management actions \cite{Buckland2007}.
 
-Many cities in Brazil generate the necessary vector surveillance data through the use of the MI-Dengue system, implemented by the company Ecovec.
-This system deploys a city-wide grid of mosquito sticky traps (MosquiTRAPs) that can be used to target control in highly infested areas \cite{Eiras2009}.
-This system has been shown to be effective, preventing an estimated 27,191 cases of dengue fever from 2007 - 2011 across the cities in which it was deployed \cite{Pepin2013}.
-Despite this, the mosquito indices and thresholds used to prioritize control efforts remain somewhat \emph{ad-hoc}.
+In order to meet the needs of mosquito control programs, however, these mechanistic models need to be embedded in a framework that can learn from commonly available surveillance data, while also accounting for multiple sources of uncertainty.
+Mosquito and clinical surveillance provide information about the underlying processes, but are observed with noise (CITATIONS).
+Differential equation models provide a natural way to encode the relevant biological mechanisms, though it is important to also account for sources of uncertainty and stochasticity in those models \cite{Hotelling1927, Wikle2010}.
+Moreover, the values of parameters are also often uncertain, which can lead to large uncertainty about the effects of management actions (CITATIONS).
+Bayesian hierarchical modeling provides a coherent framework to account for and integrate this uncertainty across the three levels of the model (data, process, and parameters \cite{Berliner1996, Cressie2009}).
 
@@
 
-Vitoria is a coastal city and the capital of the state of Espirito Santo, Brazil, with a population of 327,801.
-Since 2008, the Ecovec has monitored mosquito abundance for the city using approximately 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a grid across the city.
+Vitoria is a coastal city and the capital of the state of Espirito Santo, Brazil, with a population of 327,801 \cite{vitpop}.
+Since 2008, the company Ecovec has monitored mosquito abundance for the city using approximately 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a grid across the city.
 Each trap is checked weekly and the mosquitoes inside counted and identified, providing us with 243 weeks (2008 through week 34 of 2012) of counts of gravid female \emph{Aedes aegypti}.
 Dengue is a mandatory notifiable disease, and thus the city's Ministry of Health Secretary maintains a database of weekly notified probable dengue cases (i.e., medical care sought for dengue-like symptoms) for this same time period.
+
+To try to prevent these regular outbreaks, Vitoria and many similar cities routinely conduct mosquito control activities, often guided by mosquito surveillance data.
+One source for such surveillance data is the MI-Dengue system, implemented by the company Ecovec.
+This system deploys a city-wide grid of mosquito sticky traps (MosquiTRAPs) that can be used to target control in highly infested areas \cite{Eiras2009}.
+This system has been shown to be effective, preventing an estimated 27,191 cases of dengue fever from 2007 - 2011 across the cities in which it was deployed \cite{Pepin2013}.
+Despite this, the mosquito indices and thresholds used to prioritize control efforts remain somewhat \emph{ad-hoc}.
 
@@
 $r_0$ & Initial mosquito population growth rate & 0 & \\
+$V_{N0}$ & Initial mosquito population size & $2N$ & \\
 $\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & \\

### 2018-01-04 Manuscript/ms.tex
Cleaning up system and data

@@
 Since 2008, the company Ecovec has monitored mosquito abundance for the city using approximately 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a grid across the city.
-Each trap is checked weekly and the mosquitoes inside counted and identified, providing us with 243 weeks (2008 through week 34 of 2012) of counts of gravid female \emph{Aedes aegypti}.
-Dengue is a mandatory notifiable disease, and thus the city's Ministry of Health Secretary maintains a database of weekly notified probable dengue cases (i.e., medical care sought for dengue-like symptoms) for this same time period.
-
-To try to prevent these regular outbreaks, Vitoria and many similar cities routinely conduct mosquito control activities, often guided by mosquito surveillance data.
-One source for such surveillance data is the MI-Dengue system, implemented by the company Ecovec.
-This system deploys a city-wide grid of mosquito sticky traps (MosquiTRAPs) that can be used to target control in highly infested areas \cite{Eiras2009}.
-This system has been shown to be effective, preventing an estimated 27,191 cases of dengue fever from 2007 - 2011 across the cities in which it was deployed \cite{Pepin2013}.
-Despite this, the mosquito indices and thresholds used to prioritize control efforts remain somewhat \emph{ad-hoc}.
+Each trap is checked weekly and the mosquitoes inside counted and identified, with the results sent to a central database that city managers then use to map mosquito infestations and target control.
+These data provide us with 243 weeks (2008 through week 34 of 2012) of total city-wide counts of trapped gravid female \emph{Aedes aegypti}.
+It is important to note that this time series reflects both natural fluctuations in mosquito density and fluctuations driven by the city's existing mosquito control program.
+In addition, dengue fever is a mandatory notifiable disease, and thus the city's Ministry of Health Secretary maintains a database of weekly notified probable dengue cases (i.e., medical care sought for dengue-like symptoms) for this same time period.
 

### 2018-01-04 Manuscript/ms.tex
Adding the full hierarchical statement
And tweaking the process model discussion

@@
 It is important to note that this time series reflects both natural fluctuations in mosquito density and fluctuations driven by the city's existing mosquito control program.
-In addition, dengue fever is a mandatory notifiable disease, and thus the city's Ministry of Health Secretary maintains a database of weekly notified probable dengue cases (i.e., medical care sought for dengue-like symptoms) for this same time period.
+In addition, dengue fever is a mandatory notifiable disease, and thus the city's Ministry of Health Secretary maintains a database of weekly notified probable dengue cases (i.e., medical care sought for dengue-like symptoms) for the same time period.
 
@@
 
-Dengue epidemiology is complicated considerably by the presence of four simultaneously circulating serotypes (referred to as DENV-1, DENV-2, DENV-3, DENV-4).
-Infection with one serotype confers life-long immunity to that serotype, along with temporary immunity to other serotypes.  
-As this cross-immunity wanes, antibodies from the previous infection can result in antibody-dependent enhancement (ADE), wherein human hosts are more susceptible to infection with the other serotypes and more likely to develop severe symptoms (i.e., dengue haemorrhagic fever or dengue shock syndrome).
+Dengue epidemiology is complicated considerably by the presence of four simultaneously circulating serotypes.
+Infection with one serotype confers life-long immunity to that serotype, along with temporary immunity to other serotypes \cite{Wearing2006}.  
+As this cross-immunity wanes, antibodies from the previous infection can result in antibody-dependent enhancement (ADE), wherein human hosts are more susceptible to infection with the other serotypes and more likely to develop severe symptoms (i.e., dengue haemorrhagic fever or dengue shock syndrome)\cite{Wearing2006}.
 The strength and duration of these different inter-serotype interactions are not well understood, though different models suggest that temporary cross-immunity alone (without ADE) is sufficient to reproduce observed multi-annual dynamics in Thailand \cite{Wearing2006,Reich2013}.
@@
 Moreover, because dengue case reports do not include information on serotype, we do not have enough information to inform the dynamics of individual serotypes.
-As such, we simplify the model of \cite{Wearing2006} to an SEIRS framework, which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
+As such, we develop an SEIRS compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010} which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
 In this framework, susceptible humans ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
@@
 
-Sampling from the posterior distribution in non-linear differential equation models is often very difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations among parameters \cite{Reilly2005, Girolami2008, Calderhead2011}.
+Combining the data, process, and parameter models \cite{Berliner1996}, we can summarize the full hierarchical model as:
+\begin{align}
+y_t | \cdot & \sim \text{NegBin}(\phi_y (C(t) - C(t-1)), \eta_y),
+\\
+q_t | \cdot &\sim \text{NegBin}(V_{C}(t) - V_{C}(t-1), \eta_q)\\
+(\mathbf{C}, \mathbf{V_C}) & = \mathcal{M}(\boldsymbol{\epsilon_r}, \boldsymbol{\epsilon_{\nu}},\boldsymbol{\theta})\\
+\epsilon_{rt} & \sim \text{Normal}(0, \sigma^2_r)\\
+\epsilon_{\nu t} & \sim \text{Normal}(0, \sigma^2_{\nu})\\
+\boldsymbol{\theta} & \sim [\boldsymbol{\theta}]
+\end{align}
+where $\boldsymbol{\theta}$ is a vector of all the model parameters and initial conditions, and $\mathcal{M}(\boldsymbol{\epsilon_r}, \boldsymbol{\epsilon_{\nu}},\boldsymbol{\theta})$ represents the (numeric) solution to the differential equation as a function of $\boldsymbol{\theta}$ and the weekly stochastic forcing terms ($\boldsymbol{\epsilon_{r}}, \boldsymbol{\epsilon_{\nu}}$).
+Sampling from the posterior distribution of the parameters and forcing terms in this model is difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations induced by the nonlinearity of the the differential equation model \cite{Reilly2005, Girolami2008, Calderhead2011}.
 In addition, the computational burden of numerically solving the differential equation model at each iteration means that sampling efficiency is crucial.

### 2018-02-04 Manuscript/ms.tex
Initial scaffolding of discussion edits

@@
 
+\subsection*{Summary}
+
 Targeted mosquito control can be an effective tool in reducing the disease burden within a city.
@@
 Targeting the mosquito population early in the year as dengue just begins to spread can disrupt the transmission cycle and prevent further amplification and feedback later in the year.
+
+\subsection*{Model validation and interpretation of $d_v$}
+
+The dynamics of infectious diseases, and ecological systems more broadly, are often driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
+We capture the interplay of these factors primarily through our specification of the mosquito mortality process.
+As implemented, mosquito mortality is the only source of uncertainty/stochasticity in the transmission process.
+As a result, the estimated mosquito mortality could be soaking up other sources of stochasticity or model misspecification.
+Our implementation of the mortality process is conceptually similar to the empirical forcing functions developed by Hooker and Ellner \cite{Hooker2015}.
+In that framework, Hooker and Ellner \cite{Hooker2015} estimated nonparametric functions in time that modify either the differential equations or specific model parameters to provide a good fit to the data.
+These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables.
+We do not employ the same explicit testing framework as \cite{Hooker2015}, but we treat the estimated mosquito mortality time series in the same spirit.
+In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission, either through the death rate itself or through another process.
+Despite this, our estimated mosquito mortalities generally fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
+The fact that our estimated mortality rate varied seasonally and increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is perhaps more extreme than estimated in most laboratory studies.
+These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.
+
+\subsection*{Implications for control policy}
 
@@
 
-As implemented, mosquito mortality is the only source of uncertainty/stochasticity in the transmission process.
-As a result, the estimated mosquito mortality could be soaking up other sources of stochasticity or model misspecification.
-Our implementation of the mortality process is conceptually similar to the empirical forcing functions developed by Hooker and Ellner \cite{Hooker2015}.
-In that framework, Hooker and Ellner \cite{Hooker2015} estimated nonparametric functions in time that modify either the differential equations or specific model parameters to provide a good fit to the data.
-These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables.
-We do not employ the same explicit testing framework as \cite{Hooker2015}, but we treat the estimated mosquito mortality time series in the same spirit.
-In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission, either through the death rate itself or through another process.
-Despite this, our estimated mosquito mortalities generally fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
-The fact that our estimated mortality rate varied seasonally and increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is perhaps more extreme than estimated in most laboratory studies.
-These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.
+\subsection*{Caveats and extensions}
 
@@
 In fact, \cite{Dye1986, Hasibeder1988} found that introducing heterogeneous mixing to models of vector-borne disease spread always increases $R_0$ relative to models with homogeneous mixing.
-
-This suggests that our model may overestimate the efficacy of mosquito control, but it also points a way toward more efficient and effective control methods.
-Our results suggest that applying control early in the year is critical to disrupting city-wide disease spread, but targeting the city's entire mosquito population is likely infeasible and spatial prioritization of control is also necessary \cite{Pepin2015}.
-If disease risk is driven by high levels of contact with a relatively small proportion of the mosquito population \cite{Yoon2012}, control policies that preferentially target these subsets of the mosquito population are likely to be very effective.
-However, these subsets can be difficult to identify in practice.
-One possible strategy is to employ contact-tracing, which can identify and target specific premises that have been directly implicated in transmission \cite{Vazquez-Prokopec2017}.
-Though effective, this strategy is extremely labor-intensive and would be difficult to implement in endemic areas.
-
-Due to these difficulties, new methods are needed that can make use of currently available surveillance data to more effectively guide spatial prioritization of control \cite{Pepin2015}.
-MosquiTRAP surveillance data are available at the level of the individual trap, and case reports for the city of Vitoria are available for each of the city's 75 neighborhoods.
-To interface with these data, spatial structure and human movement could be incorporated into models of dengue transmission  (e.g., through the methods outlined in \cite{Cosner2015}) but at the cost of substantial statistical and computational burden.
-For instance, embedding the model presented here into a metapopulation framework based on the city's neighborhoods leads to a very large, unwieldy mechanistic model.
-Moreover, without external data on human movement (possible sources of which are discussed in \cite{Stoddard2009}), it would be very difficult to identify its role in disease dynamics.
-These dimensionality and identifiability issues would likely exacerbate existing difficulties inherent in fitting ODE models to data \cite{Girolami2008, Calderhead2011}.
-Developing such spatial models and the methods necessary to deal with them is a promising area of future research.
-In the absence of such methods, spatial targeting of control based on more traditional methods (e.g., local entomological thresholds) could be combined with temporal targeting based on city-scale mechanistic modeling.
-This would allow mosquito control efforts to exploit the critical timing identified by the mechanistic modeling, while also taking advantage of the spatial richness of the surveillance data to allocate limited resources within that window.
 

### 2018-02-04 Analysis/control.r
Control simulations using thresholds

@@ -98,6 +98,120 @@ reduction <- sum(data$obs) - reduction
 saveRDS(reduction, "Results/control.rds")
 
 #===============================================================================
+# At what case threshold is control most effective?
+
+thresholds <- seq(20, 200, by = 10)
+
+# Setting up parallel
+cl <- makeCluster(3, type = "SOCK")
+registerDoParallel(cl)
+
+# Parallel for-loop over mcmc iterations
+reduction <- foreach(k = 1:nmcmc, .combine = "cbind", .packages = c("rstan", "magrittr", "plyr")) %dopar% {
+  
+  init <- list(lapply(samples, extract_sample, k))
+  
+  cases <- matrix(NA, nrow = length(thresholds), ncol = 1)
+  
+  for(i in 1:length(thresholds)){
+    
+    data$above <- data$obs > thresholds[i]
+    
+    weeks <- ddply(subset(data, year > 2008 & year < 2012), .(year), summarise, 
+                   week = tot.week[which(above)[1]])
+    
+    control <- matrix(1, nrow = 243, ncol = 3)
+    control[weeks$week, 1] <- 1.05
+    
+    dat.stan <- list(T = 243,
+                     T_pred = 0,
+                     steps = 56,
+                     y =data$obs,
+                     q = data$q,
+                     tau = data$tau,
+                     rov = rov,
+                     control = control,
+                     pop = pop)
+    
+    sim <- sampling(model,
+                    data = dat.stan, 
+                    init = init, 
+                    iter = 1, 
+                    chains = 1,
+                    warmup = 0,
+                    algorithm = "Fixed_param")
+    
+    cases[i] <- rstan::extract(sim, "y_hat", permute = F)[, 1, ] %>% sum()
+  }
+  
+  return(cases)
+}
+
+stopCluster(cl)
+
+reduction <- sum(data$obs) - reduction
+
+saveRDS(reduction, "Results/case_control.rds")
+
+#===============================================================================
+# At what mosquito/trap threshold is control most effective?
+
+thresholds <- seq(0.25, 1, by = 0.04)
+
+data <- mutate(data, qt = q/tau)
+
+# Setting up parallel
+cl <- makeCluster(2, type = "SOCK")
+registerDoParallel(cl)
+
+# Parallel for-loop over mcmc iterations
+reduction <- foreach(k = 1:nmcmc, .combine = "cbind", .packages = c("rstan", "magrittr", "plyr")) %dopar% {
+  
+  init <- list(lapply(samples, extract_sample, k))
[…]

### 2018-02-04 Manuscript/ms.tex
Moving discussion paragraphs around
Without fully stitching them together again.

@@
 When faced with limited resources with which to deploy that control, these results suggest that focusing efforts early in the year, specifically in the second to fourth weeks, will lead to the largest reductions in disease.
-Because adult control (increasing the mortality rate) and larval control (decreasing the birth rate) produce nearly identical reductions in disease, the effect of control is likely driven by reducing the standing population of adult mosquitoes, rather than by limiting the transmission potential of exposed or infectious mosquitoes.
 The lag between optimal control and both mosquito population dynamics and human disease highlights the importance of early intervention and the need to account for the nonlinear processes governing transmission.
@@
 
-The dynamics of infectious diseases, and ecological systems more broadly, are often driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
-We capture the interplay of these factors primarily through our specification of the mosquito mortality process.
+The dynamics of infectious diseases, and ecological systems more broadly, are often driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity \cite{Ellner1998,Koelle2004,Bjornstad2001}.
+We capture the interplay of these factors largely through our specification of the mosquito mortality process.
+As discussed above, the race between mosquito mortality and incubation period introduces a critical bottleneck and non-linearity into the transmission process and the influence of mosquito abundance on human disease.
+Moreover, both the incubation period and mosquito lifespan vary seasonally as a result of environmental forcing.
+The mechanistic modeling framework allows us to incorporate these critical processes.
+
 As implemented, mosquito mortality is the only source of uncertainty/stochasticity in the transmission process.
-As a result, the estimated mosquito mortality could be soaking up other sources of stochasticity or model misspecification.
+This reflects the central role that mosquito mortality likely plays in driving dengue dynamics, but it also means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.
 Our implementation of the mortality process is conceptually similar to the empirical forcing functions developed by Hooker and Ellner \cite{Hooker2015}.
@@
 These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables.
-We do not employ the same explicit testing framework as \cite{Hooker2015}, but we treat the estimated mosquito mortality time series in the same spirit.
+
+We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect the estimated mosquito mortality time series in the same spirit.
 In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission, either through the death rate itself or through another process.
 Despite this, our estimated mosquito mortalities generally fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
-The fact that our estimated mortality rate varied seasonally and increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is perhaps more extreme than estimated in most laboratory studies.
+The fact that our estimated mortality rate varies seasonally and increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is generally more extreme than estimated in laboratory studies.
 These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.
+Something about the idea that stochastic fluctuations around a seasonally varying mortality rate conspire to drive disease dynamics.
 
@@
 
+The importance of mosquito longevity in driving disease dynamics suggests the potential for control efforts that reduce the lifespan of adult mosquitoes to disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state.
+In fact, this forms the basis for much of the theory of adult control \cite{Burattini2008, Morrison2008, Smith2012}.
+The fact that adult control (increasing the mortality rate) and larval control (decreasing the birth rate) produce nearly identical reductions in disease in our simulations suggests that the effect of control is likely driven by reducing the standing population of adult mosquitoes, rather than by limiting the transmission potential of exposed or infectious mosquitoes.
+
 Discussion of the effectiveness of adult control often focuses on the fact that it produces only a brief decline in mosquito abundance \cite{Newton1992, Esu2010}.
 When applied at the peak of an epidemic, this reduction in abundance is likely to have a relatively small effect, because the population rebounds while there are still a large number of infectious human hosts available \cite{Newton1992, Burattini2008}.
-However, adult control can also have an effect by shortening mosquito lifespan and preventing mosquitoes from living long enough to progress through the extrinsic incubation period and become infectious.
-This forms the basis for much of the theory of adult control \cite{Burattini2008, Morrison2008, Smith2012}, but modeling this effect in a standard differential equation framework is difficult.
-In particular the assumption of exponential wait times implicit in a compartmental model is perhaps ill-suited to modeling the race between mosquito lifespan and incubation period.
-The long tails of the exponential distribution may overestimate the probability that a mosquito survives long enough to become infectious and underestimate the effect of decreasing mosquito lifespan.
-More accurate models of the extrinsic incubation period could incorporate explicit delays (i.e., assuming a fixed incubation period \cite{Burattini2008}), or additional chained exposed compartments (i.e., assuming the incubation period is gamma-distributed, \cite{Lloyd2001}).
-Including such mechanisms in our model would allow us to better differentiate the effects of larval and adult control and provide a more accurate picture of the effectiveness of adult control.
-In fact, modeling studies that assume exponential wait times suggest that adult control is likely to be ineffective when applied during an epidemic \cite{Newton1992, Pinho2010}, while similar models that include explicit delays for the extrinsic incubation period show adult control being much more effective \cite{Burattini2008}.
-
-Thus, our model may underestimate the effect of adult mosquito control applied later in the year when infectious mosquitoes are prevalent.
-However, our model nonetheless shows that reducing mosquito abundance early in the year can be highly effective at preventing dengue cases later in the year.
+However, our model demonstrates that reducing mosquito abundance early in the year can be highly effective at preventing dengue cases later in the year.
 This effectiveness is driven primarily by the timing of the reduction in mosquito abundance and could be achieved by either adult or larval control strategies.
@@
 \subsection*{Caveats and extensions}
+
+It is important to note that the control we simulate is implemented over the top of the control policies already enacted by the city (and on which data are unavailable).
+Also, the case data represents reports of "dengue-like illness", and as such could include cases of other diseases with similar symptoms (e.g. chikungunya).
+Given the high underreporting rate expected for dengue fever, we expect unreported cases to far outnumber misreported cases.
+Moreover, many of the pathogens that cause disease that might be mistaken for dengue are also vectored by \emph{Aedes aegypti}, and control policies driven by dengue fever are likely to aid in the control of other pathogens as well.
+
+In addition, our model does not explicitly include all four serotypes.  
+It is difficult to speculate about how important this might be.
+
+The assumption of exponential wait times implicit in a compartmental model is perhaps ill-suited to modeling the race between mosquito lifespan and incubation period.
+The long tails of the exponential distribution may overestimate the probability that a mosquito survives long enough to become infectious and underestimate the effect of decreasing mosquito lifespan.
+More accurate models of the extrinsic incubation period could incorporate explicit delays (i.e., assuming a fixed incubation period \cite{Burattini2008}), or additional chained exposed compartments (i.e., assuming the incubation period is gamma-distributed, \cite{Lloyd2001}).
+Including such mechanisms in our model would allow us to better differentiate the effects of larval and adult control and provide a more accurate picture of the effectiveness of adult control.
+In fact, modeling studies that assume exponential wait times suggest that adult control is likely to be ineffective when applied during an epidemic \cite{Newton1992, Pinho2010}, while similar models that include explicit delays for the extrinsic incubation period show adult control being much more effective \cite{Burattini2008}.
 

### 2018-02-04 Manuscript/ms.tex
Cleaning up discussion of mortality

@@
 Targeted mosquito control can be an effective tool in reducing the disease burden within a city.
-When faced with limited resources with which to deploy that control, these results suggest that focusing efforts early in the year, specifically in the second to fourth weeks, will lead to the largest reductions in disease.
+When faced with limited resources with which to deploy that control, our results suggest that focusing efforts early in the year, specifically in the second to fourth weeks, will lead to the largest reductions in disease.
 The lag between optimal control and both mosquito population dynamics and human disease highlights the importance of early intervention and the need to account for the nonlinear processes governing transmission.
@@
 
-The dynamics of infectious diseases, and ecological systems more broadly, are often driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity \cite{Ellner1998,Koelle2004,Bjornstad2001}.
-We capture the interplay of these factors largely through our specification of the mosquito mortality process.
-As discussed above, the race between mosquito mortality and incubation period introduces a critical bottleneck and non-linearity into the transmission process and the influence of mosquito abundance on human disease.
-Moreover, both the incubation period and mosquito lifespan vary seasonally as a result of environmental forcing.
-The mechanistic modeling framework allows us to incorporate these critical processes.
+As discussed above, the race between mosquito mortality and incubation period introduces a critical bottleneck and non-linearity into the transmission process and the effect of mosquito abundance on human disease.
+Given the central role that this process plays in driving dengue dynamics, we model fluctuations in the mosquito mortality rate using a flexible, stochastically forced harmonic oscillator.
+The fact that this is the only source of stochasticity in the transmission process, however,  means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.
 
-As implemented, mosquito mortality is the only source of uncertainty/stochasticity in the transmission process.
-This reflects the central role that mosquito mortality likely plays in driving dengue dynamics, but it also means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.
-Our implementation of the mortality process is conceptually similar to the empirical forcing functions developed by Hooker and Ellner \cite{Hooker2015}.
+Our implementation of the mosquito mortality process is conceptually similar to the empirical forcing functions developed by Hooker and Ellner \cite{Hooker2015}.
 In that framework, Hooker and Ellner \cite{Hooker2015} estimated nonparametric functions in time that modify either the differential equations or specific model parameters to provide a good fit to the data.
 These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables.
-
 We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect the estimated mosquito mortality time series in the same spirit.
@@
 Despite this, our estimated mosquito mortalities generally fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
-The fact that our estimated mortality rate varies seasonally and increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is generally more extreme than estimated in laboratory studies.
+The fact that our estimated mortality rate increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is generally more extreme than estimated in laboratory studies.
+
 These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.
-Something about the idea that stochastic fluctuations around a seasonally varying mortality rate conspire to drive disease dynamics.
+Further, this emphasizes that the dynamics of dengue fever, like many other infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
+Indeed, the driving force of these interacting processes is channelled largely through mosquito mortality.
 

### 2018-03-01 Manuscript/ms.tex
Retooling the discussionn

@@
 
-As discussed above, the race between mosquito mortality and incubation period introduces a critical bottleneck and non-linearity into the transmission process and the effect of mosquito abundance on human disease.
-Given the central role that this process plays in driving dengue dynamics, we model fluctuations in the mosquito mortality rate using a flexible, stochastically forced harmonic oscillator.
+Given the central role that mosquito lifespan plays in driving dengue dynamics, we model fluctuations in the mosquito mortality rate using a flexible, stochastically forced harmonic oscillator.
 The fact that this is the only source of stochasticity in the transmission process, however,  means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.
-
 Our implementation of the mosquito mortality process is conceptually similar to the empirical forcing functions developed by Hooker and Ellner \cite{Hooker2015}.
@@
 We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect the estimated mosquito mortality time series in the same spirit.
-In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission, either through the death rate itself or through another process.
+In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission, either through the mortality rate itself or through another process.
+
 Despite this, our estimated mosquito mortalities generally fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
-The fact that our estimated mortality rate increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is generally more extreme than estimated in laboratory studies.
+The fact that our estimated mortality rate increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is generally more extreme than estimated in laboratory studies, possibly as a result of assuming exponentially distributed extrinsic incubation period.
+In particular, the assumption that the extrinsic incubation period is exponentially distributed, and thus that many mosquitoes experience a very brief incubation period, may overestimate the probability that a mosquito survives long enough to become infectious.
+More accurate models of the extrinsic incubation period could incorporate explicit delays (i.e., assuming a fixed incubation period \cite{Burattini2008}), or additional chained exposed compartments (i.e., assuming the incubation period is gamma-distributed, \cite{Lloyd2001}).
+These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.
 
-These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.
-Further, this emphasizes that the dynamics of dengue fever, like many other infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
-Indeed, the driving force of these interacting processes is channelled largely through mosquito mortality.
+\subsection*{Processes driving dynamics and implications for control}
 
-\subsection*{Implications for control policy}
+Infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are often driven by the combined efforts of intrinsic non-linearities and feedbacks, seasonality, and stochasticity.
+In this paper, we have highlighted the importance of these processes in driving the dynamics of dengue fever and shaping the relationship between mosquito abundance and human disease risk.
 
-The importance of mosquito longevity in driving disease dynamics suggests the potential for control efforts that reduce the lifespan of adult mosquitoes to disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state.
+\subsubsection*{Adult control}
+
+The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state.
 In fact, this forms the basis for much of the theory of adult control \cite{Burattini2008, Morrison2008, Smith2012}.
-The fact that adult control (increasing the mortality rate) and larval control (decreasing the birth rate) produce nearly identical reductions in disease in our simulations suggests that the effect of control is likely driven by reducing the standing population of adult mosquitoes, rather than by limiting the transmission potential of exposed or infectious mosquitoes.
+Our results suggest that such control efforts (i.e., those that remove adult mosquitoes from the population) are most effective when applied late in the outbreak (week 24), after the number of cases has peaked.
+If applied at the peak of the epidemic, the this control is likely less effective, as there are still a large number of infectious human hosts available \cite{Newton1992, Burattini2008}.
+Control applied late in the outbreak pushes the system in the direction it is already heading.
+Thus control that interacts with the existing seasonality is more effective.
+In measles, this is also generally when the effect of stochasticity is most limited and the dynamics are pulled most strongly towards the deterministic attractor.
+This is also generally when immunity is highest.
 
-Discussion of the effectiveness of adult control often focuses on the fact that it produces only a brief decline in mosquito abundance \cite{Newton1992, Esu2010}.
-When applied at the peak of an epidemic, this reduction in abundance is likely to have a relatively small effect, because the population rebounds while there are still a large number of infectious human hosts available \cite{Newton1992, Burattini2008}.
-However, our model demonstrates that reducing mosquito abundance early in the year can be highly effective at preventing dengue cases later in the year.
+\subsubsection*{Larval controls}
+
+In contrast, larval control is most effective early in the year, right at the beginning of the epidemic.
+Intervention early in the year helps to prevent amplification and feedback.
+Also often corresponds to the local peak in susceptibility.
 This effectiveness is driven primarily by the timing of the reduction in mosquito abundance and could be achieved by either adult or larval control strategies.
-Although the mechanisms above might suggest further advantages for adult control, a mixed strategy of adult and larval control may be most effective in the long-term \cite{Burattini2008, Pinho2010}.
-Fully exploring such combined strategies requires a more complete model of mosquito life history and the lags in population dynamics induced by the mosquito's aquatic stage.
-These lags likely mean larval control should be applied prior to week 3 to achieve the desired reduction in the adult population in week 3.
-Models of mosquito development can be very complex (e.g.
-\cite{Magori2009}) but our model could be extended to account for these lags through a compartment for the aquatic phase (as in \cite{Burattini2008, Pinho2010}), though at the cost of introducing several new parameters.
+
+The timing of the effectiveness of both of these control strategies is driven largely by the interplay of the intrinsic feedbacks and seasonality of the system, and in particular by the fluctuations in human immunity and mosquito abundance.
+It is also interesting that neither strategy is most effective at the height of the outbreak, which highlights the fact that control strategies need to consider effects beyond the current season.
 
@@
 It is important to note that the control we simulate is implemented over the top of the control policies already enacted by the city (and on which data are unavailable).
+
 Also, the case data represents reports of "dengue-like illness", and as such could include cases of other diseases with similar symptoms (e.g. chikungunya).
@@
 Moreover, many of the pathogens that cause disease that might be mistaken for dengue are also vectored by \emph{Aedes aegypti}, and control policies driven by dengue fever are likely to aid in the control of other pathogens as well.
-
-In addition, our model does not explicitly include all four serotypes.  
-It is difficult to speculate about how important this might be.
-
-The assumption of exponential wait times implicit in a compartmental model is perhaps ill-suited to modeling the race between mosquito lifespan and incubation period.
-The long tails of the exponential distribution may overestimate the probability that a mosquito survives long enough to become infectious and underestimate the effect of decreasing mosquito lifespan.
-More accurate models of the extrinsic incubation period could incorporate explicit delays (i.e., assuming a fixed incubation period \cite{Burattini2008}), or additional chained exposed compartments (i.e., assuming the incubation period is gamma-distributed, \cite{Lloyd2001}).
-Including such mechanisms in our model would allow us to better differentiate the effects of larval and adult control and provide a more accurate picture of the effectiveness of adult control.
-In fact, modeling studies that assume exponential wait times suggest that adult control is likely to be ineffective when applied during an epidemic \cite{Newton1992, Pinho2010}, while similar models that include explicit delays for the extrinsic incubation period show adult control being much more effective \cite{Burattini2008}.
 

### 2018-03-02 Analysis/control.r
Cleaning up control code

@@ -3,8 +3,6 @@ library(doParallel)
 library(plyr)
 library(magrittr)
 library(lubridate)
-library(ggplot2)
-library(gridExtra)
 
 # Loading and processing mosquito and case data
 
@@ -52,7 +50,7 @@ model <- stan_model(file = "Code/seirs.stan")
 # In which week is adult control most effective?
 
 # Setting up parallel
-cl <- makeCluster(2, type = "SOCK")
+cl <- makeCluster(4, type = "SOCK")
 registerDoParallel(cl)
 
 # Parallel for-loop over mcmc iterations
@@ -65,12 +63,11 @@ reduction <- foreach(k = 1:nmcmc, .combine = "cbind", .packages = c("rstan", "ma
   for(i in 1:52){
     
     control <- matrix(1, nrow = 243, ncol = 3)
-    control[(data$week == i & data$year > 2008 & data$year < 2012), 3] <- 0.95
+    control[(data$week == i & data$year > 2008 & data$year < 2012), 3] <- 0.905
     
     first <- which(rowSums(control) != 3)[1]
     
     dat.stan <- list(T = 243,
-                     T_pred = 0,
                      steps = 7,
                      y = data$obs,
                      q = data$q,
@@ -87,7 +84,10 @@ reduction <- foreach(k = 1:nmcmc, .combine = "cbind", .packages = c("rstan", "ma
                     warmup = 0,
                     algorithm = "Fixed_param")
     
-    cases[i] <- rstan::extract(sim, "y_hat", permute = F)[, 1, ] %>%
+    sys <- rstan::extract(sim, "state", permute = T)[[1]]
+    lines(sys[1, , 6])
+    
+    cases[i] <- rstan::extract(sim, "y_meas", permute = F)[, 1, ] %>%
       extract(first:243) %>% 
       sum() %>% 
       divide_by(sum(data$obs[first:243]))
@@ -104,7 +104,7 @@ saveRDS(reduction, "Results/adult_control.rds")
 # In which week is larval control most effective?
 
 # Setting up parallel
-cl <- makeCluster(2, type = "SOCK")
+cl <- makeCluster(4, type = "SOCK")
 registerDoParallel(cl)
 
 # Parallel for-loop over mcmc iterations
@@ -117,12 +117,11 @@ reduction <- foreach(k = 1:nmcmc, .combine = "cbind", .packages = c("rstan", "ma
   for(i in 1:52){
     
[…]

### 2018-05-11 Analysis/plots.r
Updating control plots

@@ -16,7 +16,13 @@ post <- ddply(tseries, .(tot.week), summarise,
               yobs = sum(Cases, na.rm = T),
               tau = sum(Trap, na.rm = T),
               week = unique(Week),
-              year = unique(Year))
+              year = unique(Year),
+              epiweek = 0,
+              epiyear = 0)
+
+# Generating epidemic weeks/years that count from peak to peak (week 16)
+post$epiweek[16:243] <- c(1:52)
+post$epiyear <- cumsum(post$epiweek == 1)
 
 # Population size
 pop <- 327801
@@ -121,88 +127,55 @@ grid.arrange(fig2.a, fig2.b, ncol = 2)
 dev.off()
 
 
-# Figure 3: Effect of control implemented in different weeks ===================
+# Figure 3: Effect of adult control implemented in different weeks ===================
 
-adult_reduction <- readRDS("Results/adult_control.rds") %>% 
-  subtract(1) %>% multiply_by(-1) %>% 
-  adply(1, quantile, probs = c(0.1, 0.5, 0.9)) %>% 
-  mutate(week = as.numeric(X1), yweek = c(16:52, 1:15), control = "adult")
+adult <- readRDS("Results/adult_control.rds")
 
-larval_reduction <- readRDS("Results/larval_control.rds") %>% 
-  subtract(1) %>% multiply_by(-1) %>% 
-  adply(1, quantile, probs = c(0.1, 0.5, 0.9)) %>% 
-  mutate(week = as.numeric(X1), yweek = c(16:52, 1:15), control = "larval")
+adult_best <- adult %>% 
+  ddply(.(rep, year), summarise, 
+        best = which.min(reduction),
+        prevented =  1 - min(reduction))
 
-names(adult_reduction) <- c("X1", "min", "med", "max", "week", "yweek", "control")
-names(larval_reduction) <- c("X1", "min", "med", "max", "week", "yweek", "control")
+delta <- 97 * rstan::extract(sim, "delta_c", permute = T)[[1]][1:1000]
+adult_best$delta <- rep(delta, each = 3)
 
-reduction <- rbind(adult_reduction, larval_reduction)
+adult_best %>% 
+  ggplot(aes(delta, prevented)) + 
+  geom_point(size = 0.5) + 
+  facet_grid(.~year) + 
+  theme_classic() + 
[…]

### 2018-05-11 Manuscript/ms.tex
Writing up new control scheme

@@
 
-Given samples from the posterior distribution as obtained above, we can simulate the efficacy of different control strategies.
-In particular, we investigated the optimal timing of a single pulse of control within the year (e.g., if the city only has enough resource to fog for adult mosquitoes once a year, in what week is that control going to prevent the most cases?).
-To obtain the posterior distribution of the number of cases prevented by implementing control in week $i$ each year, for each posterior sample $k$, we simulate an increase in the latent mosquito death rate during each week $i$ (e.g.,
-$\hat{d}^{(k)}_{ji} = (1 + \Delta) d^{(k)}_{ji}$ where $j$ indexes the year, and $i$ indexes the week of the year).
-Then we summed the total number of cases reported in this scenario and compare to the data to get an estimate of the number of cases prevented.
-To keep our simulations conservative relative to field estimates of the mortality induced by spraying \cite{Esu2010}, and to avoid pushing the model into the unrealistic range of dengue eradication, we set $\Delta = 0.05$.
-We also only simulated the effect of control in the middle three years, because the effectiveness of control in the first year is strongly affected by initial conditions, and we lack complete data for the last year.
+Given samples from the posterior distribution as obtained above, we simulated the efficacy of different control interventions.
+To do so, we broke our time series into epidemiological years that roughly cover the span between outbreak peaks, from week 16 of one year to week 15 of the next.
+We then asked, in which week of a given epidemiological year is a single pulse of control most effective in preventing cases in the following (and slightly overlapping) calendar year?
+Structuring the simulations this way standardized the comparison of the effectiveness of control implemented at different times, as we evaluated the number of cases prevented in a fixed window (avoiding artifacts introduced when the window used to evaluate a given control intervention moves with the timing of that control).
+This set-up also roughly aligns with the timeline of decisions faced by actual city managers, e.g., the window within which a city might plan for the 2009 dengue season extends from roughly peak incidence in 2008 through to peak incidence in 2009 (any earlier and the 2008 outbreak is still the primary focus; any later and the 2009 outbreak is already winding down).
+Further, by investigating each year separately (omitting 2008 and 2012 due to edge effects and incomplete data, respectively), we also avoid the possibly overly optimistic predictions of sequential control.
+
+To obtain the posterior distribution of the proportion of cases prevented in calendar year $j$ by implementing control in week $i$ of the preceeding epidemiological year, for each posterior sample $k$, we simulated the dynamics resulting from a $5\%$ reduction of the mosquito population (affecting susceptible, exposed, and infectious mosquitoes equally).
+We then compared the number of cases produced during year $j$ in the control scenario to the same number in the uncontrolled scenario.
+A $5\%$ reduction in mosquito abundance was chosen to keep our simulations conservative relative to field estimates of the mortality induced by spraying \cite{Esu2010}, and to avoid pushing the model into the unrealistic range of dengue eradication.
 

### 2018-05-16 Manuscript/ms.tex
Updating figure captions

@@
 \includegraphics[angle = 270, scale = 1]{figures/fig2.eps}
-\caption{{\bf Estimated latent mosquito population growth and death rates.}
-Median posterior estimate (line) and 80\% credible interval (gray band).  A: weekly mosquito population growth rate ($r_t$). B: weekly mosquito death rate ($d_t$).
+\caption{{\bf Estimated latent mosquito mortality rate.}
+A: Time series of mortality rate, showing median posterior estimate (line) and 80\% credible interval (gray band). B: weekly mosquito mortality rate as a function of weekly mean temperature..
 }
-\label{latent}
+\label{mortality}
 \end{figure}
@@
 \includegraphics[angle = 270, scale = 1]{figures/fig3.eps}
-\caption{{\bf Estimated mosquito death rate as a function of temperature.}
-Weekly median posterior death rate ($d_t$) plotted against weekly mean temperature (Celsius) in Vitoria, Brazil.
+\caption{{\bf The effect of cross-immunity on effect and timing of control.}
+Each point represents a sample from the posterior and the results of the corresponding control simulation, in which we evaluated the effects of a $5\%$ reduction in mosquito abundance at each of the 52 weeks from week 16 of the previous year to week 15 of the target year.
+A, B, C: the maximum proportion of cases prevented target years 2009, 2010, and 2011, respectively.
+D, E, F: the epidemiological week in which control achieved the maximum reduction in cases in 2009, 2010, and 2011, respectively.
 }
-\label{temp}
+\label{immunity}
 \end{figure}
@@
 \includegraphics[angle = 270, scale = 1]{figures/fig4.eps}
-\caption{{\bf Cases prevented by mosquito control.}
-A: Number of reported cases of dengue fever prevented by increasing the mosquito death rate by 5\% during a given week each year in 2009, 2010, and 2011.
-Median posterior prediction (line) and 80\% credible interval.
-B: Weekly predicted case reports resulting from implementing control in week 3 of each year (timing indicated by vertical red lines).
-Median posterior case reports from the model without (dashed line) and with control (solid line) and 80\% credible interval of controlled dynamics (gray band).
-}
-\label{control}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/fig5.eps}
 \caption{{\bf Timing of optimal control relative to disease and mosquito dynamics.}
-The vertical black line indicates week 3, during which mosquito control produces the largest decline in predicted dengue case reports.
-Each colored line represents a different year: dark green = 2008, orange = 2009, purple = 2010, pink = 2011, light green = 2012.
+The vertical gray lines indicate the median week at which control was most effective in preventing cases in the following calendar year.
 A: posterior median estimate of new dengue cases per week.
 B: posterior median estimate of number of mosquitoes per person each week.
-C: posterior median estimate of weekly mosquito death rate.
+C: posterior median estimate of the proportion of susceptible individuals.
 }

### 2018-05-16 Analysis/control.r
Plotting gamma EIP control results

@@ -43,9 +43,9 @@ rov <- 7 * exp(0.2 * covars$temp - 8)
 #===============================================================================
 # Loading mcmc samples
 
-fit <- readRDS("Results/euler7.rds")
+fit <- readRDS("Results/gamma_eip.rds")
 samples <- rstan::extract(fit)[1:16]
-fitcases <- rstan::extract(fit, "system", permute = T)[[1]] %>% extract(, ,  8)
+fitcases <- rstan::extract(fit, "state", permute = T)[[1]] %>% extract(, ,  11)
 
 nmcmc <- length(samples[[1]])
 
@@ -54,21 +54,26 @@ extract_sample <- function(x, k){
   else return(x[k, ])
 }
 
-model <- stan_model(file = "Code/seirs.stan")
+model <- stan_model(file = "Code/gammaeip.stan")
 
 #===============================================================================
 # In which week is adult control most effective?
 
 # Setting up parallel
-cl <- makeCluster(10, type = "SOCK")
+cl <- makeCluster(3, type = "SOCK")
 registerDoParallel(cl)
 
 # Parallel for-loop over mcmc iterations
 reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "magrittr")) %dopar% {
   
-  init <- list(lapply(samples, extract_sample, k))
+  init <- list(lapply(samples, extract_sample, iters[k]))
   
-  cases <- data.frame(rep = k, control = rep(1:52, each = 3), year = rep(control_years, 52), reduction = 0)
+  cases <- data.frame(rep = k, 
+                      control = rep(1:52, each = 3), 
+                      year = rep(control_years, 52), 
+                      reduction = 0, 
+                      S = 0,
+                      I = 0)
   
   for(i in 1:52){
     for(j in 1:3){
@@ -93,11 +98,14 @@ reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "ma
                       warmup = 0,
                       algorithm = "Fixed_param")
       
-      cases$reduction[(i - 1) * 3 + j] <- rstan::extract(sim, "state", permute = T)[[1]][1, , 8] %>% 
+      state <- rstan::extract(sim, "state", permute = T)[[1]][1, 2:244, ] %>% 
+        extract(range(data$tot.week[data$year == control_years[j]]), )
+      
+      cases$S[(i - 1) * 3 + j] <- state[1, 1]
[…]

### 2018-05-16 Analysis/plots.r

@@ -7,6 +7,7 @@ library(magrittr)
 library(lubridate)
 library(ggplot2)
 library(gridExtra)
+library(patchwork)
 
 # Loading and processing mosquito and case data
 tseries <- read.csv("Data/Vitoria.data.csv")
@@ -40,13 +41,13 @@ post$temp <- temp$temp
 post$rov <- 7 * exp(0.2 * temp$temp - 8)
 
 # Loading MCMC results
-sim <- readRDS("Results/euler7.rds")
+sim <- readRDS("Results/gamma_eip.rds")
 
 # Figure 1: Observed and estimated time series =================================
 
 # Computing posterior median and 80% credible interval
-yhat <- rstan::extract(sim, c("y_hat"), permute = F) %>% apply(3, quantile, c(0.1, 0.5, 0.9))
-qhat <- rstan::extract(sim, c("q_hat"), permute = F) %>% apply(3, quantile, c(0.1, 0.5, 0.9))
+yhat <- rstan::extract(sim, c("y_meas"), permute = F) %>% apply(3, quantile, c(0.1, 0.5, 0.9))
+qhat <- rstan::extract(sim, c("q_meas"), permute = F) %>% apply(3, quantile, c(0.1, 0.5, 0.9))
 
 # Assembling data frame for ggplot
 post$yhat <- yhat[2, ]
@@ -89,9 +90,9 @@ dev.off()
 
 # Figure 2: Estimates of latent mosquito mortality rate ========================
 
-system <- rstan::extract(sim, "system", permute = T)[[1]]
+system <- rstan::extract(sim, "state", permute = T)[[1]]
 
-dv <- system[, , 9] %>% + 0.39 %>% exp() %>% 
+dv <- system[, 2:244, 12] %>% + 0.39 %>% exp() %>% 
   adply(2, quantile, c(0.1, 0.5, 0.9), .id = NULL)
 names(dv) <- c("dvmin", "dvmed", "dvmax")
 
@@ -129,20 +130,24 @@ dev.off()
 
 # Figure 3: Effect of adult control implemented in different weeks ===================
 
-adult <- readRDS("Results/adult_control.rds")
+adult <- readRDS("Results/gamma_control.rds")
 
 adult_best <- adult %>% 
   ddply(.(rep, year), summarise, 
         best = which.min(reduction),
-        prevented =  1 - min(reduction))
+        prevented =  1 - min(reduction),
[…]

### 2018-05-17 Analysis/citystan.r
Tweaking initial values to work with gamma EIP

@@ -56,7 +56,7 @@ inits = list(list(S0 = 0.4,
                   ro_c = 0.8,
                   gamma_c = 1,
                   delta_c = 0.8,
-                  logNv = 0.7,
+                  logNv = 1.5,
                   dv0 = 0,
                   rv0 = 0,
                   sigmadv = 0.2,
@@ -72,7 +72,7 @@ inits = list(list(S0 = 0.4,
                   ro_c = 1,
                   gamma_c = 1.2,
                   delta_c = 1,
-                  logNv = 1,
+                  logNv = 2,
                   dv0 = 0,
                   rv0 = 0,
                   sigmadv = 0.5,
@@ -85,10 +85,10 @@ inits = list(list(S0 = 0.4,
                   log_phi_q = -12.5,
                   eta_inv_y = 5,
                   eta_inv_q = 5,
-                  ro_c = 1,
-                  gamma_c = 1,
-                  delta_c = 1,
-                  logNv = 0.7,
+                  ro_c = 0.9,
+                  gamma_c = 0.5,
+                  delta_c = 0.6,
+                  logNv = 1,
                   dv0 = 0,
                   rv0 = 0,
                   sigmadv = 0.01,

### 2018-05-18 Analysis/plots.r
Adding more posteriors to supplemental plots

@@ -104,7 +104,7 @@ ggplot(post, aes(tot.week, dvmed)) +
   geom_ribbon(aes(ymin = dvmin, ymax = dvmax), fill = "grey70") +
   geom_line() + 
   theme_classic() +
-  scale_y_continuous(expand = c(0.05, 0), limits = c(0.6, 2.5)) + 
+  scale_y_continuous(expand = c(0.05, 0), limits = c(0.3, 2.2)) + 
   scale_x_continuous(expand = c(0, 1)) +
   ylab("mosquito mortality rate") +
   xlab("week") +
@@ -113,7 +113,7 @@ ggplot(post, aes(tot.week, dvmed)) +
 ggplot(post, aes(temp, dvmed)) +
   geom_point() + 
   theme_classic() + 
-  scale_y_continuous(expand = c(0.05, 0), limits = c(0.6, 2.5)) + 
+  scale_y_continuous(expand = c(0.05, 0), limits = c(0.3, 2.2)) + 
   scale_x_continuous(expand = c(0, 1)) +
   ylab("mosquito mortality rate") +
   xlab("weekly mean temperature") + 
@@ -237,6 +237,7 @@ postscript("Manuscript/figures/figS1.eps",
 ggplot(eps_dv, aes(week, med)) + 
   geom_ribbon(aes(ymin = min, ymax = max), fill = "grey70") +
   geom_line() + 
+  geom_hline(yintercept = 0) + 
   theme_classic() + 
   scale_x_continuous(expand = c(0, 1)) +
   ylab(expression(epsilon[d])) +
@@ -268,7 +269,7 @@ dev.off()
 
 # Figure S3: epidemiological parameters ===================================================
 
-epiparams <- rstan::extract(sim, c("ro_c", "gamma", "delta_c"), permute = T)
+epiparams <- rstan::extract(sim, c("ro_c", "gamma_c", "delta_c"), permute = T)
 
 postscript("Manuscript/figures/figS3.eps",
            width = 6, height = 3,
@@ -276,40 +277,72 @@ postscript("Manuscript/figures/figS3.eps",
 
 par(mfrow = c(1, 3))
 
-hist(0.87 * epiparams[[1]], main = "", xlab = "latent period", freq = F)
-abline(v = 0.87, lwd = 2)
+hist(epiparams[[1]], main = "", xlab = "scaled latent period", freq = F)
[…]

### 2018-05-18 Analysis/citystan.r
Estimating mean mortality rate

@@ -56,6 +56,7 @@ inits = list(list(S0 = 0.4,
                   ro_c = 0.8,
                   gamma_c = 1,
                   delta_c = 0.8,
+                  dvmu_c = 1, 
                   logNv = 1.5,
                   dv0 = 0,
                   rv0 = 0,
@@ -72,7 +73,8 @@ inits = list(list(S0 = 0.4,
                   ro_c = 1,
                   gamma_c = 1.2,
                   delta_c = 1,
-                  logNv = 2,
+                  dvmu_c = 0.6,
+                  logNv = 0.7,
                   dv0 = 0,
                   rv0 = 0,
                   sigmadv = 0.5,
@@ -88,6 +90,7 @@ inits = list(list(S0 = 0.4,
                   ro_c = 0.9,
                   gamma_c = 0.5,
                   delta_c = 0.6,
+                  dvmu_c = 1,
                   logNv = 1,
                   dv0 = 0,
                   rv0 = 0,
@@ -99,7 +102,7 @@ inits = list(list(S0 = 0.4,
 fit <- stan(file = "Code/gammaeip.stan", 
             data = dat.stan, 
             init = inits, 
-            iter = 2000, 
+            iter = 2000,
             chains = 3,
             control = list(adapt_delta = 0.99,
                            max_treedepth = 15))

### 2018-05-18 Analysis/control.r

@@ -44,7 +44,7 @@ rov <- 7 * exp(0.2 * covars$temp - 8)
 # Loading mcmc samples
 
 fit <- readRDS("Results/gamma_eip.rds")
-samples <- rstan::extract(fit)[1:16]
+samples <- rstan::extract(fit)[1:17]
 fitcases <- rstan::extract(fit, "state", permute = T)[[1]] %>% extract(, ,  11)
 
 nmcmc <- length(samples[[1]])
@@ -66,7 +66,7 @@ registerDoParallel(cl)
 # Parallel for-loop over mcmc iterations
 reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "magrittr")) %dopar% {
   
-  init <- list(lapply(samples, extract_sample, iters[k]))
+  init <- list(lapply(samples, extract_sample, k))
   
   cases <- data.frame(rep = k, 
                       control = rep(1:52, each = 3), 
@@ -105,7 +105,7 @@ reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "ma
       cases$I[(i - 1) * 3 + j] <- state[1, 3]
       cases$reduction[(i - 1) * 3 + j] <- state[, 11] %>% 
         diff() %>% 
-        divide_by(sum(fitcases[iters[k], data$year == control_years[j]]))
+        divide_by(sum(fitcases[k, data$year == control_years[j]]))
     }
   }
   

### 2018-05-22 Analysis/control.r
Restructuring control simulations to extend window further into past

@@ -13,17 +13,10 @@ data <- ddply(tseries, .(tot.week), summarise,
               q = sum(Mosquitoes, na.rm = T),
               tau = sum(Trap, na.rm = T),
               year = unique(Year),
-              week = unique(Week),
-              epiweek = 0,
-              epiyear = 0
+              week = unique(Week)
               )
 
-# Generating epidemic weeks/years that count from peak to peak (week 16)
-data$epiweek[16:243] <- c(1:52)
-data$epiyear <- cumsum(data$epiweek == 1)
-
-# Specifying the year windows over which we evaluate control
-control_years = c(2009, 2010, 2011)
+windows <- ddply(data, .(year), summarise, first = min(tot.week), last = max(tot.week))[2:4, ]
 
 # Population size
 pop <- 327801
@@ -64,22 +57,15 @@ cl <- makeCluster(3, type = "SOCK")
 registerDoParallel(cl)
 
 # Parallel for-loop over mcmc iterations
-reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "magrittr")) %dopar% {
-  
-  init <- list(lapply(samples, extract_sample, k))
-  
-  cases <- data.frame(rep = k, 
-                      control = rep(1:52, each = 3), 
-                      year = rep(control_years, 52), 
-                      reduction = 0, 
-                      S = 0,
-                      I = 0)
-  
-  for(i in 1:52){
-    for(j in 1:3){
+reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "magrittr")) %:% 
+  foreach(j = iter(windows, by = "row"), .combine = "rbind") %:% 
+    foreach(i = 1:(j$first - 1), .combine = "rbind") %dopar% {
+
+      init <- list(lapply(samples, extract_sample, k))
+      init[[1]]$dvmu_c <- 1
       
       control <- matrix(1, nrow = 243, ncol = 3)
-      control[(data$epiweek == i & data$epiyear == j), 3] <- 0.95
+      control[j$first - i, 3] <- 0.95
       
       dat.stan <- list(T = 243,
                        steps = 7,
@@ -99,17 +85,16 @@ reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "ma
                       algorithm = "Fixed_param")
       
       state <- rstan::extract(sim, "state", permute = T)[[1]][1, 2:244, ] %>% 
[…]

### 2018-05-22 Analysis/plots.r
Updating timing plot for gamma eip indexing

@@ -136,7 +136,9 @@ adult_best <- adult %>%
   ddply(.(rep, year), summarise, 
         best = which.min(reduction),
         prevented =  1 - min(reduction),
-        range = max(reduction) - min(reduction))
+        range = max(reduction) - min(reduction),
+        S = S[best],
+        I = I[best])
 
 delta <- 97 * rstan::extract(sim, "delta_c", permute = T)[[1]]
 adult_best$delta <- rep(delta, each = 3)
@@ -184,12 +186,13 @@ for(i in 1:3){
 }
 
 post <- mutate(post, 
-               ymu = system[, , 8] %>% 
+               ymu = system[, ,11] %>% 
+                 apply(1, diff) %>% 
                  multiply_by(pop) %>% 
+                 apply(1, median),
+               qmu = system[, 2:244, 9] %>% 
                  apply(2, median),
-               qmu = system[, , 6] %>% 
-                 apply(2, median),
-               Smu = system[, , 1] %>% 
+               Smu = system[, 2:244, 1] %>% 
                  apply(2, median))
 
 stacked <- melt(post, id.vars = c("tot.week", "epiweek", "year"), measure.vars = c("ymu", "qmu", "Smu"))

### 2018-05-24 Analysis/control.r
Writing control code to avoid redundancy

@@ -16,7 +16,7 @@ data <- ddply(tseries, .(tot.week), summarise,
               week = unique(Week)
               )
 
-windows <- ddply(data, .(year), summarise, first = min(tot.week), last = max(tot.week))[2:4, ]
+ends <- daply(data, .(year), function(df) max(df$tot.week))
 
 # Population size
 pop <- 327801
@@ -37,7 +37,7 @@ rov <- 7 * exp(0.2 * covars$temp - 8)
 # Loading mcmc samples
 
 fit <- readRDS("Results/gamma_eip.rds")
-samples <- rstan::extract(fit)[1:17]
+samples <- rstan::extract(fit)[1:16]
 fitcases <- rstan::extract(fit, "state", permute = T)[[1]] %>% extract(, ,  11)
 
 nmcmc <- length(samples[[1]])
@@ -53,48 +53,46 @@ model <- stan_model(file = "Code/gammaeip.stan")
 # In which week is adult control most effective?
 
 # Setting up parallel
-cl <- makeCluster(3, type = "SOCK")
+cl <- makeCluster(10, type = "SOCK")
 registerDoParallel(cl)
 
 # Parallel for-loop over mcmc iterations
 reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "magrittr")) %:% 
-  foreach(j = iter(windows, by = "row"), .combine = "rbind") %:% 
-    foreach(i = 1:(j$first - 1), .combine = "rbind") %dopar% {
+  foreach(j = 1:210, .combine = "rbind") %dopar% {
 
-      init <- list(lapply(samples, extract_sample, k))
-      init[[1]]$dvmu_c <- 1
-      
-      control <- matrix(1, nrow = 243, ncol = 3)
-      control[j$first - i, 3] <- 0.95
-      
-      dat.stan <- list(T = 243,
-                       steps = 7,
-                       y = data$obs,
-                       q = data$q,
-                       tau = data$tau,
-                       rov = rov,
-                       control = control,
-                       pop = pop)
-      
-      sim <- sampling(model,
-                      data = dat.stan, 
-                      init = init, 
-                      iter = 1, 
-                      chains = 1,
-                      warmup = 0,
-                      algorithm = "Fixed_param")
-      
-      state <- rstan::extract(sim, "state", permute = T)[[1]][1, 2:244, ] %>% 
-        extract(c(j$first, j$last), )
-      
-      data.frame(rep = k,
[…]

### 2018-05-24 Manuscript/supplement.tex
Updating supplement

@@
 \gamma & = 3.5\gamma'\\
+\nu_0 & \sim \text{Normal}(0, 0.5)\\
+r_0 & \sim \text{Normal}(0, 0.5)\\
+\log(V_{N0}) & \sim \text{Normal}(0.7, 0.3) \\ 
 S_0' & \sim \text{Beta}(4, 6)\\
@@
 Assuming 12 unreported cases for every reported case, there should be 163,080 total cases, or about 50\% of the population, immune to serotype 2 at the start of 2008.
+\\
+\item \textbf{Initial conditions} ($\nu_0, r_0$): Since the harmonic oscillators on $\nu$ and $r$ are centered at zero, and we do not have any prior knowledge of where the oscillator should be initially, we center the priors of their initial conditions at zero as well.
+Further, since $r$ is a population growth rate for a relatively stable (i.e., not explosive) population, it should not be too far from zero.
+Similarly, $\nu$, as an exponential multiplier on the mean per-capita mortality, should not vary too far from zero either.
+As such, we use flexible, but still conservative prior standard deviations of 0.5 for both values.  
 \\
@@
 \caption{
-Posterior estimates of process noise forcing the harmonic oscillators for mosquito mortality (A) and growth (B) rates.  Median posterior estimate (black line) and 80\% credible interval (gray band).
+Posterior estimates of forcing function on the harmonic oscillators for mosquito mortality (A) and growth (B) rates.  Median posterior estimate (black line) and 80\% credible interval (gray band).
 }
@@
 \caption{
-Posterior distribution of process noise variances for mosquito mortality (A) and growth (B) rates.
+Posterior distribution of variances on $\epsilon$ forcing terms for mosquito mortality (A) and growth (B) rates.
 }
@@
 Posterior distribution of epidemiological parameters.
-The black line indicates the prior mean, as set from the literature.
+The black line indicates the prior density.
 A: latent period in human host ($1/\rho$), in weeks.
@@
 \caption{Posterior distribution of initial conditions.
+The black line indicates the prior density.
 A: initial proportion of the human population that is susceptible.
@@
 C: initial number of infectious individuals.
+D: log initial number of mosquitoes-per-person.
+E: initial value of centered mosquito mortality rate oscillator.
+F: initial value of mosquito population growth rate oscillator.
 }
@@
 \includegraphics[angle = 270, scale = 1]{figures/figS5.eps}
+\caption{Posterior distribution of measurement paramters.
+A: log of the mosquito trap capture rate (which is strongly correlated with initial mosquito abundance).
+The black line shows the prior density.
+B: mosquito trap count measurement over-dispersion.
+C: case report measurement over-dispersion.}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/figS6.eps}
 \caption{Posterior check of autocorrelation structure in the case reports (A) and trapped mosquitoes (B) time series.
@@
 \begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS6.eps}
+\includegraphics[angle = 270, scale = 1]{figures/figS7.eps}
 \caption{
@@
 \begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS7.eps}
+\includegraphics[angle = 270, scale = 1]{figures/figS8.eps}
 \caption{
@@
 
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS8.eps}
-\caption{
-Estimated number of (reported) cases prevented by larval control (i.e decreasing the birth rate, $b_t = r_t + d_t$, by 5\%) applied in a given week.
-Posterior median (black line) and 80\% credible interval (gray band).
-Note that this is almost identical to the results from adult control.
-}
-\end{figure}
-
 \end{document}

### 2018-05-25 Manuscript/ms.tex
Working on caveats in discussion

@@
 
-Given samples from the posterior distribution as obtained above, we simulated the efficacy of different control interventions.
-To do so, we broke our time series into epidemiological years that roughly cover the span between outbreak peaks, from week 16 of one year to week 15 of the next.
-We then asked, in which week of a given epidemiological year is a single pulse of control most effective in preventing cases in the following (and slightly overlapping) calendar year?
-Structuring the simulations this way standardized the comparison of the effectiveness of control implemented at different times, as we evaluated the number of cases prevented in a fixed window (avoiding artifacts introduced when the window used to evaluate a given control intervention moves with the timing of that control).
-This set-up also roughly aligns with the timeline of decisions faced by actual city managers, e.g., the window within which a city might plan for the 2009 dengue season extends from roughly peak incidence in 2008 through to peak incidence in 2009 (any earlier and the 2008 outbreak is still the primary focus; any later and the 2009 outbreak is already winding down).
-Further, by investigating each year separately (omitting 2008 and 2012 due to edge effects and incomplete data, respectively), we also avoid the possibly overly optimistic predictions of sequential control.
-
-To obtain the posterior distribution of the proportion of cases prevented in calendar year $j$ by implementing control in week $i$ of the preceeding epidemiological year, for each posterior sample $k$, we simulated the dynamics resulting from a $5\%$ reduction of the mosquito population (affecting susceptible, exposed, and infectious mosquitoes equally).
+Given samples from the posterior distribution as obtained above, we simulated the effects of a single pulse of control applied in each week of the first three years of the time series.
+Since the city already implements responsive, targeted control with the aim of reducing the disease burden of an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in both the current and future years.
+To obtain the posterior distribution of the proportion of cases prevented in calendar year $j$ by implementing control in week $i$, for each posterior sample $k$, we simulated the dynamics resulting from reducing the mosquito population by $5\%$ at the beginning of week $i$ (affecting susceptible, exposed, and infectious mosquitoes equally, and leaving the estimated mosquito birth and mortality rates unchanged) .
 We then compared the number of cases produced during year $j$ in the control scenario to the same number in the uncontrolled scenario.
@@
 
-It is important to note that the control we simulate is implemented over the top of the control policies already enacted by the city (and on which data are unavailable).
+The case data to which we fit the model represents reports of "dengue-like illness", without laboratory confirmation, and as such could include cases of other diseases with similar symptoms (e.g.
+chikungunya or Zika).
+However, neither chikungunya nor Zika had emerged as substantial public health threats in Brazil by the end of our time series in late 2012 \cite{PAHO2014, PAHO2015}.
+In addition, given the high underreporting rate expected for dengue fever \cite{Silva2016}, and the uncertainty incorporated into the measurement model, we expect misreported cases to have a small effect on our analyses.
 
-Also, the case data represents reports of "dengue-like illness", and as such could include cases of other diseases with similar symptoms (e.g. chikungunya).
-Given the high underreporting rate expected for dengue fever, we expect unreported cases to far outnumber misreported cases.
-Moreover, many of the pathogens that cause disease that might be mistaken for dengue are also vectored by \emph{Aedes aegypti}, and control policies driven by dengue fever are likely to aid in the control of other pathogens as well.
+In addition, it is important to note that that the data to which we fit our model implicitly reflect the control efforts already enacted by the city.
+These control efforts are guided by the MI-Dengue system, which uses the trap-level mosquito surveillance data to target areas of high mosquito infestation for control (including source reduction, larvacide, and adulticide)\cite{Eiras2009}.
+These local interventions are effective at reducing disease burden in the cities in which they are implemented \cite{Pepin2013}, but long-term epidemiological processes (e.g.
+fluctuations in human immunity) still seem to be driving disease spread.
+Our control simulations account for these processes and offer a suggestion for how an additional city-wide pulse of control might be deployed to supplement existing efforts to reduce mosquito density throughout the city.
 
-Even though the estimated mosquito mortality process seems biologically reasonable, we are still employing it to link mosquito abundance and human disease at a relatively large spatial scale, assuming homogeneous mixing between mosquitoes and humans throughout the city.
-However, structured human movement within the city is likely to induce heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009}.
+That said, we do not anticipate that a single, simultaneous, city-wide pulse of control could be feasibly implemented on top of the existing control efforts.
+However, contrary to our assumption of homogeneous mixing between mosquitoes and humans, all mosquitoes may not contribute equally to dengue spread.
+In fact, a relatively small proportion of the mosquito population may be responsible for a large proportion of disease spread \cite{Yoon2012}
+In particular, structured human movement within the city is likely to induce heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009}.
 In addition, spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
-This heterogeneity can have substantial consequences for disease dynamics.
-In fact, \cite{Dye1986, Hasibeder1988} found that introducing heterogeneous mixing to models of vector-borne disease spread always increases $R_0$ relative to models with homogeneous mixing.
+The mechanistic model presented here may be able to suggest when an intervention is likely to be effective, but in order to make the best use of limited resources, spatial prioritization may still be necessary.
+This should include not just prioritization of areas with high mosquito abundance, but also areas with disproprtionate risk of disease spread, either due to hetereogeneities in the mixing process or in the distribution of human immunity.
 

### 2018-05-29 Analysis/control.r
Tweaking control to use estimated dvmu

@@ -36,8 +36,8 @@ rov <- 7 * exp(0.2 * covars$temp - 8)
 #===============================================================================
 # Loading mcmc samples
 
-fit <- readRDS("Results/gamma_eip.rds")
-samples <- rstan::extract(fit)[1:16]
+fit <- readRDS("Results/gamma_eip_dv0.rds")
+samples <- rstan::extract(fit)[1:17]
 fitcases <- rstan::extract(fit, "state", permute = T)[[1]] %>% extract(, ,  11)
 
 nmcmc <- length(samples[[1]])
@@ -61,8 +61,7 @@ reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "ma
   foreach(j = 1:210, .combine = "rbind") %dopar% {
 
     init <- list(lapply(samples, extract_sample, k))
-    init[[1]]$dvmu_c <- 1
-    
+
     control <- matrix(1, nrow = 243, ncol = 3)
     control[j, 3] <- 0.95
     

### 2018-05-29 Manuscript/ms.tex
Updating results for gamma eip

@@
 The model captured the observed dynamics of both case reports and mosquito trap counts (Figure \ref{timeseries}).
-The estimated posterior median case reports explained 91\% of the variation in the observed time series, while the posterior median mosquito trap counts explained 45\% of the variation in the observed time series.
-In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 5 - 7).
-The posterior distributions of the three epidemiological parameters ($\rho, \gamma, \delta$) differ from their priors, though they do still cover the literature values used to set their prior means (Supplemental Figure 3).
-The estimated latent period in a human host ($1/\rho$) is influenced most strongly by the data, with a posterior mean of 1.4 weeks compared to a prior mean of 0.87 weeks.
+The estimated posterior median case reports explained 90\% of the variation in the observed time series, while the posterior median mosquito trap counts explained 46\% of the variation in the observed time series.
+In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 6 -8).
+The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) do not differ substantially from their priors(Supplemental Figure 3).
+The estimated latent period in a human host ($1/\rho$, the expected time it takes for an exposed human to become infectious to biting mosquitoes) is influenced more strongly by the data, with a posterior mean of 1.53 weeks compared to a prior mean of 0.87 weeks.
 
-The estimated latent mosquito demographic rates are smooth and strongly seasonal, with fairly narrow credible intervals (Figure \ref{latent}).
-Mosquito mortality rates fluctuate within the range estimated by \cite{Brady2013} and generally increase with temperature (Figure \ref{temp}).
-The underlying posterior process standard deviations are small ($E(\sigma_{\nu}|\mathbf{y}) = 0.01$, $E(\sigma_r|\mathbf{y}) = 0.0005$, Supplemental Figure 2), and there is little residual structure in the process noise series, though there is a weak periodic signal in the $\epsilon_{\nu}$ (Supplemental Figure 1).
-
-Modifying these estimated demographic rates to simulate targeted control efforts, we found that control applied in the third week of the year (in January) is the most effective at preventing cases of dengue fever (though weeks 2 and 4 are also very close, Figure \ref{control}A).
-In fact, increasing the mosquito mortality rate by just 5\% in the third week of 2009, 2010, and 2011 would prevent 13,988 (reported) cases of dengue fever (with 80\% credible interval of 11,130 to 15,811).
-Most of these gains come by way of substantially reducing the size of the large 2011 outbreak, while also further damping the smaller 2010 and 2012 outbreaks (Figure \ref{control}B).
-Targeting larvae instead of adults (simulated by reducing the birth rate ($b_t = r_t + d_t$) by 5\%) produces nearly identical results (Supplemental Figure 8).
-
-The third week of the year is generally very early in the dengue season, well before cases peak (Figure \ref{timing}).
-In the two years with large outbreaks (2009 and 2011), the third week of the year is near the inflection point where the disease burden begins to grow rapidly.  
-The third week is also generally before the estimated seasonal peak in mosquito abundance, which occurs between weeks 6 and 9 (Figure \ref{timing}).
-Annual peaks in estimated mosquito mortality rate are more variable, but tend to fall between weeks 2 and 7.
+The estimated weekly mosquito mortality rate varied seasonally, and was positively correlated with temperature \ref{mortality}.
+The oscillations were more complex than purely sinusoidal, with variations in shape from year to year.
+The posterior mean of the baseline mortality rate, around which the time series oscillated, was 0.7/week, roughly half the prior mean of 1.47/week.
+The posterior means of the $\epsilon_{nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (Supplementary Figure 1), though the full posterior of each $\epsilon_{nu t}$ overlapped zero.
+In addition, the standard deviation of the mortality forcing terms was small ($E(\sigma_{\nu}|\mathbf{y}) = 0.01$, Supplemental Figure 2) relative to the weak prior.
 

### 2018-05-30 Manuscript/ms.tex
Discussion of mosquito mortality rates

@@
 In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 6 -8).
-The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) do not differ substantially from their priors(Supplemental Figure 3).
+The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that these data contain little additional information about these parameters (Supplemental Figure 3).
 The estimated latent period in a human host ($1/\rho$, the expected time it takes for an exposed human to become infectious to biting mosquitoes) is influenced more strongly by the data, with a posterior mean of 1.53 weeks compared to a prior mean of 0.87 weeks.
@@
 The estimated weekly mosquito mortality rate varied seasonally, and was positively correlated with temperature \ref{mortality}.
-The oscillations were more complex than purely sinusoidal, with variations in shape from year to year.
+The estimated oscillations in mosquito mortality were more complex than purely sinusoidal, with variations in shape from year to year.
 The posterior mean of the baseline mortality rate, around which the time series oscillated, was 0.7/week, roughly half the prior mean of 1.47/week.
-The posterior means of the $\epsilon_{nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (Supplementary Figure 1), though the full posterior of each $\epsilon_{nu t}$ overlapped zero.
-In addition, the standard deviation of the mortality forcing terms was small ($E(\sigma_{\nu}|\mathbf{y}) = 0.01$, Supplemental Figure 2) relative to the weak prior.
+The posterior means of the $\epsilon_{\nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (Supplementary Figure 1), though the full posterior distribution of each $\epsilon_{\nu t}$ overlapped zero.
+In addition, the standard deviation of the mortality forcing terms was small relative to the weak prior ($E(\sigma_{\nu}|\mathbf{y}) = 0.01$, Supplemental Figure 2).
 
@@
 \begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/fig3.eps}
+%\includegraphics[angle = 270, scale = 1]{figures/fig3.eps}
 \caption{{\bf The effect of cross-immunity on effect and timing of control.}
@@
 \begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/fig4.eps}
+%\includegraphics[angle = 270, scale = 1]{figures/fig4.eps}
 \caption{{\bf Timing of optimal control relative to disease and mosquito dynamics.}
@@
 
-Given the central role that mosquito lifespan plays in driving dengue dynamics, we model fluctuations in the mosquito mortality rate using a flexible, stochastically forced harmonic oscillator.
-The fact that this is the only source of stochasticity in the transmission process, however,  means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.
-Our implementation of the mosquito mortality process is conceptually similar to the empirical forcing functions developed by Hooker and Ellner \cite{Hooker2015}.
-In that framework, Hooker and Ellner \cite{Hooker2015} estimated nonparametric functions in time that modify either the differential equations or specific model parameters to provide a good fit to the data.
+In addition, there is relatively little uncertainty on our estimates of weekly mosquito mortality (Fig.
+\ref{mortality}) and the variance of the forcing terms is small, indicating that mosquito mortality is tightly constrained by the case reports.
+
+Given the central role that mosquito lifespan plays in driving dengue dynamics, we modeled fluctuations in the mosquito mortality rate using a flexible, forced harmonic oscillator.
+This encodes our prior expectation that the mosquito mortality rate should be driven by seasonal environmental forcing (i.e., it should oscillate with a period of one year), while allowing the flexibility to diverge from the simple oscillator and learn from the case reports data.
+The fact that the $\epsilon_{\nu}$ forcing terms are the only source of variability in the transmission process, however, means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.
+Hooker and Ellner \cite{Hooker2015} provide a framework for diagnosing such model misspecification in differential equation models using forcing functions similar to our implementation of the $\epsilon_{\nu}$.
+In that framework, Hooker and Ellner \cite{Hooker2015} estimate nonparametric forcing functions that modify a fitted differential equation model to provide a good fit to the data.
 These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables.
-We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect the estimated mosquito mortality time series in the same spirit.
-In particular, the weak periodic structure in the estimated $\epsilon_{dt}$ suggests that there may be some unmodeled, higher frequency process influencing transmission, either through the mortality rate itself or through another process.
+We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect our estimated $\epsilon_{\nu}$ forcing terms in the same spirit.
 
-Despite this, our estimated mosquito mortalities generally fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
-The fact that our estimated mortality rate increases with temperature above 20 degrees also agrees with the empirical literature on mosquito survival \cite{Yang2009}, though our estimated relationship is generally more extreme than estimated in laboratory studies, possibly as a result of assuming exponentially distributed extrinsic incubation period.
-In particular, the assumption that the extrinsic incubation period is exponentially distributed, and thus that many mosquitoes experience a very brief incubation period, may overestimate the probability that a mosquito survives long enough to become infectious.
-More accurate models of the extrinsic incubation period could incorporate explicit delays (i.e., assuming a fixed incubation period \cite{Burattini2008}), or additional chained exposed compartments (i.e., assuming the incubation period is gamma-distributed, \cite{Lloyd2001}).
-These observations suggest that our estimated mosquito mortality rates may actually be capturing the patterns of mosquito mortality in this system and are not obviously influenced by other sources of uncertainty or misspecification.
+In particular, the periodic structure in the posterior means of the $\epsilon_{\nu t}$ suggests that these residuals are accounting for more than just noise, and there may be some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
+Following Hooker and Ellner, we can explore whether this process results from misspecification of the rates of change of the existing state variables, or from missing state variables altogether. 
+The lack of any apparent relationship between any of the state variables and the forcing function, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
+These unmodeled components could include additional mosquito population dynamic processes (e.g.
+aquatic stage dynamics, environmental drivers, or control interventions), or epidiemiological processes (e.g.
+multiple circulationg serotypes of the dengue virus).
+Additional work developing and testing new models would be needed to distinguish among these.
+
+Despite this suggestion of potential model misspecification, our estimated mosquito mortalities nonetheless fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
+Moreover, the fact that our estimated mortality rate increased with temperature also broadly agrees with the empirical literature on mosquito survival \cite{Yang2009, Brady2013}.
+This suggests that whatever structure remains unexplained in the forcing terms, the pattern of case reports is still very well described by realistic fluctuations in the mosquito mortality rate.
+Given the substantial increase in model complexity that, say, resolving serotype dynamics would entail, we suggest that allowing a relatively small amount of phenomenological forcing to persist is a fair price to pay for parsimony.
 
@@
 This is also generally when immunity is highest.
-
-\subsubsection*{Larval controls}
-
-In contrast, larval control is most effective early in the year, right at the beginning of the epidemic.
-Intervention early in the year helps to prevent amplification and feedback.
-Also often corresponds to the local peak in susceptibility.
-This effectiveness is driven primarily by the timing of the reduction in mosquito abundance and could be achieved by either adult or larval control strategies.
-
-The timing of the effectiveness of both of these control strategies is driven largely by the interplay of the intrinsic feedbacks and seasonality of the system, and in particular by the fluctuations in human immunity and mosquito abundance.
-It is also interesting that neither strategy is most effective at the height of the outbreak, which highlights the fact that control strategies need to consider effects beyond the current season.
 

### 2018-05-30 Manuscript/ms.tex
Equations for gamma EIP and estimating dv0

@@
 \frac{dV_N}{dt} & = r(t) V_N - \phi_q \tau(t) V_N \\
-\frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_E\\
-\frac{dV_I}{dt} &= \rho_{v}(t) V_E - (d(t) + \phi_q \tau(t)) V_I\\
+\frac{dV_{E1}}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E1}\\
+\frac{dV_{E2}}{dt} &= \rho_{v}(t) V_{E1} - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E2}\\
+\frac{dV_{E3}}{dt} &= \rho_{v}(t) V_{E2}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E3}\\
+\frac{dV_{E4}}{dt} &= \rho_{v}(t) V_{E3}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E4}\\
+\frac{dV_I}{dt} &= \rho_{v}(t) V_{E4} - (d(t) + \phi_q \tau(t)) V_I\\
 \frac{dV_C}{dt} & = \phi_q \tau(t) V_N\\
@@
 
-The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
+The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$, $d_0$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
 Where possible, we specified informative priors on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).
@@
 $1/\rho_{vt}$ & Extrinsic incubation period & $7\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
-$d_0$ & Mean mosquito mortality rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
 $V_{E0}$ & Initial exposed mosquitoes &  0 & \\
@@
 \hline
+$d_0$ & Baseline mosquito mortality rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
 $1/\rho$ & Latent period in host & 0.87 weeks  & \cite{Chan2012}\\
@@
 Given samples from the posterior distribution as obtained above, we simulated the effects of a single pulse of control applied in each week of the first three years of the time series.
-Since the city already implements responsive, targeted control with the aim of reducing the disease burden of an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in both the current and future years.
+Since the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in both the current and future years.
 To obtain the posterior distribution of the proportion of cases prevented in calendar year $j$ by implementing control in week $i$, for each posterior sample $k$, we simulated the dynamics resulting from reducing the mosquito population by $5\%$ at the beginning of week $i$ (affecting susceptible, exposed, and infectious mosquitoes equally, and leaving the estimated mosquito birth and mortality rates unchanged) .

### 2018-05-30 Manuscript/supplement.tex
Equations for gamma EIP and estimating dv0

@@
 \\
+\item \textbf{Baselin mosquito mortality rate} ($d_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
+They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $d_0$.  
+We use nonlinear least squares to fit the exponential cdf with rate $d_0$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d_0 = 1.47$.
+We specify a gamma prior on the multiplicative scale, with parameters chosen to keep the prior mass with a factor of 2 of the mean:
+\begin{equation}
+d_0' \sim \text{Gamma}(10, 10)
+\end{equation}
+where $d_0 = 1.47d_0'$.
+\\
 \item \textbf{Initial conditions} ($S_0, E_0, I_0, R_0$): The SEIRS model assumes a constant population size and is scaled so that $S_0+E_0+I_0+R_0 = 1$.
@@
 \begin{itemize}
-\item \textbf{Mean mosquito mortality rate} ($d_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
-They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $d_0$.  
-We use nonlinear least squares to fit the exponential cdf with rate $d_0$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d_0 = 1.47$.
-\\
 \item \textbf{Transmission rate} ($\lambda$): The transmission rate only appears multiplied by a term from the mosquito model (either $V_I$ or $V_S$) and is thus effectively non-identifiable when we are also estimating the overall size of the mosquito population.

### 2018-05-30 Manuscript/ms.tex
Working on introduction

@@
 Because of this, there is a growing recognition that effective control needs to be guided by high quality vector surveillance, together with quantitative tools that synthesize vector surveillance with clinical surveillance, account for local epidemiology, and connect easily to local decision making \cite{Morrison2008, Scott2010b}.
-Moreover, mosquito control needs to be guided by an understanding of the link between mosquito surveillance abundance and disease risk so that the mosquitoes most responsible for transmission can be targeted \cite{Scott2010a, Scott2010b}.
+Moreover, mosquito control needs to be guided by an understanding of the link between mosquito abundance and disease risk so that the mosquitoes most responsible for transmission can be targeted \cite{Scott2010a, Scott2010b}.
 
-Many of the attempts to establish this link have found a weak relationship between mosquito surveillance and incidence of dengue fever\cite{Bowman2014, Pepin2015, Cromwell2017}.
+Many of the attempts to establish this link have found a weak relationship between mosquito surveillance and incidence of dengue fever \cite{Bowman2014, Pepin2015, Cromwell2017}.
 However, these attempts often do not account for the complex, non-linear interactions that mediate the relationship between mosquito abundance and human disease.
@@
 
-If the goal of modern dengue control is to deploy limited mosquito control resources when and where they will be most effective at preventing disease \cite{Scott2010a, Scott2010b}, then the quantitative tools used to guide that deployment need to account for the influence of the critical processes governing transmission.
-This includes the intrinsic human epidemiological and mosquito demographic processes above, but infectious disease dynamics are often further influenced by seasonality and stochastic forcing \cite{Ellner1998,Koelle2004, Grenfell2002}.
-In fact, the interplay of these factors is one of the unique features of ecological dynamics and is often responsible for generating the complex inter-annual cycles observed therein \cite{Bjornstad2001}.
+The importance of intrinsic nonlinearities, potentially alongside seasonality and stochastic forcing \cite{Ellner1998, Koelle2004, Grenfell2002}, in governing human disease risk highlight the need to integrate mechanistic modeling into the quantitative tools used to understand the effects of control interventions.
 Such mechanistic models can often perform better than complex autoregressive statistical models in describing and forecasting population dynamics \cite{Reilly2005}.
@@
 
-In order to meet the needs of mosquito control programs, however, these mechanistic models need to be embedded in a framework that can learn from commonly available surveillance data, while also accounting for multiple sources of uncertainty.
-Mosquito and clinical surveillance provide information about the underlying processes, but are observed with noise (CITATIONS).
-Differential equation models provide a natural way to encode the relevant biological mechanisms, though it is important to also account for sources of uncertainty and stochasticity in those models \cite{Hotelling1927, Wikle2010}.
-Moreover, the values of parameters are also often uncertain, which can lead to large uncertainty about the effects of management actions (CITATIONS).
+Differential equation models provide a natural way to encode the relevant biological mechanisms, though it is important to also account for sources of uncertainty in those models \cite{Hotelling1927, Wikle2010}.
+In particular, the values of parameters (e.g.
+the length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
+In addition, the structure of the processes themselves can be uncertain \cite{Ellner1998}, and need to be informed by available, often noisy, data.
 Bayesian hierarchical modeling provides a coherent framework to account for and integrate this uncertainty across the three levels of the model (data, process, and parameters \cite{Berliner1996, Cressie2009}).
 
-In this paper, we develop a modeling framework to integrate MI-Dengue and clinical surveillance from the city of Vitoria, Brazil to provide data-driven and mechanistically-grounded guidance for targeted mosquito control.
-Specifically, we develop a mechanistic differential equation model of dengue transmission that we embed in a hierarchical Bayesian statistical framework.
-This allows us to estimate latent time series of mosquito demographic parameters from available time series of mosquito trap counts and reported cases of dengue fever.
-We then use the fitted model to explore the optimal timing of adult mosquito control interventions.
+In this paper, we aim to integrate these elements -- a detailed mechanistic model of dengue transmission with a full Bayesian accounting of uncertainty -- in order to better understand the interplay of forces governing dengue dynamics and their interaction with potential vector control interventions.
+We apply and test this framework on clinical and entomological surveillance data from the city of Vitoria, Brazil.
+These data allow us to estimate a latent time series of mosquito mortality rates that modulate the transmission process and link between mosquito abundance to human disease.  
+We then use the fitted model to explore how perturbations to the mosquito population propagate and interact with the complex epidemiology and nonlinearities involved in dengue transmission.
 

### 2018-06-04 Analysis/plots.r
Updating figures

@@ -35,13 +35,14 @@ weather <- read.csv("Data/Vitoria.weather.csv") %>%
 weather <- subset(weather, date < date[1] + weeks(243))
 weather$tot.week <- rep(c(1:243), each = 7)
 
-temp <- ddply(weather, .(tot.week), summarise, temp = mean(Mean.TemperatureC, na.rm = T))
+temp <- ddply(weather, .(tot.week), summarise, temp = mean(Mean.TemperatureC, na.rm = T),
+              humid = mean(Mean.Humidity, na.rm = T))
 
 post$temp <- temp$temp
 post$rov <- 7 * exp(0.2 * temp$temp - 8)
 
 # Loading MCMC results
-sim <- readRDS("Results/gamma_eip.rds")
+sim <- readRDS("Results/gamma_eip_dv0.rds")
 
 # Figure 1: Observed and estimated time series =================================
 
@@ -90,10 +91,11 @@ dev.off()
 
 # Figure 2: Estimates of latent mosquito mortality rate ========================
 
+dvmu <- 1.47 * rstan::extract(sim, "dvmu_c", permute = T)[[1]]
 system <- rstan::extract(sim, "state", permute = T)[[1]]
 
-dv <- system[, 2:244, 12] %>% + 0.39 %>% exp() %>% 
-  adply(2, quantile, c(0.1, 0.5, 0.9), .id = NULL)
+dv <- system[, 2:244, 12] %>% exp() %>% t() %>% multiply_by_matrix(diag(dvmu)) %>% 
+  adply(1, quantile, c(0.1, 0.5, 0.9), .id = NULL)
 names(dv) <- c("dvmin", "dvmed", "dvmax")
 
 post <- cbind(post, dv)
@@ -104,7 +106,7 @@ ggplot(post, aes(tot.week, dvmed)) +
   geom_ribbon(aes(ymin = dvmin, ymax = dvmax), fill = "grey70") +
   geom_line() + 
   theme_classic() +
-  scale_y_continuous(expand = c(0.05, 0), limits = c(0.3, 2.2)) + 
+  scale_y_continuous(expand = c(0.05, 0), limits = c(0.25, 1.35)) + 
   scale_x_continuous(expand = c(0, 1)) +
   ylab("mosquito mortality rate") +
   xlab("week") +
@@ -113,7 +115,7 @@ ggplot(post, aes(tot.week, dvmed)) +
 ggplot(post, aes(temp, dvmed)) +
[…]

### 2018-06-04 Manuscript/ms.tex
Adding new control results

@@
 
+Mosquito control (i.e.
+the removal of $5\%$ of the adult population in a given week), had the largest effect on the number of dengue cases in a given year when implemented between 30 and 40 weeks prior to the start of that year (Figure \ref{control}).
+Our simulations predicted that control implemented during this time could prevent between $20$ and $30\%$ of the cases in the following year, while control implemented during the target outbreak was much less effective.
+
+The effect of a given control intervention seemed to be driven largely by its interaction with, and timing relative to, fluctuations in human immunity.
+In particular, mosquito control seemed to be most effective when applied at or near a local minimum in the number of suceptible humans (Figure \ref{timing}).
+In addition, human immunity, as controlled by the period of cross-immunity ($1/\delta$), was aslo a major driver of uncertainty in the effect of a given control intervention.
+Specifically, we observed a strong negative posterior correlation between the maximum proportion of cases that could be prevented in a given year (i.e., the number of cases prevented by control implemented in the optimal week) and the period of cross-immunity (Figure \ref{immunity}).
+As the period of cross-immunity increases, the susceptible pool is replenished more slowly, and the proportion of the population that is susceptible decreases.
+The negative correlation between the effectiveness of control and the period of cross-immunity thus suggests that mosquito control becomes less effective as the availability of susceptibles  becomes more limiting.
+
 \begin{figure}[!h]
@@
 \caption{{\bf Estimated latent mosquito mortality rate.}
-A: Time series of mortality rate, showing median posterior estimate (line) and 80\% credible interval (gray band). B: weekly mosquito mortality rate as a function of weekly mean temperature..
+A: Time series of mortality rate, showing median posterior estimate (line) and 80\% credible interval (gray band). B: weekly mosquito mortality rate as a function of weekly mean temperature.
 }
@@
 \begin{figure}[!h]
-%\includegraphics[angle = 270, scale = 1]{figures/fig3.eps}
-\caption{{\bf The effect of cross-immunity on effect and timing of control.}
-Each point represents a sample from the posterior and the results of the corresponding control simulation, in which we evaluated the effects of a $5\%$ reduction in mosquito abundance at each of the 52 weeks from week 16 of the previous year to week 15 of the target year.
-A, B, C: the maximum proportion of cases prevented target years 2009, 2010, and 2011, respectively.
-D, E, F: the epidemiological week in which control achieved the maximum reduction in cases in 2009, 2010, and 2011, respectively.
+\includegraphics[angle = 270, scale = 1]{figures/fig3.eps}
+\caption{{\bf The effect of mosquito control as a function of the week in which it was applied.}
+Summary of posterior predicted distribution of the proportion of cases that would be prevented in each year (top to bottom, 2009 to 2012) by control enacted in a given week relative to the start of that year.
+Black lines indicate the posterior median, while gray ribbons indicate the $80\%$ credible interval.
 }
-\label{immunity}
+\label{control}
 \end{figure}
@@
 \begin{figure}[!h]
-%\includegraphics[angle = 270, scale = 1]{figures/fig4.eps}
+\includegraphics[angle = 270, scale = 1]{figures/fig4.eps}
 \caption{{\bf Timing of optimal control relative to disease and mosquito dynamics.}
@@
 A: posterior median estimate of new dengue cases per week.
-B: posterior median estimate of number of mosquitoes per person each week.
-C: posterior median estimate of the proportion of susceptible individuals.
+B: posterior median estimate of the proportion of susceptible individuals.
+C: posterior median estimate of number of mosquitoes per person each week.
 }
 \label{timing}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 0.8]{figures/fig5.eps}
+\caption{{\bf The effect of cross-immunity on effect of control.}
+The posteriod distribution of the maximum proportion of cases prevented by control in each year (left to right, 2009 to 2012), as a function of the period of cross-immunity ($1/\delta$).
+The shading indicates posterior density, with darkers shading indicating a larger density of posterior samples in that bin.
+}
+\label{immunity}
 \end{figure}
@@
 
-Targeted mosquito control can be an effective tool in reducing the disease burden within a city.
-When faced with limited resources with which to deploy that control, our results suggest that focusing efforts early in the year, specifically in the second to fourth weeks, will lead to the largest reductions in disease.
-The lag between optimal control and both mosquito population dynamics and human disease highlights the importance of early intervention and the need to account for the nonlinear processes governing transmission.
-Targeting the mosquito population early in the year as dengue just begins to spread can disrupt the transmission cycle and prevent further amplification and feedback later in the year.
-
 \subsection*{Model validation and interpretation of $d_v$}
-
-In addition, there is relatively little uncertainty on our estimates of weekly mosquito mortality (Fig.
-\ref{mortality}) and the variance of the forcing terms is small, indicating that mosquito mortality is tightly constrained by the case reports.
 
@@
 
-The case data to which we fit the model represents reports of "dengue-like illness", without laboratory confirmation, and as such could include cases of other diseases with similar symptoms (e.g.
+The case data to which we fit the model represent reports of "dengue-like illness", without laboratory confirmation, and as such could include cases of other diseases with similar symptoms (e.g.
 chikungunya or Zika).
@@
 These control efforts are guided by the MI-Dengue system, which uses the trap-level mosquito surveillance data to target areas of high mosquito infestation for control (including source reduction, larvacide, and adulticide)\cite{Eiras2009}.
+These control efforts may affect our estimates of the mosquito demographic rates, but should not influence our understanding of dengue dynamics, since we are not interested in the factors driving mosquito dynamics, but rather in how the realized mosquito dynamics (whether forced by human intervention of environmental factors) fit into the broader epidemiological picture.
+
 These local interventions are effective at reducing disease burden in the cities in which they are implemented \cite{Pepin2013}, but long-term epidemiological processes (e.g.
@@
 
-Mosquito surveillance is a valuable tool in managing dengue fever, but is most powerful when integrated into a larger modeling framework.
+Efforts to connect mosquito abundance to human disease are often hampered by the confounding influence of human immunity, and mosquito survival.
+The challenges presented by these confounding factors highlight the value of mechanistic information in studying the effect of mosquito control on disease spread.
+In the Bayesian context we have deployed here, this mechanistic knowledge, as formalized in the specification of a differential equation model, can be viewed as part of the prior on the relationship between mosquito abundance and human disease.
+As such, we should neither ignore it, nor let it swamp the signal in our data.
+
 We have developed a simple yet realistic mechanistic model of dengue fever spread that allows us to combine mosquito and clinical surveillance data to estimate the latent mosquito demographic rates that connect mosquito abundance and human disease.

### 2018-06-06 Manuscript/ms.tex
Discussion of control

@@
 
+Infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are often driven by the combined efforts of intrinsic non-linearities and feedbacks, seasonality, and stochasticity.
+In this paper, we have highlighted the importance of these processes in driving the dynamics of dengue fever and shaping the relationship between mosquito abundance and human disease risk.
+
 \subsection*{Model validation and interpretation of $d_v$}
@@
 
-Infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are often driven by the combined efforts of intrinsic non-linearities and feedbacks, seasonality, and stochasticity.
-In this paper, we have highlighted the importance of these processes in driving the dynamics of dengue fever and shaping the relationship between mosquito abundance and human disease risk.
-
-\subsubsection*{Adult control}
+\subsubsection{Timing of control}
 
@@
 In fact, this forms the basis for much of the theory of adult control \cite{Burattini2008, Morrison2008, Smith2012}.
-Our results suggest that such control efforts (i.e., those that remove adult mosquitoes from the population) are most effective when applied late in the outbreak (week 24), after the number of cases has peaked.
-If applied at the peak of the epidemic, the this control is likely less effective, as there are still a large number of infectious human hosts available \cite{Newton1992, Burattini2008}.
-Control applied late in the outbreak pushes the system in the direction it is already heading.
-Thus control that interacts with the existing seasonality is more effective.
-In measles, this is also generally when the effect of stochasticity is most limited and the dynamics are pulled most strongly towards the deterministic attractor.
-This is also generally when immunity is highest.
+Despite this, our simulations suggest that control is likelyl to be relatively ineffective when implemented during an outbreak, likely due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
+Instead, our results suggest that adult mosquito control efforts are most effective at reducing the disease burden within a given year when applied roughly 30 weeks prior to the start of that year, on the downswing of the previous year's epidemic and seasonal mosquito dynamics.
+This stands in contrast to the conclusions of Stoddard \emph{et al.} \cite{Stoddard2014}, who found that the effects of control implemented during the decline phase of an outbreak were difficult to detect in the presence of natually fading transmission intensity.
+While it may be true that it is difficult to detect the immediate effects of control implemented at this time, our results nonetheless indicate long-term benefits to controlling mosquitos during the decline phase.
+
+\subsubsection{Role of human immunity in effect of control}
+
+The mismatch between the timing of the most effective control and the peak in mosquito abundance suggets that mosquito abundance alone is an insufficient predictor of the effect of mosquito control, at least at the city-scale.
+Instead, the effectiveness of mosquito control depends in large part on the intrinsic long-term feedbacks and lags driven by the dynamics of temporary human immunity.
+In fact, control implemented at a 30 week lag is effective largely due to the way it interacts with the dynamics of the susceptible pool. 
+Implementing control just as the susceptible population begins to grow (when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) allows for additional immunizing infections as the current outbreak subsides, while also disrupting transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
+The fact that effect of mosquito control is mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g.
+cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities.
+
+\subsubsection{Uncertainty and role of cross-immunity}
+
+The Bayesian framework allows for a full accounting of uncertainty across the data, process, and parameter levels of our model \cite{Berliner1996}.
+This then allows us to carry this uncertainty through to our simulations of control interventions \cite{Elderd2006}.
+We found that there can be substantial uncertainty in the proportion of cases prevented by a control intervention (i.e., the width of the ribbons in Figure \ref{control}).
+Much of this uncertainty could be attributed to uncertainty in the period of cross-immunity (Figure \ref{immunity}), the posterior distribution of which differed little from our prior based on the estimate of Reich \emph{et al.} \cite{Reich2013}.
+Specifically, we found that increasing the period of cross-immunity decreases the proportion of cases prevented by mosquito control.
+When the period of cross-immmunity is long, the susceptible pool replenishes slowly and transmission becomes limited by the availability of susceptible hosts.
+In the face of this limitation, manipulating mosquito abundance has a relatively small effect.
+Conversely, when the period of cross-immunity is short, susceptibles are abundant and mosquito abundance is limiting, in which case reducing mosquito abundance further has a larger effect on transmission.
+This interaction between the efficacy of control and human immunity was also demonstrated by ten Bosch \emph{et al.} \cite{TenBosch2016} in their investigation of serotype interactions.
+In particular, they found that adding cross-immunity to models with antibody dependent enhancement increased the resistance to control (as implemented by reducing the transmission rate).
+
+\subsubsection{Implications for policy}
+
+This notion that mosquito control becomes less effective as the susceptible pool becomes more limiting has potential long-term consequences for the management of dengue fever.
+In particular, as immunity becomes widespread, either due to continued transmission or vaccination, mosquito abundance will no longer be limiting, and broad-scale mosquito control may become less effective.
+At the same time, the replenishment of susceptibles, from both births and waning immunity, means that sustaining effective control over the long-term is likely to be difficult \cite{Scott2010b} and will thus require continued monitoring and management.
+As such, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance to monitor and estimate immunity \cite{Morrison2008}.
+In the absence of, or as a supplement to, such data, mechanistic models like the one developed here, or so-called TSIR frameworks that reconstruct the dynamics of the susceptible class \cite{Finkenstadt2000, Reich2013}, need to be further developed to better inform mosquito control efforts.
 

### 2018-06-06 Manuscript/ms.tex
Cleaning up conclusion paragraphs

@@
 
-Efforts to connect mosquito abundance to human disease are often hampered by the confounding influence of human immunity, and mosquito survival.
+Efforts to connect mosquito abundance to human disease are often hampered by the confounding influences of human immunity, and mosquito survival.
 The challenges presented by these confounding factors highlight the value of mechanistic information in studying the effect of mosquito control on disease spread.
-In the Bayesian context we have deployed here, this mechanistic knowledge, as formalized in the specification of a differential equation model, can be viewed as part of the prior on the relationship between mosquito abundance and human disease.
-As such, we should neither ignore it, nor let it swamp the signal in our data.
+In the Bayesian context we have deployed here, this mechanistic knowledge, as formalized in the specification of a differential equation model, can be viewed as part of the prior on the relationship between mosquito abundance and human disease \cite{Ellner1998, Wikle2010}.
+As such, we should neither ignore it, nor encode it so rigidly that it swamps the signal in our data.
 
-We have developed a simple yet realistic mechanistic model of dengue fever spread that allows us to combine mosquito and clinical surveillance data to estimate the latent mosquito demographic rates that connect mosquito abundance and human disease.
-The mechanistic framework allows us to capture important lags and feedbacks in the transmission process and to identify critical intervention points that would not be apparent otherwise.
+We have developed a simple yet realistic mechanistic model of dengue fever spread that represents the fundamental elements of our prior understanding of dengue epidemiology, while also allowing for uncertainty and flexibility in the fluctuations of mosquito demographic rates.
+This mechanistic framework allows us to capture the critical role of human immunity in mediating the transmission process and to identify critical intervention points that would not be apparent otherwise.
 The fully hierarchical Bayesian framework in which we embedded the mechanistic model allows for a thorough accounting of uncertainty that is carried through to the evaluation of different control strategies.

### 2018-06-06 Manuscript/ms.tex
Cleaning up caveats

@@
 These control efforts may affect our estimates of the mosquito demographic rates, but should not influence our understanding of dengue dynamics, since we are not interested in the factors driving mosquito dynamics, but rather in how the realized mosquito dynamics (whether forced by human intervention of environmental factors) fit into the broader epidemiological picture.
+That said, our control simulations should be interepreted as exploring the effect of an additional pulse of city-wide control on top of the existing control activities.
 
-These local interventions are effective at reducing disease burden in the cities in which they are implemented \cite{Pepin2013}, but long-term epidemiological processes (e.g.
-fluctuations in human immunity) still seem to be driving disease spread.
-Our control simulations account for these processes and offer a suggestion for how an additional city-wide pulse of control might be deployed to supplement existing efforts to reduce mosquito density throughout the city.
-
-That said, we do not anticipate that a single, simultaneous, city-wide pulse of control could be feasibly implemented on top of the existing control efforts.
-However, contrary to our assumption of homogeneous mixing between mosquitoes and humans, all mosquitoes may not contribute equally to dengue spread.
-In fact, a relatively small proportion of the mosquito population may be responsible for a large proportion of disease spread \cite{Yoon2012}
+We view our simulations of this additional pulse of control primarily as a tool to explore and exploit the long-term feedbacks in the city's epidemiological system, whereas existing control is deployed largely to reduce local (in space and time) disease risk.
+This difference in scale highlights the potential for spatial heterogeneities in disease risk within a city.
 In particular, structured human movement within the city is likely to induce heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009}.
 In addition, spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
-The mechanistic model presented here may be able to suggest when an intervention is likely to be effective, but in order to make the best use of limited resources, spatial prioritization may still be necessary.
-This should include not just prioritization of areas with high mosquito abundance, but also areas with disproprtionate risk of disease spread, either due to hetereogeneities in the mixing process or in the distribution of human immunity.
+Though our model does not account for these heterogeneities, it is nontheless able to capture the city-wide dynamics reasonably well, suggesting that there may be sufficient mixing to appear homogeneous at the city scale.
+Further, though our mechanistic model may be able to suggest when a city-wide intervention is likely to be effective, in order to make the best use of limited resources, spatial prioritization may still be necessary.
+Our results emphasize that this should include not just prioritization based on the distribution of mosquito abundance, but also on the distribution of human immunity.
 

### 2018-06-07 Manuscript/ms.tex
Intro paragraph of discussion

@@
 
-Infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are often driven by the combined efforts of intrinsic non-linearities and feedbacks, seasonality, and stochasticity.
-In this paper, we have highlighted the importance of these processes in driving the dynamics of dengue fever and shaping the relationship between mosquito abundance and human disease risk.
+Infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are often driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
+In the case of dengue fever, these processes can obscure the relationship between mosquito abundance and human disease.
+We found that modeling temporary human immunity and seasonal variation in the mosquito mortality rate was able to effectively fit case reports and mosquito surveillance data from Vitoria, Brazil.
+Control simulations using this model indicated that mosquito control was most effective at reducing the disease burden in a given year when implemented approximately 30 weeks prior to the start of that year, when the susceptible pool tends to be at a minimum.
+Moreover, the effectiveness of control was strongly correlated with the period of cross-immunity, and the rate of turnover of the susceptible population.
+Thus, human immunity appears to be a critical driver of the effectiveness of mosquito control.
 

### 2018-06-07 Manuscript/ms.tex
Gamma EIP in methods

@@
 
-Similarly, susceptible mosquitoes ($V_S$) become exposed by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
+Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_S$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
+Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implement a gamma-distributed incubation period by chaining multiple exposed classes together (taking advantage of the fact that the sum of a collection of exponential random variables with the same rate parameter is gamma-distributed) \cite{Lloyd2001}.
 Total mosquito population size ($V_N$) is controlled by stochastic, seasonally varying growth rate ($r(t)$) and death rate ($d(t)$).
@@
 $\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
-$1/\rho_{vt}$ & Extrinsic incubation period & $7\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
+$1/\rho_{vt}$ & Extrinsic incubation period & $28\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
 $V_{E0}$ & Initial exposed mosquitoes &  0 & \\

### 2018-06-07 Manuscript/ms.tex
Updating the abstract

@@
 We developed a differential equation model of dengue transmission and embedded it in a Bayesian hierarchical framework that allowed us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vitoria, Brazil.
-We used the fitted model to explore the optimal timing of targeted adult control during the year.
-We found that control was most effective when deployed in the third week of the year, as mosquito abundance neared its peak and dengue transmission began to ramp up.
-Intervening at this time disrupts transmission and prevents the amplification and feedback that leads to large outbreaks of disease, an insight not possible without the underlying mechanistic model.
-Though the ability of mosquito surveillance to predict human disease is often limited, this highlights the utility of this surveillance when integrated into a larger modeling framework.
-Grounding this modeling famework in the actual mechanisms of transmission will help to establish more effective and efficient dengue control policies that allow us to better target mosquitoes that are most responsible for continuing disease spread.
+We used the fitted model to explore the effect of adult mosquito control on the disease burden in each year, while taking advantage of the Bayesian framework to carry through uncertainty from the posterior distributions of model parameters and mosquito demographic rates.
+We found that control deployed between 30 and 40 weeks prior to the start of the year was most effective at preventing cases in that year. 
+This timing generally coincided with a local minimum in the size of the susceptible pool.
+Uncertainty in the effectiveness of control interventions was driven largely by uncertainty in the period of cross-immunity, with longer periods of cross-immunity leading to less effective control.
+Understanding the dynamics of temporary human immunity is thus critical for understanding the effect of mosquito control efforts on human disease.
 

### 2018-06-07 Manuscript/ms.tex
Cleaning up the ms

@@
 In particular, host immunity is a key intrinsic driver of infectious disease dynamics, and conditions favorable for transmission can only lead to an outbreak of disease when there is a sufficiently large population of susceptible hosts \cite{Koelle2004, Koelle2005}.
-As such, the ability of mosquitoes to contribute to DENV transmission depends critically on the level of prior immunity in the human population \cite{Scott2010a}.
+As such, the ability of mosquitoes to contribute to DENV transmission depends critically on the level of immunity in the human population \cite{Scott2010a}.
 Further, the cycle of transmission between humans and mosquitoes is influenced not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period in mosquitoes \cite{Smith2012}.
@@
 the length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
-In addition, the structure of the processes themselves can be uncertain \cite{Ellner1998}, and need to be informed by available, often noisy, data.
+The structure of the processes themselves can be uncertain \cite{Ellner1998}, and need to be informed by available, often noisy, data.
 Bayesian hierarchical modeling provides a coherent framework to account for and integrate this uncertainty across the three levels of the model (data, process, and parameters \cite{Berliner1996, Cressie2009}).
@@
 In this paper, we aim to integrate these elements -- a detailed mechanistic model of dengue transmission with a full Bayesian accounting of uncertainty -- in order to better understand the interplay of forces governing dengue dynamics and their interaction with potential vector control interventions.
-We apply and test this framework on clinical and entomological surveillance data from the city of Vitoria, Brazil.
-These data allow us to estimate a latent time series of mosquito mortality rates that modulate the transmission process and link between mosquito abundance to human disease.  
+We apply this framework to clinical and entomological surveillance data from the city of Vitoria, Brazil.
+These data allow us to estimate a latent time series of mosquito mortality rates that modulate the transmission process and link mosquito abundance to human disease.  
 We then use the fitted model to explore how perturbations to the mosquito population propagate and interact with the complex epidemiology and nonlinearities involved in dengue transmission.
@@
 
-Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_S$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
-Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implement a gamma-distributed incubation period by chaining multiple exposed classes together (taking advantage of the fact that the sum of a collection of exponential random variables with the same rate parameter is gamma-distributed) \cite{Lloyd2001}.
-Total mosquito population size ($V_N$) is controlled by stochastic, seasonally varying growth rate ($r(t)$) and death rate ($d(t)$).
+Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_E$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
+Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implement a gamma-distributed incubation period by chaining multiple exposed classes together (taking advantage of the fact that a gamma-distributed random variable can be generated through the sum of a collection of exponential random variables with the same rate parameter) \cite{Lloyd2001}.
+Total mosquito population size ($V_N$) is controlled by a forced, seasonally varying growth rate ($r(t)$) and death rate ($d(t)$).
 Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.
@@
 $S_0$ & Proportion initially susceptible & 0.4 & \cite{Cardoso2011a} \\
-$E_0$ & Proportion initially exposed & $8\times 10 ^ {-5}$ & \\
-$I_0$ & Proportion initially infectious & $8\times 10 ^ {-5}$ & \\
+$E_0$ & Number initially exposed & $100$ & \\
+$I_0$ & Number initially infectious & $60$ & \\
 $r_0$ & Initial mosquito population growth rate & 0 & \\
+$\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & \\
 $V_{N0}$ & Initial mosquito population size & $2N$ & \\
-$\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & \\
 $\phi_q$ & Per-trap mosquito capture rate & $2 \times 10^{-6}$ week$^{-1}$ & \\
@@
 Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.  
+Timesteps as small as $1/8$ of a day were explored and did not qualitatively change our results.
 
@@
 
-Given samples from the posterior distribution as obtained above, we simulated the effects of a single pulse of control applied in each week of the first three years of the time series.
-Since the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in both the current and future years.
+Given samples from the posterior distribution as obtained above, we simulated the effects of a single pulse of control applied in each week of the first four years of the time series.
+Since the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in future years.
 To obtain the posterior distribution of the proportion of cases prevented in calendar year $j$ by implementing control in week $i$, for each posterior sample $k$, we simulated the dynamics resulting from reducing the mosquito population by $5\%$ at the beginning of week $i$ (affecting susceptible, exposed, and infectious mosquitoes equally, and leaving the estimated mosquito birth and mortality rates unchanged) .
@@
 In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 6 -8).
-The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that these data contain little additional information about these parameters (Supplemental Figure 3).
+The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that the Vitoria data contained little additional information about these parameters (Supplemental Figure 3).
 The estimated latent period in a human host ($1/\rho$, the expected time it takes for an exposed human to become infectious to biting mosquitoes) is influenced more strongly by the data, with a posterior mean of 1.53 weeks compared to a prior mean of 0.87 weeks.
@@
 In particular, mosquito control seemed to be most effective when applied at or near a local minimum in the number of suceptible humans (Figure \ref{timing}).
-In addition, human immunity, as controlled by the period of cross-immunity ($1/\delta$), was aslo a major driver of uncertainty in the effect of a given control intervention.
+In addition, human immunity, as controlled by the period of cross-immunity ($1/\delta$), was a major driver of uncertainty in the effect of a given control intervention.
 Specifically, we observed a strong negative posterior correlation between the maximum proportion of cases that could be prevented in a given year (i.e., the number of cases prevented by control implemented in the optimal week) and the period of cross-immunity (Figure \ref{immunity}).
-As the period of cross-immunity increases, the susceptible pool is replenished more slowly, and the proportion of the population that is susceptible decreases.
+In simulations with a short period of cross-immunity, the susceptible pool replenished quickly, leading to a relatively large standing crop of susceptibles at any time.
+In simulations with a long period of cross-immunity, the susceptible pool replenished more slowly, leaving the population with relatively few susceptibles.
 The negative correlation between the effectiveness of control and the period of cross-immunity thus suggests that mosquito control becomes less effective as the availability of susceptibles  becomes more limiting.
@@
 
-\subsection*{Summary}
+\subsection*{Model validation and interpretation of $d_v$}
 
@@
 In the case of dengue fever, these processes can obscure the relationship between mosquito abundance and human disease.
-We found that modeling temporary human immunity and seasonal variation in the mosquito mortality rate was able to effectively fit case reports and mosquito surveillance data from Vitoria, Brazil.
-Control simulations using this model indicated that mosquito control was most effective at reducing the disease burden in a given year when implemented approximately 30 weeks prior to the start of that year, when the susceptible pool tends to be at a minimum.
-Moreover, the effectiveness of control was strongly correlated with the period of cross-immunity, and the rate of turnover of the susceptible population.
-Thus, human immunity appears to be a critical driver of the effectiveness of mosquito control.
-
-\subsection*{Model validation and interpretation of $d_v$}
+We found that a mechanistic model with temporary human immunity and seasonal variation in the mosquito mortality rate was able to effectively fit case reports and mosquito surveillance data from Vitoria, Brazil.
 
@@
 In that framework, Hooker and Ellner \cite{Hooker2015} estimate nonparametric forcing functions that modify a fitted differential equation model to provide a good fit to the data.
-These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables.
+These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables \cite{Hotelling1927, Hooker2015}.
 We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect our estimated $\epsilon_{\nu}$ forcing terms in the same spirit.
 
-In particular, the periodic structure in the posterior means of the $\epsilon_{\nu t}$ suggests that these residuals are accounting for more than just noise, and there may be some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
-Following Hooker and Ellner, we can explore whether this process results from misspecification of the rates of change of the existing state variables, or from missing state variables altogether. 
-The lack of any apparent relationship between any of the state variables and the forcing function, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
+The periodic structure in the posterior means of the $\epsilon_{\nu t}$ suggests that these terms are accounting for more than just noise, and there may be some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
+Following Hooker and Ellner, we can explore whether this process is likely to result from misspecification of the rates of change of the existing state variables, or from missing state variables altogether.
+The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
 These unmodeled components could include additional mosquito population dynamic processes (e.g.
@@
 This suggests that whatever structure remains unexplained in the forcing terms, the pattern of case reports is still very well described by realistic fluctuations in the mosquito mortality rate.
-Given the substantial increase in model complexity that, say, resolving serotype dynamics would entail, we suggest that allowing a relatively small amount of phenomenological forcing to persist is a fair price to pay for parsimony.
 
-\subsection*{Processes driving dynamics and implications for control}
-
-\subsubsection{Timing of control}
+\subsection*{Processes driving effect of mosquito control}
 
@@
 
-\subsubsection{Role of human immunity in effect of control}
-
 The mismatch between the timing of the most effective control and the peak in mosquito abundance suggets that mosquito abundance alone is an insufficient predictor of the effect of mosquito control, at least at the city-scale.
@@
 
-\subsubsection{Uncertainty and role of cross-immunity}
-
 The Bayesian framework allows for a full accounting of uncertainty across the data, process, and parameter levels of our model \cite{Berliner1996}.
@@
 Much of this uncertainty could be attributed to uncertainty in the period of cross-immunity (Figure \ref{immunity}), the posterior distribution of which differed little from our prior based on the estimate of Reich \emph{et al.} \cite{Reich2013}.
-Specifically, we found that increasing the period of cross-immunity decreases the proportion of cases prevented by mosquito control.
-When the period of cross-immmunity is long, the susceptible pool replenishes slowly and transmission becomes limited by the availability of susceptible hosts.
-In the face of this limitation, manipulating mosquito abundance has a relatively small effect.
-Conversely, when the period of cross-immunity is short, susceptibles are abundant and mosquito abundance is limiting, in which case reducing mosquito abundance further has a larger effect on transmission.
+Specifically, we found that increasing the period of cross-immunity decreased the proportion of cases prevented by mosquito control.
+When the period of cross-immmunity was long, the susceptible pool replenished slowly and transmission became limited by the availability of susceptible hosts.
+In the face of this limitation, manipulating mosquito abundance had a relatively small effect.
+Conversely, when the period of cross-immunity was short, susceptibles were abundant and mosquito abundance was instead limiting, in which case reducing mosquito abundance further had a larger effect on transmission.
 This interaction between the efficacy of control and human immunity was also demonstrated by ten Bosch \emph{et al.} \cite{TenBosch2016} in their investigation of serotype interactions.
-In particular, they found that adding cross-immunity to models with antibody dependent enhancement increased the resistance to control (as implemented by reducing the transmission rate).
-
-\subsubsection{Implications for policy}
+Specifically, they found that adding cross-immunity to models with antibody dependent enhancement increased the resistance to control (as implemented by reducing the transmission rate).
 
 This notion that mosquito control becomes less effective as the susceptible pool becomes more limiting has potential long-term consequences for the management of dengue fever.
-In particular, as immunity becomes widespread, either due to continued transmission or vaccination, mosquito abundance will no longer be limiting, and broad-scale mosquito control may become less effective.
+As immunity becomes widespread, either due to continued transmission or eventually vaccination, mosquito abundance will no longer be limiting, and broad-scale mosquito control may become less effective.
 At the same time, the replenishment of susceptibles, from both births and waning immunity, means that sustaining effective control over the long-term is likely to be difficult \cite{Scott2010b} and will thus require continued monitoring and management.
@@
 These control efforts are guided by the MI-Dengue system, which uses the trap-level mosquito surveillance data to target areas of high mosquito infestation for control (including source reduction, larvacide, and adulticide)\cite{Eiras2009}.
-These control efforts may affect our estimates of the mosquito demographic rates, but should not influence our understanding of dengue dynamics, since we are not interested in the factors driving mosquito dynamics, but rather in how the realized mosquito dynamics (whether forced by human intervention of environmental factors) fit into the broader epidemiological picture.
+These control efforts may affect our estimates of the mosquito demographic rates, but should not influence our understanding of dengue dynamics, since we are not interested in the factors driving mosquito dynamics, but rather in how the realized mosquito dynamics (whether forced by human intervention or environmental factors) fit into the broader epidemiological picture.
 That said, our control simulations should be interepreted as exploring the effect of an additional pulse of city-wide control on top of the existing control activities.
@@
 Further, though our mechanistic model may be able to suggest when a city-wide intervention is likely to be effective, in order to make the best use of limited resources, spatial prioritization may still be necessary.
-Our results emphasize that this should include not just prioritization based on the distribution of mosquito abundance, but also on the distribution of human immunity.
+Our results emphasize that work to develop the tools necessary for this prioritization should focus not just on prioritization based on the distribution of mosquito abundance, but also on the distribution of human immunity.
 
@@
 
-Efforts to connect mosquito abundance to human disease are often hampered by the confounding influences of human immunity, and mosquito survival.
+Efforts to connect mosquito abundance to human disease are often hampered by the confounding influences of human immunity and mosquito survival.
 The challenges presented by these confounding factors highlight the value of mechanistic information in studying the effect of mosquito control on disease spread.
@@
 We have developed a simple yet realistic mechanistic model of dengue fever spread that represents the fundamental elements of our prior understanding of dengue epidemiology, while also allowing for uncertainty and flexibility in the fluctuations of mosquito demographic rates.
-This mechanistic framework allows us to capture the critical role of human immunity in mediating the transmission process and to identify critical intervention points that would not be apparent otherwise.
-The fully hierarchical Bayesian framework in which we embedded the mechanistic model allows for a thorough accounting of uncertainty that is carried through to the evaluation of different control strategies.
+This mechanistic framework allowed us to capture the critical role of human immunity in mediating the transmission process and to identify critical intervention points that would not be apparent otherwise.
+The fully hierarchical Bayesian framework in which we embedded the mechanistic model allowed for a thorough accounting of uncertainty that was carried through to the evaluation of different control strategies.
 This combination of model features helps to meet the need for more effective, biologically grounded, and data-driven dengue control policies and offers a building block on which these tools can be further developed in the future.

### 2018-06-07 Manuscript/ms.tex
More cleaning

@@
 Our ability to effectively prevent the transmission of the dengue virus through targeted control of its vector, \emph{Aedes aegypti}, depends critically on our understanding of the link between mosquito abundance and human disease risk.
-The mosquito and clinical surveillance data necessary to elucidate this link are widely collected, but have yet to be coupled with a modeling framework that accounts for the complex non-linear mechanisms involved in transmission, in particular the critical bottleneck imposed by mosquito mortality.
+The mosquito and clinical surveillance data necessary to elucidate this link are widely collected, but have yet to be coupled with a modeling framework that accounts for the complex non-linear mechanisms involved in transmission.
+Most critical are the bottleneck in transmission imposed by mosquito lifespan relative to the virus' incubation period, and the dynamics of human immunity.
 We developed a differential equation model of dengue transmission and embedded it in a Bayesian hierarchical framework that allowed us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vitoria, Brazil.
@@
 Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implement a gamma-distributed incubation period by chaining multiple exposed classes together (taking advantage of the fact that a gamma-distributed random variable can be generated through the sum of a collection of exponential random variables with the same rate parameter) \cite{Lloyd2001}.
-Total mosquito population size ($V_N$) is controlled by a forced, seasonally varying growth rate ($r(t)$) and death rate ($d(t)$).
+Total mosquito population size ($V_N$) is controlled by a forced, seasonally varying growth rate ($r(t)$), while the transmission bottleneck is captured with a forced, seasonally varying mortality rate ($d(t)$).
 Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.
@@
 
-The estimated weekly mosquito mortality rate varied seasonally, and was positively correlated with temperature \ref{mortality}.
+The estimated weekly mosquito mortality rate varied seasonally, and was positively correlated with temperature (Figure \ref{mortality}).
 The estimated oscillations in mosquito mortality were more complex than purely sinusoidal, with variations in shape from year to year.
@@
 \caption{{\bf The effect of cross-immunity on effect of control.}
-The posteriod distribution of the maximum proportion of cases prevented by control in each year (left to right, 2009 to 2012), as a function of the period of cross-immunity ($1/\delta$).
+The posterior distribution of the maximum proportion of cases prevented by control in each year (left to right, 2009 to 2012), as a function of the period of cross-immunity ($1/\delta$).
 The shading indicates posterior density, with darkers shading indicating a larger density of posterior samples in that bin.

### 2018-06-07 Manuscript/supplement.tex
More cleaning

@@
 \frac{dV_N}{dt} & = r V_N - \phi_q \tau_t V_N\\
-\frac{dV_E}{dt} &= \lambda IV_S - (\rho_{vt} + d + \phi_q \tau_t)V_E\\
-\frac{dV_I}{dt} &= \rho_{vt} V_E - (d + \phi_q \tau_t) V_I\\
+\frac{dV_{E1}}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E1}\\
+\frac{dV_{E2}}{dt} &= \rho_{v}(t) V_{E1} - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E2}\\
+\frac{dV_{E3}}{dt} &= \rho_{v}(t) V_{E2}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E3}\\
+\frac{dV_{E4}}{dt} &= \rho_{v}(t) V_{E3}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E4}\\\frac{dV_I}{dt} &= \rho_{vt} V_{E4} - (d(t) + \phi_q \tau_t) V_I\\
 \frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
@@
 \gamma & = 3.5\gamma'\\
+d_0' & \sim \text{Gamma(10, 10)}\\
+d_0 & = 1.47 d_0'\\
 \nu_0 & \sim \text{Normal}(0, 0.5)\\
@@
 \\
-\item \textbf{Baselin mosquito mortality rate} ($d_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
+\item \textbf{Baseline mosquito mortality rate} ($d_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
 They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $d_0$.  

### 2018-07-13 Manuscript/ms.tex
Jennifer's easy edits

@@
 Clinton Leach\textsuperscript{1,2*},
-Colleen Webb\textsuperscript{1},
+Jennifer Hoeting\textsuperscript{2}
 Kim Pepin\textsuperscript{3},
@@
 Mevin Hooten\textsuperscript{2},
-Jennifer Hoeting\textsuperscript{2}
+Colleen Webb\textsuperscript{1}
 
@@
 Differential equation models provide a natural way to encode the relevant biological mechanisms, though it is important to also account for sources of uncertainty in those models \cite{Hotelling1927, Wikle2010}.
-In particular, the values of parameters (e.g.
-the length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
+In particular, the values of parameters (e.g., the length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
 The structure of the processes themselves can be uncertain \cite{Ellner1998}, and need to be informed by available, often noisy, data.
@@
 
-Vitoria is a coastal city and the capital of the state of Espirito Santo, Brazil, with a population of 327,801 \cite{vitpop}.
+Vitoria is a coastal city and the capital of the state of Espirito Santo, Brazil, with a population of 327,801 as of 2010 \cite{vitpop}.
 Since 2008, the company Ecovec has monitored mosquito abundance for the city using approximately 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a grid across the city.
 Each trap is checked weekly and the mosquitoes inside counted and identified, with the results sent to a central database that city managers then use to map mosquito infestations and target control.
-These data provide us with 243 weeks (2008 through week 34 of 2012) of total city-wide counts of trapped gravid female \emph{Aedes aegypti}.
+These data provide us with 243 weeks (week 1 of 2008 through week 34 of 2012) of total city-wide counts of trapped gravid female \emph{Aedes aegypti}.
 It is important to note that this time series reflects both natural fluctuations in mosquito density and fluctuations driven by the city's existing mosquito control program.
@@
 Following a latent period ($\frac{1}{\rho}$), exposed humans become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
-Infectious humans recover at rate $\gamma$ and subsequently remain immune for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
+Infectious humans recover at rate $\gamma$ and subsequently remain immune ($R$) for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
 
@@
 \begin{align}
-\epsilon_{\nu i} & \sim \text{Normal}(0, \sigma^2_{\nu})\\
-\epsilon_{ri} & \sim \text{Normal}(0, \sigma^2_r),
+\epsilon_{\nu t} & \sim \text{Normal}(0, \sigma^2_{\nu})\\
+\epsilon_{rt} & \sim \text{Normal}(0, \sigma^2_r),
 \end{align}
-for $i = 1 \dots 243$.
+for $t = 1 \dots 243$.
 These stochastically-forced harmonic oscillators provide a flexible framework for generating smooth seasonal oscillations in the latent mosquito processes \cite{Ramsay2017}.
@@
 
-\subsection*{Control simulations}
+\subsection*{Mosquito control simulations}
 
@@
 The estimated weekly mosquito mortality rate varied seasonally, and was positively correlated with temperature (Figure \ref{mortality}).
-The estimated oscillations in mosquito mortality were more complex than purely sinusoidal, with variations in shape from year to year.
+The oscillations in estimated mosquito mortality were more complex than purely sinusoidal, with variations in shape from year to year.
 The posterior mean of the baseline mortality rate, around which the time series oscillated, was 0.7/week, roughly half the prior mean of 1.47/week.
@@
 the removal of $5\%$ of the adult population in a given week), had the largest effect on the number of dengue cases in a given year when implemented between 30 and 40 weeks prior to the start of that year (Figure \ref{control}).
-Our simulations predicted that control implemented during this time could prevent between $20$ and $30\%$ of the cases in the following year, while control implemented during the target outbreak was much less effective.
+Our simulations predicted that control implemented during this time could prevent between $20$ and $30\%$ of the dengue cases in the following year, while control implemented during the target outbreak was much less effective.
 
@@
 Specifically, we observed a strong negative posterior correlation between the maximum proportion of cases that could be prevented in a given year (i.e., the number of cases prevented by control implemented in the optimal week) and the period of cross-immunity (Figure \ref{immunity}).
-In simulations with a short period of cross-immunity, the susceptible pool replenished quickly, leading to a relatively large standing crop of susceptibles at any time.
+In simulations with a short period of cross-immunity, the susceptible pool replenished quickly, leading to a relatively large pool of susceptibles at any time.
 In simulations with a long period of cross-immunity, the susceptible pool replenished more slowly, leaving the population with relatively few susceptibles.
@@
 \caption{{\bf Estimated latent mosquito mortality rate.}
-A: Time series of mortality rate, showing median posterior estimate (line) and 80\% credible interval (gray band). B: weekly mosquito mortality rate as a function of weekly mean temperature.
+A: Time series of mortality rate, showing median posterior estimate (line) and 80\% credible interval (gray band). B: estimated weekly mosquito mortality rate as a function of weekly mean temperature.
 }
@@
 The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state.
-In fact, this forms the basis for much of the theory of adult control \cite{Burattini2008, Morrison2008, Smith2012}.
-Despite this, our simulations suggest that control is likelyl to be relatively ineffective when implemented during an outbreak, likely due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
+In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
+Despite this, our simulations suggest that control is likely to be relatively ineffective when implemented during an outbreak, likely due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
 Instead, our results suggest that adult mosquito control efforts are most effective at reducing the disease burden within a given year when applied roughly 30 weeks prior to the start of that year, on the downswing of the previous year's epidemic and seasonal mosquito dynamics.
-This stands in contrast to the conclusions of Stoddard \emph{et al.} \cite{Stoddard2014}, who found that the effects of control implemented during the decline phase of an outbreak were difficult to detect in the presence of natually fading transmission intensity.
+Stoddard \emph{et al.} \cite{Stoddard2014}, in their study of the dynamics of dengue in Iquitos, Peru, found that the effects of control implemented during the decline phase of an outbreak were difficult to detect in the presence of natually fading transmission intensity.
 While it may be true that it is difficult to detect the immediate effects of control implemented at this time, our results nonetheless indicate long-term benefits to controlling mosquitos during the decline phase.
@@
 In fact, control implemented at a 30 week lag is effective largely due to the way it interacts with the dynamics of the susceptible pool. 
-Implementing control just as the susceptible population begins to grow (when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) allows for additional immunizing infections as the current outbreak subsides, while also disrupting transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
+Waiting to implement control until the human susceptible population begins to grow (when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) allows for additional immunizing infections to build up as the current outbreak subsides, while also disrupting transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
 The fact that effect of mosquito control is mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g.
@@
 Specifically, we found that increasing the period of cross-immunity decreased the proportion of cases prevented by mosquito control.
-When the period of cross-immmunity was long, the susceptible pool replenished slowly and transmission became limited by the availability of susceptible hosts.
+When the period of cross-immmunity was long (e.g., 200 weeks), the susceptible pool replenished slowly and transmission became limited by the availability of susceptible hosts.
 In the face of this limitation, manipulating mosquito abundance had a relatively small effect.
-Conversely, when the period of cross-immunity was short, susceptibles were abundant and mosquito abundance was instead limiting, in which case reducing mosquito abundance further had a larger effect on transmission.
+Conversely, when the period of cross-immunity was short (e.g., 50 weeks), susceptibles were abundant and mosquito abundance was instead limiting, in which case reducing mosquito abundance further had a larger effect on transmission.
 This interaction between the efficacy of control and human immunity was also demonstrated by ten Bosch \emph{et al.} \cite{TenBosch2016} in their investigation of serotype interactions.
@@
 These control efforts are guided by the MI-Dengue system, which uses the trap-level mosquito surveillance data to target areas of high mosquito infestation for control (including source reduction, larvacide, and adulticide)\cite{Eiras2009}.
-These control efforts may affect our estimates of the mosquito demographic rates, but should not influence our understanding of dengue dynamics, since we are not interested in the factors driving mosquito dynamics, but rather in how the realized mosquito dynamics (whether forced by human intervention or environmental factors) fit into the broader epidemiological picture.
-That said, our control simulations should be interepreted as exploring the effect of an additional pulse of city-wide control on top of the existing control activities.
+These control efforts may affect our estimates of the mosquito demographic rates, but should not influence our understanding of dengue dynamics, since we are not focusing on the factors driving mosquito dynamics, but rather in how the realized mosquito dynamics (whether forced by human intervention or environmental factors) fit into the broader epidemiological picture.
+That said, our mosquito control simulations should be interepreted as exploring the effect of an additional pulse of city-wide control on top of the existing control activities.
 
 We view our simulations of this additional pulse of control primarily as a tool to explore and exploit the long-term feedbacks in the city's epidemiological system, whereas existing control is deployed largely to reduce local (in space and time) disease risk.
-This difference in scale highlights the potential for spatial heterogeneities in disease risk within a city.
+The frequent use of spatially-targeted mosquito control highlights the potential for spatial heterogeneities in disease risk within a city.
 In particular, structured human movement within the city is likely to induce heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009}.

### 2018-07-13 Manuscript/supplement.tex
Jennifer's easy edits

@@
 
-\section*{Full model}
+\section*{Bayesian mechanistic model for dengue and mosquito dynamics}
 
@@
 \item \textbf{Period of cross-immunity} ($T_{\delta}$): Reich \emph{et al.} \cite{Reich2013} estimate a mean of 1.88 years (97 weeks) with a 95\% confidence interval of (0.88, 4.31).
-We specify a gamma prior on the multiplicative scale to have a mean of one and 0.01 and 0.99 quantiles at roughly $0.88/1.88 = 0.47$ and $4.31/1.88 = 2.3$,, respectively:
+We specify a gamma prior on the multiplicative scale to have a mean of one and 0.01 and 0.99 quantiles at roughly $0.88/1.88 = 0.47$ and $4.31/1.88 = 2.3$, respectively:
 \begin{equation}
@@
 In reality, an individual's infectiousness decays with time, but the ODE model assumes infectiousness is constant for the (exponentially distributed) duration of the infection period.    
-To parameterize the model then, we assume a rough equivalence between the probability of infection from a single individual at a given time and the proportion of individuals still infectious at a given time.
+To parameterize the model, we assume a rough equivalence between the probability of infection from a single individual at a given time and the proportion of individuals still infectious at a given time.
 Thus we choose our prior for $\gamma$ to keep the time at which only 5\% of individuals are still infectious between 5 and 7 days, with a mean 5\% threshold of 6 days.
@@
 Since $V_{N}$ has been scaled by the human population size, we center the prior for $\log(V_{N0})$ at $\log(2) = 0.7$ and set the variance to concentrate mass between 0.5 and 5 mosquitoes per person.
-Maybe see Focks (2000) for further justification of range of mosquitoes/person.
 The per-trap mosquito capture probability, $\phi_q$ follows $V_{N0}$ closely and scales the estimated mosquito abundance down to magnitude of the trap counts.
@@
 \item \textbf{Extrinsic incubation period} ($1/\rho_{vt}$): Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito  (the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
-To account for this, we make $1/\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground.
-We use the temperature relationship from the recent meta-analysis of \cite{Chan2012}, which incorporates a large number of records and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
+To account for this, we make $1/\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground (Goiabeiras weather station, station ID: IESVITOR2).
+We use the temperature relationship from the recent meta-analysis of \cite{Chan2012}, which incorporates a large number of studies and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
 To correspond with the assumptions of the differential equation model, we use the relationship obtained under the assumption that the incubation period at any given temperature is exponentially-distributed:
@@
 The black line shows the prior density.
-B: mosquito trap count measurement over-dispersion.
-C: case report measurement over-dispersion.}
+B: mosquito trap count measurement over-dispersion (prior density too low to be visible).
+C: case report measurement over-dispersion (prior density too low to be visible).}
 \end{figure}

### 2018-07-13 Analysis/plots.r

@@ -156,16 +156,16 @@ adult_sum <- adult %>%
 # Plotting
 
 postscript("Manuscript/figures/fig3.eps",
-           width = 5.5, height = 5,
+           width = 4, height = 5,
            family = "ArialMT")
 
 subset(adult_sum, relweek > -53 & relweek < 53) %>% 
   ggplot(aes(relweek, med)) +
   geom_ribbon(aes(ymin = low, ymax = high), fill = "grey70") + 
   geom_line() + 
-  facet_grid(year~.) + 
+  facet_grid(year ~.) + 
   theme_classic() +
-  theme(strip.background = element_blank(), strip.text = element_blank()) + 
+  theme(strip.background = element_blank()) + 
   scale_x_continuous(expand = c(0, 0.5)) + 
   xlab("relative week of control") + 
   ylab("proportion of cases prevented")
@@ -184,7 +184,7 @@ adult_best <- adult %>%
 adult_best$delta <- rep(delta, each = 4)
 
 # Finding median most effective week of control for each year
-bests <- ddply(adult_best, .(year), summarise, best = median(best))
+bests <- ddply(adult_best, .(year), summarise, best = median(best), prevented = median(ratio))
 bests$week <- bests$best + firsts[2:5]
 
 # Summarizing state variables
@@ -213,7 +213,7 @@ subset(stacked, year < 2012) %>%
   facet_grid(variable ~ ., scales = "free_y", switch = "y") +
   theme_classic() +
   scale_color_brewer(palette = "Dark2", type = "qual", guide = F) +
-  scale_x_continuous(expand = c(0, 1), breaks = c(54, 106, 158), labels = c(2009, 2010, 2011)) +
+  scale_x_continuous(expand = c(0, 5), breaks = c(1, 54, 106, 158, 210), labels = c(2008, 2009, 2010, 2011, 2012)) +
   theme(strip.placement = "outside", strip.background = element_blank()) +
   ylab("") +
   xlab("year")
@@ -231,8 +231,8 @@ adult_best %>%
   geom_bin2d(binwidth = c(5, 0.01)) +
   scale_fill_gradient(low = "#E6E6E6", high = "#000000", guide = FALSE) +  facet_grid(.~year) +
   theme_classic() +
[…]

### 2018-07-16 Analysis/plots.r
Adding color to figure 5.

@@ -8,6 +8,7 @@ library(lubridate)
 library(ggplot2)
 library(gridExtra)
 library(patchwork)
+library(truncnorm)
 
 # Loading and processing mosquito and case data
 tseries <- read.csv("Data/Vitoria.data.csv")
@@ -229,7 +230,9 @@ postscript("Manuscript/figures/fig5.eps",
 adult_best %>%
   ggplot(aes(delta, ratio)) +
   geom_bin2d(binwidth = c(5, 0.01)) +
-  scale_fill_gradient(low = "#E6E6E6", high = "#000000", guide = FALSE) +  facet_grid(.~year) +
+  scale_fill_distiller(palette = "Oranges", direction = 1, guide = F) +
+  # scale_fill_gradient(low = "#E6E6E6", high = "#000000", guide = FALSE) +
+  facet_grid(.~year) +
   theme_classic() +
   theme(strip.background = element_blank()) +
   xlab("period of cross-immunity (weeks)") +
@@ -358,8 +361,10 @@ hist(meas$log_phi_q, freq = F, main = "", xlab = expression(log(phi[q])))
 curve(dnorm(x, -13, 0.5), add = T, lwd = 2)
 
 hist(meas$eta_inv_q, freq = F, main = "", xlab = expression(eta[q]))
+curve(dtruncnorm(x, a = 0, b = Inf, mean = 0, sd = 5), add = T)
 
 hist(meas$eta_inv_y, freq = F, main = "", xlab = expression(eta[y]))
+curve(dtruncnorm(x, a = 0, b = Inf, mean = 0, sd = 5), add = T)
 
 dev.off()
 

### 2018-07-16 Manuscript/ms.tex
Rearranging discussion

@@
 The posterior distribution of the maximum proportion of cases prevented by control in each year (left to right, 2009 to 2012), as a function of the period of cross-immunity ($1/\delta$).
-The shading indicates posterior density, with darkers shading indicating a larger density of posterior samples in that bin.
+The shading indicates posterior density, with darker shading indicating a larger density of posterior samples in that bin.
 }
@@
 
-\subsection*{Model validation and interpretation of $d_v$}
+\subsection*{Processes driving effect of mosquito control}
 
@@
 
-Given the central role that mosquito lifespan plays in driving dengue dynamics, we modeled fluctuations in the mosquito mortality rate using a flexible, forced harmonic oscillator.
-This encodes our prior expectation that the mosquito mortality rate should be driven by seasonal environmental forcing (i.e., it should oscillate with a period of one year), while allowing the flexibility to diverge from the simple oscillator and learn from the case reports data.
-The fact that the $\epsilon_{\nu}$ forcing terms are the only source of variability in the transmission process, however, means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.
-Hooker and Ellner \cite{Hooker2015} provide a framework for diagnosing such model misspecification in differential equation models using forcing functions similar to our implementation of the $\epsilon_{\nu}$.
-In that framework, Hooker and Ellner \cite{Hooker2015} estimate nonparametric forcing functions that modify a fitted differential equation model to provide a good fit to the data.
-These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables \cite{Hotelling1927, Hooker2015}.
-We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect our estimated $\epsilon_{\nu}$ forcing terms in the same spirit.
-
-The periodic structure in the posterior means of the $\epsilon_{\nu t}$ suggests that these terms are accounting for more than just noise, and there may be some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
-Following Hooker and Ellner, we can explore whether this process is likely to result from misspecification of the rates of change of the existing state variables, or from missing state variables altogether.
-The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
-These unmodeled components could include additional mosquito population dynamic processes (e.g.
-aquatic stage dynamics, environmental drivers, or control interventions), or epidiemiological processes (e.g.
-multiple circulationg serotypes of the dengue virus).
-Additional work developing and testing new models would be needed to distinguish among these.
-
-Despite this suggestion of potential model misspecification, our estimated mosquito mortalities nonetheless fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
-Moreover, the fact that our estimated mortality rate increased with temperature also broadly agrees with the empirical literature on mosquito survival \cite{Yang2009, Brady2013}.
-This suggests that whatever structure remains unexplained in the forcing terms, the pattern of case reports is still very well described by realistic fluctuations in the mosquito mortality rate.
-
-\subsection*{Processes driving effect of mosquito control}
-
 The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state.
 In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
-Despite this, our simulations suggest that control is likely to be relatively ineffective when implemented during an outbreak, likely due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
+Despite this, our simulations suggest that control is likely to be relatively less effective when implemented during an outbreak, likely due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
 Instead, our results suggest that adult mosquito control efforts are most effective at reducing the disease burden within a given year when applied roughly 30 weeks prior to the start of that year, on the downswing of the previous year's epidemic and seasonal mosquito dynamics.
-Stoddard \emph{et al.} \cite{Stoddard2014}, in their study of the dynamics of dengue in Iquitos, Peru, found that the effects of control implemented during the decline phase of an outbreak were difficult to detect in the presence of natually fading transmission intensity.
-While it may be true that it is difficult to detect the immediate effects of control implemented at this time, our results nonetheless indicate long-term benefits to controlling mosquitos during the decline phase.
+Though, the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicate long-term benefits that become apparent in the following outbreak.
 
@@
 In the absence of, or as a supplement to, such data, mechanistic models like the one developed here, or so-called TSIR frameworks that reconstruct the dynamics of the susceptible class \cite{Finkenstadt2000, Reich2013}, need to be further developed to better inform mosquito control efforts.
+
+\subsection*{Model validation and interpretation of $d_v$}
+
+Given the central role that mosquito lifespan plays in driving dengue dynamics, we modeled fluctuations in the mosquito mortality rate using a flexible, forced harmonic oscillator.
+This encodes our prior expectation that the mosquito mortality rate should be driven by seasonal environmental forcing (i.e., it should oscillate with a period of one year), while allowing the flexibility to diverge from the simple oscillator and learn from the case reports data.
+The fact that the $\epsilon_{\nu}$ forcing terms are the only source of variability in the transmission process, however, means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.
+Hooker and Ellner \cite{Hooker2015} provide a framework for diagnosing such model misspecification in differential equation models using forcing functions similar to our implementation of the $\epsilon_{\nu}$.
+In that framework, Hooker and Ellner \cite{Hooker2015} estimate nonparametric forcing functions that modify a fitted differential equation model to provide a good fit to the data.
+These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables \cite{Hotelling1927, Hooker2015}.
+We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect our estimated $\epsilon_{\nu}$ forcing terms in the same spirit.
+
+The periodic structure in the posterior means of the $\epsilon_{\nu t}$ suggests that these terms are accounting for more than just noise, and there may be some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
+Following Hooker and Ellner, we can explore whether this process is likely to result from misspecification of the rates of change of the existing state variables (indicated by a dependence of $\epsilon_{\nu t}$ on other state variables), or from missing state variables altogether (indicated by an additional dependence of $\epsilon_{\nu t}$ on its own lagged values).
+The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
+These unmodeled components could include additional mosquito population dynamic processes (e.g.
+aquatic stage dynamics, environmental drivers, or control interventions), or epidiemiological processes (e.g.
+multiple circulationg serotypes of the dengue virus).
+
+Despite this suggestion of potential model misspecification, our estimated mosquito mortalities nonetheless fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
+Moreover, the fact that our estimated mortality rate increased with temperature also broadly agrees with the empirical literature on mosquito survival \cite{Yang2009, Brady2013}.
+This suggests that whatever structure remains unexplained in the forcing terms, the pattern of case reports is still very well described by realistic fluctuations in the mosquito mortality rate.
 

### 2018-07-18 Manuscript/ms.tex
Mevin edits

@@
 Alvaro Eiras\textsuperscript{4},
-Mevin Hooten\textsuperscript{2},
+Mevin Hooten\textsuperscript{5,6,2},
 Colleen Webb\textsuperscript{1}
@@
 \textbf{4} Departamento de Parasitologia, Universidade Federal de Minas Gerais, Belo Horizonte, MG, Brazil
+\\
+\textbf{5} U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit
+\\
+\textbf{6} Department of Fish, Wildlife, and Conservation Biology, Colorado State University
 \bigskip
@@
 We developed a differential equation model of dengue transmission and embedded it in a Bayesian hierarchical framework that allowed us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vitoria, Brazil.
-We used the fitted model to explore the effect of adult mosquito control on the disease burden in each year, while taking advantage of the Bayesian framework to carry through uncertainty from the posterior distributions of model parameters and mosquito demographic rates.
-We found that control deployed between 30 and 40 weeks prior to the start of the year was most effective at preventing cases in that year. 
+We used the fitted model to explore the effect of adult mosquito control on the disease burden in each year.
+We found that mosquito control deployed between 30 and 40 weeks prior to the start of the year was most effective at preventing cases in that year. 
 This timing generally coincided with a local minimum in the size of the susceptible pool.
@@
 % Author Summary not valid for PLOS ONE submissions.   
-%\section*{Author summary}
+% \section*{Author summary}
 
@@
 % Use "Eq" instead of "Equation" for equation citations.
+
 \section*{Introduction}
@@
 Though there have been documented successes, there is limited evidence for the long-term sustainability and effectiveness of these control programs \cite{Morrison2008}.
-Because of this, there is a growing recognition that effective control needs to be guided by high quality vector surveillance, together with quantitative tools that synthesize vector surveillance with clinical surveillance, account for local epidemiology, and connect easily to local decision making \cite{Morrison2008, Scott2010b}.
+As a result, there is a growing recognition that effective control needs to be guided by high quality vector surveillance, together with quantitative tools that synthesize vector surveillance with clinical surveillance, account for local epidemiology, and facilitate local decision making \cite{Morrison2008, Scott2010b}.
 Moreover, mosquito control needs to be guided by an understanding of the link between mosquito abundance and disease risk so that the mosquitoes most responsible for transmission can be targeted \cite{Scott2010a, Scott2010b}.
@@
 
-Differential equation models provide a natural way to encode the relevant biological mechanisms, though it is important to also account for sources of uncertainty in those models \cite{Hotelling1927, Wikle2010}.
+Differential equation models provide a natural way to encode the relevant biological mechanisms, although it is important to also account for sources of uncertainty in those models \cite{Hotelling1927, Wikle2010}.
 In particular, the values of parameters (e.g., the length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
@@
 
-In this paper, we aim to integrate these elements -- a detailed mechanistic model of dengue transmission with a full Bayesian accounting of uncertainty -- in order to better understand the interplay of forces governing dengue dynamics and their interaction with potential vector control interventions.
+In this paper, we aim to integrate these elements -- a detailed mechanistic model of dengue transmission with a full Bayesian accounting of uncertainty --  to better understand the interplay of forces governing dengue dynamics and their interaction with potential vector control interventions.
 We apply this framework to clinical and entomological surveillance data from the city of Vitoria, Brazil.
@@
 Each trap is checked weekly and the mosquitoes inside counted and identified, with the results sent to a central database that city managers then use to map mosquito infestations and target control.
-These data provide us with 243 weeks (week 1 of 2008 through week 34 of 2012) of total city-wide counts of trapped gravid female \emph{Aedes aegypti}.
+These data comprise 243 weeks (week 1 of 2008 through week 34 of 2012) of total city-wide counts of trapped gravid female \emph{Aedes aegypti}.
 It is important to note that this time series reflects both natural fluctuations in mosquito density and fluctuations driven by the city's existing mosquito control program.
@@
 As this cross-immunity wanes, antibodies from the previous infection can result in antibody-dependent enhancement (ADE), wherein human hosts are more susceptible to infection with the other serotypes and more likely to develop severe symptoms (i.e., dengue haemorrhagic fever or dengue shock syndrome)\cite{Wearing2006}.
-The strength and duration of these different inter-serotype interactions are not well understood, though different models suggest that temporary cross-immunity alone (without ADE) is sufficient to reproduce observed multi-annual dynamics in Thailand \cite{Wearing2006,Reich2013}.
+The strength and duration of these different inter-serotype interactions are not well understood, although different models suggest that temporary cross-immunity alone (without ADE) is sufficient to reproduce observed multi-annual dynamics in Thailand \cite{Wearing2006,Reich2013}.
 Similarly, Aguiar \emph{et al.} \cite{Aguiar2013} found that capturing primary and secondary infections and the period of cross-immunity is critical, but that explicitly including all four serotypes (at great cost to model complexity) does not perform better than a two serotype model.
@@
 
-The differential equations governing the human population are then given as:
+We specified the differential equations governing the human population as:
 \begin{align} 
@@
 Given samples from the posterior distribution as obtained above, we simulated the effects of a single pulse of control applied in each week of the first four years of the time series.
-Since the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in future years.
+Because the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in future years.
 To obtain the posterior distribution of the proportion of cases prevented in calendar year $j$ by implementing control in week $i$, for each posterior sample $k$, we simulated the dynamics resulting from reducing the mosquito population by $5\%$ at the beginning of week $i$ (affecting susceptible, exposed, and infectious mosquitoes equally, and leaving the estimated mosquito birth and mortality rates unchanged) .
@@
 The estimated posterior median case reports explained 90\% of the variation in the observed time series, while the posterior median mosquito trap counts explained 46\% of the variation in the observed time series.
-In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 6 -8).
+In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 6 - 8).
 The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that the Vitoria data contained little additional information about these parameters (Supplemental Figure 3).
@@
 The posterior mean of the baseline mortality rate, around which the time series oscillated, was 0.7/week, roughly half the prior mean of 1.47/week.
-The posterior means of the $\epsilon_{\nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (Supplementary Figure 1), though the full posterior distribution of each $\epsilon_{\nu t}$ overlapped zero.
+The marginal posterior means of the $\epsilon_{\nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (Supplementary Figure 1), although the marginal posterior distribution of each $\epsilon_{\nu t}$ overlapped zero.
 In addition, the standard deviation of the mortality forcing terms was small relative to the weak prior ($E(\sigma_{\nu}|\mathbf{y}) = 0.01$, Supplemental Figure 2).
@@
 In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
-Despite this, our simulations suggest that control is likely to be relatively less effective when implemented during an outbreak, likely due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
+Despite this, our simulations suggested that control was likely to be relatively less effective when implemented during an outbreak, due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
 Instead, our results suggest that adult mosquito control efforts are most effective at reducing the disease burden within a given year when applied roughly 30 weeks prior to the start of that year, on the downswing of the previous year's epidemic and seasonal mosquito dynamics.
-Though, the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicate long-term benefits that become apparent in the following outbreak.
+While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicate long-term benefits that become apparent in the following outbreak.
 
-The mismatch between the timing of the most effective control and the peak in mosquito abundance suggets that mosquito abundance alone is an insufficient predictor of the effect of mosquito control, at least at the city-scale.
-Instead, the effectiveness of mosquito control depends in large part on the intrinsic long-term feedbacks and lags driven by the dynamics of temporary human immunity.
-In fact, control implemented at a 30 week lag is effective largely due to the way it interacts with the dynamics of the susceptible pool. 
+The mismatch between the timing of the most effective control and the peak in mosquito abundance suggets that mosquito abundance alone was an insufficient predictor of the effect of mosquito control, at least at the city-scale.
+Instead, the effectiveness of mosquito control depended, in large part, on the intrinsic long-term feedbacks and lags driven by the dynamics of temporary human immunity.
+In fact, control implemented at a 30 week lag was effective largely due to the way it interacted with the dynamics of the susceptible pool. 
 Waiting to implement control until the human susceptible population begins to grow (when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) allows for additional immunizing infections to build up as the current outbreak subsides, while also disrupting transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
-The fact that effect of mosquito control is mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g.
-cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities.
+The fact that effect of mosquito control was mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g., cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities.
 
 The Bayesian framework allows for a full accounting of uncertainty across the data, process, and parameter levels of our model \cite{Berliner1996}.
-This then allows us to carry this uncertainty through to our simulations of control interventions \cite{Elderd2006}.
+We then carry this uncertainty through to our simulations of control interventions \cite{Elderd2006}.
 We found that there can be substantial uncertainty in the proportion of cases prevented by a control intervention (i.e., the width of the ribbons in Figure \ref{control}).
@@
 The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
-These unmodeled components could include additional mosquito population dynamic processes (e.g.
-aquatic stage dynamics, environmental drivers, or control interventions), or epidiemiological processes (e.g.
-multiple circulationg serotypes of the dengue virus).
+These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control interventions), or epidiemiological processes (e.g., multiple circulating serotypes of the dengue virus).
 
@@
 Moreover, the fact that our estimated mortality rate increased with temperature also broadly agrees with the empirical literature on mosquito survival \cite{Yang2009, Brady2013}.
-This suggests that whatever structure remains unexplained in the forcing terms, the pattern of case reports is still very well described by realistic fluctuations in the mosquito mortality rate.
+This suggests that regardless of unexplained structure in the forcing terms, the pattern of case reports is still very well described by realistic fluctuations in the mosquito mortality rate.
 
@@
 
-The case data to which we fit the model represent reports of "dengue-like illness", without laboratory confirmation, and as such could include cases of other diseases with similar symptoms (e.g.
-chikungunya or Zika).
+The case data to which we fit the model represent reports of ``dengue-like illness,'' without laboratory confirmation, and as such could include cases of other diseases with similar symptoms (e.g., chikungunya or Zika).
 However, neither chikungunya nor Zika had emerged as substantial public health threats in Brazil by the end of our time series in late 2012 \cite{PAHO2014, PAHO2015}.
@@
 These control efforts are guided by the MI-Dengue system, which uses the trap-level mosquito surveillance data to target areas of high mosquito infestation for control (including source reduction, larvacide, and adulticide)\cite{Eiras2009}.
-These control efforts may affect our estimates of the mosquito demographic rates, but should not influence our understanding of dengue dynamics, since we are not focusing on the factors driving mosquito dynamics, but rather in how the realized mosquito dynamics (whether forced by human intervention or environmental factors) fit into the broader epidemiological picture.
-That said, our mosquito control simulations should be interepreted as exploring the effect of an additional pulse of city-wide control on top of the existing control activities.
+These control efforts may affect our estimates of the mosquito demographic rates, but should not influence our understanding of dengue dynamics.
+We are focused not on the factors driving mosquito dynamics, but rather on how the realized mosquito dynamics (whether forced by human intervention or environmental factors) fit into the broader epidemiological picture.
+However, our mosquito control simulations should be interepreted as exploring the effect of an additional pulse of city-wide control in addition to the existing control activities.
 
@@
 In addition, spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
-Though our model does not account for these heterogeneities, it is nontheless able to capture the city-wide dynamics reasonably well, suggesting that there may be sufficient mixing to appear homogeneous at the city scale.
-Further, though our mechanistic model may be able to suggest when a city-wide intervention is likely to be effective, in order to make the best use of limited resources, spatial prioritization may still be necessary.
+Although our model does not account for these heterogeneities, it is nontheless able to capture the city-wide dynamics reasonably well, suggesting that there may be sufficient mixing to appear homogeneous at the city scale.
+Further, although our mechanistic model may be able to suggest when a city-wide intervention is likely to be effective, to make the best use of limited resources, spatial prioritization may still be necessary.
 Our results emphasize that work to develop the tools necessary for this prioritization should focus not just on prioritization based on the distribution of mosquito abundance, but also on the distribution of human immunity.
@@
 
-Thanks!
+CBL was supported by the NSF Graduate Research Fellowship Program under grant DGE-1321845.
+Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.
 

### 2019-02-05 Manuscript/ms.tex
Kim edits

@@
 Our ability to effectively prevent the transmission of the dengue virus through targeted control of its vector, \emph{Aedes aegypti}, depends critically on our understanding of the link between mosquito abundance and human disease risk.
-The mosquito and clinical surveillance data necessary to elucidate this link are widely collected, but have yet to be coupled with a modeling framework that accounts for the complex non-linear mechanisms involved in transmission.
-Most critical are the bottleneck in transmission imposed by mosquito lifespan relative to the virus' incubation period, and the dynamics of human immunity.
+Mosquito and clinical surveillance data are widely collected, but linking them requires a modeling framework that accounts for the complex non-linear mechanisms involved in transmission.
+Most critical are the bottleneck in transmission imposed by mosquito lifespan relative to the virus' extrinsic incubation period, and the dynamics of human immunity.
 We developed a differential equation model of dengue transmission and embedded it in a Bayesian hierarchical framework that allowed us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vitoria, Brazil.
@@
 
-Many of the attempts to establish this link have found a weak relationship between mosquito surveillance and incidence of dengue fever \cite{Bowman2014, Pepin2015, Cromwell2017}.
+Many of the attempts to establish this link have found a weak relationship between mosquito abundance indices and incidence of disease in humans \cite{Bowman2014, Pepin2015, Cromwell2017}.
 However, these attempts often do not account for the complex, non-linear interactions that mediate the relationship between mosquito abundance and human disease.
@@
 where $\boldsymbol{\theta}$ is a vector of all the model parameters and initial conditions, and $\mathcal{M}(\boldsymbol{\epsilon_r}, \boldsymbol{\epsilon_{\nu}},\boldsymbol{\theta})$ represents the (numeric) solution to the differential equation as a function of $\boldsymbol{\theta}$ and the weekly stochastic forcing terms ($\boldsymbol{\epsilon_{r}}, \boldsymbol{\epsilon_{\nu}}$).
-Sampling from the posterior distribution of the parameters and forcing terms in this model is difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations induced by the nonlinearity of the the differential equation model \cite{Reilly2005, Girolami2008, Calderhead2011}.
+Sampling from the posterior distribution of the parameters and forcing terms in this model is difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations induced by the nonlinearity of the differential equation model \cite{Reilly2005, Girolami2008, Calderhead2011}.
 In addition, the computational burden of numerically solving the differential equation model at each iteration means that sampling efficiency is crucial.
@@
 We ran 3 chains with different starting values for 10,000 iterations each, discarding the first 5,000 as burn-in.
+Convergence diagnostics and mixing were evaluated using the shinystan package \cite{shinystan}.
 Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.  
@@
 In simulations with a long period of cross-immunity, the susceptible pool replenished more slowly, leaving the population with relatively few susceptibles.
-The negative correlation between the effectiveness of control and the period of cross-immunity thus suggests that mosquito control becomes less effective as the availability of susceptibles  becomes more limiting.
+The negative correlation between the effectiveness of control and the period of cross-immunity thus suggests that mosquito control becomes less effective as the overall availability of susceptibles  becomes more limiting.
 
@@
 This interaction between the efficacy of control and human immunity was also demonstrated by ten Bosch \emph{et al.} \cite{TenBosch2016} in their investigation of serotype interactions.
-Specifically, they found that adding cross-immunity to models with antibody dependent enhancement increased the resistance to control (as implemented by reducing the transmission rate).
+Specifically, they found that adding cross-immunity to models with antibody-dependent enhancement increased the resistance to control (as implemented by reducing the transmission rate).
 
@@
 
-In addition, it is important to note that that the data to which we fit our model implicitly reflect the control efforts already enacted by the city.
+In addition, it is important to note that the data to which we fit our model implicitly reflect the control efforts already enacted by the city.
 These control efforts are guided by the MI-Dengue system, which uses the trap-level mosquito surveillance data to target areas of high mosquito infestation for control (including source reduction, larvacide, and adulticide)\cite{Eiras2009}.

### 2019-02-08 Manuscript/ms.tex
Cleaning up some phrasing

@@
 
-The importance of intrinsic nonlinearities, potentially alongside seasonality and stochastic forcing \cite{Ellner1998, Koelle2004, Grenfell2002}, in governing human disease risk highlight the need to integrate mechanistic modeling into the quantitative tools used to understand the effects of control interventions.
+The importance of intrinsic nonlinearities, potentially alongside seasonality and stochastic forcing \cite{Ellner1998, Koelle2004, Grenfell2002}, in governing human disease risk highlights the need to integrate mechanistic modeling into the quantitative tools used to understand the effects of control interventions.
 Such mechanistic models can often perform better than complex autoregressive statistical models in describing and forecasting population dynamics \cite{Reilly2005}.
@@
 In particular, the values of parameters (e.g., the length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
-The structure of the processes themselves can be uncertain \cite{Ellner1998}, and need to be informed by available, often noisy, data.
+The structure of the processes themselves can be uncertain \cite{Ellner1998}, and needs to be informed by available, often noisy, data.
 Bayesian hierarchical modeling provides a coherent framework to account for and integrate this uncertainty across the three levels of the model (data, process, and parameters \cite{Berliner1996, Cressie2009}).
@@
 These data allow us to estimate a latent time series of mosquito mortality rates that modulate the transmission process and link mosquito abundance to human disease.  
-We then use the fitted model to explore how perturbations to the mosquito population propagate and interact with the complex epidemiology and nonlinearities involved in dengue transmission.
+We then use the fitted model to explore how perturbations to the mosquito population propagate and interact with the nonlinearities of dengue transmission in order to better inform mosquito control efforts.
 
@@
 Explicitly accounting for interactions between serotypes, even only two of the four, leads to a large and complex mechanistic model.
-Moreover, because dengue case reports do not include information on serotype, we do not have enough information to inform the dynamics of individual serotypes.
-As such, we develop an SEIRS compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010} which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
+Moreover, because dengue case reports do not identify serotype, there is not enough information in the data to inform the dynamics of individual serotypes.
+As such, we developed an SEIRS compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010} which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
 In this framework, susceptible humans ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
@@
 Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_E$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
-Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implement a gamma-distributed incubation period by chaining multiple exposed classes together (taking advantage of the fact that a gamma-distributed random variable can be generated through the sum of a collection of exponential random variables with the same rate parameter) \cite{Lloyd2001}.
+Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implemented a gamma-distributed incubation period by chaining multiple exposed classes together (taking advantage of the fact that a gamma-distributed random variable can be generated through the sum of a collection of exponential random variables with the same rate parameter) \cite{Lloyd2001}.
 Total mosquito population size ($V_N$) is controlled by a forced, seasonally varying growth rate ($r(t)$), while the transmission bottleneck is captured with a forced, seasonally varying mortality rate ($d(t)$).
@@
 
-We model the centered and log-transformed mosquito mortality rate ($\nu$) and the per-capita mosquito growth rate ($r$) as forced harmonic oscillators with natural periods of one year:
+We modeled the centered and log-transformed mosquito mortality rate ($\nu$) and the per-capita mosquito growth rate ($r$) as forced harmonic oscillators with natural periods of one year:
 \begin{align}
@@
 
-To connect the differential equation model to the observed case reports, we add an extra state, $C$, that collects the cumulative number of transitions from the exposed to infectious class (assuming that case reporting coincides with the onset of symptoms).
-We then model the number of new cases reported in week $t$ ($y_t$) as:
+To connect the differential equation model to the observed case reports, we added an extra state, $C$, that collects the cumulative number of transitions from the exposed to infectious class (assuming that case reporting coincides with the onset of symptoms).
+We then modeled the number of new cases reported in week $t$ ($y_t$) as:
 \begin{equation}
@@
 
-We similarly model the number of mosquitoes trapped in week $t$ ($q_t$) as:
+We similarly modeled the number of mosquitoes trapped in week $t$ ($q_t$) as:
 \begin{equation}
@@
 Several of the parameters in this model are assumed to be fixed and known (Table 1).
-The human population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
-To maintain identifiability, the transmission rate ($\lambda$) and case reporting probability ($\phi_y$) are also fixed at literature values.
-Lastly, the extrinsic incubation period in mosquitoes is modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground.
+The human population size and average life span (which we use to parameterize the birth/death rate) for Vitoria were taken from the 2010 census.
+To maintain identifiability, the transmission rate ($\lambda$) and case reporting probability ($\phi_y$) were also fixed at literature values.
+Lastly, the extrinsic incubation period in mosquitoes was modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground.
 
@@
 \begin{center}
-\caption{Model parameters and their values.  Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the prior mean.}
+\caption{Model parameters and their values.
+Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the prior mean.
+See Supplemental Material for a full description of all priors.}
 \begin{tabular}{llll}
@@
 The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that the Vitoria data contained little additional information about these parameters (Supplemental Figure 3).
-The estimated latent period in a human host ($1/\rho$, the expected time it takes for an exposed human to become infectious to biting mosquitoes) is influenced more strongly by the data, with a posterior mean of 1.53 weeks compared to a prior mean of 0.87 weeks.
+The estimated latent period in a human host ($1/\rho$, the expected time it takes for an exposed human to become infectious to biting mosquitoes) was influenced more strongly by the data, with a posterior mean of 1.53 weeks compared to a prior mean of 0.87 weeks.
 
@@
 In simulations with a long period of cross-immunity, the susceptible pool replenished more slowly, leaving the population with relatively few susceptibles.
-The negative correlation between the effectiveness of control and the period of cross-immunity thus suggests that mosquito control becomes less effective as the overall availability of susceptibles  becomes more limiting.
+The negative correlation between the effectiveness of control and the period of cross-immunity thus suggests that mosquito control was less effective when the overall availability of susceptibles was more limiting.
 
@@
 
-Infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are often driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
-In the case of dengue fever, these processes can obscure the relationship between mosquito abundance and human disease.
-We found that a mechanistic model with temporary human immunity and seasonal variation in the mosquito mortality rate was able to effectively fit case reports and mosquito surveillance data from Vitoria, Brazil.
+The dynamics of dengue fever, like those of many infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
+In the dengue system, these drivers include waning human immunity, the race between the extrinsic incubation period of the virus and mosquito lifespan, and the seasonal and stochastic variation in these rates.
+These processes are particularly important for establishing the relationship between mosquito abundance -- and efforts to control that abundance -- and human disease.
 
@@
 In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
-Despite this, our simulations suggested that control was likely to be relatively less effective when implemented during an outbreak, due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
-Instead, our results suggest that adult mosquito control efforts are most effective at reducing the disease burden within a given year when applied roughly 30 weeks prior to the start of that year, on the downswing of the previous year's epidemic and seasonal mosquito dynamics.
-While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicate long-term benefits that become apparent in the following outbreak.
+Despite this, our simulations suggest that control was relatively less effective when implemented during an outbreak, likely due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
+Instead, we found that adult mosquito control efforts were most effective at reducing the disease burden within a given year when applied roughly 30 weeks prior to the start of that year, on the downswing of the previous year's epidemic and seasonal mosquito dynamics.
+While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicated long-term benefits that carried through to the following outbreak.
 
@@
 In fact, control implemented at a 30 week lag was effective largely due to the way it interacted with the dynamics of the susceptible pool. 
-Waiting to implement control until the human susceptible population begins to grow (when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) allows for additional immunizing infections to build up as the current outbreak subsides, while also disrupting transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
+Implementing control just as the human susceptible population begins to grow (when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) allows for additional immunizing infections to build up as the current outbreak subsides, while also disrupting transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
 The fact that effect of mosquito control was mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g., cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities.
 
-The Bayesian framework allows for a full accounting of uncertainty across the data, process, and parameter levels of our model \cite{Berliner1996}.
-We then carry this uncertainty through to our simulations of control interventions \cite{Elderd2006}.
+The Bayesian framework allowed for a full accounting of uncertainty across the data, process, and parameter levels of our model \cite{Berliner1996}.
+We then carried this uncertainty through to our simulations of control interventions \cite{Elderd2006}.
 We found that there can be substantial uncertainty in the proportion of cases prevented by a control intervention (i.e., the width of the ribbons in Figure \ref{control}).
@@
 Given the central role that mosquito lifespan plays in driving dengue dynamics, we modeled fluctuations in the mosquito mortality rate using a flexible, forced harmonic oscillator.
-This encodes our prior expectation that the mosquito mortality rate should be driven by seasonal environmental forcing (i.e., it should oscillate with a period of one year), while allowing the flexibility to diverge from the simple oscillator and learn from the case reports data.
-The fact that the $\epsilon_{\nu}$ forcing terms are the only source of variability in the transmission process, however, means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.
+This encoded our prior expectation that the mosquito mortality rate should be driven by seasonal environmental forcing (i.e., it should oscillate with a period of one year), while allowing the flexibility to diverge from the simple oscillator and learn from the case reports data.
+The fact that the $\epsilon_{\nu}$ forcing terms were the only source of variability in the transmission process, however, means that the estimated mosquito mortality time series could have been soaking up other sources of stochasticity or model misspecification.
 Hooker and Ellner \cite{Hooker2015} provide a framework for diagnosing such model misspecification in differential equation models using forcing functions similar to our implementation of the $\epsilon_{\nu}$.
@@
 
-The periodic structure in the posterior means of the $\epsilon_{\nu t}$ suggests that these terms are accounting for more than just noise, and there may be some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
+The periodic structure in the posterior means of the $\epsilon_{\nu t}$ suggests that these terms were accounting for more than just noise, and there may have been some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
 Following Hooker and Ellner, we can explore whether this process is likely to result from misspecification of the rates of change of the existing state variables (indicated by a dependence of $\epsilon_{\nu t}$ on other state variables), or from missing state variables altogether (indicated by an additional dependence of $\epsilon_{\nu t}$ on its own lagged values).
@@
 Moreover, the fact that our estimated mortality rate increased with temperature also broadly agrees with the empirical literature on mosquito survival \cite{Yang2009, Brady2013}.
-This suggests that regardless of unexplained structure in the forcing terms, the pattern of case reports is still very well described by realistic fluctuations in the mosquito mortality rate.
+This suggests that regardless of unexplained structure in the forcing terms, the pattern of case reports was still very well described by realistic fluctuations in the mosquito mortality rate.
 
@@
 In addition, spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
-Although our model does not account for these heterogeneities, it is nontheless able to capture the city-wide dynamics reasonably well, suggesting that there may be sufficient mixing to appear homogeneous at the city scale.
+Although our model does not account for these heterogeneities, it was nontheless able to capture the city-wide dynamics reasonably well, suggesting that there may be sufficient mixing to appear homogeneous at the city scale.
 Further, although our mechanistic model may be able to suggest when a city-wide intervention is likely to be effective, to make the best use of limited resources, spatial prioritization may still be necessary.

### 2019-02-26 Manuscript/ms.tex
Small changes

@@
 Dengue fever is a massive global public health burden, with millions of cases per year \cite{Bhatt2013}.  
-Because the dengue virus (DENV) is vectored by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
+Because the dengue virus (DENV) is transmitted by the mosquito \textit{Aedes aegypti}, dengue fever is prevented primarily through mosquito control programs \cite{Achee2015}.
 Though there have been documented successes, there is limited evidence for the long-term sustainability and effectiveness of these control programs \cite{Morrison2008}.
@@
 CBL was supported by the NSF Graduate Research Fellowship Program under grant DGE-1321845.
+AEE was supported by the Brazilian National Council for Scientific and Technological Development (CNPq).
 Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

### 2019-02-28 Manuscript/ms.tex
Adding author summary

@@
 % Author Summary not valid for PLOS ONE submissions.   
-% \section*{Author summary}
+\section*{Author summary}
+
+The contribution of the mosquito vector \emph{Aedes aegypti} to the spread of dengue fever depends not only on their abundance, but also on the likelihood of an exposed mosquito living long enough to incubate the dengue virus and subsequently it transmit to a susceptible human host.
+We developed a mechanistic model that accounts for the role of this process in the dynamics of dengue fever and fit the model to a time series of human case reports and mosquito trap counts from the city of Vitoria, Brazil.
+We then used this fitted model to simulate the effect of mosquito control implemented at different times of the year and found that  mosquito control leads to the largest reduction in human dengue cases during a given year when applied at the local minimum of the previous year’s human immunity trajectory.
+Further, the effectiveness of mosquito control was strongly correlated with how quickly previously infected humans become susceptible again, with mosquito control making the largest impact when immunity wanes quickly.
+Thus, the effects of mosquito control efforts are tightly coupled with the dynamics of human immunity and the long-term lags and feedbacks generated by the epidemiological system.
 

### 2019-03-14 Manuscript/ms.tex
Changing some phrasing and adding citations

@@
 We developed a differential equation model of dengue transmission and embedded it in a Bayesian hierarchical framework that allowed us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vitoria, Brazil.
-We used the fitted model to explore the effect of adult mosquito control on the disease burden in each year.
-We found that mosquito control deployed between 30 and 40 weeks prior to the start of the year was most effective at preventing cases in that year. 
-This timing generally coincided with a local minimum in the size of the susceptible pool.
-Uncertainty in the effectiveness of control interventions was driven largely by uncertainty in the period of cross-immunity, with longer periods of cross-immunity leading to less effective control.
-Understanding the dynamics of temporary human immunity is thus critical for understanding the effect of mosquito control efforts on human disease.
+We used the fitted model to explore how the timing of a pulse of adult mosquito control influences its effect on the human disease burden in the following year.
+We found that control implemented roughly 30 to 40 weeks prior to the start of the year produced the largest reduction in predicted human case reports for that year.
+This highlights the potential long-term utility of broad, off-peak-season mosquito control in addition to existing, locally targeted within-season efforts.
+Human immunity played an important role in mediating the effect of control, with the  30 to 40 week lag generally aligning with the estimated local minimum in the size of the human susceptible pool.
+Further, uncertainty in the effectiveness of control interventions was driven largely by uncertainty in the period of cross-immunity, with longer periods of cross-immunity leading to less effective control.
+Understanding the dynamics of temporary human immunity is thus critical for understanding how the effects of mosquito control propagate through the dengue system.
 
@@
 the removal of $5\%$ of the adult population in a given week), had the largest effect on the number of dengue cases in a given year when implemented between 30 and 40 weeks prior to the start of that year (Figure \ref{control}).
-Our simulations predicted that control implemented during this time could prevent between $20$ and $30\%$ of the dengue cases in the following year, while control implemented during the target outbreak was much less effective.
+Our simulations predicted that a pulse of control implemented during this time reduced the disease burden in the following year by $20$ to $30\%$, while perturbing mosquito abundance during the target outbreak was much less effective.
 
@@
 In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
-Despite this, our simulations suggest that control was relatively less effective when implemented during an outbreak, likely due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
+Despite this, our simulations suggest that a single pulse of control was relatively less effective when implemented during an outbreak, likely due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
 Instead, we found that adult mosquito control efforts were most effective at reducing the disease burden within a given year when applied roughly 30 weeks prior to the start of that year, on the downswing of the previous year's epidemic and seasonal mosquito dynamics.
-While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicated long-term benefits that carried through to the following outbreak.
+Barsante \emph{et al.} \cite{Barsante2015}, similarly found that control was most effective when applied during the dry season (roughly matching our 30 week lag) than during the rainy season.
+While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicated that disrupting inter-seasonal transmission can be an effective longer-term strategy \cite{Hladish2018}.
 
@@
 
-CBL was supported by the NSF Graduate Research Fellowship Program under grant DGE-1321845.
-AEE was supported by the Brazilian National Council for Scientific and Technological Development (CNPq).
 Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

### 2019-03-27 Manuscript/ms.tex
Updating figure 5 results

@@
 The oscillations in estimated mosquito mortality were more complex than purely sinusoidal, with variations in shape from year to year.
-The posterior mean of the baseline mortality rate, around which the time series oscillated, was 0.7/week, roughly half the prior mean of 1.47/week.
+The posterior distribution of the baseline mortality rate ($d_0$) had a mean of 0.7/week, roughly half the prior mean, and was negatively correlated with the period of cross immunity ($1/\delta$, Figure \ref{immunity}).
+This negative correltaion indicates that similar dynamics could be achieved with either rapidly waning human immunity and short-lived mosquitoes or with long-lasting human immunity and long-lived mosquitoes.
 The marginal posterior means of the $\epsilon_{\nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (Supplementary Figure 1), although the marginal posterior distribution of each $\epsilon_{\nu t}$ overlapped zero.
@@
 In particular, mosquito control seemed to be most effective when applied at or near a local minimum in the number of suceptible humans (Figure \ref{timing}).
-In addition, human immunity, as controlled by the period of cross-immunity ($1/\delta$), was a major driver of uncertainty in the effect of a given control intervention.
-Specifically, we observed a strong negative posterior correlation between the maximum proportion of cases that could be prevented in a given year (i.e., the number of cases prevented by control implemented in the optimal week) and the period of cross-immunity (Figure \ref{immunity}).
-In simulations with a short period of cross-immunity, the susceptible pool replenished quickly, leading to a relatively large pool of susceptibles at any time.
-In simulations with a long period of cross-immunity, the susceptible pool replenished more slowly, leaving the population with relatively few susceptibles.
-The negative correlation between the effectiveness of control and the period of cross-immunity thus suggests that mosquito control was less effective when the overall availability of susceptibles was more limiting.
+In addition, the average effectiveness of optimally timed mosquito control was driven largely by the period of cross-immunity ($1/\delta$) and the mean mosquito mortality rate ($d_0$, Figure \ref{immunity}).
+Specifically, we observed that mosquito control was able to prevent the largest number of human dengue cases, averaged over all years, in simulations with a low period of cross-immunity and a high mosquito mortality rate (Figure \ref{immunity}).
+Thus mosquito control was more effective at reducing disease burden in systems where short mosquito lifespan was already limiting the abundance of infectious mosquitoes and susceptible hosts were relatively abundant.
 
@@
 \begin{figure}[!h]
-\includegraphics[angle = 270, scale = 0.8]{figures/fig5.eps}
-\caption{{\bf The effect of cross-immunity on effect of control.}
-The posterior distribution of the maximum proportion of cases prevented by control in each year (left to right, 2009 to 2012), as a function of the period of cross-immunity ($1/\delta$).
-The shading indicates posterior density, with darker shading indicating a larger density of posterior samples in that bin.
+\includegraphics[angle = 270, scale = 1]{figures/fig5.eps}
+\caption{{\bf The relationship among the period of cross-immunity, mosquito mortality rate, and effectiveness of control.}
+The maximum proportion of cases prevented by control, averaged over all years and all posterior samples within each cell.
+Cells with darker shading indicate combinations of the period of cross-immunity and mosquito mortality that produce a system more amenable to control by adult mosquito removal.
 }

### 2019-03-27 Analysis/plots.r

@@ -183,6 +183,7 @@ adult_best <- adult %>%
         ratio = max(prevented))
 
 adult_best$delta <- rep(delta, each = 4)
+adult_best$dvmu <- rep(dvmu, each = 4)
 
 # Finding median most effective week of control for each year
 bests <- ddply(adult_best, .(year), summarise, best = median(best), prevented = median(ratio))
@@ -224,19 +225,19 @@ dev.off()
 # Figure 5: effect of delta on control =========================================
 
 postscript("Manuscript/figures/fig5.eps",
-           width = 6, height = 4,
+           width = 5, height = 3,
            family = "ArialMT")
 
-adult_best %>%
-  ggplot(aes(delta, ratio)) +
-  geom_bin2d(binwidth = c(5, 0.01)) +
-  scale_fill_distiller(palette = "Oranges", direction = 1, guide = F) +
-  # scale_fill_gradient(low = "#E6E6E6", high = "#000000", guide = FALSE) +
-  facet_grid(.~year) +
+adult_best %>% 
+  ggplot(aes(dvmu, delta, z = ratio)) + 
+  stat_summary_2d(binwidth = c(0.025, 10)) +
+  scale_fill_distiller(palette = "Oranges", direction = 1) +
+  labs(fill = "average \nproportion \nof cases \nprevented") + 
   theme_classic() +
-  theme(strip.background = element_blank()) +
-  xlab("period of cross-immunity (weeks)") +
-  ylab("maximum proportion of cases prevented")
+  theme(strip.background = element_blank(),
+        text = element_text(size = 10)) +
+  ylab("period of cross-immunity (weeks)") +
+  xlab("mosquito mortality rate")
 
 dev.off()
 

### 2019-04-16 Analysis/predict.r
Deleting deprecated code

@@ -1,163 +1,0 @@
-library(rstan)
-library(plyr)
-library(magrittr)
-library(lubridate)
-library(doParallel)
-
-rstan_options(auto_write = TRUE)
-options(mc.cores = parallel::detectCores())
-
-#===============================================================================
-# Loading data and setting up model
-
-tseries <- read.csv("Data/Vitoria.data.csv")
-
-obs <- tapply(tseries$Cases, list(tseries$tot.week), sum, na.rm=TRUE)
-
-# Population size
-pop <- 327801
-
-# Pull out mosquito counts and traps inspected
-q <- tapply(tseries$Mosquitoes, list(tseries$tot.week), sum, na.rm = TRUE)
-tau <- tapply(tseries$Trap, list(tseries$tot.week), sum, na.rm = TRUE)
-
-# Wrangling weather data and computing EIP
-weather <- read.csv("Data/Vitoria.weather.csv") %>% 
-  mutate(date = ymd(BRST), year = year(date), week = week(date))
-
-start <- weather$date[1]
-
-weather <- subset(weather, date < date[1] + weeks(243))
-weather$tot.week <- rep(c(1:243), each = 7)
-
-temp <- ddply(weather, .(tot.week), summarise, temp = mean(Mean.TemperatureC, na.rm = T))
-
-rov <- 7 * exp(0.2 * temp$temp - 8)
-#===============================================================================
-# Fitting to first four years
-
-T_pred <- 0
-T_fit <- 191
-
-dat.stan <- list(T = T_fit,
-                 T_pred = T_pred,
-                 y = head(obs, T_fit),
-                 q = head(q, T_fit),
-                 tau = head(tau, T_fit),
-                 rov = head(rov, T_fit),
-                 control = matrix(1, nrow = 191, ncol = 3),
-                 pop = pop)
-
-inits = list(list(S0 = 0.4,
-                  E0 = 80,
-                  I0 = 40,
-                  log_phi_q = -13,
-                  eta_inv_y = 0.1,
-                  eta_inv_q = 0.1,
-                  ro_c = 0.8,
-                  gamma_c = 1,
-                  delta_c = 0.8,
-                  logNv = 0.7,
-                  dv0 = 0,
-                  rv0 = 0,
-                  sigmadv = 0.2,
-                  sigmarv = 0.2,
-                  eps_dv = rep(0, T_fit),
-                  eps_rv = rep(0, T_fit)),
-             list(S0 = 0.3,
-                  E0 = 100,
-                  I0 = 80,
[…]

### 2019-04-17 Analysis/citystan.r
Better documentation of scripts

@@ -1,8 +1,12 @@
+# Script to fit an SEIRS-SEI dengue transmission model to weekly trap counts
+# and reports of "dengue-like illness" using Stan.
+
 library(rstan)
 library(plyr)
 library(magrittr)
 library(lubridate)
 
+# Setting Stan options
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
 
@@ -29,14 +33,18 @@ weather <- read.csv("Data/Vitoria.weather.csv") %>%
 weather <- subset(weather, date < date[1] + weeks(243))
 weather$tot.week <- rep(c(1:243), each = 7)
 
+# Weekly mean temperature
 covars <- ddply(weather, .(tot.week), summarise, 
                 temp = mean(Mean.TemperatureC, na.rm = T),
                 humid = mean(Mean.Humidity, na.rm = T))
 
+# Computing inverse of extrinsic incubation period
 rov <- 7 * exp(0.2 * covars$temp - 8)
+
 #===============================================================================
 # Running stan
 
+# Wrapping up all the data
 dat.stan <- list(T = 243,
                  T_pred = 0,
                  steps = 7,
@@ -47,6 +55,7 @@ dat.stan <- list(T = 243,
                  control = matrix(1, nrow = 243, ncol = 3),
                  pop = pop)
 
+# Specifying initial conditions for 3 chains
 inits = list(list(S0 = 0.4,
                   E0 = 80,
                   I0 = 40,
@@ -98,6 +107,9 @@ inits = list(list(S0 = 0.4,
                   sigmarv = 0.01,
                   eps_dv = rep(0, 243),
                   eps_rv = rep(0, 243)))
+
+# Running HMC (note non-default options for adapt_delta and max_treedepth, which 
+# help ensure good mixing and avoid divergent transitions)
 
 fit <- stan(file = "Code/gammaeip.stan", 
             data = dat.stan, 

### 2019-04-17 Analysis/control.r

@@ -1,3 +1,5 @@
+# Script to simulate control interventions in fitted dengue transmission model
+
 library(rstan)
 library(doParallel)
 library(plyr)
@@ -50,7 +52,12 @@ extract_sample <- function(x, k){
 model <- stan_model(file = "Code/gammaeip.stan")
 
 #===============================================================================
-# In which week is adult control most effective?
+# Adult control simulations
+
+# Loops over each HMC iteration and simulates a 5% decresase in total mosquito 
+# population size in each of the first 210 weeks of the time series.
+# Summarizes the difference in number of cases observed in each year between the
+# control and uncontrolled time series.
 
 # Setting up parallel
 cl <- makeCluster(10, type = "SOCK")
@@ -60,11 +67,14 @@ registerDoParallel(cl)
 reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "magrittr")) %:% 
   foreach(j = 1:210, .combine = "rbind") %dopar% {
 
+    # Extract parameter values in kth iteration
     init <- list(lapply(samples, extract_sample, k))
 
+    # Set-up control structure
     control <- matrix(1, nrow = 243, ncol = 3)
     control[j, 3] <- 0.95
     
+    # Stan data object
     dat.stan <- list(T = 243,
                      steps = 7,
                      y = data$obs,
@@ -74,6 +84,7 @@ reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "ma
                      control = control,
                      pop = pop)
     
+    # Simulate trajectory with control
     sim <- sampling(model,
                     data = dat.stan, 
                     init = init, 
@@ -82,8 +93,10 @@ reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "ma
                     warmup = 0,
                     algorithm = "Fixed_param")
     
+    # Extract values of the state variables at the beginning of each year
     state <- rstan::extract(sim, "state", permute = T)[[1]][1, ends + 1, ]
     
+    # Assemble results
     data.frame(rep = k,
                control = j,
                year = c(2009, 2010, 2011, 2012),
@@ -96,138 +109,4 @@ reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "ma
 
 stopCluster(cl)
 
-saveRDS(reduction, "Results/adult_control.rds")
-
[…]

### 2019-04-17 Analysis/citystan.r
Updating scripts to match changes to stan files

@@ -46,7 +46,6 @@ rov <- 7 * exp(0.2 * covars$temp - 8)
 
 # Wrapping up all the data
 dat.stan <- list(T = 243,
-                 T_pred = 0,
                  steps = 7,
                  y =data$obs,
                  q = data$q,

### 2019-04-17 Analysis/stepsize.r

@@ -1,3 +1,6 @@
+# Script to explore the numerical consquences of changes the Euler integration
+# step size.
+
 library(rstan)
 library(doParallel)
 library(plyr)
@@ -15,8 +18,7 @@ data <- ddply(tseries, .(tot.week), summarise,
               q = sum(Mosquitoes, na.rm = T),
               tau = sum(Trap, na.rm = T),
               year = unique(Year),
-              week = unique(Week)
-)
+              week = unique(Week))
 
 # Population size
 pop <- 327801
@@ -36,29 +38,31 @@ rov <- 7 * exp(0.2 * covars$temp - 8)
 #===============================================================================
 # Loading mcmc samples
 
-fit <- readRDS("Results/oscillator.rds")
+fit <- readRDS("Results/gamma_eip_dv0.rds")
 
-samples <- rstan::extract(fit)[1:16]
+samples <- rstan::extract(fit)[1:17]
 
 extract_sample <- function(x, k){
   if(length(dim(x)) == 1) return(x[k])
   else return(x[k, ])
 }
 
+# Selecting a sample to run the simulation with
 init <- list(lapply(samples, extract_sample, 1))
 
-model <- stan_model(file = "Code/seirs.stan")
+# Loading Stan code
+model <- stan_model(file = "Code/gammaeip.stan")
 
 #===============================================================================
 # Simulating dynamics with smaller and smaller Euler steps
 
+# Number of Euler steps per week
 steps <- c(7, 14, 28, 56, 112, 224)
 
-# Parallel for-loop over mcmc iterations
+# Loop over each of the step sizes and simulate dynamics
 runs <- foreach(k = 1:6, .packages = c("rstan", "magrittr"), .combine = "rbind") %do% {
   
   dat.stan <- list(T = 243,
-                   T_pred = 0,
                    steps = steps[k],
                    y =data$obs,
                    q = data$q,
@@ -75,13 +79,14 @@ runs <- foreach(k = 1:6, .packages = c("rstan", "magrittr"), .combine = "rbind")
                   warmup = 0,
                   algorithm = "Fixed_param")
   
-  cases <- rstan::extract(sim, "system", permute = T)[[1]][1, , 8]
+  cases <- diff(rstan::extract(sim, "state", permute = T)[[1]][1, , 11])
   
   return(cases)
 }
 
 
 #===============================================================================
+# Compare number of cases per week for the different step sizes
 
 plot(runs[6, ])
 points(runs[5, ], pch = 20, col = "blue")

### 2019-04-17 Analysis/citystan.r
Runtime warning

@@ -107,13 +107,15 @@ inits = list(list(S0 = 0.4,
                   eps_dv = rep(0, 243),
                   eps_rv = rep(0, 243)))
 
-# Running HMC (note non-default options for adapt_delta and max_treedepth, which 
-# help ensure good mixing and avoid divergent transitions)
+# Running HMC
+# Note non-default options for adapt_delta and max_treedepth, which 
+# help ensure good mixing and avoid divergent transitions
+# Also note that this will take on the order of 1 week to run
 
 fit <- stan(file = "Code/gammaeip.stan", 
             data = dat.stan, 
             init = inits, 
-            iter = 2000,
+            iter = 5000,
             chains = 3,
             control = list(adapt_delta = 0.99,
                            max_treedepth = 15))

### 2019-04-17 Manuscript/ms.tex
Linking to github repro in text
Manuscript updates

@@
 Most critical are the bottleneck in transmission imposed by mosquito lifespan relative to the virus' extrinsic incubation period, and the dynamics of human immunity.
-We developed a differential equation model of dengue transmission and embedded it in a Bayesian hierarchical framework that allowed us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vitoria, Brazil.
+We developed a differential equation model of dengue transmission and embedded it in a Bayesian hierarchical framework that allowed us to estimate latent time series of mosquito demographic rates from mosquito trap counts and dengue case reports from the city of Vit\'oria, Brazil.
 We used the fitted model to explore how the timing of a pulse of adult mosquito control influences its effect on the human disease burden in the following year.
@@
 Human immunity played an important role in mediating the effect of control, with the 30 to 40 week lag generally aligning with the estimated local minimum in the size of the human susceptible pool.
-Further, uncertainty in the effectiveness of control interventions was driven largely by uncertainty in the period of cross-immunity, with longer periods of cross-immunity leading to less effective control.
-Understanding the dynamics of temporary human immunity is thus critical for understanding how the effects of mosquito control propagate through the dengue system.
+Further, uncertainty in the effectiveness of control interventions was driven largely by posterior variation in the period of cross-immunity and the average mosquito mortality rate, with shorter periods of cross-immunity and higher mosquito mortality leading to more effective control.
+Understanding the dynamics of human immunity is thus critical for understanding how the effects of mosquito control propagate through the dengue system.
 
@@
 
-The contribution of the mosquito vector \emph{Aedes aegypti} to the spread of dengue fever depends not only on their abundance, but also on the likelihood of an exposed mosquito living long enough to incubate the dengue virus and subsequently it transmit to a susceptible human host.
-We developed a mechanistic model that accounts for the role of this process in the dynamics of dengue fever and fit the model to a time series of human case reports and mosquito trap counts from the city of Vitoria, Brazil.
+The contribution of the mosquito vector \emph{Aedes aegypti} to the spread of dengue fever depends not only on their abundance, but also on the likelihood of an exposed mosquito living long enough to incubate the dengue virus and subsequently transmit it to a susceptible human host.
+We developed a mechanistic model that accounts for the role of this process in the dynamics of dengue fever and fit the model to a time series of human case reports and mosquito trap counts from the city of Vit\'oria, Brazil.
 We then used this fitted model to simulate the effect of mosquito control implemented at different times of the year and found that mosquito control leads to the largest reduction in human dengue cases during a given year when applied at the local minimum of the previous year’s human immunity trajectory.
-Further, the effectiveness of mosquito control was strongly correlated with how quickly previously infected humans become susceptible again, with mosquito control making the largest impact when immunity wanes quickly.
+Further, the effectiveness of mosquito control was strongly correlated with how quickly previously infected humans become susceptible again and with the average mosquito mortality rate.
 Thus, the effects of mosquito control efforts are tightly coupled with the dynamics of human immunity and the long-term lags and feedbacks generated by the epidemiological system.
@@
 In this paper, we aim to integrate these elements -- a detailed mechanistic model of dengue transmission with a full Bayesian accounting of uncertainty --  to better understand the interplay of forces governing dengue dynamics and their interaction with potential vector control interventions.
-We apply this framework to clinical and entomological surveillance data from the city of Vitoria, Brazil.
+We apply this framework to clinical and entomological surveillance data from the city of Vit\'oria, Brazil.
 These data allow us to estimate a latent time series of mosquito mortality rates that modulate the transmission process and link mosquito abundance to human disease.  
@@
 
-Vitoria is a coastal city and the capital of the state of Espirito Santo, Brazil, with a population of 327,801 as of 2010 \cite{vitpop}.
+Vit\'oria is a coastal city and the capital of the state of Esp\'irito Santo, Brazil, with a population of 327,801 as of 2010 \cite{vitpop}.
 Since 2008, the company Ecovec has monitored mosquito abundance for the city using approximately 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a grid across the city.
@@
 To maintain identifiability, the transmission rate ($\lambda$) and case reporting probability ($\phi_y$) were also fixed at literature values.
-Lastly, the extrinsic incubation period in mosquitoes was modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground.
+Lastly, the extrinsic incubation period in mosquitoes was modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground \cite{weather}.
 
@@
 Samples from the posterior distribution were thus generated using HMC implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R \cite{R2016}. 
-We ran 3 chains with different starting values for 10,000 iterations each, discarding the first 5,000 as burn-in.
+We ran 3 chains with different starting values for 5,000 iterations each, discarding the first 2,500 as burn-in.
 Convergence diagnostics and mixing were evaluated using the shinystan package \cite{shinystan}.
 Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.  
-Timesteps as small as $1/8$ of a day were explored and did not qualitatively change our results.
+Timesteps as small as $1/8$ of a day were explored and did not qualitatively change the modeled dynamics.
+Code is available from \href{https://github.com/clint-leach/mosquito-recon}{https://github.com/clint-leach/mosquito-recon}.
 
@@
 In particular, mosquito control seemed to be most effective when applied at or near a local minimum in the number of suceptible humans (Figure \ref{timing}).
-In addition, the average effectiveness of optimally timed mosquito control was driven largely by the period of cross-immunity ($1/\delta$) and the mean mosquito mortality rate ($d_0$, Figure \ref{immunity}).
-Specifically, we observed that mosquito control was able to prevent the largest number of human dengue cases, averaged over all years, in simulations with a low period of cross-immunity and a high mosquito mortality rate (Figure \ref{immunity}).
+In addition, the average effectiveness of mosquito control applied in this optimal window (i.e.
+at 30 weeks prior to the start of the year) was driven largely by the period of cross-immunity ($1/\delta$) and the mean mosquito mortality rate ($d_0$, Figure \ref{immunity}).
+Specifically, we observed that mosquito control was able to prevent the largest number of human dengue cases, averaged over all years, in simulations with a low period of cross-immunity and a high average mosquito mortality rate (Figure \ref{immunity}).
 Thus mosquito control was more effective at reducing disease burden in systems where short mosquito lifespan was already limiting the abundance of infectious mosquitoes and susceptible hosts were relatively abundant.
@@
 \caption{{\bf Timing of optimal control relative to disease and mosquito dynamics.}
-The vertical gray lines indicate the median week at which control was most effective in preventing cases in the following calendar year.
+The vertical gray lines indicate the week at which control generated the largest median number of prevented cases in the following calendar year.
 A: posterior median estimate of new dengue cases per week.
@@
 \includegraphics[angle = 270, scale = 1]{figures/fig5.eps}
-\caption{{\bf The relationship among the period of cross-immunity, mosquito mortality rate, and effectiveness of control.}
-The maximum proportion of cases prevented by control, averaged over all years and all posterior samples within each cell.
-Cells with darker shading indicate combinations of the period of cross-immunity and mosquito mortality that produce a system more amenable to control by adult mosquito removal.
+\caption{{\bf The relationship among the period of cross-immunity, average mosquito mortality rate, and effectiveness of control.}
+The proportion of cases prevented in a year by control implemented at 30 weeks prior to the start of the year, averaged over all years and all posterior samples within each cell.
+Cells with darker shading indicate combinations of the period of cross-immunity and average mosquito mortality that produce a system more amenable to control by adult mosquito removal.
 }
@@
 Implementing control just as the human susceptible population begins to grow (when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) allows for additional immunizing infections to build up as the current outbreak subsides, while also disrupting transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
-The fact that effect of mosquito control was mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g., cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities.
+The fact that the effect of mosquito control was mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g., cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities.
 
 The Bayesian framework allowed for a full accounting of uncertainty across the data, process, and parameter levels of our model \cite{Berliner1996}.
-We then carried this uncertainty through to our simulations of control interventions \cite{Elderd2006}.
-We found that there can be substantial uncertainty in the proportion of cases prevented by a control intervention (i.e., the width of the ribbons in Figure \ref{control}).
-Much of this uncertainty could be attributed to uncertainty in the period of cross-immunity (Figure \ref{immunity}), the posterior distribution of which differed little from our prior based on the estimate of Reich \emph{et al.} \cite{Reich2013}.
-Specifically, we found that increasing the period of cross-immunity decreased the proportion of cases prevented by mosquito control.
-When the period of cross-immmunity was long (e.g., 200 weeks), the susceptible pool replenished slowly and transmission became limited by the availability of susceptible hosts.
-In the face of this limitation, manipulating mosquito abundance had a relatively small effect.
-Conversely, when the period of cross-immunity was short (e.g., 50 weeks), susceptibles were abundant and mosquito abundance was instead limiting, in which case reducing mosquito abundance further had a larger effect on transmission.
-This interaction between the efficacy of control and human immunity was also demonstrated by ten Bosch \emph{et al.} \cite{TenBosch2016} in their investigation of serotype interactions.
-Specifically, they found that adding cross-immunity to models with antibody-dependent enhancement increased the resistance to control (as implemented by reducing the transmission rate).
+We carried this uncertainty through to our simulations of control interventions \cite{Elderd2006} and found that there was substantial uncertainty in the proportion of cases prevented by a control intervention (i.e., the width of the ribbons in Figure \ref{control}).
+Much of this uncertainty could be attributed to uncertainty in and covariation between the period of cross-immunity and the average mosquito mortality rate (Figure \ref{immunity}).
+Specifically, we found that control implemented at 30 weeks prior to the start of the year (approximately the average optimum timing) had the largest impact in simulations with a low period of cross-immunity and a high average mosquito mortality rate.
+In these simulations, the rapid loss of immunity meant that human susceptibles were abundant and mosquito survival was instead limiting.
+Given this limitation, removing adult mosquitoes further tightened the bottleneck imposed by the race between mosquito survival and the extrinsic incubation period, resulting in a larger disruption to transmission.
+Hladish \emph{et al.} \cite{Hladish2018} similarly found that simulated indoor residual spraying was more effective when the modeled mosquito abundance was already low.
+This suggests that control efforts that reduce the ability of mosquitoes to transmit DENV are going to be most effective in situations in which mosquitoes are already the limiting component to maintaining transmission.
 
-This notion that mosquito control becomes less effective as the susceptible pool becomes more limiting has potential long-term consequences for the management of dengue fever.
-As immunity becomes widespread, either due to continued transmission or eventually vaccination, mosquito abundance will no longer be limiting, and broad-scale mosquito control may become less effective.
-At the same time, the replenishment of susceptibles, from both births and waning immunity, means that sustaining effective control over the long-term is likely to be difficult \cite{Scott2010b} and will thus require continued monitoring and management.
-As such, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance to monitor and estimate immunity \cite{Morrison2008}.
-In the absence of, or as a supplement to, such data, mechanistic models like the one developed here, or so-called TSIR frameworks that reconstruct the dynamics of the susceptible class \cite{Finkenstadt2000, Reich2013}, need to be further developed to better inform mosquito control efforts.
+While mosquito population dynamics play an important part in determining the effectiveness of control interventions, their role is also closely coupled with human immune processes.
+In addition to controlling how strongly a mosquito control intervention propagates into the following year, human immune processes also influence the overall vulnerability of the system to control.
+ten Bosch \emph{et al.} \cite{TenBosch2016} observed similar behavior in their models of dengue dynamics, and found that longer periods of cross-immunity generated systems in which transmission was more difficult to disrupt with control actions.
+Accurate prediction of the effects and efficient deployment of mosquito control is likely to depend on our ability to monitor, predict, and manipulate the dynamics of human immunity.
+As such, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance \cite{Morrison2008} and tighter estimates of the period of cross-immunity \cite{TenBosch2016}.
+In the absence of, or as a supplement to, such data, mechanistic models like the one developed here, or so-called TSIR frameworks that reconstruct the dynamics of the susceptible class \cite{Finkenstadt2000, Reich2013}, need to be further developed to better inform and understand mosquito control efforts.
 
@@
 \label{S1}
-{Full description of model and priors, along with all supplemental plots.}
+{Full description of model and priors, along with all supplemental figures.}
 
@@
 
+CBL would like to thank members of the Webb Lab for feedback and support through innumerable previous iterations of this project.
 Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

### 2019-04-17 Analysis/plots.r
Updates to figs 4 and 5

@@ -144,7 +144,8 @@ adult <- readRDS("Results/gamma_control.rds")
 adult <- mutate(adult, 
                 prevented = 1 - reduction, 
                 relweek = control - firsts[as.character(year)],
-                delta = rep(delta, each = 840))
+                delta = rep(delta, each = 840),
+                dv = rep(dvmu, each = 840))
 
 # Summarizing adult control simulations
 adult_sum <- adult %>%
@@ -175,18 +176,11 @@ dev.off()
 
 # Figure 4: Timing of control ==================================================
 
-# Finding week when control is most effective for each rep
-adult_best <- adult %>%
-  subset(relweek > -53 & relweek < 53) %>% 
-  ddply(.(rep, year), summarise,
-        best = relweek[which.min(reduction)],
-        ratio = max(prevented))
+# Finding week when median number of cases prevented is largest
+bests <- ddply(adult_sum, .(year), summarise, 
+               best = relweek[which.max(med)],
+               ratio = max(med))
 
-adult_best$delta <- rep(delta, each = 4)
-adult_best$dvmu <- rep(dvmu, each = 4)
-
-# Finding median most effective week of control for each year
-bests <- ddply(adult_best, .(year), summarise, best = median(best), prevented = median(ratio))
 bests$week <- bests$best + firsts[2:5]
 
 # Summarizing state variables
@@ -228,9 +222,9 @@ postscript("Manuscript/figures/fig5.eps",
            width = 5, height = 3,
            family = "ArialMT")
 
-adult_best %>% 
-  ggplot(aes(dvmu, delta, z = ratio)) + 
-  stat_summary_2d(binwidth = c(0.025, 10)) +
+subset(adult, relweek == -30) %>% 
+  ggplot(aes(dv, delta, z = prevented)) + 
+  stat_summary_2d(binwidth = c(0.025, 5)) +
   scale_fill_distiller(palette = "Oranges", direction = 1) +
   labs(fill = "average \nproportion \nof cases \nprevented") + 
   theme_classic() +

### 2019-04-17 Manuscript/ms.tex
A few discussion tweaks

@@
 
-While mosquito population dynamics play an important part in determining the effectiveness of control interventions, their role is also closely coupled with human immune processes.
-In addition to controlling how strongly a mosquito control intervention propagates into the following year, human immune processes also influence the overall vulnerability of the system to control.
+The posterior correlation between the effectiveness of control, the period of cross-immunity, and the average mosquito mortality rate emphasizes that the impact of mosquito control is jointly regulated by both mosquito population dynamics and human immune processes.
 ten Bosch \emph{et al.} \cite{TenBosch2016} observed similar behavior in their models of dengue dynamics, and found that longer periods of cross-immunity generated systems in which transmission was more difficult to disrupt with control actions.
-Accurate prediction of the effects and efficient deployment of mosquito control is likely to depend on our ability to monitor, predict, and manipulate the dynamics of human immunity.
-As such, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance \cite{Morrison2008} and tighter estimates of the period of cross-immunity \cite{TenBosch2016}.
+As a result of these relationships, efficient deployment of mosquito control, and accurate prediction of its effects, is likely to depend on our ability to monitor, predict, and manipulate the dynamics of human immunity.
+To meet these needs, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance \cite{Morrison2008} and tighter estimates of the period of cross-immunity \cite{TenBosch2016}.
 In the absence of, or as a supplement to, such data, mechanistic models like the one developed here, or so-called TSIR frameworks that reconstruct the dynamics of the susceptible class \cite{Finkenstadt2000, Reich2013}, need to be further developed to better inform and understand mosquito control efforts.

### 2019-05-24 Analysis/plots.r
Changes x-axis to year in figs 1 and 2

@@ -66,9 +66,9 @@ ggplot(post, aes(tot.week, yobs)) +
   geom_line(aes(tot.week, yhat)) + 
   theme_classic() +
   scale_y_continuous(expand = c(0, 0)) + 
-  scale_x_continuous(expand = c(0, 1)) +
+  scale_x_continuous(expand = c(0, 1), breaks = c(1, 54, 106, 158, 210), labels = c(2008, 2009, 2010, 2011, 2012)) +
   ylab("case reports") + 
-  xlab("week") +
+  xlab("year") +
   ggtitle("A") -> fig1.a
 
 ggplot(post, aes(tot.week, qobs)) + 
@@ -77,9 +77,9 @@ ggplot(post, aes(tot.week, qobs)) +
   geom_line(aes(tot.week, qhat)) + 
   theme_classic() +
   scale_y_continuous(expand = c(0.05, 0)) + 
-  scale_x_continuous(expand = c(0, 1)) +
+  scale_x_continuous(expand = c(0, 1), breaks = c(1, 54, 106, 158, 210), labels = c(2008, 2009, 2010, 2011, 2012)) +
   ylab("mosquitoes trapped") +
-  xlab("week") +
+  xlab("year") +
   ggtitle("B") -> fig1.b
 
 postscript("Manuscript/figures/fig1.eps",
@@ -108,16 +108,16 @@ ggplot(post, aes(tot.week, dvmed)) +
   geom_line() + 
   theme_classic() +
   scale_y_continuous(expand = c(0.05, 0), limits = c(0.25, 1.35)) + 
-  scale_x_continuous(expand = c(0, 1)) +
+  scale_x_continuous(expand = c(0, 1), breaks = c(1, 54, 106, 158, 210), labels = c(2008, 2009, 2010, 2011, 2012)) +
   ylab("mosquito mortality rate") +
-  xlab("week") +
+  xlab("year") +
   ggtitle("A") -> fig2.a
 
 ggplot(post, aes(temp, dvmed)) +
   geom_point() + 
   theme_classic() + 
   scale_y_continuous(expand = c(0.05, 0), limits = c(0.25, 1.35)) + 
-  scale_x_continuous(expand = c(0, 1)) +
+  scale_x_continuous(expand = c(0, 1)) + 
   ylab("mosquito mortality rate") +
   xlab("weekly mean temperature") + 
   ggtitle("B") -> fig2.b

### 2019-05-25 Manuscript/ms.tex
Adding middle initials to author list
Clarity edits

@@
 \\
-Clinton Leach\textsuperscript{1,2*},
-Jennifer Hoeting\textsuperscript{2}
-Kim Pepin\textsuperscript{3},
-Alvaro Eiras\textsuperscript{4},
-Mevin Hooten\textsuperscript{5,6,2},
-Colleen Webb\textsuperscript{1}
+Clinton B. Leach\textsuperscript{1,2*},
+Jennifer A. Hoeting\textsuperscript{2}
+Kim M. Pepin\textsuperscript{3},
+Alvaro E. Eiras\textsuperscript{4},
+Mevin B. Hooten\textsuperscript{5,6,2},
+Colleen T. Webb\textsuperscript{1}
 
@@
 The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$, $d_0$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
-Where possible, we specified informative priors on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).
+Where possible, we specified informative prior distributions on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).
 
@@
 \caption{Model parameters and their values.
-Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the prior mean.
-See Supplemental Material for a full description of all priors.}
+Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the mean of the prior distribution.
+See Supplemental Material for a full description of all prior distributions.}
 \begin{tabular}{llll}
@@
 \end{align}
-where $\boldsymbol{\theta}$ is a vector of all the model parameters and initial conditions, and $\mathcal{M}(\boldsymbol{\epsilon_r}, \boldsymbol{\epsilon_{\nu}},\boldsymbol{\theta})$ represents the (numeric) solution to the differential equation as a function of $\boldsymbol{\theta}$ and the weekly stochastic forcing terms ($\boldsymbol{\epsilon_{r}}, \boldsymbol{\epsilon_{\nu}}$).
+where $\boldsymbol{\theta}$ is a vector of all the model parameters and initial conditions, and $\mathcal{M}(\boldsymbol{\epsilon_r}, \boldsymbol{\epsilon_{\nu}},\boldsymbol{\theta})$ represents the (numeric) solution to the differential equation model (Eq.
+1-14) as a function of $\boldsymbol{\theta}$ and the weekly stochastic forcing terms ($\boldsymbol{\epsilon_{r}}, \boldsymbol{\epsilon_{\nu}}$).
 Sampling from the posterior distribution of the parameters and forcing terms in this model is difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations induced by the nonlinearity of the differential equation model \cite{Reilly2005, Girolami2008, Calderhead2011}.
@@
 These problems can be mitigated somewhat by introducing stochasticity through the $\epsilon_t$ in the mosquito demographic processes.
-This allows the data to pull the latent states closer even when the parameters are far from optimal values \cite{Leander2014}.
+This flexibility allows the latent states to hew closer to the data, even when the parameters are far from optimal values \cite{Leander2014}.
 This reduces multimodality and allows gradient-based methods like Hamiltonain Monte Carlo (HMC) to more easily and efficiently traverse the posterior.
@@
 Because the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in future years.
-To obtain the posterior distribution of the proportion of cases prevented in calendar year $j$ by implementing control in week $i$, for each posterior sample $k$, we simulated the dynamics resulting from reducing the mosquito population by $5\%$ at the beginning of week $i$ (affecting susceptible, exposed, and infectious mosquitoes equally, and leaving the estimated mosquito birth and mortality rates unchanged) .
-We then compared the number of cases produced during year $j$ in the control scenario to the same number in the uncontrolled scenario.
+For each week and each posterior sample, we simulated the dynamics resulting from a $5\%$ reduction in the mosquito population implemented at the beginning of that week (affecting susceptible, exposed, and infectious mosquitoes equally, and leaving the estimated mosquito birth and mortality rates unchanged).
+We then compared the number of cases produced during each year in the control scenario to the same number in the uncontrolled scenario.
 A $5\%$ reduction in mosquito abundance was chosen to keep our simulations conservative relative to field estimates of the mortality induced by spraying \cite{Esu2010}, and to avoid pushing the model into the unrealistic range of dengue eradication.
@@
 
+\begin{figure}[!h]
+%\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
+\caption{{\bf Vitoria data and model estimates.}
+Weekly observations from 2008 to week 34 of 2012 (points), with corresponding median posterior prediction (line) and 80\% credible interval (gray band).
+A: reported dengue cases.
+B: total number of trapped mosquitoes.
+}
+\label{timeseries}
+\end{figure}
+
 The estimated weekly mosquito mortality rate varied seasonally, and was positively correlated with temperature (Figure \ref{mortality}).
@@
 The posterior distribution of the baseline mortality rate ($d_0$) had a mean of 0.7/week, roughly half the prior mean, and was negatively correlated with the period of cross immunity ($1/\delta$, Figure \ref{immunity}).
-This negative correltaion indicates that similar dynamics could be achieved with either rapidly waning human immunity and short-lived mosquitoes or with long-lasting human immunity and long-lived mosquitoes.
+This negative correlation indicates that similar dynamics could be achieved with either rapidly waning human immunity and short-lived mosquitoes or with long-lasting human immunity and long-lived mosquitoes.
 The marginal posterior means of the $\epsilon_{\nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (Supplementary Figure 1), although the marginal posterior distribution of each $\epsilon_{\nu t}$ overlapped zero.
 In addition, the standard deviation of the mortality forcing terms was small relative to the weak prior ($E(\sigma_{\nu}|\mathbf{y}) = 0.01$, Supplemental Figure 2).
+
+\begin{figure}[!h]
+%\includegraphics[angle = 270, scale = 1]{figures/fig2.eps}
+\caption{{\bf Estimated mosquito mortality rate.}
+A: Time series of mortality rate, showing median posterior estimate (line) and 80\% credible interval (gray band). B: estimated weekly mosquito mortality rate as a function of weekly mean temperature.
+}
+\label{mortality}
+\end{figure}
 
@@
 Our simulations predicted that a pulse of control implemented during this time reduced the disease burden in the following year by $20$ to $30\%$, while perturbing mosquito abundance during the target outbreak was much less effective.
+
+\begin{figure}[!h]
+%\includegraphics[angle = 270, scale = 1]{figures/fig3.eps}
+\caption{{\bf The effect of mosquito control as a function of the week in which it was applied.}
+Summary of posterior predicted distribution of the proportion of cases that would be prevented in each year (top to bottom, 2009 to 2012) by control enacted in a given week relative to the start of that year.
+Black lines indicate the posterior median, while gray ribbons indicate the $80\%$ credible interval.
+For example, mosquito control applied 25 weeks prior to January 1, 2009 would have prevented about 20\% of the human cases in 2009, and control applied 25 weeks prior to January 1, 2010 would have prevented about 30\% of the human cases in 2010.
+}
+\label{control}
+\end{figure}
 
@@
 \begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
-\caption{{\bf Vitoria data and model fits.}
-Weekly observations from 2008 to week 34 of 2012 (points), with corresponding median posterior prediction (line) and 80\% credible interval (gray band).
-A: reported dengue cases.
-B: total number of trapped mosquitoes.
-}
-\label{timeseries}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/fig2.eps}
-\caption{{\bf Estimated latent mosquito mortality rate.}
-A: Time series of mortality rate, showing median posterior estimate (line) and 80\% credible interval (gray band). B: estimated weekly mosquito mortality rate as a function of weekly mean temperature.
-}
-\label{mortality}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/fig3.eps}
-\caption{{\bf The effect of mosquito control as a function of the week in which it was applied.}
-Summary of posterior predicted distribution of the proportion of cases that would be prevented in each year (top to bottom, 2009 to 2012) by control enacted in a given week relative to the start of that year.
-Black lines indicate the posterior median, while gray ribbons indicate the $80\%$ credible interval.
-}
-\label{control}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/fig4.eps}
+%\includegraphics[angle = 270, scale = 1]{figures/fig4.eps}
 \caption{{\bf Timing of optimal control relative to disease and mosquito dynamics.}
@@
 \begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/fig5.eps}
+%\includegraphics[angle = 270, scale = 1]{figures/fig5.eps}
 \caption{{\bf The relationship among the period of cross-immunity, average mosquito mortality rate, and effectiveness of control.}
@@
 Instead, we found that adult mosquito control efforts were most effective at reducing the disease burden within a given year when applied roughly 30 weeks prior to the start of that year, on the downswing of the previous year's epidemic and seasonal mosquito dynamics.
-Barsante \emph{et al.} \cite{Barsante2015}, similarly found that control was most effective when applied during the dry season (roughly matching our 30 week lag) than during the rainy season.
+Barsante \emph{et al.} \cite{Barsante2015} similarly found that control was most effective when applied during the dry season (roughly matching our 30 week lag) than during the rainy season.
 While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicated that disrupting inter-seasonal transmission can be an effective longer-term strategy \cite{Hladish2018}.
@@
 In fact, control implemented at a 30 week lag was effective largely due to the way it interacted with the dynamics of the susceptible pool. 
-Implementing control just as the human susceptible population begins to grow (when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) allows for additional immunizing infections to build up as the current outbreak subsides, while also disrupting transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
+Implementing control when human susceptibility is near a local minimum (i.e., when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) disrupts transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
 The fact that the effect of mosquito control was mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g., cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities.
 
-The Bayesian framework allowed for a full accounting of uncertainty across the data, process, and parameter levels of our model \cite{Berliner1996}.
+The Bayesian framework allowed us to account for uncertainty across the data, process, and parameter levels of our model \cite{Berliner1996}.
 We carried this uncertainty through to our simulations of control interventions \cite{Elderd2006} and found that there was substantial uncertainty in the proportion of cases prevented by a control intervention (i.e., the width of the ribbons in Figure \ref{control}).
@@
 To meet these needs, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance \cite{Morrison2008} and tighter estimates of the period of cross-immunity \cite{TenBosch2016}.
-In the absence of, or as a supplement to, such data, mechanistic models like the one developed here, or so-called TSIR frameworks that reconstruct the dynamics of the susceptible class \cite{Finkenstadt2000, Reich2013}, need to be further developed to better inform and understand mosquito control efforts.
+In the absence of, or as a supplement to, such data, mechanistic models like the one developed here, or so-called TSIR (Time-series Susceptiple-Infected-Recovered) frameworks that reconstruct the dynamics of the susceptible class \cite{Finkenstadt2000, Reich2013}, need to be further developed to better inform and understand mosquito control efforts.
 
-\subsection*{Model validation and interpretation of $d_v$}
+\subsection*{Interpretation of the estimated mosquito mortality rate}
 
@@
 This encoded our prior expectation that the mosquito mortality rate should be driven by seasonal environmental forcing (i.e., it should oscillate with a period of one year), while allowing the flexibility to diverge from the simple oscillator and learn from the case reports data.
-The fact that the $\epsilon_{\nu}$ forcing terms were the only source of variability in the transmission process, however, means that the estimated mosquito mortality time series could have been soaking up other sources of stochasticity or model misspecification.
+The fact that the mosquito mortality forcing terms ($\epsilon_{\nu}$, Eq.
+13 and 16) were the only source of variability in the transmission process, however, means that the estimated mosquito mortality time series could have been soaking up other sources of stochasticity or model misspecification.
 Hooker and Ellner \cite{Hooker2015} provide a framework for diagnosing such model misspecification in differential equation models using forcing functions similar to our implementation of the $\epsilon_{\nu}$.
@@
 
-The periodic structure in the posterior means of the $\epsilon_{\nu t}$ suggests that these terms were accounting for more than just noise, and there may have been some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
+The periodic structure in the time series of the posterior means of the $\epsilon_{\nu}$ (Supplemental Figure 1) suggests that these terms were accounting for more than just noise, and there may have been some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
 Following Hooker and Ellner, we can explore whether this process is likely to result from misspecification of the rates of change of the existing state variables (indicated by a dependence of $\epsilon_{\nu t}$ on other state variables), or from missing state variables altogether (indicated by an additional dependence of $\epsilon_{\nu t}$ on its own lagged values).
@@
 The challenges presented by these confounding factors highlight the value of mechanistic information in studying the effect of mosquito control on disease spread.
-In the Bayesian context we have deployed here, this mechanistic knowledge, as formalized in the specification of a differential equation model, can be viewed as part of the prior on the relationship between mosquito abundance and human disease \cite{Ellner1998, Wikle2010}.
-As such, we should neither ignore it, nor encode it so rigidly that it swamps the signal in our data.
+In the Bayesian context we have deployed here, this mechanistic knowledge, as formalized in the specification of a differential equation model, can be viewed as part of the prior knowledge on the relationship between mosquito abundance and human disease \cite{Ellner1998, Wikle2010}.
+As such, we should neither ignore this mechanistic information, nor encode it so rigidly that it swamps the signal in our data.
 
@@
 \label{S1}
-{Full description of model and priors, along with all supplemental figures.}
+{Full description of model and prior distributions, along with all supplemental figures.}
 

### 2019-05-25 Analysis/plots.r

@@ -261,7 +261,7 @@ ggplot(eps_dv, aes(week, med)) +
   geom_hline(yintercept = 0, linetype = 2) + 
   theme_classic() + 
   scale_x_continuous(expand = c(0, 1)) +
-  ylab(expression(epsilon[d])) +
+  ylab(expression(epsilon[nu])) +
   ggtitle("A") -> figs1.a
 
 ggplot(eps_rv, aes(week, med)) + 

### 2020-02-14 Manuscript/ms.tex
Revisions

@@
 
-Differential equation models provide a natural way to encode the relevant biological mechanisms, although it is important to also account for sources of uncertainty in those models \cite{Hotelling1927, Wikle2010}.
+Differential equation models provide a natural way to describe mechanistic processes, but that description must also account for sources of uncertainty \cite{Hotelling1927, Wikle2010}.
 In particular, the values of parameters (e.g., the length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
@@
 Vit\'oria is a coastal city and the capital of the state of Esp\'irito Santo, Brazil, with a population of 327,801 as of 2010 \cite{vitpop}.
-Since 2008, the company Ecovec has monitored mosquito abundance for the city using approximately 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a grid across the city.
+Since 2008, the company Ecovec has monitored mosquito abundance for the city using approximately 1327 sticky traps (MosquiTRAP, \cite{Eiras2009}) arranged in a roughly 250m grid across the city \cite{Pepin2015, Lana2018}.
 Each trap is checked weekly and the mosquitoes inside counted and identified, with the results sent to a central database that city managers then use to map mosquito infestations and target control.
@@
 The strength and duration of these different inter-serotype interactions are not well understood, although different models suggest that temporary cross-immunity alone (without ADE) is sufficient to reproduce observed multi-annual dynamics in Thailand \cite{Wearing2006,Reich2013}.
-Similarly, Aguiar \emph{et al.} \cite{Aguiar2013} found that capturing primary and secondary infections and the period of cross-immunity is critical, but that explicitly including all four serotypes (at great cost to model complexity) does not perform better than a two serotype model.
 
-Explicitly accounting for interactions between serotypes, even only two of the four, leads to a large and complex mechanistic model.
+Explicitly capturing the cross-immune interactions among all four serotypes, or even only two of the four \cite{Aguiar2013}, leads to a large and complex mechanistic model.
 Moreover, because dengue case reports do not identify serotype, there is not enough information in the data to inform the dynamics of individual serotypes.
-As such, we developed an SEIRS compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010} which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
-In this framework, susceptible humans ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
+As such, we captured temporary cross-immunity, and the potential for multiple sequential infections, as simply and tractably as possible in a susceptible-exposed-infectious-recovered-susceptible (SEIRS) compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010}.
+Though it captures the critical influence of temporary immunity, this framework does not account for the potential relationship between an individual's infection history and the likelihood that a new infection with be symptomatic (and thus reported).
+In particular, secondary infections appear more likely to be symptomatic than primary infections \cite{Imai2016,Clapham2017}, while third and fourth infections appear much less likely to be symptomatic \cite{Olkowski2013} (but see \cite{Montoya2013} who found similar rates of symptomatic cases across infection number).
+Despite these potential differences in reporting rate, modeling work has suggested that the dynamics of primary and secondary infections are closely coupled (and thus not dynamically distinct) under many conditions \cite{Schwartz2005}.
+Moreover, given the short time scale of our data (5 years) relative to the period of cross-immunity (roughly 2 years \cite{Reich2013}), we expect that third and fourth infections will be relatively rare.
+We thus expect the SEIRS framework to be sufficient for capturing the dengue dynamics of Vit\'oria and that reporting differences across infection number are unlikely to have a large impact on the inferred relationship between mosquito abundance and human disease.
+
+In the SEIRS framework, susceptible humans ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
 Following a latent period ($\frac{1}{\rho}$), exposed humans become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
 Infectious humans recover at rate $\gamma$ and subsequently remain immune ($R$) for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
-
 Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_E$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
@@
 \frac{dV_N}{dt} & = r(t) V_N - \phi_q \tau(t) V_N \\
-\frac{dV_{E1}}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E1}\\
-\frac{dV_{E2}}{dt} &= \rho_{v}(t) V_{E1} - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E2}\\
-\frac{dV_{E3}}{dt} &= \rho_{v}(t) V_{E2}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E3}\\
-\frac{dV_{E4}}{dt} &= \rho_{v}(t) V_{E3}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E4}\\
-\frac{dV_I}{dt} &= \rho_{v}(t) V_{E4} - (d(t) + \phi_q \tau(t)) V_I\\
+\frac{dV_{E1}}{dt} &= \lambda \frac{I}{N} V_S - (\nu\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E1}\\
+\frac{dV_{E2}}{dt} &= \nu\rho_{v}(t) V_{E1} - (\nu\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E2}\\
+\frac{dV_{E3}}{dt} &= \nu\rho_{v}(t) V_{E2}  - (\nu\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E3}\\
+\frac{dV_{E4}}{dt} &= \nu\rho_{v}(t) V_{E3}  - (\nu\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E4}\\
+\frac{dV_I}{dt} &= \nu\rho_{v}(t) V_{E4} - (d(t) + \phi_q \tau(t)) V_I\\
 \frac{dV_C}{dt} & = \phi_q \tau(t) V_N\\
@@
 \end{align}
-for $t = 1 \dots 243$.
+for each week $t = 1 \dots 243$.
 These stochastically-forced harmonic oscillators provide a flexible framework for generating smooth seasonal oscillations in the latent mosquito processes \cite{Ramsay2017}.
@@
 $\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
-$1/\rho_{vt}$ & Extrinsic incubation period & $28\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
+$1/\rho_{v}(t)$ & Extrinsic incubation period & $\exp\left(7.9 - 0.21 T(t) \right)$ days & \cite{Chan2012}\\
 $V_{E0}$ & Initial exposed mosquitoes &  0 & \\
@@
 1-14) as a function of $\boldsymbol{\theta}$ and the weekly stochastic forcing terms ($\boldsymbol{\epsilon_{r}}, \boldsymbol{\epsilon_{\nu}}$).
-Sampling from the posterior distribution of the parameters and forcing terms in this model is difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations induced by the nonlinearity of the differential equation model \cite{Reilly2005, Girolami2008, Calderhead2011}.
-In addition, the computational burden of numerically solving the differential equation model at each iteration means that sampling efficiency is crucial.
-These problems can be mitigated somewhat by introducing stochasticity through the $\epsilon_t$ in the mosquito demographic processes.
-This flexibility allows the latent states to hew closer to the data, even when the parameters are far from optimal values \cite{Leander2014}.
-This reduces multimodality and allows gradient-based methods like Hamiltonain Monte Carlo (HMC) to more easily and efficiently traverse the posterior.
+Sampling from the posterior distribution of the parameters in mechanistic models is difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations induced by the nonlinearity of the differential equation model \cite{Reilly2005, Girolami2008, Calderhead2011}.
+However, the variable $\epsilon_t$ introduce flexibility to the mechanistic model that 
+soaks up lack-of-fit when the process parameters are far from optimal \cite{Leander2014}, thereby reducing multimodality and helping to smooth the posterior surface.
+Gradient-based methods like Hamiltonain Monte Carlo (HMC) can then more easily and efficiently traverse the posterior.
 Samples from the posterior distribution were thus generated using HMC implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R \cite{R2016}. 
@@
 Code is available from \href{https://github.com/clint-leach/mosquito-recon}{https://github.com/clint-leach/mosquito-recon}.
+
+In addition, the computational burden of numerically solving the differential equation model at each iteration means that sampling efficiency is crucial.
 
@@
 
-The mismatch between the timing of the most effective control and the peak in mosquito abundance suggets that mosquito abundance alone was an insufficient predictor of the effect of mosquito control, at least at the city-scale.
+The mismatch between the timing of the most effective control and the peak in mosquito abundance suggests that mosquito abundance alone was an insufficient predictor of the effect of mosquito control, at least at the city-scale.
 Instead, the effectiveness of mosquito control depended, in large part, on the intrinsic long-term feedbacks and lags driven by the dynamics of temporary human immunity.
@@
 Implementing control when human susceptibility is near a local minimum (i.e., when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) disrupts transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
-The fact that the effect of mosquito control was mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g., cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities.
+The fact that the effect of mosquito control was mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g., cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities \cite{Wearing2006, Reich2013}.
 
@@
 The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
-These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control interventions), or epidiemiological processes (e.g., multiple circulating serotypes of the dengue virus).
+These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control interventions), or epidemiological processes (e.g., multiple circulating serotypes of the dengue virus).
 

### 2020-02-14 Manuscript/supplement.tex
Revisions

@@
 To account for this, we make $1/\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground (Goiabeiras weather station, station ID: IESVITOR2).
-We use the temperature relationship from the recent meta-analysis of \cite{Chan2012}, which incorporates a large number of studies and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
-To correspond with the assumptions of the differential equation model, we use the relationship obtained under the assumption that the incubation period at any given temperature is exponentially-distributed:
+We use the temperature relationship from the meta-analysis of Chan annd Johansson \cite{Chan2012}, which incorporates a large number of studies and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
+Chan \textit{et al.}find that the log-normal and gamma distributions are nearly indistinguishable in providingthe best fit to the EIP data.
+To capture a gamma-distributed EIP in the differential equation model, we follow \cite{Lloyd2001} and chain together $\nu$ exposed compartments, each with an out-flow rate of $\nu\rho_v(t)$ to generate
 \begin{equation}
-\rho_v = 7\exp \left(0.2T - 8\right)\\
+EIP \sim \text{Gamma}(\nu, \nu \rho_v(t))
+\end{equation}
+such that $E[EIP] = 1/ \rho_v(t)$.
+From \cite{Chan2012}, we let $\nu = 4$ and 
+\begin{equation}
+\rho_v(t) = 7\exp \left(0.21T(t) - 7.9\right)\\
 \end{equation}

### 2020-04-27 Manuscript/ms.tex
Updating figures and results

@@
 
-Given samples from the posterior distribution as obtained above, we simulated the effects of a single pulse of control applied in each week of the first four years of the time series.
+Given samples from the posterior distribution as obtained above, we simulated the effects of a single pulse of mosquito control applied in each week of the first four years of the time series.
 Because the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in future years.
-For each week and each posterior sample, we simulated the dynamics resulting from a $5\%$ reduction in the mosquito population implemented at the beginning of that week (affecting susceptible, exposed, and infectious mosquitoes equally, and leaving the estimated mosquito birth and mortality rates unchanged).
+For each week and each posterior sample, we simulated the dynamics resulting from a $5\%$ reduction in the mosquito population implemented at the beginning of that week (affecting susceptible, exposed, and infectious mosquitoes equally).
+To capture the likely rapid rebound in mosquito abundance following a single pulse of control \cite{Focks1987, Burattini2008}, we also simulated a $5\%$ increase in mosquito birth rate in the following week to return mosquito abundance to its previous trajectory (without this, the $5\%$ reduction in abundance persists indefinitely).
 We then compared the number of cases produced during each year in the control scenario to the same number in the uncontrolled scenario.
@@
 The model captured the observed dynamics of both case reports and mosquito trap counts (Figure \ref{timeseries}).
-The estimated posterior median case reports explained 90\% of the variation in the observed time series, while the posterior median mosquito trap counts explained 46\% of the variation in the observed time series.
+The estimated posterior median case reports explained 91\% of the variation in the observed time series, while the posterior median mosquito trap counts explained 46\% of the variation in the observed time series.
 In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 6 - 8).
 The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that the Vitoria data contained little additional information about these parameters (Supplemental Figure 3).
-The estimated latent period in a human host ($1/\rho$, the expected time it takes for an exposed human to become infectious to biting mosquitoes) was influenced more strongly by the data, with a posterior mean of 1.53 weeks compared to a prior mean of 0.87 weeks.
+The estimated latent period in a human host ($1/\rho$, the expected time it takes for an exposed human to become infectious to biting mosquitoes) was influenced more strongly by the data, with a posterior mean of 1.73 weeks compared to a prior mean of 0.87 weeks.
+Further, the posterior mean case reporting rate ($\phi$) was 0.14, larger than the prior mean of 0.08.
 
 \begin{figure}[!h]
-%\includegraphics[angle = 270, scale = 1]{figures/fig1.eps}
+\includegraphics[scale = 1]{figures/fig1.eps}
 \caption{{\bf Vitoria data and model estimates.}
-Weekly observations from 2008 to week 34 of 2012 (points), with corresponding median posterior prediction (line) and 80\% credible interval (gray band).
-A: reported dengue cases.
-B: total number of trapped mosquitoes.
+A: weekly observed case reports (points), with corresponding posterior median (black line) and 80\% posterior credible interval (gray band).
+B: weekly mosquito trap counts (points), with posteriod median (black line) and 80\% posterior credible interval (gray band).
+C: extrinsic incubation period (weeks), computed from weekly mean temperature data.
+D: estimated weekly mosquito mortality rate, with the posterior median (black line) and the 80\% posterior credible interval (gray band).
 }
@@
 
-The estimated weekly mosquito mortality rate varied seasonally, and was positively correlated with temperature (Figure \ref{mortality}).
-The oscillations in estimated mosquito mortality were more complex than purely sinusoidal, with variations in shape from year to year.
-The posterior distribution of the baseline mortality rate ($d_0$) had a mean of 0.7/week, roughly half the prior mean, and was negatively correlated with the period of cross immunity ($1/\delta$, Figure \ref{immunity}).
-This negative correlation indicates that similar dynamics could be achieved with either rapidly waning human immunity and short-lived mosquitoes or with long-lasting human immunity and long-lived mosquitoes.
+The estimated weekly mosquito mortality rate varied seasonally, with generally high mortality early in the year and low mortality in August to October (Figure \ref{timeseries}).
+This seasonal trend broadly tracked seasonal variation in temperature (Figure \ref{mortality}) and mosquito abundance (Figure \ref{timeseries}, though the shape of the annual trajectory differed from year to year.
+The posterior distribution of the baseline mortality rate ($d_0$) had a mean of 0.88/week, roughly $60\%$ of the prior mean.
 The marginal posterior means of the $\epsilon_{\nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (Supplementary Figure 1), although the marginal posterior distribution of each $\epsilon_{\nu t}$ overlapped zero.
@@
 \begin{figure}[!h]
-%\includegraphics[angle = 270, scale = 1]{figures/fig2.eps}
-\caption{{\bf Estimated mosquito mortality rate.}
-A: Time series of mortality rate, showing median posterior estimate (line) and 80\% credible interval (gray band). B: estimated weekly mosquito mortality rate as a function of weekly mean temperature.
+\includegraphics[scale = 1]{figures/fig2.eps}
+\caption{{\bf Mosquito mortality and temperature.}
+Posterior median mosquito mortality rate as a function of weekly mean temperature (Celsius).
 }
@@
 
-Mosquito control (i.e.
-the removal of $5\%$ of the adult population in a given week), had the largest effect on the number of dengue cases in a given year when implemented between 30 and 40 weeks prior to the start of that year (Figure \ref{control}).
-Our simulations predicted that a pulse of control implemented during this time reduced the disease burden in the following year by $20$ to $30\%$, while perturbing mosquito abundance during the target outbreak was much less effective.
+The effect of a given mosquito control intervention (i.e.
+the temporary removal of $5\%$ of the adult population in a given week) on the number of cases in the following year (relative to no control) varied both seasonally and interannually (Figure \ref{control}).
+This variation in the effect of control was tightly correlated with the estimated mosquito mortality, with a median posterior correlation between the two time series of 0.96.
+As such, the seasonal variation in the effectiveness of control followed the same trend as mosquito mortality rate (though the annual minima in the case ratio time series generally fell 1 to 2 weeks before the minima in the mosquito mortality time series), with the largest reductions in case load coming from interventions during the dengue off-season (early September for 2008 and 2009, and mid-July for 2010).
+Summing over the interannual variation to compute the overall effect of control implemented in a given week of the year, we found that mosquito control was most effective when implemented in week 34 (late August), reducing the estimated case load by roughly $14\%$ (Figure \ref{control}, bottom panel).
+This broadly corresponds to the time of year when both dengue case reports and mosquito trap counts are low.
 
 \begin{figure}[!h]
-%\includegraphics[angle = 270, scale = 1]{figures/fig3.eps}
+\includegraphics[scale = 1]{figures/fig3.eps}
 \caption{{\bf The effect of mosquito control as a function of the week in which it was applied.}
-Summary of posterior predicted distribution of the proportion of cases that would be prevented in each year (top to bottom, 2009 to 2012) by control enacted in a given week relative to the start of that year.
+Summary of the posterior predicted effect of mosquito control implemented in a given week of the year on the number of cases in the following year (relative to the number of cases expected without control).
 Black lines indicate the posterior median, while gray ribbons indicate the $80\%$ credible interval.
-For example, mosquito control applied 25 weeks prior to January 1, 2009 would have prevented about 20\% of the human cases in 2009, and control applied 25 weeks prior to January 1, 2010 would have prevented about 30\% of the human cases in 2010.
+The first three panels show the results for control implemented in the years 2008-2010, and the last panel shows the overall effect of control impelmented in a given week of the year, summing over all three years.
+For example, mosquito control applied in week 37 of 2008 would have prevented about 13\% of the human cases over the following year (i.e., the caseload would have been 87\% of the expectation without control).
 }
@@
 
-The effect of a given control intervention seemed to be driven largely by its interaction with, and timing relative to, fluctuations in human immunity.
-In particular, mosquito control seemed to be most effective when applied at or near a local minimum in the number of suceptible humans (Figure \ref{timing}).
-In addition, the average effectiveness of mosquito control applied in this optimal window (i.e.
-at 30 weeks prior to the start of the year) was driven largely by the period of cross-immunity ($1/\delta$) and the mean mosquito mortality rate ($d_0$, Figure \ref{immunity}).
-Specifically, we observed that mosquito control was able to prevent the largest number of human dengue cases, averaged over all years, in simulations with a low period of cross-immunity and a high average mosquito mortality rate (Figure \ref{immunity}).
-Thus mosquito control was more effective at reducing disease burden in systems where short mosquito lifespan was already limiting the abundance of infectious mosquitoes and susceptible hosts were relatively abundant.
+The variation around the predicted effect of control implemented in week 34 of the year (i.e., the width of ribbon in Figure \ref{control}) was correlated with the baseline mosquito mortality rate ($d_0$) and the case reporting probability ($\phi$).
+Simulated mosquito control created the largest reduction in case reports in posterior samples with low baseline mosquito mortality rate and/or high reporting probability, while control was relatively less effective in simulations from samples with high mosquito mortality or low reporting probability (Figure \ref{correlations}).
+Thus mosquito control was more effective at reducing disease burden in simulations with long average mosquito lifespans (i.e., low mosquito mortality rates) or low overall prevalence (i.e., fewer undetected cases).
 
 \begin{figure}[!h]
-%\includegraphics[angle = 270, scale = 1]{figures/fig4.eps}
-\caption{{\bf Timing of optimal control relative to disease and mosquito dynamics.}
-The vertical gray lines indicate the week at which control generated the largest median number of prevented cases in the following calendar year.
-A: posterior median estimate of new dengue cases per week.
-B: posterior median estimate of the proportion of susceptible individuals.
-C: posterior median estimate of number of mosquitoes per person each week.
+\includegraphics[scale = 0.9]{figures/fig4.eps}
+\caption{{\bf Posterior correlation between system parameters and the effectiveness of control.}
+The y-axis represents the effectiveness of optimally timed control, i.e., the effect of control implmented in the 34th week of the year on the relative number of cases in the following year, summed over 2008, 2009, and 2010.
+Each point represets a single sample from the posterior distribution, giving the number of cases in the controlled simulation (relative to the number of cases expected without control) as a function of A: the mean mosquito mortality rate, $d_0$, and B: the case reporting probability, $\phi$, from that posterior sample.
 }
-\label{timing}
-\end{figure}
-
-\begin{figure}[!h]
-%\includegraphics[angle = 270, scale = 1]{figures/fig5.eps}
-\caption{{\bf The relationship among the period of cross-immunity, average mosquito mortality rate, and effectiveness of control.}
-The proportion of cases prevented in a year by control implemented at 30 weeks prior to the start of the year, averaged over all years and all posterior samples within each cell.
-Cells with darker shading indicate combinations of the period of cross-immunity and average mosquito mortality that produce a system more amenable to control by adult mosquito removal.
-}
-\label{immunity}
+\label{correlations}
 \end{figure}

### 2020-04-27 Analysis/plots.r

@@ -32,9 +32,9 @@ weather <- read.csv("Data/Vitoria.weather.csv") %>%
   mutate(date = ymd(BRST), year = year(date), week = week(date))
 
 weather <- subset(weather, date < date[1] + weeks(243))
-weather$tot.week <- rep(c(1:243), each = 7)
+weather$control <- rep(c(1:243), each = 7)
 
-temp <- ddply(weather, .(tot.week), summarise, temp = mean(Mean.TemperatureC, na.rm = T),
+temp <- ddply(weather, .(control), summarise, temp = mean(Mean.TemperatureC, na.rm = T),
               humid = mean(Mean.Humidity, na.rm = T))
 
 post$temp <- temp$temp
@@ -44,113 +44,160 @@ post$rov <- 7 * exp(0.21 * temp$temp - 7.9)
 sim <- readRDS("Results/chain.rds")
 
 # Loading control simulations
-control <- readRDS("Results/control_rebound.rds") %>% 
-  mutate(ratio = cases / cases0,
-         diff = cases0 - cases)
+# control <- readRDS("Results/control_rebound.rds") 
+
+# Loading control summaries (fixed and moving window)
+
+# Effect of timing of control on the number of cases in a given year
+# annual <- readRDS("Results/annual_control.rds")
+
+# Effect of control in the number of cases in the following year
+moving <- readRDS("Results/moving_control.rds")
 
 # Figure 1: Observed and estimated time series =================================
 
-# Computing posterior median and 80% credible interval
-yhat <- rstan::extract(sim, c("y_meas"), permute = F) %>% apply(3, quantile, c(0.1, 0.5, 0.9))
-qhat <- rstan::extract(sim, c("q_meas"), permute = F) %>% apply(3, quantile, c(0.1, 0.5, 0.9))
+# Case reports
+yhat <- rstan::extract(sim, c("y_meas"), permute = T)[[1]] %>% 
+  reshape2::melt(varnames = c("rep", "week"), value.name = "yhat") %>% 
+  ddply(.(week), summarise,
+        ymed = median(yhat),
+        ymin = quantile(yhat, 0.1), 
+        ymax = quantile(yhat, 0.9))
+  
+# Trapped mosquitoes
+qhat <- rstan::extract(sim, c("q_meas"), permute = T)[[1]] %>%
[…]

### 2020-05-01 Manuscript/ms.tex
More revisions
largely discussion

@@
 We used the fitted model to explore how the timing of a pulse of adult mosquito control influences its effect on the human disease burden in the following year.
-We found that control implemented roughly 30 to 40 weeks prior to the start of the year produced the largest reduction in predicted human case reports for that year.
+We found that control was generally more effective when implemented in periods of relatively low mosquito mortality (when mosquito abundance was also generally low).
+In particular, control implemented in early September (week 34 of the year) produced the largest reduction in predicted human case reports over the following year.
 This highlights the potential long-term utility of broad, off-peak-season mosquito control in addition to existing, locally targeted within-season efforts.
-Human immunity played an important role in mediating the effect of control, with the 30 to 40 week lag generally aligning with the estimated local minimum in the size of the human susceptible pool.
-Further, uncertainty in the effectiveness of control interventions was driven largely by posterior variation in the period of cross-immunity and the average mosquito mortality rate, with shorter periods of cross-immunity and higher mosquito mortality leading to more effective control.
-Understanding the dynamics of human immunity is thus critical for understanding how the effects of mosquito control propagate through the dengue system.
+Further, uncertainty in the effectiveness of control interventions was driven largely by posterior variation in the case reporting probability (closely tied to overall prevalence and immunity) and the average mosquito mortality rate (closely tied to total mosquito abundance) with higher reporting rates and lower mosquito mortality generating systems more vulnerable to control.
+Broadly, these correlations suggest that mosquito control is most effective in situations in which transmission is already limited by mosquito abundance.
 
@@
 We developed a mechanistic model that accounts for the role of this process in the dynamics of dengue fever and fit the model to a time series of human case reports and mosquito trap counts from the city of Vit\'oria, Brazil.
-We then used this fitted model to simulate the effect of mosquito control implemented at different times of the year and found that mosquito control leads to the largest reduction in human dengue cases during a given year when applied at the local minimum of the previous year’s human immunity trajectory.
-Further, the effectiveness of mosquito control was strongly correlated with how quickly previously infected humans become susceptible again and with the average mosquito mortality rate.
-Thus, the effects of mosquito control efforts are tightly coupled with the dynamics of human immunity and the long-term lags and feedbacks generated by the epidemiological system.
+We then used this fitted model to simulate the effect of mosquito control implemented at different times of the year and found that mosquito control leads to the largest reduction in human dengue cases over the following year when implemented in early September, during the dengue off-season.
+Further, the effectiveness of mosquito control was strongly negatively correlated with the overall average abundance of mosquitoes.
+Together with the timing of effective control, these results suggest that mosquito control is most effective when mosquitoes are already limiting to transmission.
 
@@
 
-The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$, $d_0$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
+The remaining parameters include the epidemiological parameters controlling the average latent, infectious, and immune periods ($\rho$, $\gamma$, $\delta$) and average mosquito lifespan ($d_0$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
 Where possible, we specified informative prior distributions on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).
@@
 See Supplemental Material for a full description of all prior distributions.}
-\begin{tabular}{llll}
-Parameter & Description & Value & Citation\\
+\begin{tabular}{lp{6cm}lll}
+Parameter & Description & Prior mean & Posterior mean & Citation\\
 \hline
-$N$ & Human population size & 327801 & \cite{vitpop} \\
-$1/d$ & Human life-span & 76 years & \cite{vitlong} \\
-$\lambda$ & Transmission rate & 4.87 week$^{-1}$ & \cite{Scott2000}\\
-$\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
-$1/\rho_{v}(t)$ & Extrinsic incubation period & $\exp\left(7.9 - 0.21 T(t) \right)$ days & \cite{Chan2012}\\
-$V_{E0}$ & Initial exposed mosquitoes &  0 & \\
-$V_{I0}$ & Initial infectious mosquitoes & 0 & \\
+$N$ & Human population size & 327801 & & \cite{vitpop} \\
+$1/d$ & Human life-span & 76 years & &\cite{vitlong} \\
+$\lambda$ & Transmission rate & 4.87 week$^{-1}$ & & \cite{Scott2000}\\
+$1/\rho_{v}(t)$ & Extrinsic incubation period & $\exp\left(7.9 - 0.21 T(t) \right)$ days & & \cite{Chan2012}\\
+$V_{E0}$ & Initial exposed mosquitoes &  0 & &\\
+$V_{I0}$ & Initial infectious mosquitoes & 0 & &\\
 \hline
-$d_0$ & Baseline mosquito mortality rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
-$1/\rho$ & Latent period in host & 0.87 weeks  & \cite{Chan2012}\\
-$\gamma$ & Rate of loss of infectiousness & 3.5 week$^{-1}$ & \cite{Nguyet2013}\\
-$1/\delta$ & Period of cross-immunity & 97 weeks &  \cite{Reich2013}\\
-$\sigma_r$ & Standard deviation of mosquito growth rate forcing & $0^*$ & \\
-$\sigma_{\nu}$ & Standard deviation of mosquito mortality rate forcing & $0^*$ & \\
-$S_0$ & Proportion initially susceptible & 0.4 & \cite{Cardoso2011a} \\
-$E_0$ & Number initially exposed & $100$ & \\
-$I_0$ & Number initially infectious & $60$ & \\
-$r_0$ & Initial mosquito population growth rate & 0 & \\
-$\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & \\
-$V_{N0}$ & Initial mosquito population size & $2N$ & \\
-$\phi_q$ & Per-trap mosquito capture rate & $2 \times 10^{-6}$ week$^{-1}$ & \\
-$\eta_y$ & Overdispersion of case reports & $0^*$ & \\
-$\eta_q$ & Overdispersion of mosquito trap counts & $0^*$ & \\
+$d_0$ & Baseline mosquito mortality rate & 1.47 week$^{-1}$ & 0.88 (0.7, 1.1) &  \cite{Brady2013} \\
+$1/\rho$ & Latent period in host & 0.87 weeks  & 1.72 (1.2, 2.3) &  \cite{Chan2012}\\
+$\gamma$ & Rate of loss of infectiousness & 3.5 week$^{-1}$ & 3.6 (3.2, 4.1) & \cite{Nguyet2013}\\
+$1/\delta$ & Period of cross-immunity & 97 weeks &  114 (72, 160) & \cite{Reich2013}\\
+$\sigma_r$ & Standard deviation of mosquito growth rate forcing & $0^*$ & 0.013 (0.01, 0.02) &\\
+$\sigma_{\nu}$ & Standard deviation of mosquito mortality rate forcing & $0^*$ & 0.0005 (0.0004, 0.0008) & \\
+$S_0$ & Proportion initially susceptible & 0.4 & 0.42 (0.28, 0.57)& \cite{Cardoso2011a} \\
+$E_0$ & Number initially exposed & $100$ & 148 (104, 196) & \\
+$I_0$ & Number initially infectious & $60$ & 79 (45, 116) &\\
+$r_0$ & Initial mosquito population growth rate & 0 & 0.008 (-0.01, 0.02) &\\
+$\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & -0.16 (-0.4, 0.08) &\\
+$V_{N0}$ & Initial mosquito population size & $2N$ & $1.5N$ $(1.1N, 1.9N)$ &\\
+$\phi_y$ & Reporting probability & 0.083 & 0.14  (0.1, 0.18) &\cite{Silva2016}\\
+$\log(\phi_q)$ & Log per-trap mosquito capture rate & $-13$  & -13.2 (-13.6, -13)  &\\
+$\eta_y$ & Overdispersion of case reports & $0^*$ & 0.12 (0.1, 0.13) &\\
+$\eta_q$ & Overdispersion of mosquito trap counts & $0^*$ & 0.14 (0.12, 0.16)&\\
 \end{tabular}
@@
 Code is available from \href{https://github.com/clint-leach/mosquito-recon}{https://github.com/clint-leach/mosquito-recon}.
-
-In addition, the computational burden of numerically solving the differential equation model at each iteration means that sampling efficiency is crucial.
 
@@
 The dynamics of dengue fever, like those of many infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
-In the dengue system, these drivers include waning human immunity, the race between the extrinsic incubation period of the virus and mosquito lifespan, and the seasonal and stochastic variation in these rates.
-These processes are particularly important for establishing the relationship between mosquito abundance -- and efforts to control that abundance -- and human disease.
+%In the dengue system, these drivers include waning human immunity, the race between the extrinsic incubation period of the virus and mosquito lifespan, and the seasonal and stochastic variation in these rates.
+Seasonality in particular is an important factor in capturing the annual cycle of dengue outbreaks \cite{Wearing2006,Aguiar2011,Reich2013}.
+However, the observed seasonality in transmission likely emerges from the combined effects of multiple seasonally-varying components that may be driven by different environmental factors and oscillate in different phases (e.g., mosquito abundance seems to lag slightly behind temperature-driven variation in extrinsic incubation period, Figure \ref{timeseries}).
+Integrating these seasonally-varying components into a synthetic measure of transmission potential (e.g., a temperature-dependent effective reproduction number, \cite{Codeco2018}), or more specifically, a measure of the transmission potential of mosquitoes, is difficult.
 
-The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state.
+\textbf{Mosquito mortality modulates the transmission potential generated by other mosquito and epidemiological processes}
+Here, we positioned the latent mosquito mortality as the link between mosquito abundance, the extrinsic incubation period, and human cases.
+In this way, the estimated mosquito mortality rate serves as an index of transmission potential, opening or closing the mosquito life history bottleneck \cite{Smith2012} as necessary to fit to the case reports data.
+The resulting seasonality in the estimated trajectory suggests that the seasonality in mosquito abundance and the extrinsic incubation period was not sufficient to capture the observed case reports.
+In particular, when mosquitoes were relatively scarce and transmission limited, we estimated a  relatively low mosquito mortality, suggesting that long-lived mosquitoes are required to maintain observed levels of transmission through the off-season.
+On the other hand, when mosquitoes were abundant, we estimated relatively high mosquito mortality rates, suggesting that transmission needed to be damped.
+
+The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state and then bite a susceptible human.
 In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
-Despite this, our simulations suggest that a single pulse of control was relatively less effective when implemented during an outbreak, likely due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
-Instead, we found that adult mosquito control efforts were most effective at reducing the disease burden within a given year when applied roughly 30 weeks prior to the start of that year, on the downswing of the previous year's epidemic and seasonal mosquito dynamics.
-Barsante \emph{et al.} \cite{Barsante2015} similarly found that control was most effective when applied during the dry season (roughly matching our 30 week lag) than during the rainy season.
+The tight correlation between our estimated mosquito mortality and the effect of control fits with this theory, suggesting that control is most effective when it targets long-lived mosquitoes.  
+Specifically, our simulated mosquito control interventions were most effective at reducing the disease burden when applied around week 34 (i.e., early September), in the dengue off-season.
+The effectiveness of this control was likely driven by the fact that transmission during the off-season was already limited by relatively few mosquitoes and a relatively high extrinsic incubation period.
+Given that limitation, transmission was maintained by relatively few long-lived mosquitoes, making the system vulnerable to perturbation.
+
+On the other hand, we found that a single pulse of control was relatively less effective when implemented during an outbreak, when mosquitoes were abundant (and unlikely to be limiting transmission) but short-lived.
+Given the relatively high mosquito mortality rates during this time, exposed and infectious mosquitoes were already fairly ephemeral, such that the relatively small disruption induced by control likely made little difference.
+Moreover, due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population, the population of exposed mosquitoes likely recovered quickly \cite{Newton1992, Burattini2008}.
+Barsante \emph{et al.} \cite{Barsante2015} and Oki \emph{et al.} \cite{Oki2011} similarly found that control was most effective when applied well before peak prevalence, either during the dry season \cite{Barsante2015} or early in the rainy season \cite{Oki2011} (September is near the end of the dry season in Vit\'oria).
 While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicated that disrupting inter-seasonal transmission can be an effective longer-term strategy \cite{Hladish2018}.
-
-The mismatch between the timing of the most effective control and the peak in mosquito abundance suggests that mosquito abundance alone was an insufficient predictor of the effect of mosquito control, at least at the city-scale.
-Instead, the effectiveness of mosquito control depended, in large part, on the intrinsic long-term feedbacks and lags driven by the dynamics of temporary human immunity.
-In fact, control implemented at a 30 week lag was effective largely due to the way it interacted with the dynamics of the susceptible pool. 
-Implementing control when human susceptibility is near a local minimum (i.e., when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) disrupts transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
-The fact that the effect of mosquito control was mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g., cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities \cite{Wearing2006, Reich2013}.
 
@@
 We carried this uncertainty through to our simulations of control interventions \cite{Elderd2006} and found that there was substantial uncertainty in the proportion of cases prevented by a control intervention (i.e., the width of the ribbons in Figure \ref{control}).
-Much of this uncertainty could be attributed to uncertainty in and covariation between the period of cross-immunity and the average mosquito mortality rate (Figure \ref{immunity}).
-Specifically, we found that control implemented at 30 weeks prior to the start of the year (approximately the average optimum timing) had the largest impact in simulations with a low period of cross-immunity and a high average mosquito mortality rate.
-In these simulations, the rapid loss of immunity meant that human susceptibles were abundant and mosquito survival was instead limiting.
-Given this limitation, removing adult mosquitoes further tightened the bottleneck imposed by the race between mosquito survival and the extrinsic incubation period, resulting in a larger disruption to transmission.
+Much of this uncertainty could be attributed to posterior uncertainty in the case reporting rate ($\phi$) and the average mosquito mortality rate ($d_0$).
+Specifically, we found that control implemented at the overall optimum (week 34) had the largest impact (i.e., the lowest case ratio) in simulations with a low average  mosqutio mortality and/or a high case reporting rate (Figure \ref{immunity}).
+Mosquito mortality and the case reporting rate were correlated with the overall level of mosquito abundance and the overall size of the susceptible population, respectively, suggesting that control was most effective in simulations with fewer mosquitoes and more susceptible humans.
 Hladish \emph{et al.} \cite{Hladish2018} similarly found that simulated indoor residual spraying was more effective when the modeled mosquito abundance was already low.
-This suggests that control efforts that reduce the ability of mosquitoes to transmit DENV are going to be most effective in situations in which mosquitoes are already the limiting component to maintaining transmission.
+This suggests that control efforts that reduce the ability of mosquitoes to transmit DENV are going to be most effective in situations in which mosquito abundance is already the limiting component to maintaining transmission (relative to other factors like human immunity).
 
-The posterior correlation between the effectiveness of control, the period of cross-immunity, and the average mosquito mortality rate emphasizes that the impact of mosquito control is jointly regulated by both mosquito population dynamics and human immune processes.
-ten Bosch \emph{et al.} \cite{TenBosch2016} observed similar behavior in their models of dengue dynamics, and found that longer periods of cross-immunity generated systems in which transmission was more difficult to disrupt with control actions.
-As a result of these relationships, efficient deployment of mosquito control, and accurate prediction of its effects, is likely to depend on our ability to monitor, predict, and manipulate the dynamics of human immunity.
-To meet these needs, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance \cite{Morrison2008} and tighter estimates of the period of cross-immunity \cite{TenBosch2016}.
+The posterior correlation between the effectiveness of control, the case reporting rate, and the average mosquito mortality rate emphasizes that the impact of mosquito control is jointly regulated by both mosquito population dynamics and human immune processes.
+ten Bosch \emph{et al.} \cite{TenBosch2016} observed similar behavior in their models of dengue dynamics, and found that longer periods of cross-immunity (and thus fewer susceptibles) generated systems in which transmission was more difficult to disrupt with control actions.
+As a result of these relationships, efficient deployment of mosquito control, and accurate prediction of its effects, is likely to depend on our ability to monitor and predict the dynamics of human immunity.
+To meet these needs, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance \cite{Morrison2008} and tighter estimates of the period of cross-immunity \cite{TenBosch2016} and the number of unreported cases \cite{Silva2016}..
 In the absence of, or as a supplement to, such data, mechanistic models like the one developed here, or so-called TSIR (Time-series Susceptiple-Infected-Recovered) frameworks that reconstruct the dynamics of the susceptible class \cite{Finkenstadt2000, Reich2013}, need to be further developed to better inform and understand mosquito control efforts.
@@
 
-Given the central role that mosquito lifespan plays in driving dengue dynamics, we modeled fluctuations in the mosquito mortality rate using a flexible, forced harmonic oscillator.
-This encoded our prior expectation that the mosquito mortality rate should be driven by seasonal environmental forcing (i.e., it should oscillate with a period of one year), while allowing the flexibility to diverge from the simple oscillator and learn from the case reports data.
-The fact that the mosquito mortality forcing terms ($\epsilon_{\nu}$, Eq.
-13 and 16) were the only source of variability in the transmission process, however, means that the estimated mosquito mortality time series could have been soaking up other sources of stochasticity or model misspecification.
+The tight correlation between the estimated mosquito mortality rate and the simulated effect of control suggests that mosquito control was most effective when it targeted long-lived mosquitoes during the inter-epidemic periods when mosquito abundance was low.
+However, the fact that the mosquito mortality forcing terms ($\epsilon_{\nu}$, Eq.
+13 and 16) were the only source of variability in the transmission process, means that the estimated mosquito mortality time series could have been soaking up other sources of stochasticity or model misspecification.
 Hooker and Ellner \cite{Hooker2015} provide a framework for diagnosing such model misspecification in differential equation models using forcing functions similar to our implementation of the $\epsilon_{\nu}$.
@@
 The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
-These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control interventions), or epidemiological processes (e.g., multiple circulating serotypes of the dengue virus).
+These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control intervention), or epidemiological processes (e.g., multiple circulating serotypes of the dengue virus, subsets of the population with different risk levels).
 
-Despite this suggestion of potential model misspecification, our estimated mosquito mortalities nonetheless fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
-Moreover, the fact that our estimated mortality rate increased with temperature also broadly agrees with the empirical literature on mosquito survival \cite{Yang2009, Brady2013}.
-This suggests that regardless of unexplained structure in the forcing terms, the pattern of case reports was still very well described by realistic fluctuations in the mosquito mortality rate.
+Despite this suggestion of potential sources of model misspecification, our estimated mosquito mortalities nonetheless fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
+Moreover, the fact that our estimated mortality rate increased with temperature also broadly agrees with the empirical literature on mosquito survival \cite{Yang2009, Brady2013, Morin2013}.
+This suggests that regardless of unexplained structure in the forcing terms, the pattern of case reports was still very well described by realistic seasonal fluctuations in the mosquito mortality rate.
+Given the importance of seasonality in understanding dengue epidemiology \cite{TenBosch2016}, and the potential importance of mosquito mortality in driving the effect of control interventions, future work should focus on developing a more complete, predictive understanding of the seasonal drivers of mosquito mortality (including, potentially, control itself).
 
@@
 
+In addition to epidemiological complexity, dengue dynamics are further complicated by the reporting process.
+We estimated a relatively long average intrinsic incubation period ($1/\rho$, posterior mean of 1.7 weeks) relative to our prior mean (0.87 weeks), suggesting possible reporting delays \cite{Reich2016}.
 The case data to which we fit the model represent reports of ``dengue-like illness,'' without laboratory confirmation, and as such could include cases of other diseases with similar symptoms (e.g., chikungunya or Zika).
@@
 
-In addition, it is important to note that the data to which we fit our model implicitly reflect the control efforts already enacted by the city.
+Mosquito control interventions can prevent cases by acting on any of the components of vectorial capacity.
+We focused on the direct effect of killing adult mosquitoes on transmission, but adult control can also act by reducing egg laying and the number of mosquitoes in the next generation \cite{Brady2016}.
+Given the relative simplicity of our mosquito model, we were unable to eplore the feedbacks that adult control may induce in mosquito population dynamics, and we instead assume that mosquito populations quickly rebound from any perturbations \cite{Focks1987}.
+We expect that capturing these feedbacks would likely reinforce our conclusions about the utility of off-season control, as the disruption to mosquito population dynamics would more strongly limit the ability of the mosquito population to maintain transmission through the off-season.
+
+Although our results suggest that a pulse of adult control in the off-season may be an effective tool for preventing human cases, achieving particular control thresholds or policy goals will likely require deploying a combination of control interventions \cite{Brady2016}.
+In fact, it is important to note that the data to which we fit our model implicitly reflect the control efforts already enacted by the city (the effects of which may have influenced our estimates of mosquito demographic rates).
+Thus, our mosquito control simulations should be interepreted as exploring the effect of an additional pulse of city-wide control in addition to the existing control activities.
 These control efforts are guided by the MI-Dengue system, which uses the trap-level mosquito surveillance data to target areas of high mosquito infestation for control (including source reduction, larvacide, and adulticide)\cite{Eiras2009}.
-These control efforts may affect our estimates of the mosquito demographic rates, but should not influence our understanding of dengue dynamics.
-We are focused not on the factors driving mosquito dynamics, but rather on how the realized mosquito dynamics (whether forced by human intervention or environmental factors) fit into the broader epidemiological picture.
-However, our mosquito control simulations should be interepreted as exploring the effect of an additional pulse of city-wide control in addition to the existing control activities.
+Although these targeted, reactive interventions likely help reduce local disease risk at times and locations of high mosqutio abundance \cite{Pepin2013}, our results suggest the utility of an additional pulse of proactive control in the off-season when mosquitoes are less abundant.
 
-We view our simulations of this additional pulse of control primarily as a tool to explore and exploit the long-term feedbacks in the city's epidemiological system, whereas existing control is deployed largely to reduce local (in space and time) disease risk.
 The frequent use of spatially-targeted mosquito control highlights the potential for spatial heterogeneities in disease risk within a city.
@@
 Further, although our mechanistic model may be able to suggest when a city-wide intervention is likely to be effective, to make the best use of limited resources, spatial prioritization may still be necessary.
-Our results emphasize that work to develop the tools necessary for this prioritization should focus not just on prioritization based on the distribution of mosquito abundance, but also on the distribution of human immunity.
 
@@
 We have developed a simple yet realistic mechanistic model of dengue fever spread that represents the fundamental elements of our prior understanding of dengue epidemiology, while also allowing for uncertainty and flexibility in the fluctuations of mosquito demographic rates.
-This mechanistic framework allowed us to capture the critical role of human immunity in mediating the transmission process and to identify critical intervention points that would not be apparent otherwise.
+This mechanistic framework allowed us to capture the critical contribution of long-lived, off-season mosquitoes to the maintenance of transmission and to identify critical intervention points that would not be apparent otherwise.
 The fully hierarchical Bayesian framework in which we embedded the mechanistic model allowed for a thorough accounting of uncertainty that was carried through to the evaluation of different control strategies.

### 2020-05-01 Manuscript/supplement.tex
Updating the supplement

@@
 -\log(\phi_q) & \sim \text{Half-normal}(13, 0.5)\\
+\phi & \sim \text{Beta}(7, 77)\\
 \eta_y &\sim \text{Half-normal}(0, 5)\\
@@
 Since $ E[q_t] = \phi_q \tau V_{Nt} N$, we set the prior on $\log(\phi_q)$ to have a mean of $\log(q_1) - \log(\tau_1) - \log(V_{N0}) - \log(N) = -13$ and a variance to match the prior on $\log(V_{N0})$.
+\\
+\item \textbf{Case reporting probability} ($\phi$): Silva \emph{et al.} \cite{Silva2016} estimate a reporting rate of 1/12 (0.083) in Brazil.
+Because the reporting rate is bound between 0 and 1, we used a Beta-distribution for our prior, parameterized to have a mean of 1/12, with most of the mass between 1/6 and 1/24:
+\begin{equation}
+\phi \sim \text{Beta}(7, 77)
+\end{equation} 
 \\
@@
 
+\begin{figure}[!h]
+\includegraphics[angle = 270, scale = 1]{figures/figS9.eps}
+\caption{
+Posterior estimates of the mosquito mortality rate and the probability of a mosquito surving the external incubation period.
+The black line indicates the posterior median, while the gray ribbon shows the 90$\%$ credible interval.
+The probability of surviving the EIP was computed by taking the ratio of the number of mosquitoes that entered the infectious class in a week over the total number of mosquitoes that exited the exposed class that week.
+That is, the probability of surviving the EIP represents the faction of mosquitoes leaving the exposed class that enter the infectious class.
+}
+\end{figure}
+
 \end{document}

### 2020-05-01 Analysis/plots.r

@@ -85,12 +85,6 @@ dv <- system[, 2:244, 12] %>%
   mutate(dvnat = dvnat * dvmu) %>% 
   join(temp)
 
-moving %>% 
-  ggplot(aes(week, dvnat)) + 
-  stat_summary(fun.y = "median", geom = "line") + 
-  stat_summary(aes(week, mosq), fun.y = "median", geom = "line", color = "dodgerblue") +
-  facet_grid(year ~.)
-
 # Plotting
 yhat %>% 
   ggplot(aes(week, ymin = ymin, ymax = ymax)) +
@@ -516,3 +510,25 @@ hist(apply(qrep, 1, min), main = "", xlab = "minimum weekly trap count", freq = 
 abline(v = min(post$qobs), lwd = 2)
 
 dev.off()
+
+# Figure S9: Probability of surviving EIP ======================================
+
+postscript("Manuscript/figures/figS9.eps",
+           width = 5, height = 3,
+           family = "ArialMT")
+
+moving %>% 
+  ggplot(aes(control, psurv)) + 
+  stat_summary(fun.ymin = function(x) quantile(x, 0.05),
+               fun.ymax = function(x) quantile(x, 0.95),
+               geom = "ribbon", fill = "gray70") +
+  stat_summary(fun.y = "median", geom = "line") + 
+  theme_classic() + 
+  scale_x_continuous(expand = c(0, 1), breaks = c(1, 54, 106, 158, 210), labels = c(2008, 2009, 2010, 2011, 2012)) +
+  xlab("year") + 
+  ylab("probability of surviving EIP") -> figS9a
+
+fig1d + ggtitle("") + figS9a + plot_layout(ncol = 1)
+  
+dev.off()
+

### 2020-05-02 Manuscript/ms.tex
Tidying

@@
 Differential equation models provide a natural way to describe mechanistic processes, but that description must also account for sources of uncertainty \cite{Hotelling1927, Wikle2010}.
-In particular, the values of parameters (e.g., the length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
+In particular, the values of parameters (e.g., the average length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
 The structure of the processes themselves can be uncertain \cite{Ellner1998}, and needs to be informed by available, often noisy, data.
@@
 The human population size and average life span (which we use to parameterize the birth/death rate) for Vitoria were taken from the 2010 census.
-To maintain identifiability, the transmission rate ($\lambda$) and case reporting probability ($\phi_y$) were also fixed at literature values.
+To maintain identifiability, the transmission rate ($\lambda$) was also fixed at literature values.
 Lastly, the extrinsic incubation period in mosquitoes was modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground \cite{weather}.
 
-The remaining parameters include the epidemiological parameters controlling the average latent, infectious, and immune periods ($\rho$, $\gamma$, $\delta$) and average mosquito lifespan ($d_0$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).
+The remaining parameters include the epidemiological parameters controlling the average latent, infectious, and immune periods ($\rho$, $\gamma$, $\delta$) and average mosquito lifespan ($d_0$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_y$, $\phi_q$, $\eta_y$, $\eta_q$).
 Where possible, we specified informative prior distributions on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).
@@
 \caption{Model parameters and their values.
-Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the mean of the prior distribution.
+Parameters above the rule are fixed, while parameters below the rule are random.
+The values in parentheses after the posterior means give the 80$\%$ credible interval.
 See Supplemental Material for a full description of all prior distributions.}
@@
 Samples from the posterior distribution were thus generated using HMC implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R \cite{R2016}. 
-We ran 3 chains with different starting values for 5,000 iterations each, discarding the first 2,500 as burn-in.
+We ran 3 chains with different starting values for 4,000 iterations each, discarding the first 2,000 as burn-in.
 Convergence diagnostics and mixing were evaluated using the shinystan package \cite{shinystan}.
@@
 
-Given samples from the posterior distribution as obtained above, we simulated the effects of a single pulse of mosquito control applied in each week of the first four years of the time series.
-Because the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in future years.
+Given a subset of the samples from the posterior distribution as obtained above (2000, taken to reduce computation time), we simulated the effects of a single pulse of mosquito control applied in each week of the first three years of the time series.
+Because the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden over the following year.
 For each week and each posterior sample, we simulated the dynamics resulting from a $5\%$ reduction in the mosquito population implemented at the beginning of that week (affecting susceptible, exposed, and infectious mosquitoes equally).
 To capture the likely rapid rebound in mosquito abundance following a single pulse of control \cite{Focks1987, Burattini2008}, we also simulated a $5\%$ increase in mosquito birth rate in the following week to return mosquito abundance to its previous trajectory (without this, the $5\%$ reduction in abundance persists indefinitely).
-We then compared the number of cases produced during each year in the control scenario to the same number in the uncontrolled scenario.
+We then compared the number of cases produced over the year following the control intervention in the control scenario to the same number in the uncontrolled scenario.
 A $5\%$ reduction in mosquito abundance was chosen to keep our simulations conservative relative to field estimates of the mortality induced by spraying \cite{Esu2010}, and to avoid pushing the model into the unrealistic range of dengue eradication.
@@
 \caption{{\bf Mosquito mortality and temperature.}
-Posterior median mosquito mortality rate as a function of weekly mean temperature (Celsius).
+Posterior median mosquito mortality rate as a function of weekly mean temperature (degrees Celsius).
 }
@@
 As such, the seasonal variation in the effectiveness of control followed the same trend as mosquito mortality rate (though the annual minima in the case ratio time series generally fell 1 to 2 weeks before the minima in the mosquito mortality time series), with the largest reductions in case load coming from interventions during the dengue off-season (early September for 2008 and 2009, and mid-July for 2010).
-Summing over the interannual variation to compute the overall effect of control implemented in a given week of the year, we found that mosquito control was most effective when implemented in week 34 (late August), reducing the estimated case load by roughly $14\%$ (Figure \ref{control}, bottom panel).
-This broadly corresponds to the time of year when both dengue case reports and mosquito trap counts are low.
+Summing over the interannual variation to compute the overall effect of control implemented in a given week of the year, we found that mosquito control was most effective when implementedaround  week 34 (late August/early September), reducing the estimated case load by roughly $14\%$ (Figure \ref{control}, bottom panel).
+This broadly corresponds to the end of the dry season in Vit\'oria, when both dengue case reports and mosquito trap counts are low.
 
@@
 
-The variation around the predicted effect of control implemented in week 34 of the year (i.e., the width of ribbon in Figure \ref{control}) was correlated with the baseline mosquito mortality rate ($d_0$) and the case reporting probability ($\phi$).
+The variation around the posterior predicted effect of control implemented in week 34 of the year (i.e., the width of ribbon in Figure \ref{control}) was correlated with the baseline mosquito mortality rate ($d_0$) and the case reporting probability ($\phi$).
 Simulated mosquito control created the largest reduction in case reports in posterior samples with low baseline mosquito mortality rate and/or high reporting probability, while control was relatively less effective in simulations from samples with high mosquito mortality or low reporting probability (Figure \ref{correlations}).
@@
 The dynamics of dengue fever, like those of many infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
-%In the dengue system, these drivers include waning human immunity, the race between the extrinsic incubation period of the virus and mosquito lifespan, and the seasonal and stochastic variation in these rates.
 Seasonality in particular is an important factor in capturing the annual cycle of dengue outbreaks \cite{Wearing2006,Aguiar2011,Reich2013}.
@@
 
-\textbf{Mosquito mortality modulates the transmission potential generated by other mosquito and epidemiological processes}
 Here, we positioned the latent mosquito mortality as the link between mosquito abundance, the extrinsic incubation period, and human cases.
@@
 The resulting seasonality in the estimated trajectory suggests that the seasonality in mosquito abundance and the extrinsic incubation period was not sufficient to capture the observed case reports.
-In particular, when mosquitoes were relatively scarce and transmission limited, we estimated a  relatively low mosquito mortality, suggesting that long-lived mosquitoes are required to maintain observed levels of transmission through the off-season.
+In particular, when mosquitoes were relatively scarce and transmission limited, we estimated a  relatively low mosquito mortality, suggesting that long-lived mosquitoes were required to maintain observed levels of transmission through the off-season.
 On the other hand, when mosquitoes were abundant, we estimated relatively high mosquito mortality rates, suggesting that transmission needed to be damped.
@@
 Specifically, our simulated mosquito control interventions were most effective at reducing the disease burden when applied around week 34 (i.e., early September), in the dengue off-season.
-The effectiveness of this control was likely driven by the fact that transmission during the off-season was already limited by relatively few mosquitoes and a relatively high extrinsic incubation period.
+The effectiveness of this control was likely driven by the fact that transmission during the off-season was already limited by low mosquito abundance and a relatively high extrinsic incubation period.
 Given that limitation, transmission was maintained by relatively few long-lived mosquitoes, making the system vulnerable to perturbation.
@@
 While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicated that disrupting inter-seasonal transmission can be an effective longer-term strategy \cite{Hladish2018}.
+Further, although large pulses of imported cases could potentially swamp the effects of early control, in additional simulations we found that our results were robust to the import 10 infectious humans just before the annual outbreak.
 
@@
 The posterior correlation between the effectiveness of control, the case reporting rate, and the average mosquito mortality rate emphasizes that the impact of mosquito control is jointly regulated by both mosquito population dynamics and human immune processes.
-ten Bosch \emph{et al.} \cite{TenBosch2016} observed similar behavior in their models of dengue dynamics, and found that longer periods of cross-immunity (and thus fewer susceptibles) generated systems in which transmission was more difficult to disrupt with control actions.
-As a result of these relationships, efficient deployment of mosquito control, and accurate prediction of its effects, is likely to depend on our ability to monitor and predict the dynamics of human immunity.
-To meet these needs, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance \cite{Morrison2008} and tighter estimates of the period of cross-immunity \cite{TenBosch2016} and the number of unreported cases \cite{Silva2016}..
+ten Bosch \emph{et al.} \cite{TenBosch2016} observed similar behavior in their models of dengue dynamics, and found that longer periods of cross-immunity (such that susceptibles replenished more slowly) generated systems in which transmission was more difficult to disrupt with control actions.
+As a result of these relationships, efficient deployment of mosquito control, and accurate prediction of its effects, is likely to depend in part on our ability to monitor and predict the dynamics of human immunity.
+To meet these needs, existing mosquito monitoring efforts need to be paired with more detailed clinical surveillance \cite{Morrison2008} and tighter estimates of the period of cross-immunity \cite{TenBosch2016} and the number of unreported cases \cite{Silva2016}.
 In the absence of, or as a supplement to, such data, mechanistic models like the one developed here, or so-called TSIR (Time-series Susceptiple-Infected-Recovered) frameworks that reconstruct the dynamics of the susceptible class \cite{Finkenstadt2000, Reich2013}, need to be further developed to better inform and understand mosquito control efforts.
@@
 The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
-These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control intervention), or epidemiological processes (e.g., multiple circulating serotypes of the dengue virus, subsets of the population with different risk levels).
+These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control intervention), or epidemiological processes (e.g., multiple circulating serotypes of the dengue virus, subsets of the population with different mixing or risk levels).
 
@@
 We estimated a relatively long average intrinsic incubation period ($1/\rho$, posterior mean of 1.7 weeks) relative to our prior mean (0.87 weeks), suggesting possible reporting delays \cite{Reich2016}.
-The case data to which we fit the model represent reports of ``dengue-like illness,'' without laboratory confirmation, and as such could include cases of other diseases with similar symptoms (e.g., chikungunya or Zika).
+Moreover, the case data to which we fit the model represent reports of ``dengue-like illness,'' without laboratory confirmation, and as such could include cases of other diseases with similar symptoms (e.g., chikungunya or Zika).
 However, neither chikungunya nor Zika had emerged as substantial public health threats in Brazil by the end of our time series in late 2012 \cite{PAHO2014, PAHO2015}.
@@
 We focused on the direct effect of killing adult mosquitoes on transmission, but adult control can also act by reducing egg laying and the number of mosquitoes in the next generation \cite{Brady2016}.
-Given the relative simplicity of our mosquito model, we were unable to eplore the feedbacks that adult control may induce in mosquito population dynamics, and we instead assume that mosquito populations quickly rebound from any perturbations \cite{Focks1987}.
+Given the relative simplicity of our mosquito model, we were unable to eplore the feedbacks that adult control may induce in mosquito population dynamics, and instead assumed that mosquito populations quickly rebound from any perturbations \cite{Focks1987}.
 We expect that capturing these feedbacks would likely reinforce our conclusions about the utility of off-season control, as the disruption to mosquito population dynamics would more strongly limit the ability of the mosquito population to maintain transmission through the off-season.

### 2020-05-02 Manuscript/ms.tex
More tidying

@@
 
-The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state and then bite a susceptible human.
+The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state and bite a susceptible human.
 In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
@@
 While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicated that disrupting inter-seasonal transmission can be an effective longer-term strategy \cite{Hladish2018}.
-Further, although large pulses of imported cases could potentially swamp the effects of early control, in additional simulations we found that our results were robust to the import 10 infectious humans just before the annual outbreak.
+Further, although large pulses of imported cases could potentially swamp the effects of early control, in additional simulations we found that our results were robust to the import of 10 infectious humans just before the annual outbreak.
 
@@
 Much of this uncertainty could be attributed to posterior uncertainty in the case reporting rate ($\phi$) and the average mosquito mortality rate ($d_0$).
-Specifically, we found that control implemented at the overall optimum (week 34) had the largest impact (i.e., the lowest case ratio) in simulations with a low average  mosqutio mortality and/or a high case reporting rate (Figure \ref{immunity}).
+Specifically, we found that control implemented at the overall optimum (week 34) had the largest impact (i.e., the lowest case ratio) in simulations with a low average mosqutio mortality and/or a high case reporting rate (Figure \ref{correlations}).
 Mosquito mortality and the case reporting rate were correlated with the overall level of mosquito abundance and the overall size of the susceptible population, respectively, suggesting that control was most effective in simulations with fewer mosquitoes and more susceptible humans.

### 2020-05-02 Manuscript/supplement.tex
More tidying

@@
 \\
-\item \textbf{Reporting probability} ($\phi_y$): Silva \textit{et al.} \cite{Silva2016} estimate that there are 12 cases of dengue for every case reported to Brazil's Notifiable Diseases Information System (SINAN, the source of our data), so we fix $\phi_y = 1/12$.
-\\
 \item \textbf{Extrinsic incubation period} ($1/\rho_{vt}$): Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito  (the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
@@
 Posterior estimates of the mosquito mortality rate and the probability of a mosquito surving the external incubation period.
-The black line indicates the posterior median, while the gray ribbon shows the 90$\%$ credible interval.
+The black line indicates the posterior median, while the gray ribbon shows the 80$\%$ credible interval.
 The probability of surviving the EIP was computed by taking the ratio of the number of mosquitoes that entered the infectious class in a week over the total number of mosquitoes that exited the exposed class that week.

### 2020-05-02 Analysis/plots.r

@@ -202,8 +202,8 @@ postscript("Manuscript/figures/fig3.eps",
 overall %>% 
   subset(year %in% c("2008", "2009", "2010", "overall") & week < 53) %>%
   ggplot(aes(week, ratio)) +
-  stat_summary(fun.ymin = function(x) quantile(x, 0.05),
-               fun.ymax = function(x) quantile(x, 0.95),
+  stat_summary(fun.ymin = function(x) quantile(x, 0.1),
+               fun.ymax = function(x) quantile(x, 0.9),
                geom = "ribbon", fill = "gray70") +
   stat_summary(fun.y = "median", geom = "line") +
   facet_grid(year ~.) +
@@ -514,13 +514,13 @@ dev.off()
 # Figure S9: Probability of surviving EIP ======================================
 
 postscript("Manuscript/figures/figS9.eps",
-           width = 5, height = 3,
+           width = 6, height = 5,
            family = "ArialMT")
 
 moving %>% 
   ggplot(aes(control, psurv)) + 
-  stat_summary(fun.ymin = function(x) quantile(x, 0.05),
-               fun.ymax = function(x) quantile(x, 0.95),
+  stat_summary(fun.ymin = function(x) quantile(x, 0.1),
+               fun.ymax = function(x) quantile(x, 0.9),
                geom = "ribbon", fill = "gray70") +
   stat_summary(fun.y = "median", geom = "line") + 
   theme_classic() + 

### 2020-05-04 Manuscript/ms.tex
Cleaning up

@@
 As such, the ability of mosquitoes to contribute to DENV transmission depends critically on the level of immunity in the human population \cite{Scott2010a}.
-Further, the cycle of transmission between humans and mosquitoes is influenced not just by mosquito abundance, but also by mosquito survival relative to the virus' incubation period in mosquitoes \cite{Smith2012}.
+Further, the cycle of transmission between humans and mosquitoes is influenced not just by mosquito abundance, but also by mosquito survival relative to the virus incubation period in mosquitoes \cite{Smith2012}.
 In fact, whether or not an exposed mosquito will survive long enough to become infectious represents a critical bottleneck in the transmission process and leads to nonlinear dependence of transmission on mosquito survival \cite{Smith2012}.
@@
 
-In this paper, we aim to integrate these elements -- a detailed mechanistic model of dengue transmission with a full Bayesian accounting of uncertainty --  to better understand the interplay of forces governing dengue dynamics and their interaction with potential vector control interventions.
+In what follows, we integrate these elements -- a detailed mechanistic model of dengue transmission with a full Bayesian accounting of uncertainty --  to better understand the interplay of forces governing dengue dynamics and their interaction with potential vector control interventions.
 We apply this framework to clinical and entomological surveillance data from the city of Vit\'oria, Brazil.
 These data allow us to estimate a latent time series of mosquito mortality rates that modulate the transmission process and link mosquito abundance to human disease.  
-We then use the fitted model to explore how perturbations to the mosquito population propagate and interact with the nonlinearities of dengue transmission in order to better inform mosquito control efforts.
+We then use the fitted model to explore how perturbations to the mosquito population propagate and interact with the nonlinearities of dengue transmission to better inform mosquito control efforts.
 
@@
 Moreover, because dengue case reports do not identify serotype, there is not enough information in the data to inform the dynamics of individual serotypes.
-As such, we captured temporary cross-immunity, and the potential for multiple sequential infections, as simply and tractably as possible in a susceptible-exposed-infectious-recovered-susceptible (SEIRS) compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010}.
-Though it captures the critical influence of temporary immunity, this framework does not account for the potential relationship between an individual's infection history and the likelihood that a new infection with be symptomatic (and thus reported).
+As such, we captured temporary cross-immunity, and the potential for multiple sequential infections, as simply and tractably as possible in a susceptible-exposed-infectious-recovered-susceptible (SEIRS) compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010, Hooten2010}.
+Although it captures the critical influence of temporary immunity, this framework does not account for the potential relationship between an individual's infection history and the likelihood that a new infection with be symptomatic (and thus reported).
 In particular, secondary infections appear more likely to be symptomatic than primary infections \cite{Imai2016,Clapham2017}, while third and fourth infections appear much less likely to be symptomatic \cite{Olkowski2013} (but see \cite{Montoya2013} who found similar rates of symptomatic cases across infection number).
@@
 Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_E$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
-Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implemented a gamma-distributed incubation period by chaining multiple exposed classes together (taking advantage of the fact that a gamma-distributed random variable can be generated through the sum of a collection of exponential random variables with the same rate parameter) \cite{Lloyd2001}.
+Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implemented a gamma-distributed incubation period by chaining together multiple exposed classes ($V_{Ej}$, taking advantage of the fact that a gamma-distributed random variable can be generated through the sum of exponential random variables with the same rate parameter) \cite{Lloyd2001}.
 Total mosquito population size ($V_N$) is controlled by a forced, seasonally varying growth rate ($r(t)$), while the transmission bottleneck is captured with a forced, seasonally varying mortality rate ($d(t)$).
@@
 \end{align}
-for each week $t = 1 \dots 243$.
+for each week $t = 1, \dots, 243$.
 These stochastically-forced harmonic oscillators provide a flexible framework for generating smooth seasonal oscillations in the latent mosquito processes \cite{Ramsay2017}.
@@
 Several of the parameters in this model are assumed to be fixed and known (Table 1).
-The human population size and average life span (which we use to parameterize the birth/death rate) for Vitoria were taken from the 2010 census.
+The human population size and average life span (which we use to parameterize the birth/death rate) for Vit\'oria were taken from the 2010 census.
 To maintain identifiability, the transmission rate ($\lambda$) was also fixed at literature values.
@@
 The remaining parameters include the epidemiological parameters controlling the average latent, infectious, and immune periods ($\rho$, $\gamma$, $\delta$) and average mosquito lifespan ($d_0$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_y$, $\phi_q$, $\eta_y$, $\eta_q$).
-Where possible, we specified informative prior distributions on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).
+Where possible, we specified informative prior distributions for these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).
 
@@
 
-Combining the data, process, and parameter models \cite{Berliner1996}, we can summarize the full hierarchical model as:
+Combining the data, process, and parameter models \cite{Berliner1996}, we summarize the full hierarchical model as:
 \begin{align}
@@
 Sampling from the posterior distribution of the parameters in mechanistic models is difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations induced by the nonlinearity of the differential equation model \cite{Reilly2005, Girolami2008, Calderhead2011}.
-However, the variable $\epsilon_t$ introduce flexibility to the mechanistic model that 
-soaks up lack-of-fit when the process parameters are far from optimal \cite{Leander2014}, thereby reducing multimodality and helping to smooth the posterior surface.
+However, the variable $\epsilon_t$ introduces flexibility to the mechanistic model that remedies lack-of-fit when the process parameters are far from optimal \cite{Leander2014}, thereby reducing multimodality and helping to smooth the posterior surface.
 Gradient-based methods like Hamiltonain Monte Carlo (HMC) can then more easily and efficiently traverse the posterior.
-Samples from the posterior distribution were thus generated using HMC implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R \cite{R2016}. 
+Samples from the posterior distribution were generated using HMC implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R \cite{R2016}. 
 We ran 3 chains with different starting values for 4,000 iterations each, discarding the first 2,000 as burn-in.
 Convergence diagnostics and mixing were evaluated using the shinystan package \cite{shinystan}.
-Within stan, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.  
+In our implementation, the solution to the differential equation model was approximated with an Euler scheme with a time step of 1 day.  
 Timesteps as small as $1/8$ of a day were explored and did not qualitatively change the modeled dynamics.
@@
 In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 6 - 8).
-The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that the Vitoria data contained little additional information about these parameters (Supplemental Figure 3).
+The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that the Vit\'oria data contained little additional information about these parameters (Supplemental Figure 3).
 The estimated latent period in a human host ($1/\rho$, the expected time it takes for an exposed human to become infectious to biting mosquitoes) was influenced more strongly by the data, with a posterior mean of 1.73 weeks compared to a prior mean of 0.87 weeks.
@@
 \includegraphics[scale = 1]{figures/fig1.eps}
-\caption{{\bf Vitoria data and model estimates.}
+\caption{{\bf Vit\'oria data and model estimates.}
 A: weekly observed case reports (points), with corresponding posterior median (black line) and 80\% posterior credible interval (gray band).
@@
 
-The effect of a given mosquito control intervention (i.e.
-the temporary removal of $5\%$ of the adult population in a given week) on the number of cases in the following year (relative to no control) varied both seasonally and interannually (Figure \ref{control}).
+The effect of a given mosquito control intervention (i.e., the temporary removal of $5\%$ of the adult population in a given week) on the number of cases in the following year (relative to no control) varied both seasonally and interannually (Figure \ref{control}).
 This variation in the effect of control was tightly correlated with the estimated mosquito mortality, with a median posterior correlation between the two time series of 0.96.
-As such, the seasonal variation in the effectiveness of control followed the same trend as mosquito mortality rate (though the annual minima in the case ratio time series generally fell 1 to 2 weeks before the minima in the mosquito mortality time series), with the largest reductions in case load coming from interventions during the dengue off-season (early September for 2008 and 2009, and mid-July for 2010).
-Summing over the interannual variation to compute the overall effect of control implemented in a given week of the year, we found that mosquito control was most effective when implementedaround  week 34 (late August/early September), reducing the estimated case load by roughly $14\%$ (Figure \ref{control}, bottom panel).
+As such, the seasonal variation in the effectiveness of control followed the same trend as mosquito mortality rate (although the annual minima in the case ratio time series generally fell 1 to 2 weeks before the minima in the mosquito mortality time series), with the largest reductions in case load resulting from interventions during the dengue off-season (early September for 2008 and 2009, and mid-July for 2010).
+Summing over the interannual variation to compute the overall effect of control implemented in a given week of the year, we found that mosquito control was most effective when implemented around  week 34 (late August/early September), reducing the estimated case load by roughly $14\%$ (Figure \ref{control}, bottom panel).
 This broadly corresponds to the end of the dry season in Vit\'oria, when both dengue case reports and mosquito trap counts are low.
@@
 
-The variation around the posterior predicted effect of control implemented in week 34 of the year (i.e., the width of ribbon in Figure \ref{control}) was correlated with the baseline mosquito mortality rate ($d_0$) and the case reporting probability ($\phi$).
+The variation associated with the posterior predicted effect of control implemented in week 34 of the year (i.e., the width of ribbon in Figure \ref{control}) was correlated with the baseline mosquito mortality rate ($d_0$) and the case reporting probability ($\phi$).
 Simulated mosquito control created the largest reduction in case reports in posterior samples with low baseline mosquito mortality rate and/or high reporting probability, while control was relatively less effective in simulations from samples with high mosquito mortality or low reporting probability (Figure \ref{correlations}).
@@
 The dynamics of dengue fever, like those of many infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
-Seasonality in particular is an important factor in capturing the annual cycle of dengue outbreaks \cite{Wearing2006,Aguiar2011,Reich2013}.
+Seasonality, in particular, is an important factor in capturing the annual cycle of dengue outbreaks \cite{Wearing2006,Aguiar2011,Reich2013}.
 However, the observed seasonality in transmission likely emerges from the combined effects of multiple seasonally-varying components that may be driven by different environmental factors and oscillate in different phases (e.g., mosquito abundance seems to lag slightly behind temperature-driven variation in extrinsic incubation period, Figure \ref{timeseries}).
@@
 
-Here, we positioned the latent mosquito mortality as the link between mosquito abundance, the extrinsic incubation period, and human cases.
+We positioned the latent mosquito mortality as the link between mosquito abundance, the extrinsic incubation period, and human cases.
 In this way, the estimated mosquito mortality rate serves as an index of transmission potential, opening or closing the mosquito life history bottleneck \cite{Smith2012} as necessary to fit to the case reports data.
@@
 In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
-The tight correlation between our estimated mosquito mortality and the effect of control fits with this theory, suggesting that control is most effective when it targets long-lived mosquitoes.  
+The tight correlation between our estimated mosquito mortality and the effect of control confirms this theory, suggesting that control is most effective when it targets long-lived mosquitoes.  
 Specifically, our simulated mosquito control interventions were most effective at reducing the disease burden when applied around week 34 (i.e., early September), in the dengue off-season.
@@
 Much of this uncertainty could be attributed to posterior uncertainty in the case reporting rate ($\phi$) and the average mosquito mortality rate ($d_0$).
-Specifically, we found that control implemented at the overall optimum (week 34) had the largest impact (i.e., the lowest case ratio) in simulations with a low average mosqutio mortality and/or a high case reporting rate (Figure \ref{correlations}).
+Specifically, we found that control implemented at the overall optimum (week 34) had the largest impact (i.e., the lowest case ratio) in simulations with a low average mosquito mortality and/or a high case reporting rate (Figure \ref{correlations}).
 Mosquito mortality and the case reporting rate were correlated with the overall level of mosquito abundance and the overall size of the susceptible population, respectively, suggesting that control was most effective in simulations with fewer mosquitoes and more susceptible humans.
 Hladish \emph{et al.} \cite{Hladish2018} similarly found that simulated indoor residual spraying was more effective when the modeled mosquito abundance was already low.
-This suggests that control efforts that reduce the ability of mosquitoes to transmit DENV are going to be most effective in situations in which mosquito abundance is already the limiting component to maintaining transmission (relative to other factors like human immunity).
+This suggests that control efforts that reduce the ability of mosquitoes to transmit DENV are most effective for situations in which mosquito abundance is already the limiting component to maintaining transmission (relative to other factors like human immunity).
 
@@
 However, the fact that the mosquito mortality forcing terms ($\epsilon_{\nu}$, Eq.
-13 and 16) were the only source of variability in the transmission process, means that the estimated mosquito mortality time series could have been soaking up other sources of stochasticity or model misspecification.
+13 and 16) were the only source of variability in the transmission process implies that the estimated mosquito mortality time series could have absorbed other sources of stochasticity or model misspecification.
 Hooker and Ellner \cite{Hooker2015} provide a framework for diagnosing such model misspecification in differential equation models using forcing functions similar to our implementation of the $\epsilon_{\nu}$.
@@
 
-Despite this suggestion of potential sources of model misspecification, our estimated mosquito mortalities nonetheless fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
+Despite these potential sources of model misspecification, our estimated mosquito mortalities nonetheless fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
 Moreover, the fact that our estimated mortality rate increased with temperature also broadly agrees with the empirical literature on mosquito survival \cite{Yang2009, Brady2013, Morin2013}.
@@
 
-\subsection*{Caveats and extensions}
+\subsection*{Additional considerations and extensions}
 
@@
 These control efforts are guided by the MI-Dengue system, which uses the trap-level mosquito surveillance data to target areas of high mosquito infestation for control (including source reduction, larvacide, and adulticide)\cite{Eiras2009}.
-Although these targeted, reactive interventions likely help reduce local disease risk at times and locations of high mosqutio abundance \cite{Pepin2013}, our results suggest the utility of an additional pulse of proactive control in the off-season when mosquitoes are less abundant.
+Although these targeted, reactive interventions likely help reduce local disease risk at times and locations of high mosquito abundance \cite{Pepin2013}, our results suggest the utility of an additional pulse of proactive control in the off-season when mosquitoes are less abundant.
 
@@
 In addition, spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
-Although our model does not account for these heterogeneities, it was nontheless able to capture the city-wide dynamics reasonably well, suggesting that there may be sufficient mixing to appear homogeneous at the city scale.
+Although our model does not account for these heterogeneities, it was nontheless able to capture the city-wide dynamics well, suggesting that there may be sufficient mixing to appear homogeneous at the city scale.
 Further, although our mechanistic model may be able to suggest when a city-wide intervention is likely to be effective, to make the best use of limited resources, spatial prioritization may still be necessary.
@@
 The challenges presented by these confounding factors highlight the value of mechanistic information in studying the effect of mosquito control on disease spread.
-In the Bayesian context we have deployed here, this mechanistic knowledge, as formalized in the specification of a differential equation model, can be viewed as part of the prior knowledge on the relationship between mosquito abundance and human disease \cite{Ellner1998, Wikle2010}.
-As such, we should neither ignore this mechanistic information, nor encode it so rigidly that it swamps the signal in our data.
+In the Bayesian context we deployed, this mechanistic knowledge, as formalized in the specification of a differential equation model, can be viewed as part of the prior knowledge on the relationship between mosquito abundance and human disease \cite{Ellner1998, Wikle2010}.
+As such, we should neither ignore this mechanistic information, nor encode it so rigidly that it overwhelms the signal in our data.
 
-We have developed a simple yet realistic mechanistic model of dengue fever spread that represents the fundamental elements of our prior understanding of dengue epidemiology, while also allowing for uncertainty and flexibility in the fluctuations of mosquito demographic rates.
+We developed a simple yet realistic mechanistic model of dengue fever spread that represents the fundamental elements of our prior understanding of dengue epidemiology, while also allowing for uncertainty and flexibility in the fluctuations of mosquito demographic rates.
 This mechanistic framework allowed us to capture the critical contribution of long-lived, off-season mosquitoes to the maintenance of transmission and to identify critical intervention points that would not be apparent otherwise.

### 2020-06-17 Manuscript/ms.tex
Typos and small clarity edits

@@
 Moreover, given the short time scale of our data (5 years) relative to the period of cross-immunity (roughly 2 years \cite{Reich2013}), we expect that third and fourth infections will be relatively rare.
-We thus expect the SEIRS framework to be sufficient for capturing the dengue dynamics of Vit\'oria and that reporting differences across infection number are unlikely to have a large impact on the inferred relationship between mosquito abundance and human disease.
+We thus expect the SEIRS framework, and the assumption of equal symptomatic rates across infection number, to be sufficient for capturing the dengue dynamics of Vit\'oria and the relationship between mosquito abundance and human disease.
 
@@
 However, the variable $\epsilon_t$ introduces flexibility to the mechanistic model that remedies lack-of-fit when the process parameters are far from optimal \cite{Leander2014}, thereby reducing multimodality and helping to smooth the posterior surface.
-Gradient-based methods like Hamiltonain Monte Carlo (HMC) can then more easily and efficiently traverse the posterior.
+Gradient-based methods like Hamiltonian Monte Carlo (HMC) can then more easily and efficiently traverse the posterior.
 Samples from the posterior distribution were generated using HMC implemented in the rstan package \cite{Carpenter2016, Rstan2017} for R \cite{R2016}. 
@@
 B: weekly mosquito trap counts (points), with posteriod median (black line) and 80\% posterior credible interval (gray band).
-C: extrinsic incubation period (weeks), computed from weekly mean temperature data.
+C: extrinsic incubation period (EIP; weeks), computed from weekly mean temperature data.
 D: estimated weekly mosquito mortality rate, with the posterior median (black line) and the 80\% posterior credible interval (gray band).
@@
 The estimated weekly mosquito mortality rate varied seasonally, with generally high mortality early in the year and low mortality in August to October (Figure \ref{timeseries}).
-This seasonal trend broadly tracked seasonal variation in temperature (Figure \ref{mortality}) and mosquito abundance (Figure \ref{timeseries}, though the shape of the annual trajectory differed from year to year.
+This seasonal trend broadly tracked seasonal variation in temperature (Figure \ref{mortality}) and mosquito trap counts (Figure \ref{timeseries}, though the shape of the annual trajectory differed from year to year.
 The posterior distribution of the baseline mortality rate ($d_0$) had a mean of 0.88/week, roughly $60\%$ of the prior mean.
@@
 While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicated that disrupting inter-seasonal transmission can be an effective longer-term strategy \cite{Hladish2018}.
-Further, although large pulses of imported cases could potentially swamp the effects of early control, in additional simulations we found that our results were robust to the import of 10 infectious humans just before the annual outbreak.
+Further, although large pulses of imported cases could potentially swamp the effects of early control, in additional simulations we found that our results were robust to the import of 10 infectious humans (roughly the same order as the number of locally reported cases) just before the annual outbreak.
 
@@
 The posterior correlation between the effectiveness of control, the case reporting rate, and the average mosquito mortality rate emphasizes that the impact of mosquito control is jointly regulated by both mosquito population dynamics and human immune processes.
-ten Bosch \emph{et al.} \cite{TenBosch2016} observed similar behavior in their models of dengue dynamics, and found that longer periods of cross-immunity (such that susceptibles replenished more slowly) generated systems in which transmission was more difficult to disrupt with control actions.
+Similar observations were made by ten Bosch \emph{et al.} \cite{TenBosch2016} who found that models with longer periods of cross-immunity (such that susceptibles replenished more slowly) generated systems in which transmission was more difficult to disrupt with control actions.
 As a result of these relationships, efficient deployment of mosquito control, and accurate prediction of its effects, is likely to depend in part on our ability to monitor and predict the dynamics of human immunity.
@@
 We focused on the direct effect of killing adult mosquitoes on transmission, but adult control can also act by reducing egg laying and the number of mosquitoes in the next generation \cite{Brady2016}.
-Given the relative simplicity of our mosquito model, we were unable to eplore the feedbacks that adult control may induce in mosquito population dynamics, and instead assumed that mosquito populations quickly rebound from any perturbations \cite{Focks1987}.
+Given the relative simplicity of our mosquito model, we were unable to explore the feedbacks that adult control may induce in mosquito population dynamics, and instead assumed that mosquito populations quickly rebound from any perturbations \cite{Focks1987}.
 We expect that capturing these feedbacks would likely reinforce our conclusions about the utility of off-season control, as the disruption to mosquito population dynamics would more strongly limit the ability of the mosquito population to maintain transmission through the off-season.

### 2020-06-17 Manuscript/supplement.tex
Typos and small clarity edits

@@
 \\
-\item \textbf{Decay rate of infectiousness} ($\gamma$): Nguyet \emph{et al.} \cite{Nguyet2013} find that there is very little transmission of dengue from humans to mosquitoes after 5-6 days since the onset of fever (with roughly an extra of infectiousness prior to onset of fever).
+\item \textbf{Decay rate of infectiousness} ($\gamma$): Nguyet \emph{et al.} \cite{Nguyet2013} find that there is very little transmission of dengue from humans to mosquitoes after 5-6 days since the onset of fever (with roughly an extra day of infectiousness prior to onset of fever).
 In reality, an individual's infectiousness decays with time, but the ODE model assumes infectiousness is constant for the (exponentially distributed) duration of the infection period.    
@@
 The black line indicates the prior density.
-A: latent period in human host ($1/\rho$), in weeks.
-B: rate of infectious decay in human host ($\gamma$) in weeks $^{-1}$.
-C: period of cross-immunity ($1/\delta$) in weeks.
+A: scaled latent period in human host ($1/\rho$).
+B: scaled rate of infectious decay in human host ($\gamma$).
+C: scaled period of cross-immunity ($1/\delta$).
+D: scaled baseline mosquito mortality rate ($d_0$).
 }
@@
 \caption{Posterior distribution of measurement paramters.
+The black line shows the prior density.
 A: log of the mosquito trap capture rate (which is strongly correlated with initial mosquito abundance).
-The black line shows the prior density.
-B: mosquito trap count measurement over-dispersion (prior density too low to be visible).
-C: case report measurement over-dispersion (prior density too low to be visible).}
+B: human case reporting probability.
+C: mosquito trap count measurement over-dispersion (prior density too low to be visible).
+D: case report measurement over-dispersion (prior density too low to be visible).}
 \end{figure}
@@
 \caption{
-Posterior distribution of the total number of cases reported (A) and the total number of mosquitoes captured (B).  Thick line indicates the observed totals.
+Posterior distribution of the total number of cases reported (A) and the total number of mosquitoes captured (B).  The thick line indicates the observed totals.
 }
@@
 \caption{
-Posterior distributions of the minimum and maximum weekly reported cases and trapped mosquitoes. Thick line indicates the observed minima and maxima.
+Posterior distributions of the minimum and maximum weekly reported cases (A and C) and trapped mosquitoes (B and D). The thick line indicates the observed minima and maxima.
 }
@@
 \caption{
-Posterior estimates of the mosquito mortality rate and the probability of a mosquito surving the external incubation period.
+Posterior estimates of the mosquito mortality rate (top panel) and the probability of a mosquito surving the external incubation period (bottom panel).
 The black line indicates the posterior median, while the gray ribbon shows the 80$\%$ credible interval.
-The probability of surviving the EIP was computed by taking the ratio of the number of mosquitoes that entered the infectious class in a week over the total number of mosquitoes that exited the exposed class that week.
-That is, the probability of surviving the EIP represents the faction of mosquitoes leaving the exposed class that enter the infectious class.
+The probability of surviving the extrinsic incubation period was computed by taking the ratio of the number of mosquitoes that entered the infectious class in a week over the total number of mosquitoes that exited the exposed class that week.
+That is, the probability of surviving the extrinsic incubatio period represents the faction of mosquitoes leaving the exposed class that enter the infectious class.
 }

### 2020-06-17 Analysis/plots.r

@@ -331,8 +331,8 @@ postscript("Manuscript/figures/figS2.eps",
            family = "ArialMT")
 
 par(mfrow = c(1, 2))
-rstan::extract(sim, "sigmadv", permute = T)[[1]] %>% hist(main = "", xlab = expression(sigma[d]), freq = F)
-rstan::extract(sim, "sigmarv", permute = T)[[1]] %>% hist(main = "", xlab = expression(sigma[r]), freq = F)
+rstan::extract(sim, "sigmadv", permute = T)[[1]] %>% hist(main = "A", xlab = expression(sigma[d]), freq = F)
+rstan::extract(sim, "sigmarv", permute = T)[[1]] %>% hist(main = "B", xlab = expression(sigma[r]), freq = F)
 
 dev.off()
 
@@ -346,16 +346,16 @@ postscript("Manuscript/figures/figS3.eps",
 
 par(mfrow = c(2, 2))
 
-hist(epiparams[[1]], main = "", xlab = "scaled latent period", freq = F, breaks = 30, ylim = c(0, 1.2))
+hist(epiparams[[1]], main = "A", xlab = "scaled latent period", freq = F, breaks = 30, ylim = c(0, 1.2))
 curve(dgamma(x, 8.3, 8.3), add = T, lwd = 2)
 
-hist(epiparams[[2]], main = "", xlab = "scaled rate of infectious decay", freq = F, breaks = 30)
+hist(epiparams[[2]], main = "B", xlab = "scaled rate of infectious decay", freq = F, breaks = 30)
 curve(dgamma(x, 100, 100), add = T, lwd = 2)
 
-hist(epiparams[[3]], main = "", xlab = "scaled period of cross-immunity", freq = F, breaks = 30, ylim = c(0, 1.3))
+hist(epiparams[[3]], main = "C", xlab = "scaled period of cross-immunity", freq = F, breaks = 30, ylim = c(0, 1.3))
 curve(dgamma(x, 10, 10), add = T, lwd = 2)
 
-hist(epiparams[[4]], main = "", xlab = "scaled mosquito mortality rate", freq = F, breaks = 30)
+hist(epiparams[[4]], main = "D", xlab = "scaled mosquito mortality rate", freq = F, breaks = 30)
 curve(dgamma(x, 100, 10), add = T, lwd = 2)
 
 dev.off()
[…]

### 2020-06-30 Manuscript/ms.tex
Final resubmission edits

@@
 This highlights the potential long-term utility of broad, off-peak-season mosquito control in addition to existing, locally targeted within-season efforts.
-Further, uncertainty in the effectiveness of control interventions was driven largely by posterior variation in the case reporting probability (closely tied to overall prevalence and immunity) and the average mosquito mortality rate (closely tied to total mosquito abundance) with higher reporting rates and lower mosquito mortality generating systems more vulnerable to control.
+Further, uncertainty in the effectiveness of control interventions was driven largely by posterior variation in the average mosquito mortality rate (closely tied to total mosquito abundance) with lower mosquito mortality generating systems more vulnerable to control.
 Broadly, these correlations suggest that mosquito control is most effective in situations in which transmission is already limited by mosquito abundance.
@@
 Explicitly capturing the cross-immune interactions among all four serotypes, or even only two of the four \cite{Aguiar2013}, leads to a large and complex mechanistic model.
-Moreover, because dengue case reports do not identify serotype, there is not enough information in the data to inform the dynamics of individual serotypes.
+Moreover, because dengue case reports do not identify serotype, there is not enough information in our data to inform the dynamics of individual serotypes.
 As such, we captured temporary cross-immunity, and the potential for multiple sequential infections, as simply and tractably as possible in a susceptible-exposed-infectious-recovered-susceptible (SEIRS) compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010, Hooten2010}.
@@
 
-In the SEIRS framework, susceptible humans ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
+In the SEIRS framework, the total human population of Vit\'oria ($N$) is divided into susceptible ($S$), exposed ($E$), infectious ($I$), and immune ($R$) classes. 
+Susceptible humans ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
 Following a latent period ($\frac{1}{\rho}$), exposed humans become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
@@
 \hline
-$N$ & Human population size & 327801 & & \cite{vitpop} \\
+$N$ & Human population size in Vit\'oria, Brazil & 327801 & & \cite{vitpop} \\
 $1/d$ & Human life-span & 76 years & &\cite{vitlong} \\
@@
 1-14) as a function of $\boldsymbol{\theta}$ and the weekly stochastic forcing terms ($\boldsymbol{\epsilon_{r}}, \boldsymbol{\epsilon_{\nu}}$).
-Sampling from the posterior distribution of the parameters in mechanistic models is difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations induced by the nonlinearity of the differential equation model \cite{Reilly2005, Girolami2008, Calderhead2011}.
+Sampling from the posterior distribution of the parameters in a mechanistic model is difficult due to multimodality, variable parameter sensitivities (e.g., small changes in one parameter may lead to large changes in output, while similar changes in another parameter may have little effect), and potentially strong posterior correlations induced by the nonlinearity of the differential equation model \cite{Reilly2005, Girolami2008, Calderhead2011}.
 However, the variable $\epsilon_t$ introduces flexibility to the mechanistic model that remedies lack-of-fit when the process parameters are far from optimal \cite{Leander2014}, thereby reducing multimodality and helping to smooth the posterior surface.
@@
 \begin{figure}[!h]
-\includegraphics[scale = 1]{figures/fig1.eps}
+%\includegraphics[scale = 1]{figures/fig1.eps}
 \caption{{\bf Vit\'oria data and model estimates.}
@@
 The estimated weekly mosquito mortality rate varied seasonally, with generally high mortality early in the year and low mortality in August to October (Figure \ref{timeseries}).
-This seasonal trend broadly tracked seasonal variation in temperature (Figure \ref{mortality}) and mosquito trap counts (Figure \ref{timeseries}, though the shape of the annual trajectory differed from year to year.
+This seasonal trend broadly tracked seasonal variation in temperature (Figure \ref{mortality}, posterior mean correlation coefficient of 0.64) and mosquito trap counts (Figure \ref{timeseries}, though the shape of the annual trajectory differed from year to year.
 The posterior distribution of the baseline mortality rate ($d_0$) had a mean of 0.88/week, roughly $60\%$ of the prior mean.
@@
 \begin{figure}[!h]
-\includegraphics[scale = 1]{figures/fig2.eps}
+%\includegraphics[scale = 1]{figures/fig2.eps}
 \caption{{\bf Mosquito mortality and temperature.}
@@
 \begin{figure}[!h]
-\includegraphics[scale = 1]{figures/fig3.eps}
+%\includegraphics[scale = 1]{figures/fig3.eps}
 \caption{{\bf The effect of mosquito control as a function of the week in which it was applied.}
@@
 
-The variation associated with the posterior predicted effect of control implemented in week 34 of the year (i.e., the width of ribbon in Figure \ref{control}) was correlated with the baseline mosquito mortality rate ($d_0$) and the case reporting probability ($\phi$).
+The variation associated with the posterior predicted effect of control implemented in week 34 of the year (i.e., the width of ribbon in Figure \ref{control}) was correlated with the baseline mosquito mortality rate ($d_0$, posterior correlation coefficient of 0.65) and the case reporting probability ($\phi$, posterior correlation coefficient of -0.19).
 Simulated mosquito control created the largest reduction in case reports in posterior samples with low baseline mosquito mortality rate and/or high reporting probability, while control was relatively less effective in simulations from samples with high mosquito mortality or low reporting probability (Figure \ref{correlations}).
@@
 \begin{figure}[!h]
-\includegraphics[scale = 0.9]{figures/fig4.eps}
+%\includegraphics[scale = 0.9]{figures/fig4.eps}
 \caption{{\bf Posterior correlation between system parameters and the effectiveness of control.}
@@
 In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
-The tight correlation between our estimated mosquito mortality and the effect of control confirms this theory, suggesting that control is most effective when it targets long-lived mosquitoes.  
+The high correlation between our estimated mosquito mortality and the effect of control confirms this theory, suggesting that control is most effective when it targets long-lived mosquitoes.  
 Specifically, our simulated mosquito control interventions were most effective at reducing the disease burden when applied around week 34 (i.e., early September), in the dengue off-season.
@@
 
-The tight correlation between the estimated mosquito mortality rate and the simulated effect of control suggests that mosquito control was most effective when it targeted long-lived mosquitoes during the inter-epidemic periods when mosquito abundance was low.
+The positive correlation between the estimated mosquito mortality rate and the simulated effect of control suggests that mosquito control was most effective when it targeted long-lived mosquitoes during the inter-epidemic periods when mosquito abundance was low.
 However, the fact that the mosquito mortality forcing terms ($\epsilon_{\nu}$, Eq.
@@
 These control efforts are guided by the MI-Dengue system, which uses the trap-level mosquito surveillance data to target areas of high mosquito infestation for control (including source reduction, larvacide, and adulticide)\cite{Eiras2009}.
-Although these targeted, reactive interventions likely help reduce local disease risk at times and locations of high mosquito abundance \cite{Pepin2013}, our results suggest the utility of an additional pulse of proactive control in the off-season when mosquitoes are less abundant.
+Although these targeted, reactive interventions are necessary to help reduce local disease risk at times and locations of high mosquito abundance \cite{Pepin2013}, our results suggest that an additional pulse of proactive control in the off-season when mosquitoes are less abundant would minimize human cases.
 

### 2020-06-30 Analysis/plots.r

@@ -227,7 +227,7 @@ overall%>%
   labs(x = "mosquito mortality rate", y = "case ratio") -> fig4a
 
 overall%>% 
-  subset(week == 36 & year == "overall") %>% 
+  subset(week == 34 & year == "overall") %>% 
   ggplot(aes(phi, ratio)) + 
   geom_point(size = 0.5) + 
   theme_classic() + 

### 2020-06-30 Manuscript/ms.tex
Cleaning up notation and typos

@@
 \frac{dV_N}{dt} & = r(t) V_N - \phi_q \tau(t) V_N \\
-\frac{dV_{E1}}{dt} &= \lambda \frac{I}{N} V_S - (\nu\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E1}\\
-\frac{dV_{E2}}{dt} &= \nu\rho_{v}(t) V_{E1} - (\nu\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E2}\\
-\frac{dV_{E3}}{dt} &= \nu\rho_{v}(t) V_{E2}  - (\nu\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E3}\\
-\frac{dV_{E4}}{dt} &= \nu\rho_{v}(t) V_{E3}  - (\nu\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E4}\\
-\frac{dV_I}{dt} &= \nu\rho_{v}(t) V_{E4} - (d(t) + \phi_q \tau(t)) V_I\\
+\frac{dV_{E1}}{dt} &= \lambda \frac{I}{N} V_S - (4\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E1}\\
+\frac{dV_{E2}}{dt} &= 4\rho_{v}(t) V_{E1} - (4\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E2}\\
+\frac{dV_{E3}}{dt} &= 4\rho_{v}(t) V_{E2}  - (4\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E3}\\
+\frac{dV_{E4}}{dt} &= 4\rho_{v}(t) V_{E3}  - (4\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E4}\\
+\frac{dV_I}{dt} &= 4\rho_{v}(t) V_{E4} - (d(t) + \phi_q \tau(t)) V_I\\
 \frac{dV_C}{dt} & = \phi_q \tau(t) V_N\\
@@
 $\lambda$ & Transmission rate & 4.87 week$^{-1}$ & & \cite{Scott2000}\\
-$1/\rho_{v}(t)$ & Extrinsic incubation period & $\exp\left(7.9 - 0.21 T(t) \right)$ days & & \cite{Chan2012}\\
+$1/\rho_{v}(t)$ & Extrinsic incubation period & $\frac{1}{7}\exp\left(7.9 - 0.21 T(t) \right)$ weeks & & \cite{Chan2012}\\
 $V_{E0}$ & Initial exposed mosquitoes &  0 & &\\
@@
 The estimated weekly mosquito mortality rate varied seasonally, with generally high mortality early in the year and low mortality in August to October (Figure \ref{timeseries}).
-This seasonal trend broadly tracked seasonal variation in temperature (Figure \ref{mortality}, posterior mean correlation coefficient of 0.64) and mosquito trap counts (Figure \ref{timeseries}, though the shape of the annual trajectory differed from year to year.
+This seasonal trend broadly tracked seasonal variation in temperature (Figure \ref{mortality}, correlation coefficient of 0.64) and mosquito trap counts (Figure \ref{timeseries}), though the shape of the annual trajectory differed from year to year.
 The posterior distribution of the baseline mortality rate ($d_0$) had a mean of 0.88/week, roughly $60\%$ of the prior mean.

### 2020-06-30 Analysis/citystan.r
Setting code back to defaults (i.e., no imports)
and dropping unused output

@@ -47,7 +47,7 @@ rov <- 7 * exp(0.21 * covars$temp - 7.9)
 # Setting imports (if any)
 firsts <- daply(data, .(year), function(df) min(df$tot.week))[2:5]
 import <- rep(0, 243)
-import[firsts - 8] <- 10 / pop
+# import[firsts - 8] <- 10 / pop
 
 # Wrapping up all the data
 dat.stan <- list(T = 243,
@@ -129,4 +129,4 @@ fit <- stan(file = "Code/gammaeip.stan",
             control = list(adapt_delta = 0.99,
                            max_treedepth = 15))
 
-saveRDS(fit, "Results/import_chain.rds")
+saveRDS(fit, "Results/chain.rds")

### 2020-06-30 Analysis/control.r

@@ -38,14 +38,14 @@ rov <- 7 * exp(0.21 * covars$temp - 7.9)
 # Setting imports (if any)
 firsts <- daply(data, .(year), function(df) min(df$tot.week))[2:5]
 import <- rep(0, 243)
-import[firsts - 8] <- 10 / pop
+# import[firsts - 8] <- 10 / pop
 
 model <- stan_model(file = "Code/gammaeip.stan", verbose = T, save_dso = TRUE, auto_write = TRUE)
 
 #===============================================================================
 # Loading mcmc samples
 
-chain <- readRDS("Results/import_chain.rds")
+chain <- readRDS("Results/chain.rds")
 samples <- rstan::extract(chain, permute = T)[1:18] 
 
 fitcases <- rstan::extract(chain, "state", permute = T)[[1]] %>% 
@@ -122,19 +122,7 @@ reduction <- foreach(k = 1:nmcmc, .combine = "rbind", .packages = c("rstan", "ma
 
 stopCluster(cl)
 
-saveRDS(reduction, "Results/import_control.rds")
-
-# Effect of timing of control on the number of cases in a given year
-annual <- ddply(control, .(year, rep, control), summarise,
-                tcases = sum(cases),
-                tcases0 = sum(cases0),
-                ratio = tcases / tcases0,
-                dvmu = mean(dvmu),
-                delta = mean(delta),
-                phi = mean(phi),
-                relweek = (control - firsts[as.character(year)])[1])
-
-saveRDS(annual, "Results/import_annual.rds")
+saveRDS(reduction, "Results/control.rds")
 
 # Effect of control in the number of cases in the following year
 moving <- ddply(control, .(rep, control), summarise,
@@ -146,4 +134,4 @@ moving <- ddply(control, .(rep, control), summarise,
                 delta = mean(delta),
                 phi = mean(phi))
 
-saveRDS(moving, "Results/import_moving.rds")
+saveRDS(moving, "Results/moving_control.rds")

### 2020-06-30 Analysis/plots.r

@@ -43,14 +43,6 @@ post$rov <- 7 * exp(0.21 * temp$temp - 7.9)
 # Loading MCMC results
 sim <- readRDS("Results/chain.rds")
 
-# Loading control simulations
-# control <- readRDS("Results/control_rebound.rds") 
-
-# Loading control summaries (fixed and moving window)
-
-# Effect of timing of control on the number of cases in a given year
-# annual <- readRDS("Results/annual_control.rds")
-
 # Effect of control in the number of cases in the following year
 moving <- readRDS("Results/moving_control.rds")
 

### 2020-08-17 Manuscript/ms.tex
Adding citation

@@
 This suggests that regardless of unexplained structure in the forcing terms, the pattern of case reports was still very well described by realistic seasonal fluctuations in the mosquito mortality rate.
-Given the importance of seasonality in understanding dengue epidemiology \cite{TenBosch2016}, and the potential importance of mosquito mortality in driving the effect of control interventions, future work should focus on developing a more complete, predictive understanding of the seasonal drivers of mosquito mortality (including, potentially, control itself).
+As demonstrated by Reiner \emph{et al.} \cite{Reiner2015} for malaria transmission, estimates of transmission potential can be sensitive to fluctuations in mosquito abundance and age structure.
+Moreover, given the broad importance of seasonality in understanding dengue epidemiology \cite{TenBosch2016}, and the role of mosquito mortality and age in driving the effect of control interventions, future work should focus on developing a more complete, predictive understanding of the seasonal drivers of mosquito mortality (including, potentially, control itself).
 

### 2020-08-17 Manuscript/ms.tex
Adding ethics statement to methods

@@
 In addition, dengue fever is a mandatory notifiable disease, and thus the city's Ministry of Health Secretary maintains a database of weekly notified probable dengue cases (i.e., medical care sought for dengue-like symptoms) for the same time period.
+We did not obtain IRB approval for this work as the data we worked with were received by us as aggregated data at the weekly and neighborhood level.  
+Hence, this research does not meet the definition of human subjects research requiring IRB approval.  
+The data were analyzed in the aggregated form, which protects the anonymity of individuals.
 

### 2020-11-09 Manuscript/ms.tex
Formatting changes

@@
 % cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
+\usepackage[nospace]{cite}
 
@@
 \bigskip
-\textbf{1} Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO, USA
+\textbf{1} Graduate Degree Program in Ecology, Colorado State University, Fort Collins, Colorado, United States of America
 \\
-\textbf{2} Department of Statistics, Colorado State University, Fort Collins, CO, USA
+\textbf{2} Department of Statistics, Colorado State University, Fort Collins, Colorado, United States of America
 \\
-\textbf{3} National Wildlife Research Center, United States Department of Agriculture, Wildlife Services, Fort Collins, CO, USA
+\textbf{3} National Wildlife Research Center, United States Department of Agriculture, Wildlife Services, Fort Collins, Colorado, United States of America
 \\
-\textbf{4} Departamento de Parasitologia, Universidade Federal de Minas Gerais, Belo Horizonte, MG, Brazil
+\textbf{4} Departamento de Parasitologia, Universidade Federal de Minas Gerais, Belo Horizonte, Minas Gerais, Brazil
 \\
-\textbf{5} U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit
+\textbf{5} U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Fort Collins, Colorado, United States of America
 \\
-\textbf{6} Department of Fish, Wildlife, and Conservation Biology, Colorado State University
+\textbf{6} Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, Colorado, United States of America
 \bigskip
@@
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* clint.leach@colostate.edu
+* clint.leach@gmail.com
 
@@
 
-Many of the attempts to establish this link have found a weak relationship between mosquito abundance indices and incidence of disease in humans \cite{Bowman2014, Pepin2015, Cromwell2017}.
+Many of the attempts to establish this link have found a weak relationship between mosquito abundance indices and incidence of disease in humans \cite{Bowman2014, Pepin2015,Cromwell2017}.
 However, these attempts often do not account for the complex, non-linear interactions that mediate the relationship between mosquito abundance and human disease.
@@
 
-The importance of intrinsic nonlinearities, potentially alongside seasonality and stochastic forcing \cite{Ellner1998, Koelle2004, Grenfell2002}, in governing human disease risk highlights the need to integrate mechanistic modeling into the quantitative tools used to understand the effects of control interventions.
+The importance of intrinsic nonlinearities, potentially alongside seasonality and stochastic forcing \cite{Ellner1998,Koelle2004,Grenfell2002}, in governing human disease risk highlights the need to integrate mechanistic modeling into the quantitative tools used to understand the effects of control interventions.
 Such mechanistic models can often perform better than complex autoregressive statistical models in describing and forecasting population dynamics \cite{Reilly2005}.
@@
 
+\subsection*{Ethics statement}
+
+We did not obtain Institutional Review Board (IRB) approval for this work as the data we worked with were received by us as aggregated data at the weekly and neighborhood level.  
+Hence, this research does not meet the definition of human subjects research requiring IRB approval.  
+The data were analyzed in the aggregated form, which protects the anonymity of individuals.
+
 \subsection*{Study system and data}
@@
 In addition, dengue fever is a mandatory notifiable disease, and thus the city's Ministry of Health Secretary maintains a database of weekly notified probable dengue cases (i.e., medical care sought for dengue-like symptoms) for the same time period.
-We did not obtain IRB approval for this work as the data we worked with were received by us as aggregated data at the weekly and neighborhood level.  
-Hence, this research does not meet the definition of human subjects research requiring IRB approval.  
-The data were analyzed in the aggregated form, which protects the anonymity of individuals.
 
-\subsection*{Process Model}
+\subsection*{Process model}
 
@@
 Infection with one serotype confers life-long immunity to that serotype, along with temporary immunity to other serotypes \cite{Wearing2006}.  
-As this cross-immunity wanes, antibodies from the previous infection can result in antibody-dependent enhancement (ADE), wherein human hosts are more susceptible to infection with the other serotypes and more likely to develop severe symptoms (i.e., dengue haemorrhagic fever or dengue shock syndrome)\cite{Wearing2006}.
+As this cross-immunity wanes, antibodies from the previous infection can result in antibody-dependent enhancement (ADE), wherein human hosts are more susceptible to infection with the other serotypes and more likely to develop severe symptoms (i.e., dengue hemorrhagic fever or dengue shock syndrome)\cite{Wearing2006}.
 The strength and duration of these different inter-serotype interactions are not well understood, although different models suggest that temporary cross-immunity alone (without ADE) is sufficient to reproduce observed multi-annual dynamics in Thailand \cite{Wearing2006,Reich2013}.
@@
 Moreover, because dengue case reports do not identify serotype, there is not enough information in our data to inform the dynamics of individual serotypes.
-As such, we captured temporary cross-immunity, and the potential for multiple sequential infections, as simply and tractably as possible in a susceptible-exposed-infectious-recovered-susceptible (SEIRS) compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010, Hooten2010}.
-Although it captures the critical influence of temporary immunity, this framework does not account for the potential relationship between an individual's infection history and the likelihood that a new infection with be symptomatic (and thus reported).
+As such, we captured temporary cross-immunity, and the potential for multiple sequential infections, as simply and tractably as possible in a susceptible--exposed--infectious--recovered--susceptible (SEIRS) compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010, Hooten2010}.
+Although it captures the critical influence of temporary immunity, this framework does not account for the potential relationship between an individual's infection history and the likelihood that a new infection will be symptomatic (and thus reported).
 In particular, secondary infections appear more likely to be symptomatic than primary infections \cite{Imai2016,Clapham2017}, while third and fourth infections appear much less likely to be symptomatic \cite{Olkowski2013} (but see \cite{Montoya2013} who found similar rates of symptomatic cases across infection number).
@@
 Infectious humans recover at rate $\gamma$ and subsequently remain immune ($R$) for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.
-Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_E$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
+Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_E$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{v}(t)}$) before becoming infectious ($V_I$).
 Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implemented a gamma-distributed incubation period by chaining together multiple exposed classes ($V_{Ej}$, taking advantage of the fact that a gamma-distributed random variable can be generated through the sum of exponential random variables with the same rate parameter) \cite{Lloyd2001}.
 Total mosquito population size ($V_N$) is controlled by a forced, seasonally varying growth rate ($r(t)$), while the transmission bottleneck is captured with a forced, seasonally varying mortality rate ($d(t)$).
-Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.
+Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau(t)$, where $\phi_q$ is the per-trap capture rate, and $\tau(t)$ is the number of traps deployed in week $t$.
 
@@
 The remaining parameters include the epidemiological parameters controlling the average latent, infectious, and immune periods ($\rho$, $\gamma$, $\delta$) and average mosquito lifespan ($d_0$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_y$, $\phi_q$, $\eta_y$, $\eta_q$).
-Where possible, we specified informative prior distributions for these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).
+Where possible, we specified informative prior distributions for these parameters based on existing laboratory and field studies (see Table 1 for means and \nameref{S1} for detailed explanations).
 
@@
 The values in parentheses after the posterior means give the 80$\%$ credible interval.
-See Supplemental Material for a full description of all prior distributions.}
+See \nameref{S1} for a full description of all prior distributions.}
 \begin{tabular}{lp{6cm}lll}
@@
 
-The model captured the observed dynamics of both case reports and mosquito trap counts (Figure \ref{timeseries}).
+The model captured the observed dynamics of both case reports and mosquito trap counts (Fig \ref{timeseries}).
 The estimated posterior median case reports explained 91\% of the variation in the observed time series, while the posterior median mosquito trap counts explained 46\% of the variation in the observed time series.
-In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 6 - 8).
-The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that the Vit\'oria data contained little additional information about these parameters (Supplemental Figure 3).
+In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, \nameref{S6_fig}, \nameref{S7_fig}, \nameref{S8_fig}).
+The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that the Vit\'oria data contained little additional information about these parameters (\nameref{S3_fig}).
 The estimated latent period in a human host ($1/\rho$, the expected time it takes for an exposed human to become infectious to biting mosquitoes) was influenced more strongly by the data, with a posterior mean of 1.73 weeks compared to a prior mean of 0.87 weeks.
@@
 A: weekly observed case reports (points), with corresponding posterior median (black line) and 80\% posterior credible interval (gray band).
-B: weekly mosquito trap counts (points), with posteriod median (black line) and 80\% posterior credible interval (gray band).
+B: weekly mosquito trap counts (points), with posterior median (black line) and 80\% posterior credible interval (gray band).
 C: extrinsic incubation period (EIP; weeks), computed from weekly mean temperature data.
@@
 
-The estimated weekly mosquito mortality rate varied seasonally, with generally high mortality early in the year and low mortality in August to October (Figure \ref{timeseries}).
-This seasonal trend broadly tracked seasonal variation in temperature (Figure \ref{mortality}, correlation coefficient of 0.64) and mosquito trap counts (Figure \ref{timeseries}), though the shape of the annual trajectory differed from year to year.
+The estimated weekly mosquito mortality rate varied seasonally, with generally high mortality early in the year and low mortality in August to October (Fig \ref{timeseries}).
+This seasonal trend broadly tracked seasonal variation in temperature (Fig \ref{mortality}, correlation coefficient of 0.64) and mosquito trap counts (Fig \ref{timeseries}), though the shape of the annual trajectory differed from year to year.
 The posterior distribution of the baseline mortality rate ($d_0$) had a mean of 0.88/week, roughly $60\%$ of the prior mean.
-The marginal posterior means of the $\epsilon_{\nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (Supplementary Figure 1), although the marginal posterior distribution of each $\epsilon_{\nu t}$ overlapped zero.
-In addition, the standard deviation of the mortality forcing terms was small relative to the weak prior ($E(\sigma_{\nu}|\mathbf{y}) = 0.01$, Supplemental Figure 2).
+The marginal posterior means of the $\epsilon_{\nu t}$ forcing the mosquito mortality process exhibited a higher-frequency periodic oscillation (\nameref{S1_fig}), although the marginal posterior distribution of each $\epsilon_{\nu t}$ overlapped zero.
+In addition, the standard deviation of the mortality forcing terms was small relative to the weak prior ($E(\sigma_{\nu}|\mathbf{y}) = 0.01$, \nameref{S2_fig}).
 
@@
 
-The effect of a given mosquito control intervention (i.e., the temporary removal of $5\%$ of the adult population in a given week) on the number of cases in the following year (relative to no control) varied both seasonally and interannually (Figure \ref{control}).
+The effect of a given mosquito control intervention (i.e., the temporary removal of $5\%$ of the adult population in a given week) on the number of cases in the following year (relative to no control) varied both seasonally and interannually (Fig \ref{control}).
 This variation in the effect of control was tightly correlated with the estimated mosquito mortality, with a median posterior correlation between the two time series of 0.96.
 As such, the seasonal variation in the effectiveness of control followed the same trend as mosquito mortality rate (although the annual minima in the case ratio time series generally fell 1 to 2 weeks before the minima in the mosquito mortality time series), with the largest reductions in case load resulting from interventions during the dengue off-season (early September for 2008 and 2009, and mid-July for 2010).
-Summing over the interannual variation to compute the overall effect of control implemented in a given week of the year, we found that mosquito control was most effective when implemented around  week 34 (late August/early September), reducing the estimated case load by roughly $14\%$ (Figure \ref{control}, bottom panel).
+Summing over the interannual variation to compute the overall effect of control implemented in a given week of the year, we found that mosquito control was most effective when implemented around  week 34 (late August/early September), reducing the estimated case load by roughly $14\%$ (Fig \ref{control}).
 This broadly corresponds to the end of the dry season in Vit\'oria, when both dengue case reports and mosquito trap counts are low.
@@
 
-The variation associated with the posterior predicted effect of control implemented in week 34 of the year (i.e., the width of ribbon in Figure \ref{control}) was correlated with the baseline mosquito mortality rate ($d_0$, posterior correlation coefficient of 0.65) and the case reporting probability ($\phi$, posterior correlation coefficient of -0.19).
-Simulated mosquito control created the largest reduction in case reports in posterior samples with low baseline mosquito mortality rate and/or high reporting probability, while control was relatively less effective in simulations from samples with high mosquito mortality or low reporting probability (Figure \ref{correlations}).
+The variation associated with the posterior predicted effect of control implemented in week 34 of the year (i.e., the width of ribbon in Fig \ref{control}) was correlated with the baseline mosquito mortality rate ($d_0$, posterior correlation coefficient of 0.65) and the case reporting probability ($\phi$, posterior correlation coefficient of -0.19).
+Simulated mosquito control created the largest reduction in case reports in posterior samples with low baseline mosquito mortality rate and/or high reporting probability, while control was relatively less effective in simulations from samples with high mosquito mortality or low reporting probability (Fig \ref{correlations}).
 Thus mosquito control was more effective at reducing disease burden in simulations with long average mosquito lifespans (i.e., low mosquito mortality rates) or low overall prevalence (i.e., fewer undetected cases).
@@
 Seasonality, in particular, is an important factor in capturing the annual cycle of dengue outbreaks \cite{Wearing2006,Aguiar2011,Reich2013}.
-However, the observed seasonality in transmission likely emerges from the combined effects of multiple seasonally-varying components that may be driven by different environmental factors and oscillate in different phases (e.g., mosquito abundance seems to lag slightly behind temperature-driven variation in extrinsic incubation period, Figure \ref{timeseries}).
+However, the observed seasonality in transmission likely emerges from the combined effects of multiple seasonally-varying components that may be driven by different environmental factors that oscillate in different phases (e.g., mosquito abundance seems to lag slightly behind temperature-driven variation in extrinsic incubation period, Fig \ref{timeseries}).
 Integrating these seasonally-varying components into a synthetic measure of transmission potential (e.g., a temperature-dependent effective reproduction number, \cite{Codeco2018}), or more specifically, a measure of the transmission potential of mosquitoes, is difficult.
@@
 The Bayesian framework allowed us to account for uncertainty across the data, process, and parameter levels of our model \cite{Berliner1996}.
-We carried this uncertainty through to our simulations of control interventions \cite{Elderd2006} and found that there was substantial uncertainty in the proportion of cases prevented by a control intervention (i.e., the width of the ribbons in Figure \ref{control}).
+We carried this uncertainty through to our simulations of control interventions \cite{Elderd2006} and found that there was substantial uncertainty in the proportion of cases prevented by a control intervention (i.e., the width of the ribbons in Fig \ref{control}).
 Much of this uncertainty could be attributed to posterior uncertainty in the case reporting rate ($\phi$) and the average mosquito mortality rate ($d_0$).
-Specifically, we found that control implemented at the overall optimum (week 34) had the largest impact (i.e., the lowest case ratio) in simulations with a low average mosquito mortality and/or a high case reporting rate (Figure \ref{correlations}).
+Specifically, we found that control implemented at the overall optimum (week 34) had the largest impact (i.e., the lowest case ratio) in simulations with a low average mosquito mortality and/or a high case reporting rate (Fig \ref{correlations}).
 Mosquito mortality and the case reporting rate were correlated with the overall level of mosquito abundance and the overall size of the susceptible population, respectively, suggesting that control was most effective in simulations with fewer mosquitoes and more susceptible humans.
@@
 
-The periodic structure in the time series of the posterior means of the $\epsilon_{\nu}$ (Supplemental Figure 1) suggests that these terms were accounting for more than just noise, and there may have been some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
+The periodic structure in the time series of the posterior means of the $\epsilon_{\nu}$ (\nameref{S1_fig}) suggests that these terms were accounting for more than just noise, and there may have been some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
 Following Hooker and Ellner, we can explore whether this process is likely to result from misspecification of the rates of change of the existing state variables (indicated by a dependence of $\epsilon_{\nu t}$ on other state variables), or from missing state variables altogether (indicated by an additional dependence of $\epsilon_{\nu t}$ on its own lagged values).
 The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
-These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control intervention), or epidemiological processes (e.g., multiple circulating serotypes of the dengue virus, subsets of the population with different mixing or risk levels).
+These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control interventions), or epidemiological processes (e.g., multiple circulating serotypes of the dengue virus, subsets of the population with different mixing or risk levels).
 
@@
 In addition, spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
-Although our model does not account for these heterogeneities, it was nontheless able to capture the city-wide dynamics well, suggesting that there may be sufficient mixing to appear homogeneous at the city scale.
+Although our model does not account for these heterogeneities, it was nonetheless able to capture the city-wide dynamics well, suggesting that there may be sufficient mixing to appear homogeneous at the city scale.
 Further, although our mechanistic model may be able to suggest when a city-wide intervention is likely to be effective, to make the best use of limited resources, spatial prioritization may still be necessary.
@@
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+\paragraph*{S1 Fig.}
+\label{S1_fig}
+{\bf Posterior estimates of harmonic oscillator forcing functions.} A: mosquito mortality forcing.
+B: mosquito growth forcing.
+Median posterior estimate (black line) and 80\% credible interval (gray band).
+
+\paragraph*{S2 Fig.}
+\label{S2_fig}
+{\bf Posterior distribution of variances on forcing terms.} A: mosquito mortality forcing. B: mosquito growth forcing.
+
+\paragraph*{S3 Fig.}
+\label{S3_fig}
+{\bf Posterior distribution of epidemiological parameters.}  The black line indicates the prior density.
+A: scaled latent period in human host ($1/\rho$).
+B: scaled rate of infectious decay in human host ($\gamma$).
+C: scaled period of cross-immunity ($1/\delta$).
+D: scaled baseline mosquito mortality rate ($d_0$).
+
+\paragraph*{S4 Fig.}
+\label{S4_fig}
+{\bf Posterior distribution of initial conditions.} The black line indicates the prior density.
+A: initial proportion of the human population that is susceptible.
+B: initial number of exposed individuals.
+C: initial number of infectious individuals.
+D: log initial number of mosquitoes-per-person.
+E: initial value of centered mosquito mortality rate oscillator.
+F: initial value of mosquito population growth rate oscillator.
+
+\paragraph*{S5 Fig.}
+\label{S5_fig}
+{\bf Posterior distribution of measurement parameters.} The black line shows the prior density.
+A: log of the mosquito trap capture rate (which is strongly correlated with initial mosquito abundance).
+B: human case reporting probability.
+C: mosquito trap count measurement over-dispersion (prior density too low to be visible).
+D: case report measurement over-dispersion (prior density too low to be visible).
+
+\paragraph*{S6 Fig.}
+\label{S6_fig}
+{\bf Posterior check of autocorrelation structure in data.} A: autocorrelation in case report time series.
+B: autocorrelation in trapped mosquito time series.
+Points indicate the observed autocorrelation at the given lag, and vertical lines give the 80\% posterior credible interval of the autocorrelation in the estimated time series.
+
+\paragraph*{S7 Fig.}
+\label{S7_fig}
+{\bf Posterior distribution of the total number of cases reported (A) and the total number of mosquitoes captured (B).}  The thick line indicates the observed totals.
+
+\paragraph*{S8 Fig.}
+\label{S8_fig}
+{\bf Posterior distributions of the minimum and maximum weekly reported cases (A and C) and trapped mosquitoes (B and D).} The thick line indicates the observed minima and maxima.
+
+\paragraph*{S9 Fig.}
+\label{S9_fig}
+{\bf Posterior estimates of the mosquito mortality rate and the probability of a mosquito surving the external incubation period.}  The black line indicates the posterior median, while the gray ribbon shows the 80$\%$ credible interval.
+The probability of surviving the extrinsic incubation period (bottom panel) was computed by taking the ratio of the number of mosquitoes that entered the infectious class in a week over the total number of mosquitoes that exited the exposed class that week.
+That is, the probability of surviving the extrinsic incubatio period represents the fraction of mosquitoes leaving the exposed class that enter the infectious class.
+
 \paragraph*{S1 Text.}
 \label{S1}
-{Full description of model and prior distributions, along with all supplemental figures.}
+{\bf Full specification of the model and prior distributions.}
 
@@
 
-\bibliographystyle{plos2015}
-\bibliography{dengue}
+%\bibliographystyle{plos2015}
+%\bibliography{dengue}
+
+\begin{thebibliography}{10}
+
+\bibitem{Bhatt2013}
+Bhatt S, Gething PW, Brady OJ, Messina JP, Farlow AW, Moyes CL, et~al.
+\newblock {The global distribution and burden of dengue.}
+\newblock Nature. 2013;496:504--507.
+\newblock doi:{10.1038/nature12060}.
+
+\bibitem{Achee2015}
+Achee NL, Gould F, Perkins TA, Reiner RC, Morrison AC, Ritchie Sa, et~al.
+\newblock {A critical assessment of vector control for dengue prevention}.
+\newblock PLoS Negl Trop Dis. 2015;9(5):e0003655.
+\newblock doi:{10.1371/journal.pntd.0003655}.
+
+\bibitem{Morrison2008}
+Morrison AC, Zielinski-Gutierrez E, Scott TW, Rosenberg R.
+\newblock {Defining challenges and proposing solutions for control of the virus
+  vector \emph{Aedes aegypti}}.
+\newblock PLoS Med. 2008;5(3):0362--0366.
+\newblock doi:{10.1371/journal.pmed.0050068}.
+
+\bibitem{Scott2010b}
+Scott TW, Morrison AC.
+\newblock {Longitudinal field studies will guide a paradigm shift in Dengue
+  prevention}.
+\newblock In: Vector Biology, Ecology, and Control; 2010. p. 139--161.
+\newblock Available from:
+  \url{http://link.springer.com/10.1007/978-90-481-2458-9}.
+
+\bibitem{Scott2010a}
+Scott TW, Morrison AC.
+\newblock {Vector dynamics and transmission of dengue virus: implications for
+  dengue surveillance and prevention strategies}.
+\newblock Dengue Virus. 2010;(338):161.
+\newblock doi:{10.1007/978-3-642-02215-9}.
+
+\bibitem{Bowman2014}
+Bowman LR, Runge-Ranzinger S, McCall PJ.
+\newblock {Assessing the relationship between vector indices and dengue
+  transmission: A systematic review of the evidence}.
+\newblock PLoS Negl Trop Dis. 2014;8(5):e2848.
+\newblock doi:{10.1371/journal.pntd.0002848}.
+
+\bibitem{Pepin2015}
+Pepin KM, Leach CB, Marques-Toledo C, Laass KH, Paixao KS, Luis AD, et~al.
+\newblock {Utility of mosquito surveillance data for spatial prioritization of
+  vector control against dengue viruses in three Brazilian cities}.
+\newblock Parasites and Vectors. 2015;8(1):1--15.
+\newblock doi:{10.1186/s13071-015-0659-y}.
+
+\bibitem{Cromwell2017}
+Cromwell EA, Stoddard ST, Barker CM, {Van Rie} A, Messer WB, Meshnick SR,
+  et~al.
+\newblock {The relationship between entomological indicators of \emph{Aedes
+  aegypti} abundance and dengue virus infection}.
+\newblock PLoS Negl Trop Dis. 2017;11(3):e0005429.
+\newblock doi:{10.1371/journal.pntd.0005429}.
+
+\bibitem{Koelle2004}
+Koelle K, Pascual M.
+\newblock {Disentangling extrinsic from intrinsic factors in disease dynamics:
+  A nonlinear time series approach with an application to cholera}.
+\newblock Am Nat. 2004;163(6):901--913.
+
+\bibitem{Koelle2005}
+Koelle K, Rod{\'{o}} X, Pascual M, Yunus M, Mostafa G.
+\newblock {Refractory periods and climate forcing in cholera dynamics.}
+\newblock Nature. 2005;436:696--700.
+\newblock doi:{10.1038/nature03820}.
+
+\bibitem{Smith2012}
+Smith DL, Battle KE, Hay SI, Barker CM, Scott TW, Mckenzie FE.
+\newblock {Ross, Macdonald, and a theory for the dynamics and control of
+  mosquito-transmitted pathogens}.
+\newblock PLoS Pathog. 2012;8(4).
+\newblock doi:{10.1371/journal.ppat.1002588}.
+
+\bibitem{Ellner1998}
+Ellner SP, Bailey Ba, Bobashev GV, Gallant aR, Grenfell BT, Nychka DW.
+\newblock {Noise and nonlinearity in measles epidemics: combining mechanistic
+  and statistical approaches to population modeling.}
+\newblock Am Nat. 1998;151(5):425--440.
+\newblock doi:{10.1086/286130}.
+
+\bibitem{Grenfell2002}
+Grenfell BT, Bj{\o}rnstad ON, Finkenst{\"{a}}dt BF.
+\newblock {Dynamics of measles epidemics: Scaling noise, determinism, and
+  predictability with the TSIR model}.
+\newblock Ecol Monogr. 2002;72(2):185--202.
+\newblock doi:{10.1890/0012-9615(2002)072[0185:DOMESN]2.0.CO;2}.
+
+\bibitem{Reilly2005}
+Reilly C, Zeringue A.
+\newblock {Improved predictions of lynx trappings using a biological model}.
+\newblock In: Gelman A, Meng XL, editors. Applied Bayesian Modeling and Causal
+  Inference from Incomplete-Data Perspectives. John Wiley {\&} Sons Ltd; 2005.
+  p. 297--308.
+
+\bibitem{Buckland2007}
+Buckland ST, Newman KB, Fern{\'{a}}ndez C, Thomas L, Harwood J.
+\newblock {Embedding population dynamics models in inference}.
+\newblock Stat Sci. 2007;22(1):44--58.
+\newblock doi:{10.1214/088342306000000673}.
+
+\bibitem{Hotelling1927}
+Hotelling H.
+\newblock {Differential equations subject to error, and population estimates}.
+\newblock J Am Stat Assoc. 1927;22(159):283--314.
+
+\bibitem{Wikle2010}
+Wikle CK, Hooten MB.
+\newblock {A general science-based framework for dynamical spatio-temporal
+  models}.
+\newblock Test. 2010;19:417--451.
+\newblock doi:{10.1007/s11749-010-0209-z}.
+
+\bibitem{Elderd2006}
+Elderd BD, Dukic VM, Dwyer G.
+\newblock {Uncertainty in predictions of disease spread and public health
+  responses to bioterrorism and emerging diseases}.
+\newblock Proc Natl Acad Sci U S A. 2006;103(42):15693--15697.
+
+\bibitem{Berliner1996}
+Berliner LM.
+\newblock {Hierarchical Bayesian time series models}.
+\newblock In: Hanson KM, Silver RN, editors. Maximum entropy and Bayesian
+  methods. Fundamental Theories of Physics (An international book series on the
+  fundamental theories of physics: Their clarification, development and
+  application). vol.~79. Springer, Dordrecht; 1996.
+
+\bibitem{Cressie2009}
+Cressie N, Calder Ca, Clark JS, {Ver Hoef} JM, Wikle CK.
+\newblock {Accounting for uncertainty in ecological analysis: the strengths and
+  limitations of hierarchical statistical modeling.}
+\newblock Ecol Appl. 2009;19(3):553--70.
+
+\bibitem{vitpop}
+Demographic Census 2000 and 2010 and population estimates 2001 to 2009 and 2011
+  to 2014.
+\newblock IBGE;.
+\newblock Available from:
+  \url{http://legado.vitoria.es.gov.br/regionais/Censo_2010/Tab17_crespopulacional.asp}.
+
+\bibitem{Eiras2009}
+Eiras {\'{A}}, Resende M.
+\newblock {Preliminary evaluation of the ``Dengue-MI" technology for
+  \emph{Aedes aegypti} monitoring and control}.
+\newblock Cadernos de Sa{\'{u}}de P{\'{u}}blica. 2009; p. 45--58.
+
+\bibitem{Lana2018}
+Lana RM, Morais MM, de~Lima TFM, Carneiro TGdS, Stolerman LM, dos Santos JPC,
+  et~al.
+\newblock {Assessment of a trap based \emph{Aedes aegypti} surveillance program
+  using mathematical modeling}.
+\newblock PLoS ONE. 2018;13(1):1--16.
+
+\bibitem{Wearing2006}
+Wearing HJ, Rohani P.
+\newblock {Ecological and immunological determinants of dengue epidemics.}
+\newblock Proc Natl Acad Sci U S A. 2006;103(31):11802--7.
+\newblock doi:{10.1073/pnas.0602960103}.
+
+\bibitem{Reich2013}
+Reich NG, Shrestha S, King Aa, Rohani P, Lessler J, Kalayanarooj S, et~al.
+\newblock {Interactions between serotypes of dengue highlight epidemiological
+  impact of cross-immunity.}
+\newblock J R Soc Interface. 2013;10(86):20130414.
+\newblock doi:{10.1098/rsif.2013.0414}.
+
+\bibitem{Aguiar2013}
+Aguiar M, Kooi BW, Rocha F, Ghaffari P, Stollenwerk N.
+\newblock {How much complexity is needed to describe the fluctuations observed
+  in dengue hemorrhagic fever incidence data?}
+\newblock Ecol Complex. 2013;16:31--40.
+\newblock doi:{10.1016/j.ecocom.2012.09.001}.
+
+\bibitem{Newton1992}
+Newton EAC, Reiter A.
+\newblock {A model of the transmission of dengue fever with an evaluation of
+  the impact of ultra-low volume (ULV) insecticide applications on dengue
+  epidemics}.
+\newblock Am J Trop Med Hyg. 1992;47(6):709--720.
+
+\bibitem{Burattini2008}
+Burattini MN, Chen M, Chow A, Coutinho FAB, Goh KT, Lopez LF, et~al.
+\newblock {Modelling the control strategies against dengue in Singapore}.
+\newblock Epidemiol Infect. 2008;136(03):309--319.
+
+\bibitem{Pinho2010}
+Pinho STR, Ferreira CP, Esteva L, Barreto FR, {Morato e Silva} VC, Teixeira
+  MGL.
+\newblock {Modelling the dynamics of dengue real epidemics}.
+\newblock Philos Trans R Soc A Math Phys Eng Sci. 2010;368(1933):5679--5693.
+\newblock doi:{10.1098/rsta.2010.0278}.
+
+\bibitem{Hooten2010}
+Hooten MB, Anderson J, Waller La.
+\newblock {Assessing North American influenza dynamics with a statistical SIRS
+  model.}
+\newblock Spatial and Spatio-temporal Epidemiology. 2010;1:177--85.
+\newblock doi:{10.1016/j.sste.2010.03.003}.
+
+\bibitem{Imai2016}
+Imai N, Dorigatti I, Cauchemez S, Ferguson NM.
+\newblock {Estimating dengue transmission intensity from case-notification data
+  from multiple countries}.
+\newblock PLoS Neglected Tropical Diseases. 2016;10(7):e0004833.
+\newblock doi:{10.1371/journal.pntd.0004833}.
+
+\bibitem{Clapham2017}
+Clapham HE, Cummings DAT, Johansson MA.
+\newblock {Immune status alters the probability of apparent illness due to
+  dengue virus infection: Evidence from a pooled analysis across multiple
+  cohort and cluster studies}.
+\newblock PLoS Neglected Tropical Diseases. 2017;11(9):e0005926.
+\newblock doi:{10.1371/journal.pntd.0005926}.
+
+\bibitem{Olkowski2013}
+Olkowski S, Forshey BM, Morrison AC, Rocha C, Vilcarromero S, Halsey ES, et~al.
+\newblock {Reduced risk of disease during postsecondary dengue virus
+  infections.}
+\newblock The Journal of Infectious Diseases. 2013;208:1026--33.
+\newblock doi:{10.1093/infdis/jit273}.
+
+\bibitem{Montoya2013}
+Montoya M, Gresh L, Mercado JC, Williams KL, Vargas MJ, Gutierrez G, et~al.
+\newblock {Symptomatic versus inapparent outcome in repeat dengue virus
+  infections is influenced by the time interval between infections and study
+  year}.
+\newblock PLoS Neglected Tropical Diseases. 2013;7(8):e2357.
+\newblock doi:{10.1371/journal.pntd.0002357}.
+
+\bibitem{Schwartz2005}
+Schwartz IB, Shaw LB, Cummings DAT, Billings L, Mccrary M, Burke DS.
+\newblock {Chaotic desynchronization of multistrain diseases}.
+\newblock Physical Review E - Statistical, Nonlinear, and Soft Matter Physics.
+  2005;72(6):066201.
+\newblock doi:{10.1103/PhysRevE.72.066201}.
+
+\bibitem{Chan2012}
+Chan M, Johansson MA.
+\newblock {The incubation periods of dengue viruses}.
+\newblock PLoS One. 2012;7(11):1--7.
+\newblock doi:{10.1371/journal.pone.0050972}.
+
+\bibitem{Lloyd2001}
+Lloyd AL.
+\newblock {Realistic distributions of infectious periods in epidemic models:
+  Changing patterns of persistence and dynamics}.
+\newblock Theor Popul Biol. 2001;60(1):59--71.
+\newblock doi:{10.1006/tpbi.2001.1525}.
+
+\bibitem{Ramsay2017}
+Ramsay J, Hooker G.
+\newblock Dynamic Data Analysis: Modeling Data with Differential Equations.
+\newblock Springer; 2017.
+
+\bibitem{weather}
+Historial weather data for station IESVITOR2, 2008-2012.
+\newblock WeatherUnderground;.
+\newblock Available from:
+  \url{https://www.wunderground.com/dashboard/pws/IESVITOR2}.
+
+\bibitem{vitlong}
+Longevity Subindex.
+\newblock IDHM;.
+\newblock Available from:
+  \url{http://legado.vitoria.es.gov.br/regionais/indicadores/esperancadevidanascer.asp}.
+
+\bibitem{Scott2000}
+Scott TW, Amerasinghe PH, Morrison AC, Lorenz LH, Clark GG, Strickman D, et~al.
+\newblock {Longitudinal studies of \emph{Aedes aegypti} (Diptera: Culicidae) in
+  Thailand and Puerto Rico: blood feeding frequency}.
+\newblock J Med Entomol. 2000;37(1):89--101.
+
+\bibitem{Brady2013}
+Brady OJ, Johansson MA, Guerra CA, Bhatt S, Golding N, Pigott DM, et~al.
+\newblock {Modelling adult \emph{Aedes aegypti} and \emph{Aedes albopictus}
+  survival at different temperatures in laboratory and field settings.}
+\newblock Parasit Vectors. 2013;6:351.
+\newblock doi:{10.1186/1756-3305-6-351}.
+
+\bibitem{Nguyet2013}
+Nguyet MN, Duong THK, Trung VT, Nguyen THQ, Tran CNB, Long VT, et~al.
+\newblock {Host and viral features of human dengue cases shape the population
+  of infected and infectious \emph{Aedes aegypti} mosquitoes.}
+\newblock Proc Natl Acad Sci U S A. 2013;110(22):9072--7.
+\newblock doi:{10.1073/pnas.1303395110}.
+
+\bibitem{Cardoso2011a}
+Cardoso I, Cabidelle A.
+\newblock {Dengue: clinical forms and risk groups in a high incidence city in
+  the southeastern region of Brazil}.
+\newblock Revista da Sociedade Brasileira de Medicina Tropical.
+  2011;44(4):430--435.
+
+\bibitem{Silva2016}
+Silva MMO, Rodrigues MS, Paploski IAD, Kikuti M, Kasper AM, Cruz JS, et~al.
+\newblock {Accuracy of dengue reporting by national surveillance system,
+  Brazil}.
+\newblock Emerg Infect Dis. 2016;22(2):336--339.
+\newblock doi:{10.3201/eid2202.150495}.
+
+\bibitem{Girolami2008}
+Girolami M.
+\newblock {Bayesian inference for differential equations}.
+\newblock Theor Comput Sci. 2008;408(1):4--16.
+\newblock doi:{10.1016/j.tcs.2008.07.005}.
+
+\bibitem{Calderhead2011}
+Calderhead B, Girolami M.
+\newblock {Statistical analysis of nonlinear dynamical systems using
+  differential geometric sampling methods}.
+\newblock Interface Focus. 2011;1(6):821--835.
+\newblock doi:{10.1098/rsfs.2011.0051}.
+
+\bibitem{Leander2014}
+Leander J, Lundh T, Jirstrand M.
+\newblock {Stochastic differential equations as a tool to regularize the
+  parameter estimation problem for continuous time dynamical systems given
+  discrete time measurements}.
+\newblock Math Biosci. 2014;251(1):54--62.
+\newblock doi:{10.1016/j.mbs.2014.03.001}.
+
+\bibitem{Carpenter2016}
+Carpenter B, Gelman A, Hoffman M, Lee D, Goodrich B, Betancourt M, et~al.
+\newblock {Stan: A Probabilistic Programming Language}.
+\newblock J Stat Softw. 2016;VV(Ii).
+\newblock doi:{10.18637/jss.v076.i01}.
+
+\bibitem{Rstan2017}
+{Stan Development Team}. Rstan: the R interface to Stan, R package version
+  2.16.2; 2017.
+
+\bibitem{R2016}
+{R Core Team}. R: A Language and Environment for Statistical Computing; 2016.
+\newblock Available from: \url{https://www.R-project.org/}.
+
+\bibitem{shinystan}
+Gabry J. shinystan: Interactive visual and numerical diagnostics and posterior
+  analysis for Bayesian models; 2018.
+\newblock Available from: \url{https://CRAN.R-project.org/package=shinystan}.
+
+\bibitem{Focks1987}
+Focks DA, Kloter KO, Carmichael GT.
+\newblock {The impact of sequential ultra-low volume ground aerosol
+  applications of malathion on the population dynamics of Aedes aegypti (L.)}.
+\newblock American Journal of Tropical Medicine and Hygiene.
+  1987;36(3):639--647.
+\newblock doi:{10.4269/ajtmh.1987.36.639}.
+
+\bibitem{Esu2010}
+Esu E, Lenhart A, Smith L, Horstick O.
+\newblock {Effectiveness of peridomestic space spraying with insecticide on
+  dengue transmission; Systematic review}.
+\newblock Trop Med Int Heal. 2010;15(5):619--631.
+\newblock doi:{10.1111/j.1365-3156.2010.02489.x}.
+
+\bibitem{Bjornstad2001}
+Bj{\o}rnstad ON, Grenfell BT.
+\newblock {Noisy clockwork: time series analysis of population fluctuations in
+  animals.}
+\newblock Science. 2001;293(5530):638--43.
+\newblock doi:{10.1126/science.1062226}.
+
+\bibitem{Aguiar2011}
+Aguiar M, Ballesteros S, Kooi BW, Stollenwerk N.
+\newblock {The role of seasonality and import in a minimalistic multi-strain
+  dengue model capturing differences between primary and secondary infections:
+  Complex dynamics and its implications for data analysis}.
+\newblock J Theor Biol. 2011;289:181--196.
+\newblock doi:{10.1016/j.jtbi.2011.08.043}.
+
+\bibitem{Codeco2018}
+Code{\c{c}}o CT, Villela DAM, Coelho FC.
+\newblock {Estimating the effective reproduction number of dengue considering
+  temperature-dependent generation intervals}.
+\newblock Epidemics. 2018;25(May):101--111.
+\newblock doi:{10.1016/j.epidem.2018.05.011}.
+
+\bibitem{Barsante2015}
+Barsante LS, Cordeiro FS, Cardoso RTN, Acebal JL, Paix{\~{a}}o KS, Eiras
+  {\'{A}}E.
+\newblock {Controle da dengue na esta{\c{c}}{\~{a}}o seca da cidade de
+  Curvelo-MG atrav{\'{e}}s de um modelo preditivo}.
+\newblock Proceedings of the Brazilian Society of Applied and Computational
+  Mathematics. 2015;3(1):1--7.
+\newblock doi:{10.5540/03.2015.003.01.0082}.
+
+\bibitem{Oki2011}
+Oki M, Sunahara T, Hashizume M, Yamamoto T.
+\newblock {Optimal timing of insecticide fogging to minimize dengue cases:
+  Modeling dengue transmission among various Seasonalities and transmission
+  intensities}.
+\newblock PLoS Neglected Tropical Diseases. 2011;5(10).
+\newblock doi:{10.1371/journal.pntd.0001367}.
+
+\bibitem{Stoddard2014}
+Stoddard ST, Wearing HJ, Jr RCR, Morrison AC, Astete H, Vilcarromero S, et~al.
+\newblock {Long-term and seasonal dynamics of dengue in Iquitos, Peru}.
+\newblock PLoS Neglected Tropical Diseases. 2014;8(7):19--21.
+\newblock doi:{10.1371/journal.pntd.0003003}.
+
+\bibitem{Hladish2018}
+Hladish TJ, Pearson CAB, {Patricia Rojas} D, Gomez-Dantes H, Halloran ME,
+  Vazquez-Prokopec GM, et~al.
+\newblock {Forecasting the effectiveness of indoor residual spraying for
+  reducing dengue burden}.
+\newblock PLoS Neglected Tropical Diseases. 2018;12(6):1--16.
+\newblock doi:{10.1371/journal.pntd.0006570}.
+
+\bibitem{TenBosch2016}
+ten Bosch QA, Singh BK, Hassan MRA, Chadee DD, Michael E.
+\newblock {The role of serotype interactions and seasonality in dengue model
+  selection and control: Insights from a pattern matching approach}.
+\newblock PLoS Neglected Tropical Diseases. 2016;10(5):e0004680.
+\newblock doi:{10.1371/journal.pntd.0004680}.
+
+\bibitem{Finkenstadt2000}
+Finkenstadt BF, Grenfell BT.
+\newblock {Time series modelling of childhood diseases: a dynamical systems
+  approach}.
+\newblock Applied Statistics. 2000;49:187--205.
+\newblock doi:{10.1111/1467-9876.00187}.
+
+\bibitem{Hooker2015}
+Hooker G, Ellner SP.
+\newblock {Goodness of fit in nonlinear dynamics: Misspecified rates or
+  misspecified states?}
+\newblock Ann Appl Stat. 2015;9(2):754--776.
+\newblock doi:{10.1214/15-AOAS828}.
+
+\bibitem{Maciel-de-Freitas2008}
+Maciel{-}de{-}Freitas R.
+\newblock {Calculating the survival rate and estimated population density of
+  gravid \emph{Aedes aegypti} (Diptera, Culicidae) in Rio de Janeiro, Brazil}.
+\newblock Cadernos de Sa{\'{u}}de P{\'{u}}blica. 2008;24(12):2747--2754.
+
+\bibitem{Yang2009}
+Yang HM, Macoris MLG, Galvani KC, Andrighetti MTM, Wanderley DMV.
+\newblock {Assessing the effects of temperature on the population of
+  \emph{Aedes aegypti}, the vector of dengue}.
+\newblock Epidemiol Infect. 2009;137(08):1188.
+\newblock doi:{10.1017/S0950268809002040}.
+
+\bibitem{Morin2013}
+Morin CW, Comrie AC, Ernst K.
+\newblock {Climate and dengue transmission: Evidence and implications}.
+\newblock Environmental Health Perspectives. 2013;121(11-12):1264--1272.
+\newblock doi:{10.1289/ehp.1306556}.
+
+\bibitem{Reiner2015}
+Reiner RC, Guerra C, Donnelly MJ, Bousema T, Drakeley C, Smith DL.
+\newblock {Estimating malaria transmission from humans to mosquitoes in a noisy
+  landscape}.
+\newblock Journal of the Royal Society Interface. 2015;12(20150478).
+\newblock doi:{10.1098/rsif.2015.0478}.
+
+\bibitem{Reich2016}
+Reich NG, Lauer SA, Sakrejda K, Iamsirithaworn S, Hinjoy S, Suangtho P, et~al.
+\newblock {Challenges in Real-Time Prediction of Infectious Disease: A Case
+  Study of Dengue in Thailand}.
+\newblock PLoS Neglected Tropical Diseases. 2016;10(6):1--17.
+\newblock doi:{10.1371/journal.pntd.0004761}.
+
+\bibitem{PAHO2014}
+{Pan American Health Organization}, {World Health Organization}.
+\newblock {Epidemiological alert: Chikungunya and dengue fever in the
+  Americas}; 2014. August.
+
+\bibitem{PAHO2015}
+{Pan American Health Organization}, {World Health Organization}.
+\newblock {Epidemiological update: Zika virus infection}; 2015. October.
+
+\bibitem{Brady2016}
+Brady OJ, Godfray HCJ, Tatem AJ, Gething PW, Cohen JM, McKenzie FE, et~al.
+\newblock {Vectorial capacity and vector control: Reconsidering sensitivity to
+  parameters for malaria elimination}.
+\newblock Transactions of the Royal Society of Tropical Medicine and Hygiene.
+  2016;110(2):107--117.
+\newblock doi:{10.1093/trstmh/trv113}.
+
+\bibitem{Pepin2013}
+Pepin KM, Marques-toledo C, Scherer L, Morais MM, Ellis B, Eiras AE.
+\newblock {Cost-effectiveness of novel system of mosquito surveillance}.
+\newblock Emerg Infect Dis. 2013;19(4).
+
+\bibitem{Adams2009}
+Adams B, Kapan DD.
+\newblock {Man bites mosquito: Understanding the contribution of human movement
+  to vector-borne disease dynamics.}
+\newblock PLoS One. 2009;4(8):e6763.
+\newblock doi:{10.1371/journal.pone.0006763}.
+
+\bibitem{Cosner2009a}
+Cosner C, Beier JC, Cantrell RS, Impoinvil D, Kapitanski L, Potts MD, et~al.
+\newblock {The effects of human movement on the persistence of vector-borne
+  diseases.}
+\newblock J Theor Biol. 2009;258(4):550--60.
+\newblock doi:{10.1016/j.jtbi.2009.02.016}.
+
+\bibitem{Stoddard2009}
+Stoddard ST, Morrison AC, Vazquez-Prokopec GM, {Paz Soldan} V, Kochel TJ,
+  Kitron U, et~al.
+\newblock {The role of human movement in the transmission of vector-borne
+  pathogens.}
+\newblock PLoS Negl Trop Dis. 2009;3(7):e481.
+\newblock doi:{10.1371/journal.pntd.0000481}.
+
+\bibitem{Mondini2008}
+Mondini A, Chiaravalloti-Neto F.
+\newblock {Spatial correlation of incidence of dengue with socioeconomic,
+  demographic and environmental variables in a Brazilian city.}
+\newblock Sci Total Environ. 2008;393(2-3):241--8.
+\newblock doi:{10.1016/j.scitotenv.2008.01.010}.
+
+\bibitem{Honorio2009}
+Hon{\'{o}}rio NA, Nogueira RMR, Code{\c{c}}o CT, Carvalho MS, Cruz OG,
+  Magalh{\~{a}}es MdAFM, et~al.
+\newblock {Spatial evaluation and modeling of Dengue seroprevalence and vector
+  density in Rio de Janeiro, Brazil.}
+\newblock PLoS Negl Trop Dis. 2009;3(11):e545.
+\newblock doi:{10.1371/journal.pntd.0000545}.
+
+\bibitem{Hu2012}
+Hu W, Clements A, Williams G, Tong S, Mengersen K.
+\newblock {Spatial patterns and socioecological drivers of dengue fever
+  transmission in Queensland, Australia.}
+\newblock Environ Health Perspect. 2012;120(2):260--6.
+\newblock doi:{10.1289/ehp.1003270}.
+
+\bibitem{DeMattosAlmeida2007}
+{de Mattos Almeida} MC, Caiaffa WT, Assun{\c{c}}{\~{a}}o RM, Proietti FA.
+\newblock {Spatial vulnerability to dengue in a Brazilian urban area during a
+  7-year surveillance.}
+\newblock J Urban Health. 2007;84(3):334--45.
+\newblock doi:{10.1007/s11524-006-9154-2}.
+
+\end{thebibliography}
 

### 2020-11-09 Manuscript/supplement.tex
Cleaning up notation and typos
Formatting changes

@@
 \usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
-\title{Supplementary material}
+\title{Supplementary text}
 \author{Clinton Leach}
@@
 \\
-\frac{dV_N}{dt} & = r V_N - \phi_q \tau_t V_N\\
-\frac{dV_{E1}}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E1}\\
-\frac{dV_{E2}}{dt} &= \rho_{v}(t) V_{E1} - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E2}\\
-\frac{dV_{E3}}{dt} &= \rho_{v}(t) V_{E2}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E3}\\
-\frac{dV_{E4}}{dt} &= \rho_{v}(t) V_{E3}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E4}\\\frac{dV_I}{dt} &= \rho_{vt} V_{E4} - (d(t) + \phi_q \tau_t) V_I\\
-\frac{dV_C}{dt} & = \phi_q \tau_t V_N\\
+\frac{dV_N}{dt} & = r V_N - \phi_q \tau(t) V_N\\
+\frac{dV_{E1}}{dt} &= \lambda I V_S - (4\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E1}\\
+\frac{dV_{E2}}{dt} &= 4\rho_{v}(t) V_{E1} - (4\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E2}\\
+\frac{dV_{E3}}{dt} &= 4\rho_{v}(t) V_{E2}  - (4\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E3}\\
+\frac{dV_{E4}}{dt} &= 4\rho_{v}(t) V_{E3}  - (4\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E4}\\\frac{dV_I}{dt} &= 4\rho_{v}(t) V_{E4} - (d(t) + \phi_q \tau(t)) V_I\\
+\frac{dV_C}{dt} & = \phi_q \tau(t) V_N\\
 V_S &= V_N - V_E - V_I\\
@@
 We use nonlinear least squares to fit the exponential cdf with rate $d_0$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d_0 = 1.47$.
-We specify a gamma prior on the multiplicative scale, with parameters chosen to keep the prior mass with a factor of 2 of the mean:
+We specify a gamma prior on the multiplicative scale, with parameters chosen to keep the prior mass within a factor of 2 of the mean:
 \begin{equation}
@@
 \\
-\item \textbf{Extrinsic incubation period} ($1/\rho_{vt}$): Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito  (the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
+\item \textbf{Extrinsic incubation period} ($1/\rho_{v}(t)$): Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito  (the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
 To account for this, we make $1/\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground (Goiabeiras weather station, station ID: IESVITOR2).
 We use the temperature relationship from the meta-analysis of Chan annd Johansson \cite{Chan2012}, which incorporates a large number of studies and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
-Chan \textit{et al.}find that the log-normal and gamma distributions are nearly indistinguishable in providingthe best fit to the EIP data.
+Chan and Johansson find that the log-normal and gamma distributions are nearly indistinguishable in providing the best fit to the EIP data.
 To capture a gamma-distributed EIP in the differential equation model, we follow \cite{Lloyd2001} and chain together $\nu$ exposed compartments, each with an out-flow rate of $\nu\rho_v(t)$ to generate
@@
 
-\newpage
-
-\section*{Supplemental figures}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS1.eps}
-\caption{
-Posterior estimates of forcing function on the harmonic oscillators for mosquito mortality (A) and growth (B) rates.  Median posterior estimate (black line) and 80\% credible interval (gray band).
-}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS2.eps}
-\caption{
-Posterior distribution of variances on $\epsilon$ forcing terms for mosquito mortality (A) and growth (B) rates.
-}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS3.eps}
-\caption{
-Posterior distribution of epidemiological parameters.
-The black line indicates the prior density.
-A: scaled latent period in human host ($1/\rho$).
-B: scaled rate of infectious decay in human host ($\gamma$).
-C: scaled period of cross-immunity ($1/\delta$).
-D: scaled baseline mosquito mortality rate ($d_0$).
-}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS4.eps}
-\caption{Posterior distribution of initial conditions.
-The black line indicates the prior density.
-A: initial proportion of the human population that is susceptible.
-B: initial number of exposed individuals.
-C: initial number of infectious individuals.
-D: log initial number of mosquitoes-per-person.
-E: initial value of centered mosquito mortality rate oscillator.
-F: initial value of mosquito population growth rate oscillator.
-}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS5.eps}
-\caption{Posterior distribution of measurement paramters.
-The black line shows the prior density.
-A: log of the mosquito trap capture rate (which is strongly correlated with initial mosquito abundance).
-B: human case reporting probability.
-C: mosquito trap count measurement over-dispersion (prior density too low to be visible).
-D: case report measurement over-dispersion (prior density too low to be visible).}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS6.eps}
-\caption{Posterior check of autocorrelation structure in the case reports (A) and trapped mosquitoes (B) time series.
-Points indicate the observed autocorrelation at the given lag, and vertical lines give the 80\% posterior credible interval of the autocorrelation in the estimated time series.
-}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS7.eps}
-\caption{
-Posterior distribution of the total number of cases reported (A) and the total number of mosquitoes captured (B).  The thick line indicates the observed totals.
-}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS8.eps}
-\caption{
-Posterior distributions of the minimum and maximum weekly reported cases (A and C) and trapped mosquitoes (B and D). The thick line indicates the observed minima and maxima.
-}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[angle = 270, scale = 1]{figures/figS9.eps}
-\caption{
-Posterior estimates of the mosquito mortality rate (top panel) and the probability of a mosquito surving the external incubation period (bottom panel).
-The black line indicates the posterior median, while the gray ribbon shows the 80$\%$ credible interval.
-The probability of surviving the extrinsic incubation period was computed by taking the ratio of the number of mosquitoes that entered the infectious class in a week over the total number of mosquitoes that exited the exposed class that week.
-That is, the probability of surviving the extrinsic incubatio period represents the faction of mosquitoes leaving the exposed class that enter the infectious class.
-}
-\end{figure}
-
 \end{document}
